### 2022-04-10 article/manuscript.tex
added tex file for article and updated gitignore accordingly

\documentclass[]{article}

\usepackage{natbib}
\usepackage{graphicx}

%opening
\title{A multi-species model of landbird detectability that considers phylogeny}
\author{Brandon P.M. Edwards}

\begin{document}

\maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}

\par Detectability is an important component of bird surveys.

\par In birds, detectability has been shown to vary at least partially based on phylogeny. \citet{johnston_species_2014} show that different species traits explain the variability in detectability among different UK birds. \citet{solymos_phylogeny_2018} show that phylogenetic structure is an important predictor of cue rate, and that species traits are more important for estimating effective detection radius.

\par The NA-POPS project has produced estimates of detectability for 338 species of North American landbirds \citep{edwards_point_nodate}, but only in a single-species modelling framework.

\par Here, we derive estimates of detectability under a multi-species framework.

\section{Methods}

\par We accessed the raw count data from the NA-POPS organization for 11 species of Turdidae thrushes \citep{edwards_point_nodate}. The species were: American Robin (\textit{Turdus migratorius}), Eastern Bluebird (\textit{Sialia sialis}), Western Bluebird (\textit{Sialia mexicana}), Mountain Bluebird \textit{Sialia currucoides}, Townsend's Solitaire (\textit{Myadestes townsendi}), Varied Thrush (\textit{Ixoreau naevius}), Wood Thrush (\textit{Hylocichla mustelina}), Swainson's Thrush (\textit{Catharus ustulatus}), Gray-cheeked Thrush (\textit{Catharus minimus}), Hermit Thrush (\textit{Catharus guttatus}), and Veery (\textit{Catharus fuscescens}).

We requested a subsetted phylogeny of the 11 Turdidae species from the birdtree.org website \citep{jetz_global_nodate}, specifying 1000 random trees from an Ericson backbone. Using the "ape" package in R \citep{paradis_ape_2019}, we generated a consensus tree from these 1000 random trees, and generated a correlation matrix following the methods in \citet{solymos_phylogeny_2018} (Figure \ref{fig:phylo_corr}).

\section{Results}

\section{Discussion}

\bibliography{references}

\section{Figures}

\begin{figure}
	\includegraphics[width=\linewidth]{../plots/turdidae_phylo_corr.png}
	\caption{TO DO}
	\label{fig:phylo_corr}
\end{figure}

\end{document}

### 2022-04-11 src/03-distance-model.R
missed two parameter initializations

@@ -144,6 +144,8 @@ stan_job <- sampling(model,
                                       "log_tau[7]" = 3,
                                       "log_tau[8]" = 3,
                                       "log_tau[9]" = 3,
+                                      "log_tau[10]" = 3,
+                                      "log_tau[11]" = 3,
                                       "mu[1]" = 3,
                                       "mu[2]" = 3,
                                       "mu[3]" = 3,
@@ -153,6 +155,8 @@ stan_job <- sampling(model,
                                       "mu[7]" = 3,
                                       "mu[8]" = 3,
                                       "mu[9]" = 3,
+                                      "mu[10]" = 3,
+                                      "mu[11]" = 3,
                                       "sigma[1]" = 1,
                                       "sigma[2]" = 1,
                                       "sigma[3]" = 1,
@@ -161,7 +165,9 @@ stan_job <- sampling(model,
                                       "sigma[6]" = 1,
                                       "sigma[7]" = 1,
                                       "sigma[8]" = 1,
-                                      "sigma[9]" = 1),
+                                      "sigma[9]" = 1,
+                                      "sigma[10]" = 1,
+                                      "sigma[11]" = 1),
                                  list("log_tau[1]" = 3,
                                       "log_tau[2]" = 3,
                                       "log_tau[3]" = 3,
@@ -171,6 +177,8 @@ stan_job <- sampling(model,
                                       "log_tau[7]" = 3,
                                       "log_tau[8]" = 3,
                                       "log_tau[9]" = 3,
+                                      "log_tau[10]" = 3,
+                                      "log_tau[11]" = 3,
                                       "mu[1]" = 3,
                                       "mu[2]" = 3,
                                       "mu[3]" = 3,
@@ -180,6 +188,8 @@ stan_job <- sampling(model,
                                       "mu[7]" = 3,
                                       "mu[8]" = 3,
                                       "mu[9]" = 3,
+                                      "mu[10]" = 3,
+                                      "mu[11]" = 3,
                                       "sigma[1]" = 1,
                                       "sigma[2]" = 1,
                                       "sigma[3]" = 1,
@@ -188,7 +198,9 @@ stan_job <- sampling(model,
                                       "sigma[6]" = 1,
                                       "sigma[7]" = 1,
                                       "sigma[8]" = 1,
-                                      "sigma[9]" = 1),
+                                      "sigma[9]" = 1,
+                                      "sigma[10]" = 1,
+                                      "sigma[11]" = 1),
[…]

### 2022-04-11 src/03-distance-model.R
drop erroneous distance method I

@@ -23,6 +23,10 @@ binomial <- read.csv("output/binomial_names.csv")
 
 # Most of this code adopted from Edwards et al. 2022
 
+# Drop method I
+dist_count_matrix <- dist_count_matrix[-which(dist_count_matrix$Distance_Method == "I"), ]
+dist_design <- dist_design[-which(dist_design$Method == "I"), ]
+
 max_bands <- ncol(dist_design) - 2
 count_names <- c("Sample_ID", "Species", "Distance_Method",
                  paste0(rep("Int", times = max_bands), 1:max_bands))

### 2022-04-11 src/03-distance-model.R
intermediate broken things

@@ -132,113 +132,14 @@ model <- stan_model(file = "models/distance.stan")
 stan_job <- sampling(model,
                      data = stan_data,
                      verbose = TRUE,
-                     chains = 4,
-                     iter = 1000,
-                     warmup = 500,
-                     cores = 4,
+                     chains = 1,
+                     iter = 100,
+                     warmup = 50,
+                     cores = 1,
                      pars = c("log_tau", "sigma", "mu"),
                      control = list(adapt_delta = 0.8,
                                     max_treedepth = 15),
                      init = list(list("log_tau[1]" = 3,
-                                      "log_tau[2]" = 3,
-                                      "log_tau[3]" = 3,
-                                      "log_tau[4]" = 3,
-                                      "log_tau[5]" = 3,
-                                      "log_tau[6]" = 3,
-                                      "log_tau[7]" = 3,
-                                      "log_tau[8]" = 3,
-                                      "log_tau[9]" = 3,
-                                      "log_tau[10]" = 3,
-                                      "log_tau[11]" = 3,
-                                      "mu[1]" = 3,
-                                      "mu[2]" = 3,
-                                      "mu[3]" = 3,
-                                      "mu[4]" = 3,
-                                      "mu[5]" = 3,
-                                      "mu[6]" = 3,
-                                      "mu[7]" = 3,
-                                      "mu[8]" = 3,
-                                      "mu[9]" = 3,
-                                      "mu[10]" = 3,
-                                      "mu[11]" = 3,
-                                      "sigma[1]" = 1,
-                                      "sigma[2]" = 1,
-                                      "sigma[3]" = 1,
-                                      "sigma[4]" = 1,
-                                      "sigma[5]" = 1,
-                                      "sigma[6]" = 1,
-                                      "sigma[7]" = 1,
-                                      "sigma[8]" = 1,
-                                      "sigma[9]" = 1,
-                                      "sigma[10]" = 1,
-                                      "sigma[11]" = 1),
-                                 list("log_tau[1]" = 3,
-                                      "log_tau[2]" = 3,
-                                      "log_tau[3]" = 3,
-                                      "log_tau[4]" = 3,
-                                      "log_tau[5]" = 3,
-                                      "log_tau[6]" = 3,
-                                      "log_tau[7]" = 3,
-                                      "log_tau[8]" = 3,
-                                      "log_tau[9]" = 3,
-                                      "log_tau[10]" = 3,
-                                      "log_tau[11]" = 3,
[…]

### 2023-05-15 src/05-distance-prior-check.R
updating distance model to reflect need for informative priors

@@ -3,7 +3,7 @@
 # Multi-species QPAD Detectability
 # 05-distance-prior-check.R
 # Created March 2023
-# Last Updated March 2023
+# Last Updated May 2023
 
 ####### Import Libraries and External Files #######
 
@@ -42,7 +42,7 @@ print(ggplot(data = data.frame(sigma), aes(x = sigma)) +
         NULL)
 dev.off()
 
-intercept <- rnorm(n = n_sims, mean = 4, sd = 0.1)
+intercept <- rnorm(n = n_sims, mean = 0, sd = 0.001)
 pdf(file = "output/prior_predictive_check/distance/intercept.pdf")
 print(ggplot(data = data.frame(intercept), aes(x = intercept)) +
         geom_histogram(bins = 20) +
@@ -55,7 +55,7 @@ mu_mig_strat <- matrix(data = NA,
                        nrow = n_sims)
 for (i in 1:distance_stan_data_pred$n_mig_strat)
 {
-  mu_mig_strat[,i] <- rnorm(n_sims, mean = 0, sd = 0.05)
+  mu_mig_strat[,i] <- rnorm(n_sims, mean = 0, sd = 0.001)
 }
 pdf(file = "output/prior_predictive_check/distance/mu_mig_strat.pdf")
 to_plot <- data.frame(Value = c(mu_mig_strat[,1],
@@ -79,7 +79,7 @@ mu_habitat <- matrix(data = NA,
                      nrow = n_sims)
 for (i in 1:distance_stan_data_pred$n_habitat)
 {
-  mu_habitat[,i] <- rnorm(n_sims, mean = 0, sd = 0.05)
+  mu_habitat[,i] <- rnorm(n_sims, mean = 0, sd = 0.001)
 }
 pdf(file = "output/prior_predictive_check/distance/mu_habitat.pdf")
 to_plot <- data.frame(Value = c(mu_habitat[,1],
@@ -98,7 +98,7 @@ for (i in unique(to_plot$Habitat))
 dev.off()
 
 # beta mass
-beta_mass <- rnorm(n = n_sims, mean = 0.01, sd = 0.005)
+beta_mass <- rnorm(n = n_sims, mean = 0, sd = 0.001)
 pdf(file = "output/prior_predictive_check/distance/beta_mass.pdf")
 
 mass_hist <- ggplot(data = data.frame(Mass_Slope = beta_mass), aes(x = Mass_Slope)) +
[…]

### 2023-05-15 article/article.tex

\documentclass[]{article}

%opening
\title{Predicting detection probabilities for rare and undersampled birds}
\author{}

\begin{document}

\maketitle

\begin{abstract}

	Insert abstract here.

\end{abstract}

\section{Introduction}

Detection probability, or detectability, is an important metric to consider in conservation management problems (Bennett et al., unpubl.). When deciding how to spend limited conservation budgets (Buxton et al., 2020), the role of detectability can vastly shape the decision on when to monitor a species and when to take conservation action (Bennett et al., unpubl.; Canessa et al., 2015). Unfortunately, detectability is often unaccounted for in conservation problems, because detection probabilities are not known for a large proportion of taxa (Bennett et al., unpubl.). By not accounting for detectability, wrong decisions about where and when to monitor can often be made, which can be perilous in species at risk of extirpation or extinction (Bennett et al., unpubl.).

\end{document}

### 2023-05-16 src/05-distance-prior-check.R
updated names of output files to be numerically ordered

@@ -36,14 +36,14 @@ for (i in 1:dim(phylo_corr_pl)[1])
 
 # Sigma
 sigma <- rexp(n = n_sims, rate = 5)
-pdf(file = "output/prior_predictive_check/distance/sigma.pdf")
+pdf(file = "output/prior_predictive_check/distance/01-sigma.pdf")
 print(ggplot(data = data.frame(sigma), aes(x = sigma)) +
         geom_histogram(bins = 20) +
         NULL)
 dev.off()
 
 intercept <- rnorm(n = n_sims, mean = 0, sd = 0.001)
-pdf(file = "output/prior_predictive_check/distance/intercept.pdf")
+pdf(file = "output/prior_predictive_check/distance/02-intercept.pdf")
 print(ggplot(data = data.frame(intercept), aes(x = intercept)) +
         geom_histogram(bins = 20) +
         NULL)
@@ -57,7 +57,7 @@ for (i in 1:distance_stan_data_pred$n_mig_strat)
 {
   mu_mig_strat[,i] <- rnorm(n_sims, mean = 0, sd = 0.001)
 }
-pdf(file = "output/prior_predictive_check/distance/mu_mig_strat.pdf")
+pdf(file = "output/prior_predictive_check/distance/03-mu_mig_strat.pdf")
 to_plot <- data.frame(Value = c(mu_mig_strat[,1],
                                 mu_mig_strat[,2]),
                       Mig_Strat = c(rep("Resident", 100),
@@ -81,7 +81,7 @@ for (i in 1:distance_stan_data_pred$n_habitat)
 {
   mu_habitat[,i] <- rnorm(n_sims, mean = 0, sd = 0.001)
 }
-pdf(file = "output/prior_predictive_check/distance/mu_habitat.pdf")
+pdf(file = "output/prior_predictive_check/distance/04-mu_habitat.pdf")
 to_plot <- data.frame(Value = c(mu_habitat[,1],
                                 mu_habitat[,2]),
                       Habitat = c(rep("Open", 100),
@@ -99,7 +99,7 @@ dev.off()
 
 # beta mass
 beta_mass <- rnorm(n = n_sims, mean = 0, sd = 0.001)
-pdf(file = "output/prior_predictive_check/distance/beta_mass.pdf")
+pdf(file = "output/prior_predictive_check/distance/05-beta_mass.pdf")
 
[…]

### 2023-05-16 src/05-distance-prior-check.R
standard normal priors on prior predictive check (distance)

@@ -42,7 +42,7 @@ print(ggplot(data = data.frame(sigma), aes(x = sigma)) +
         NULL)
 dev.off()
 
-intercept <- rnorm(n = n_sims, mean = 0, sd = 0.001)
+intercept <- rnorm(n = n_sims)
 pdf(file = "output/prior_predictive_check/distance/02-intercept.pdf")
 print(ggplot(data = data.frame(intercept), aes(x = intercept)) +
         geom_histogram(bins = 20) +
@@ -55,7 +55,7 @@ mu_mig_strat <- matrix(data = NA,
                        nrow = n_sims)
 for (i in 1:distance_stan_data_pred$n_mig_strat)
 {
-  mu_mig_strat[,i] <- rnorm(n_sims, mean = 0, sd = 0.001)
+  mu_mig_strat[,i] <- rnorm(n_sims)
 }
 pdf(file = "output/prior_predictive_check/distance/03-mu_mig_strat.pdf")
 to_plot <- data.frame(Value = c(mu_mig_strat[,1],
@@ -79,7 +79,7 @@ mu_habitat <- matrix(data = NA,
                      nrow = n_sims)
 for (i in 1:distance_stan_data_pred$n_habitat)
 {
-  mu_habitat[,i] <- rnorm(n_sims, mean = 0, sd = 0.001)
+  mu_habitat[,i] <- rnorm(n_sims)
 }
 pdf(file = "output/prior_predictive_check/distance/04-mu_habitat.pdf")
 to_plot <- data.frame(Value = c(mu_habitat[,1],
@@ -98,7 +98,7 @@ for (i in unique(to_plot$Habitat))
 dev.off()
 
 # beta mass
-beta_mass <- rnorm(n = n_sims, mean = 0, sd = 0.001)
+beta_mass <- rnorm(n = n_sims)
 pdf(file = "output/prior_predictive_check/distance/05-beta_mass.pdf")
 
 mass_hist <- ggplot(data = data.frame(Mass_Slope = beta_mass), aes(x = Mass_Slope)) +
@@ -121,7 +121,7 @@ print(mass_plot)
 dev.off()
 
 # beta pitch
-beta_pitch <- rnorm(n = n_sims, mean = 0, sd = 0.001)
+beta_pitch <- rnorm(n = n_sims)
 pdf(file = "output/prior_predictive_check/distance/06-beta_pitch.pdf")
 
[…]

### 2023-05-16 src/05-distance-prior-check.R
testing with scaled covariates, unscaled max distances

@@ -181,7 +181,7 @@ for (s in 1:distance_stan_data_pred$n_species)
 dev.off()
 
 pdf(file = "output/prior_predictive_check/distance/09-tau.pdf")
-tau <- exp(log_tau) * 100
+tau <- exp(log_tau) #* 100
 for (s in 1:distance_stan_data_pred$n_species)
 {
   to_plot <- data.frame(Value = (tau[,s]))

### 2023-05-16 src/07-distance-model.R

@@ -34,7 +34,8 @@ threads_per_chain <- 3
 distance_stan_data_pred$pitch <- distance_stan_data_pred$pitch[,1]
 distance_stan_data_pred$mass <- distance_stan_data_pred$mass[,1]
 
-
+# temporarily unscale the max_dists to be in metres, for diagnostic purposes
+distance_stan_data_pred$max_dist <- distance_stan_data_pred$max_dist * 100
 model_file <- cmdstan_model(stan_file = "models/distance.stan",
                             cpp_options = list(stan_threads = TRUE))
 

### 2023-05-16 src/05-distance-prior-check.R
reverting back to scaled distances

@@ -181,7 +181,7 @@ for (s in 1:distance_stan_data_pred$n_species)
 dev.off()
 
 pdf(file = "output/prior_predictive_check/distance/09-tau.pdf")
-tau <- exp(log_tau) #* 100
+tau <- exp(log_tau) * 100
 for (s in 1:distance_stan_data_pred$n_species)
 {
   to_plot <- data.frame(Value = (tau[,s]))

### 2023-05-16 src/07-distance-model.R

@@ -34,8 +34,6 @@ threads_per_chain <- 3
 distance_stan_data_pred$pitch <- distance_stan_data_pred$pitch[,1]
 distance_stan_data_pred$mass <- distance_stan_data_pred$mass[,1]
 
-# temporarily unscale the max_dists to be in metres, for diagnostic purposes
-distance_stan_data_pred$max_dist <- distance_stan_data_pred$max_dist * 100
 model_file <- cmdstan_model(stan_file = "models/distance.stan",
                             cpp_options = list(stan_threads = TRUE))
 

### 2023-05-16 src/04-prepare-distance-data.R
changed maximum distance value to 450m rather than 1000m

@@ -26,7 +26,7 @@ traits <- read.csv("data/raw/traits.csv")
 ####### Wrangle Data for Modelling ################
 
 # Change infinite values to some other very large number to avoid Stan issues
-dist_design <- do.call(data.frame,lapply(dist_design, function(x) replace(x, is.infinite(x),1000)))
+dist_design <- do.call(data.frame,lapply(dist_design, function(x) replace(x, is.infinite(x),450)))
 
 # Most of this code adopted from Edwards et al. 2022
 

### 2023-05-17 src/04-prepare-distance-data.R
take the log of the inputted max distances, and do not exponentiate the estimated log tau in the model

@@ -170,11 +170,13 @@ abundance_per_band_pred[is.na(abundance_per_band_pred)] <- 0
 #' Corresponds with "max_dist" in distance.stan
 max_dist_cv <- D_cv
 max_dist_cv[is.na(max_dist_cv)] <- 0
-max_dist_cv <- max_dist_cv / 100 # scale to be on scale of 100m
+max_dist_cv <- log(max_dist_cv)# / 100 # scale to be on scale of 100m
+max_dist_cv[is.infinite(max_dist_cv)] <- -1000 # change all -Inf values to -1000
 
 max_dist_pred <- D_pred
 max_dist_pred[is.na(max_dist_pred)] <- 0
-max_dist_pred <- max_dist_pred / 100 #scale to be on scale of 100m
+max_dist_pred <- log(max_dist_pred)# / 100 #scale to be on scale of 100m
+max_dist_pred[is.infinite(max_dist_pred)] <- -1000
 
 n_samples_cv <- nrow(Y_cv)
 n_samples_pred <- nrow(Y_pred)

### 2023-05-17 src/04-prepare-distance-data.R
removed scaling to 100m

@@ -170,11 +170,9 @@ abundance_per_band_pred[is.na(abundance_per_band_pred)] <- 0
 #' Corresponds with "max_dist" in distance.stan
 max_dist_cv <- D_cv
 max_dist_cv[is.na(max_dist_cv)] <- 0
-max_dist_cv <- max_dist_cv / 100 # scale to be on scale of 100m
 
 max_dist_pred <- D_pred
 max_dist_pred[is.na(max_dist_pred)] <- 0
-max_dist_pred <- max_dist_pred / 100 #scale to be on scale of 100m
 
 n_samples_cv <- nrow(Y_cv)
 n_samples_pred <- nrow(Y_pred)

### 2023-10-04 article/article.tex
wrote up derivations of the removal and distance models

@@
-\documentclass[]{article}
+\documentclass[12pt]{article}
+
+\usepackage{setspace}
+\onehalfspacing
+
+\usepackage{natbib}
+\bibliographystyle{apalike}
+
+\usepackage[margin=1.0in]{geometry}
+
+\usepackage{amsmath}
+
+\usepackage{lineno}
+\linenumbers
 
 %opening
-\title{Predicting detection probabilities for rare and undersampled birds}
-\author{}
+\title{Predicting detection probabilities for rare and undersampled North American landbirds}
+\author{
+	Edwards, Brandon P.M.\\
+	\and
+	Solymos, Peter\\
+	\and
+	Stralberg, Diana\\
+	\and
+	Michel, Nicole\\
+	\and
+	Robinson, Barry\\
+	\and
+	Gahbauer, Marcel\\
+	\and
+	Grinde, Alexis\\
+	\and
+	Hope, David\\
+	\and
+	Bennett, Joseph R.\\
+	\and
+	Smith, Adam C.\\
+}
 
@@
 
-	Insert abstract here.
-
 \end{abstract}
@@
 
-Detection probability, or detectability, is an important metric to consider in conservation management problems (Bennett et al., unpubl.).
-When deciding how to spend limited conservation budgets (Buxton et al., 2020), the role of detectability can vastly shape the decision on when to monitor a species and when to take conservation action (Bennett et al., unpubl.; Canessa et al., 2015).
-Unfortunately, detectability is often unaccounted for in conservation problems, because detection probabilities are not known for a large proportion of taxa (Bennett et al., unpubl.).
-By not accounting for detectability, wrong decisions about where and when to monitor can often be made, which can be perilous in species at risk of extirpation or extinction (Bennett et al., unpubl.).
+\section{Methods}
+
+\par The following subsections outline the modelling techniques used in this study as well as the data that were collected. 
+We start by introducing the Bayesian hierarchical QPAD removal and distance models, and then describe the input.
+
+\par For any point count that employs removal and/or distance sampling, let $Y_{sijk}$ be the observed count of species $s$ during sampling event $i$, occuring in time band $j \in [1,J]$ and/or distance band $k \in [1,K]$.
+That is, if we have a point count where removal sampling or distance sampling (but not both) is not used, then we have that either $J = 1$ or $K = 1$, respectively. 
+The following two subsections detail the Bayesian hierarchical QPAD Removal and Distance Models.
+
+\subsection{Bayesian Hierarchical QPAD Removal Model}
+
+\par For removal modelling, we are not interested in distance to the bird, and so we will sum counts from each sampling event $i$ over all distance bands. 
+Thus, we will be considering counts $Y_{sij.}$. 
+From \citet{solymos_calibrating_2013}, we model $Y_{sij.}$ as
+$$Y_{sij.} \sim multinomial\left(Y_{si..}, \Pi_{si.}\right).$$
+
+\par For sample $i$, $\Pi_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $J$ time bands; in essense, it is the mixing parameter.
+Let $t_ij$ be the maximum time for time band $j$ during sampling event $i$, and let $phi_s$ be the unknown cue rate for species $s$.
+We then have the following calculation of each component $\pi_{sij}$ of the mixing parameter $\Pi_{si.}$:
+
+\begin{equation*}
+	\pi_{sij} = 
+	\begin{cases}
+		\dfrac{\exp\left\{ -t_{i,j-1}\phi_{s} \right\} - \exp\left\{ -t_{ij}\phi_{s} \right\}}{1 - \exp\left\{ -t_{iJ}\phi_{s} \right\}} & \text{for } j > 1 \\
+		1 - \sum_{n = 2}^{J} \pi_{sin} & \text{for } j = 1
+	\end{cases}
+\end{equation*}
+
+\par We are most interested in estimating $phi_s$, the cue rate for species $s$. 
+To allow for the partial pooling of estimates for each species to be based on a phylogenic relationship, we can use Gaussian processes, which are effectively allow for "continuous" factors. 
+Let $\Phi$ be the vector of cue rates (of length $s$). 
+We then have:
+
+$$\log \Phi \sim MVN\left( \mu_{[s]}, \Sigma_{[s \times s]} \right).$$
+
+\par The covariance function $\Sigma$ (denoted with subscript $[s \times s]$ to show dimensionality) determines the strength of the partial pooling, and can be any function that contains a measure of distance.
+In this case, we created a function based on phylogenetic distance $D_{[s \times s]}$, i.e.,
+
+$$\Sigma_{[s \times s]} = D_{[s \times s]}\lambda\sigma.$$
+
+\par Here, $\lambda$ is Pagal's Lambda (CITATION), which determines the strength of the phylogenetic relationship. 
+We set $\lambda = 0.76$, consistent with previous results \cite{solymos_phylogeny_2018}. 
+We also have the extra variance term $\sigma \sim exp(5)$.
+
+\par Finally, the vector of mean cue rate $\mu_{[s]}$ is just a function of the species migratory strategy $\alpha_M$, where the migratory strategy $M$ could either be Resident or Migrant. 
+We set priors on both migratory strategies to be $\alpha_M \sim N(-1, 0.01)$, because on the log scale we would expect cue rate to be negative.
+
+\par Results of simulated priors can be found in Appendix TO DO.
+
+\subsection{Bayesian Hierarchical QPAD Distance Model}
+
+\par For distance modelling, we are not interested in the time bin that the bird was recorded in, and so we will sum counts from each sampling event $i$ over all time bands. 
+Thus, we will be considering counts $Y_{si.k}$. 
+From \citet{solymos_calibrating_2013}, we model $Y_{si.k}$ as
+$$Y_{si.k} \sim multinomial\left(Y_{si..}, \Pi_{si.}\right).$$
+
+\par For sample $i$, $\Pi_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $K$ distance bins; in essense, it is the mixing parameter.
+Let $r_ik$ be the maximum distance for distance bin $k$ during sampling event $i$, and let $tau_s$ be the unknown effective detection radius for species $s$. 
+We then have the following calculation of each component $\pi_{sik}$ of the mixing parameter $\Pi_{si.}$:
+
+%\begin{equation*}
+%	\pi_{sik} = 
+%	\begin{cases}
+%		\dfrac{\left( 1 - \exp\left\{ -\dfrac{r_{i,k}^2}{\exp\left\{\tau_s^2\right\}} \right\} \right) - \left( 1 - \exp\left\{ -\dfrac{r_{i,k-1}^2}{\exp\left\{\tau_s^2\right\}} \right\} \right)}
+%		{ 1 - \exp\left\{ -\dfrac{r_{i,K}^2}{\exp\left\{\tau_s^2\right\}} \right\} } & \text{for } k > 1 \\
+%		1 - \sum_{n = 2}^{J} \pi_{sin} & \text{for } k = 1
+%	\end{cases}
+%\end{equation*}
+
+\begin{equation*}
+	\pi_{sik} = 
+	\begin{cases}
+		\dfrac{f(r_{i,k}, \tau_s) - f(r_{i,k-1}, \tau_s)}{f(r_{i,K}, \tau_s)} & \text{for } k > 1 \\
+		1 - \sum_{n = 2}^{J} \pi_{sin} & \text{for } k = 1
+	\end{cases}
+\end{equation*}
+
+where 
+$$f(r,\tau) =  1 - \exp\left\{ -\dfrac{r^2}{\exp\left\{\tau^2\right\}} \right\} .$$
+
+\par We are most interested in estimating $\tau_s$, the effective detection radius for species $s$. 
+To allow for the partial pooling of estimates for each species to be based on traits, we set up a mixed effects model as follows:
+$$\log \tau_s \sim N(\mu_s, \sigma).$$
+
+\par In this model, mean effective detection radius for species $s$ (i.e., $\mu_s$) is a function of the following traits as described in \cite{solymos_phylogeny_2018}: an overall intercept term $\alpha_0$; migratory strategy $\alpha_M$, where $M$ can be Migratory or Resident; habitat preference $\alpha_H$, where $H$ can be Open or Closed habitat; log body size $\beta_b$; and log song pitch $\beta_p$.
+That is, we have
+$$ \mu_s = \alpha_0 + \alpha_{M_s} + \alpha_{H_s} + \beta_b \times \log BodySize_s + \beta_p \times \log SongPitch_s$$
+$$\alpha_0 \sim N(0.05, 0.1)$$
+$$ \alpha_M, \alpha_H \sim N(0, 0.05)$$
+$$ \beta_b \sim N(0.01, 0.005)$$
+$$ \beta_p \sim N(-0.01, 0.005) $$
+$$\sigma \sim exp(5)$$
+
+\par Results of simulated priors can be found in Appendix TO DO.
+
+\bibliography{refs}
 

### 2023-10-04 article/article.tex
added more to methods and added bibliography

@@
 
+\subsection{Data Collection}
+\subsubsection{Point Count Data}
+\par We used the same standardized database of point count data that was put together through the NA-POPS project \cite{edwards_point_2023}. 
+This consists of $712138$ point counts conducted across $292$ projects in Canada and the United States, $422514$ of which used removal sampling and $522820$ of which used distance sampling.
+
+\subsubsection{Phylogenetic Trees}
+\par We accessed 2000 pseudo-posterior trees from \citet{jetz_global_2012} that contains phylogeny for all species that contained sufficient data for removal modelling ($>75$ observations as per \cite{edwards_point_2023,solymos_evaluating_2018,buckland_introduction_2001}).
+We also requested that the phylogenies include the following bird species with insufficient data from NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird, and Spotted Owl.
+All seven of these bird species are listed on Partner's in Flight's watchlist for various threats (CITATION). 
+This resulted in 2000 pseudo-posterior trees that contained 323 species.
+
+\par We generated a phylogenetic correlation matrix using the same methods outlined in the Appendix of \citet{solymos_phylogeny_2018} that generate distances between species, and further processed using the R package "ape" \citet{paradis_ape_2019}.
+
+\subsubsection{Species Traits}
+\par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts of all the birds to model in this study (CITATION). 
+Body mass for each bird species was obtained form the Elton traits dataset (CITATION). 
+Where body mass was not available through the Elton traits dataset, the Birds of the World account was used instead.
+
+\subsection{Bayesian Hierarchical Models}
+
 \par For any point count that employs removal and/or distance sampling, let $Y_{sijk}$ be the observed count of species $s$ during sampling event $i$, occuring in time band $j \in [1,J]$ and/or distance band $k \in [1,K]$.
@@
 
-\subsection{Bayesian Hierarchical QPAD Removal Model}
+\subsubsection{Bayesian Hierarchical QPAD Removal Model}
 
@@
 
-\subsection{Bayesian Hierarchical QPAD Distance Model}
+\subsubsection{Bayesian Hierarchical QPAD Distance Model}
 
@@
 
+\subsection{Model Assessment}
+
 \bibliography{refs}

### 2023-10-05 src/08-removal-analysis.R
use bayesplot to plot posterior of all models, and better model differences

@@ -14,6 +14,7 @@ library(plyr)
 library(ggplot2)
 library(ggpubr)
 theme_set(theme_pubclean())
+bayesplot::color_scheme_set("red")
 
 ####### Read Data #################################
 
@@ -56,18 +57,18 @@ napops_summary <- napops::coef_removal(species = rem_summary$Code, model = 1)
 
 ####### 1-to-1 Plot ###############################
 
-to_plot <- merge(rem_summary[,c("Code", "mean")], napops_summary[, c("Species", "Intercept")],
+to_plot <- merge(rem_summary[,c("Code", "mean", "N")], napops_summary[, c("Species", "Intercept")],
                  by.x = "Code", by.y = "Species")
-names(to_plot) <- c("Species", "Multi", "Single")
+names(to_plot) <- c("Species", "Multi", "N", "Single")
 to_plot$Label <- ""
 to_plot$Multi <- exp(to_plot$Multi)
 to_plot$Single <- exp(to_plot$Single)
 to_plot$diff <- to_plot$Multi - to_plot$Single
 
-# Check for a 5% relative difference in cue rates
+# Check for a 20% relative difference in cue rates
 for (i in 1:nrow(to_plot))
 {
-  if ((abs(to_plot$diff[i]) / to_plot$Single[i]) > 0.10)
+  if ((abs(to_plot$diff[i]) / to_plot$Single[i]) > 0.20)
   {
     to_plot$Label[i] <- to_plot$Species[i]
   }
@@ -93,7 +94,23 @@ for (i in 1:nrow(to_plot))
     ylab("Difference in Cue Rate (Multi - Single Model)") +
   NULL)
 
+diff_model_data <- list(n_samples = nrow(to_plot),
+                        difference = to_plot$diff,
+                        N = to_plot$N)
+diff_model <- cmdstan_model(stan_file = "models/difference.stan")
 
+diff_model_run <- diff_model$sample(
+  data = diff_model_data,
+  iter_warmup = 1000,
+  iter_sampling = 2000,
+  chains = 4,
+  parallel_chains = 4,
+  refresh = 10
+)
+
+(modelled_difference_plot <- bayesplot::mcmc_areas(diff_model_run$draws(c("mu")),
+                                                   prob = 0.95) +
+  xlab("Modelled Difference"))
 
 ####### SD Comparison Plot ########################
 
[…]

### 2023-10-05 src/08-removal-analysis.R
added bayesplot functionality for distance modelling

@@ -28,15 +28,6 @@ load("data/generated/removal_stan_data.rda")
 # Extract log_phi summary statistics from full Stan model runs
 rem_summary <- rem_model$summary("log_phi")
 
-# Model diagnostics
-
-# diag <- rem_model$sampler_diagnostics(format = "df")
-# sigma_draws <- rem_model$draws(variables = "mu_mig_strat[1]", format = "df")
-# sigma_draws$.chain <- factor(sigma_draws$.chain, levels = c("1", "2", "3", "4"))
-# ggplot(data = sigma_draws, aes(x = .iteration, y = mu, group = (.chain), color = .chain)) +
-#   geom_line()
-# mcmc_hist(rem_model$draws("sigma")) 
-
 # Add species names to these summaries
 rem_summary$Scientific_BT <- gsub("_", " ", rownames(corr_matrix_predict))
 rem_summary <- join(rem_summary, binomial[, c("Scientific_BT", "Code")], by = "Scientific_BT")
@@ -161,14 +152,16 @@ species_vars <- to_plot$variable
 (species_prediction_plot <- bayesplot::mcmc_intervals(exp(rem_model$draws(species_vars))) + 
   xlab("Predicted Cue Rate") + 
   ylab("Species") +
-  scale_y_discrete(labels = rev(sp)) +
+  scale_y_discrete(labels = (to_plot$Code)) +
   coord_flip())
 
 ####### Output ####################################
 
 png("output/plots/removal_1vs1.png",
-    width = 12, height = 8, res = 600, units = "in")
-ggarrange(single_vs_multi_plot, modelled_difference_plot, ncol = 2, labels = c("A", "B"))
+    width = 12, height = 12, res = 600, units = "in")
+ggarrange(ggarrange(single_vs_multi_plot, modelled_difference_plot,
+                    ncol = 2,
+                    labels = c("A", "B")), difference_plot, nrow = 2, labels = c("", "C"))
 dev.off()
 
 png("output/plots/removal_sd_plot.png",

### 2023-10-05 src/09-distance-analysis.R

@@ -13,6 +13,7 @@ library(plyr)
 library(ggplot2)
 library(ggpubr)
 theme_set(theme_pubclean())
+bayesplot::color_scheme_set("red")
 
 ####### Read Data #################################
 
@@ -54,10 +55,10 @@ to_plot$Multi <- exp(to_plot$Multi) * 100
 to_plot$Single <- exp(to_plot$Single)
 to_plot$diff <- to_plot$Multi - to_plot$Single
 
-# Check for a 10% relative difference in EDR
+# Check for a 20% relative difference in EDR
 for (i in 1:nrow(to_plot))
 {
-  if ((abs(to_plot$diff[i]) / to_plot$Single[i]) > 0.10)
+  if ((abs(to_plot$diff[i]) / to_plot$Single[i]) > 0.20)
   {
     to_plot$Label[i] <- to_plot$Species[i]
   }
@@ -82,6 +83,23 @@ for (i in 1:nrow(to_plot))
           axis.ticks.x=element_blank()) +
     ylab("Difference in EDR (Multi - Single Model)") +
     NULL)
+
+diff_model_data <- list(n_samples = nrow(to_plot),
+                        difference = to_plot$diff)
+diff_model <- cmdstan_model(stan_file = "models/difference.stan")
+
+diff_model_run <- diff_model$sample(
+  data = diff_model_data,
+  iter_warmup = 1000,
+  iter_sampling = 2000,
+  chains = 4,
+  parallel_chains = 4,
+  refresh = 10
+)
+
+(modelled_difference_plot <- bayesplot::mcmc_areas(diff_model_run$draws(c("mu")),
+                                                   prob = 0.95) +
+    xlab("Modelled Difference"))
 
 ####### SD Comparison Plot ########################
 
@@ -115,13 +133,11 @@ sd_model_run <- sd_model$sample(
   refresh = 10
 )
 
-sd_comp_plot <- ggplot(data = to_plot_long, 
-                       aes(x = log(N), y = (value),
-                           group = variable, color = variable)) + 
-  geom_smooth() +
-  xlab("log(Sample Size)") +
-  ylab("Standard Deviation") +
-  NULL
+(sd_intercept_plot <- bayesplot::mcmc_areas(sd_model_run$draws(c("alpha[1]", "alpha[2]")),
+                                           prob = 0.95))
+
+(sd_slope_plot <- bayesplot::mcmc_areas(sd_model_run$draws(c("beta[1]", "beta[2]")),
[…]

### 2023-10-05 src/08-removal-analysis.R
change size of figures

@@ -165,12 +165,12 @@ ggarrange(ggarrange(single_vs_multi_plot, modelled_difference_plot,
 dev.off()
 
 png("output/plots/removal_sd_plot.png",
-    width = 6, height = 6, res = 600, units = "in")
+    width = 12, height = 12, res = 600, units = "in")
 ggarrange(sd_intercept_plot, sd_slope_plot, ncol = 2, labels = c("A", "B"))
 dev.off()
 
 png("output/plots/removal_predictions_plot.png",
-    width = 6, height = 4, res = 600, units = "in")
+    width = 12, height = 12, res = 600, units = "in")
 print(species_prediction_plot)
 dev.off()
 

### 2023-10-05 src/09-distance-analysis.R

@@ -163,12 +163,12 @@ ggarrange(ggarrange(single_vs_multi_plot, modelled_difference_plot,
 dev.off()
 
 png("output/plots/distance_sd_plot.png",
-    width = 6, height = 6, res = 600, units = "in")
+    width = 12, height = 12, res = 600, units = "in")
 ggarrange(sd_intercept_plot, sd_slope_plot, ncol = 2, labels = c("A", "B"))
 dev.off()
 
 png("output/plots/distance_predictions_plot.png",
-    width = 6, height = 4, res = 600, units = "in")
+    width = 12, height = 12, res = 600, units = "in")
 print(species_prediction_plot)
 dev.off()
 

### 2023-10-05 src/09-distance-analysis.R
change figure sizes back to 6in

@@ -156,19 +156,19 @@ species_vars <- to_plot$variable
 ####### Output ####################################
 
 png("output/plots/distance_1vs1.png",
-    width = 12, height = 12, res = 600, units = "in")
+    width = 6, height = 6, res = 600, units = "in")
 ggarrange(ggarrange(single_vs_multi_plot, modelled_difference_plot,
                     ncol = 2,
                     labels = c("A", "B")), difference_plot, nrow = 2, labels = c("", "C"))
 dev.off()
 
 png("output/plots/distance_sd_plot.png",
-    width = 12, height = 12, res = 600, units = "in")
+    width = 6, height = 6, res = 600, units = "in")
 ggarrange(sd_intercept_plot, sd_slope_plot, ncol = 2, labels = c("A", "B"))
 dev.off()
 
 png("output/plots/distance_predictions_plot.png",
-    width = 12, height = 12, res = 600, units = "in")
+    width = 6, height = 6, res = 600, units = "in")
 print(species_prediction_plot)
 dev.off()
 

### 2023-10-05 article/article.tex
finished methods section (as much as possible for now), added figures, and added predictions text

@@
 \usepackage{lineno}
+
+\usepackage{graphicx}
 \linenumbers
@@
 \section{Introduction}
+
+\subsection{Predictions}
+\par We predict that for both the Bayesian hierarchical QPAD removal model and Bayesian hierarchical QPAD distance model, estimates of cue rate and EDR, respectively, will change very little from those estimated from the original NA-POPS database \cite{edwards_point_2023}.
+This is because the species that were modelled in \citet{edwards_point_2023} had a sufficient number of data points to stand on their own, and so adding in species-level effects should have little effect on estimates.
+We also predict that a multi-species model will increase the precision of the estimates of cue rate and EDR, but only up to a certain sample size. 
+This is because there are many species for which NA-POPS has thousands to even tens of thousands of data points; given precision generally increases with sample size, after a certain sample size, a single-species vs multi-species estimate of cue rate and EDR should be similarly precise.
+Put differently, we expect that species with a smaller sample size will have precision of cue rate and EDR estimates improved with a multi-species model.
+Finally, we expect that a multi-species model will be able to provide predictions of cue rate and EDR for species with little to no data, but should be able to provide better/more precise estimates for data-sparse species compared to data-absent species.
+Additionally, we expect to see a higher precision of estimates for species where there are more "similar" species.
+That is, species with more closely-related species modelled in the multi-species model will have more precise estimates of cue rate, and species with more species with similar traits will have more precise estimates of EDR.
 
@@
 
+\par For both the Bayesian hierarchical QPAD Removal and Distance models described above, we compared the estimated cue rates and effective detection radii, respectively, to those produced within a single-species context in \citet{edwards_point_2023}.
+To do this, we modelled the difference as
+$$\Delta X \sim N(0,\sigma)$$
+$$\sigma \sim exp(1)$$
+
+where $\Delta X$ is calculated as the mean multi-species estimate of cue rate or EDR minus the mean single-species estimate of cue rate of EDR.
+If our predictions are true, we would expect to see a distribution of modelled differences centred about 0, indicating no real difference in estimates of cue rate or EDR between single-species and multi-species models, for species for which we have sufficient data.
+
+\par We were also interested in determining if a multi-species model improved the precision around the estimates of cue rate or EDR. 
+To do this, we modelled the standard deviations about each estimate of cue rate or EDR from both the multi-species and single-species models from \citet{edwards_point_2023}, and regressed them against the log sample size per species.
+That is
+$$SD \sim N(\alpha_M + \beta_M\times \log(N), \sigma)$$
+$$\alpha, \beta \sim N(0,1)$$
+$$\sigma \sim exp(1)$$
+
+where $\alpha$ is the model-specific intercept and $\beta$ is the model-specific slope.
+
+\par If our predictions are true, we would expect to see a higher value of $\alpha_1$ than for $\alpha_2$, corresponding to a higher baseline standard deviation in the single species models.
+We would also likely see a slightly smaller (i.e., more negative) value of $\beta_1$ than $\beta_2$ (which should also be negative), indicating that the standard deviation decreases with log sample size for both models, but should decrease at a slower rate with the multi-species model (which has a lower baseline standard deviation to begin with) than the single-species model.
+
+\subsection{Predicting Detection Probabilities}
+\par To predict estimates of cue rate or EDR from the multi-species models, we adopted two possible approaches.
+If the NA-POPS database contained some point-count data for the species of interest, we simply added these data into the model.
+This is different from \citet{edwards_point_2023, solymos_calibrating_2013, solymos_evaluating_2018}, where any species with less than 75 data points were removed from the analysis.
+For species that did not have any data in the NA-POPS database, we generated a row of traits data for the species and modelled the species just as a parameter.
+
+\section{Results}
+\includegraphics{../output/plots/removal_1vs1.png}
+
+\includegraphics{../output/plots/removal_sd_plot.png}
+
+\includegraphics{../output/plots/removal_predictions_plot.png}
+
+\includegraphics{../output/plots/distance_1vs1.png}
+
+\includegraphics{../output/plots/distance_sd_plot.png}
+
+\includegraphics{../output/plots/distance_predictions_plot.png}
+
 \bibliography{refs}

### 2023-10-06 src/02-prepare-removal-data.R
updated removal and distance stan data generation to include ordered list of species, and to contain just a vector of species rather than a dataframe

@@ -128,7 +128,7 @@ max_intervals_pred <- ncol(Y_pred)
 # a 1 corresponds with resident, a 2 corresponds with a migrant
 mig_strat_pred <- traits_pred$Migrant + 1
 
-removal_stan_data<- list(n_samples = n_samples_pred,
+removal_stan_data <- list(n_samples = n_samples_pred,
                          n_species = n_species_pred,
                          max_intervals = max_intervals_pred,
                          species = sp_pred_numeric$num,
@@ -136,9 +136,10 @@ removal_stan_data<- list(n_samples = n_samples_pred,
                          bands_per_sample = time_bands_per_sample_pred,
                          max_time = max_time_pred,
                          phylo_corr = corr_matrix_predict,
-                         sp_list = sp_list_pred,
+                         sp_list = sp_list_pred$Species,
                          n_mig_strat = max(mig_strat_pred),
-                         mig_strat = mig_strat_pred)
+                         mig_strat = mig_strat_pred,
+                         sp_all = species_pred_code)
 
 ####### Output ####################################
 save(removal_stan_data, file = "data/generated/removal_stan_data.rda")

### 2023-10-06 src/04-prepare-distance-data.R

@@ -3,7 +3,7 @@
 # Multi-species QPAD Detectability
 # 04-prepare-distance-data.R
 # Created August 2022
-# Last Updated May 2023
+# Last Updated October 2023
 
 ####### Import Libraries and External Files #######
 
@@ -163,14 +163,14 @@ distance_stan_data <- list(n_samples = n_samples_pred,
                            abund_per_band = abundance_per_band_pred,
                            bands_per_sample = dist_bands_per_sample_pred,
                            max_dist = max_dist_pred,
-                           phylo_corr = corr_matrix_predict,
-                           sp_list = sp_list_pred,
+                           sp_list = sp_list_pred$Species,
                            n_mig_strat = max(mig_strat_pred),
                            mig_strat = mig_strat_pred,
                            n_habitat = max(habitat_pred),
                            habitat = habitat_pred,
                            mass = mass_pred,
-                           pitch = pitch_pred)
+                           pitch = pitch_pred,
+                           sp_all = species_pred_code)
 
 ####### Output ####################################
 save(distance_stan_data, file = "data/generated/distance_stan_data.rda")

### 2023-10-06 src/functions/subset-distance-data.R

@@ -1,10 +1,9 @@
 subset_distance_data <- function(distance_stan_data = NULL,
                                  sps = NULL)
 {
-  orig_data_df <- data.frame(species = distance_stan_data$sp_list[,1],
+  orig_data_df <- data.frame(species = distance_stan_data$sp_list,
                              sp_n = distance_stan_data$species)
-  ### NOTE there are only 322 species in this list.
-  ### is n_species wrong (== 323)
+
   sp_df <- orig_data_df %>%
     group_by(species,sp_n) %>%
     summarise(n_counts = n())
@@ -37,6 +36,7 @@ subset_distance_data <- function(distance_stan_data = NULL,
 
   distance_stan_data2$n_species <- length(sps)
   distance_stan_data2$n_samples <- length(data_sel)
+  distance_stan_data2$sp_list <- distance_stan_data2$sp_list[which(distance_stan_data2$sp_list %in% sps)]
 
   return(distance_stan_data2)
 }

### 2023-10-06 src/functions/subset-distance-data.R
fixed bug in distance data subset

@@ -6,7 +6,7 @@ subset_distance_data <- function(distance_stan_data = NULL,
 
   sp_df <- orig_data_df %>%
     group_by(species,sp_n) %>%
-    summarise(n_counts = n())
+    dplyr::summarise(n_counts = n())
 
   sub_data <- data.frame(mig_strat = distance_stan_data$mig_strat,
                          habitat = distance_stan_data$habitat,

### 2023-10-06 src/generate-distance-cv-folds.R
WIP for distance cross validation, need to properly subset the data for each fold and run the model with generated quantities still

@@ -1,0 +1,62 @@
+####### Script Information ########################
+# Brandon P.M. Edwards
+# Multi-species QPAD Detectability
+# 07-generate-distance-cv-folds.R
+# Created October 2023
+# Last Updated October 2023
+
+####### Import Libraries and External Files #######
+
+library(dplyr)
+
+source("src/functions/subset-distance-data.R")
+
+####### Load Data #################################
+
+load("data/generated/distance_stan_data.rda")
+
+####### Set Constants #############################
+
+k <- 10
+
+####### Allocate CV Numbers #######################
+
+#' First we must drop all species which are just not going to be involved
+#' in the cross validation at all. This will be the species for which we are
+#' generated predictions
+pred_drops <- c("LCTH", "LEPC", "HASP", "SPOW", "KIWA", "BITH")
+dis_data <- subset_distance_data(distance_stan_data = distance_stan_data,
+                                 sps = setdiff(unique(distance_stan_data$sp_list),
+                                               pred_drops))
+rm(distance_stan_data) #clear up some mem
+
+# Create a dataframe of traits from the current distance data, including species name
+traits <- data.frame(Mig = dis_data$mig_strat,
+                     Habitat = dis_data$habitat,
+                     Mass = dis_data$mass,
+                     Pitch = dis_data$pitch,
+                     Species = unique(dis_data$sp_list))
+
+# Create an indicator for each Mig x Hab association
+traits$Mig_Hab <- paste0(traits$Mig, "-", traits$Habitat)
+counts_per_group <- data.frame(table(traits$Mig_Hab))
+
+traits$cv_fold <- NA
+
+for (i in 1:nrow(counts_per_group))
+{
+  group <- counts_per_group$Var1[i]
+  n_sp <- counts_per_group$Freq[i]
+  
+  cv_allocs <- rep(seq(1,k), times = ceiling(n_sp / k))
+  cv_allocs <- cv_allocs[1:n_sp]
+  cv_alloc_random <- sample(cv_allocs)
+  
+  traits[which(traits$Mig_Hab == group), "cv_fold"] <- cv_alloc_random
+}
+
+cv_folds <- traits[, c("Species", "cv_fold")]
+
+####### Output ####################################
+
+write.csv(cv_folds, file = "data/generated/distance_cv_folds.csv", row.names = FALSE)

### 2023-10-06 src/run-distance-cv.R

@@ -1,0 +1,94 @@
+####### Script Information ########################
+# Brandon P.M. Edwards
+# Multi-species QPAD Detectability
+# 07-run-distance-cv.R
+# Created October 2023
+# Last Updated October 2023
+
+####### Import Libraries and External Files #######
+
+library(cmdstanr)
+library(dplyr)
+
+source("src/functions/generate-distance-inits.R")
+source("src/functions/subset-distance-data.R")
+
+####### Load Data #################################
+
+load("data/generated/distance_stan_data.rda")
+cv_folds <- read.csv("data/generated/distance_cv_folds.csv")
+
+####### Set Constants #############################
+
+# Stan settings
+n_iter <- 2000
+n_warmup <- 1000
+n_chains <- 4
+refresh <- 10
+threads_per_chain <- 7
+
+####### Run Cross Validation ######################
+
+#' First we must drop all species which are just not going to be involved
+#' in the cross validation at all. This will be the species for which we are
+#' generated predictions
+pred_drops <- c("LCTH", "LEPC", "HASP", "SPOW", "KIWA", "BITH")
+dis_data <- subset_distance_data(distance_stan_data = distance_stan_data,
+                                 sps = setdiff(unique(distance_stan_data$sp_list),
+                                               pred_drops))
+rm(distance_stan_data)
+
+model_file <- cmdstan_model(stan_file = "models/distance_cp.stan",
+                            cpp_options = list(stan_threads = TRUE))
+
+for (i in 1:max(cv_folds$cv_fold))
+{
+  drops <- cv_folds[which(cv_folds$cv_fold == i), "Species"]
+  indices_to_drop <- which(dis_data$sp_list %in% drops)
+  
+}
+
+
+
+
+distance_stan_data$grainsize <- 1
+distance_stan_data$lambda <- 0
+
+distance_stan_data$sp_list <- NULL
+phylo_corr <- distance_stan_data$phylo_corr
+distance_stan_data$phylo_corr <- NULL
+
+# Scale the maximum distances for computational ease
+distance_stan_data$max_dist <- distance_stan_data$max_dist / 100
+
+# get rid of centre/scale attributes for modelling
+distance_stan_data$pitch <- distance_stan_data$pitch[,1]
+distance_stan_data$mass <- distance_stan_data$mass[,1]
+
+
+
+
+
+
+####### Run Model #################################
+
+inits <- generate_distance_inits(n_chains = n_chains,
[…]

### 2023-10-16 src/removal/removal-cv-analysis.R
WIP for analysis of removal model cross validation. Model to analyze difference seems odd, but looking at plot maybe there is no clear difference...

@@ -1,0 +1,124 @@
+####### Script Information ########################
+# Brandon P.M. Edwards
+# Multi-species QPAD Detectability
+# removal-cv-analysis.R
+# Created October 2023
+# Last Updated October 2023
+
+####### Import Libraries and External Files #######
+
+library(cmdstanr)
+library(bayesplot)
+library(ggplot2)
+library(ggpubr)
+theme_set(theme_pubclean())
+bayesplot::color_scheme_set("red")
+
+source("src/functions/rem-lik.R")
+
+####### Read Data #################################
+
+load("data/generated/removal_stan_data_cv.rda")
+cv_folds <- read.csv("data/generated/removal_cv_folds.csv")
+
+####### Main Code #################################
+
+#' Create dataframe that will connect species, data point,
+#' and lppd
+lppd <- data.frame(Species = cv_folds$Species,
+                   CV_Fold = cv_folds$cv_fold,
+                   SS_LPPD = NA,
+                   MS_LPPD = NA,
+                   Difference = NA)
+
+for (f in 1:max(cv_folds$cv_fold))
+{
+  # Extract dataframe of phi from single species and multi species model
+  ss_mod <- readRDS(paste0("output/model_runs/cv_removal/ss_fold_",
+                    f,
+                    ".RDS"))
+  ms_mod <- readRDS(paste0("output/model_runs/cv_removal/ms_fold_",
+                           f,
+                           ".RDS"))
+  
+  ss_phi_summary <- ss_mod$summary("log_phi")
+  ms_phi_summary <- ms_mod$summary("log_phi")
+  
+  # Subset the data to just the test set
+  data_indices <- which(lppd$CV_Fold == f)
+  abund <- removal_stan_data_cv$abund_per_band[data_indices, ]
+  bands <- removal_stan_data_cv$bands_per_sample[data_indices]
+  max_time <- removal_stan_data_cv$max_time[data_indices, ]
+  species <- removal_stan_data_cv$species[data_indices]
+  max_intervals <- removal_stan_data_cv$max_intervals
+
+  phi_ss <- exp(ss_phi_summary$mean[species])
+  phi_ms <- exp(ms_phi_summary$mean[species])
+  
+  # Check to see if we can drop any training columns (i.e. if bands per sample is smaller)
+  if (max(bands) < max_intervals)
+  {
+    max_intervals <- max(bands)
+    abund <- abund[, 1:max_intervals]
[…]

### 2023-10-17 src/removal/removal-cv-analysis.R
plotting some of the output from the removal cross validation

@@ -11,6 +11,8 @@ library(cmdstanr)
 library(bayesplot)
 library(ggplot2)
 library(ggpubr)
+library(ape)
+library(ggtree)
 theme_set(theme_pubclean())
 bayesplot::color_scheme_set("red")
 
@@ -20,6 +22,9 @@ source("src/functions/rem-lik.R")
 
 load("data/generated/removal_stan_data_cv.rda")
 cv_folds <- read.csv("data/generated/removal_cv_folds.csv")
+phylo_tree <- ape::read.nexus(file = "data/raw/all_species.nex")
+binomial <- read.csv("data/generated/binomial_names.csv")
+traits <- read.csv("data/raw/traits.csv")
 
 ####### Main Code #################################
 
@@ -82,43 +87,100 @@ for (f in 1:max(cv_folds$cv_fold))
 lppd$Difference <- lppd$MS_LPPD - lppd$SS_LPPD
 
 lppd_summary <- data.frame(Species = unique(lppd$Species),
-                           Median_LPPD = NA)
+                           Mean_Difference = NA,
+                           StdErr = NA,
+                           Model_Preference = NA)
 
 sp_sample_size <- data.frame(table(removal_stan_data_cv$sp_list$Species))
 names(sp_sample_size) <- c("Species", "N")
-
 lppd_summary <- dplyr::left_join(lppd_summary, sp_sample_size, by = "Species")
+
 for (i in 1:nrow(lppd_summary))
 {
   sp <- lppd_summary$Species[i]
-  lppd_summary$Median_LPPD[i] <- mean(lppd[which(lppd$Species == sp),
+  lppd_summary$Mean_Difference[i] <- mean(lppd[which(lppd$Species == sp),
                                              "Difference"])
+  lppd_summary$StdErr[i] <- sqrt(var(lppd[which(lppd$Species == sp),
+                                          "Difference"])) / lppd_summary$N[i]
+  
 }
 
-plot(lppd_summary$Median_LPPD ~ lppd_summary$N)
+lppd_summary[which((lppd_summary$Mean_Difference - lppd_summary$StdErr < 0) &
+                    (lppd_summary$Mean_Difference + lppd_summary$StdErr < 0)),
+             "Model_Preference"] <- "SS"
+
+lppd_summary[which((lppd_summary$Mean_Difference - lppd_summary$StdErr > 0) &
+                     (lppd_summary$Mean_Difference + lppd_summary$StdErr > 0)),
[…]

### 2023-10-17 article/article.tex
updated methods section of article to include cross validation approach

@@
 \subsection{Predictions}
-\par We predict that for both the Bayesian hierarchical QPAD removal model and Bayesian hierarchical QPAD distance model, estimates of cue rate and EDR, respectively, will change very little from those estimated from the original NA-POPS database \cite{edwards_point_2023}.
+\begin{enumerate}
+	\item\label{pred1} For both the Bayesian hierarchical QPAD removal model and Bayesian hierarchical QPAD distance model, estimates of cue rate and EDR, respectively, will change very little from those estimated from the original NA-POPS database \cite{edwards_point_2023}.
 	This is because the species that were modelled in \citet{edwards_point_2023} had a sufficient number of data points to stand on their own, and so adding in species-level effects should have little effect on estimates.
-We also predict that a multi-species model will increase the precision of the estimates of cue rate and EDR, but only up to a certain sample size. 
+	\item\label{pred2} A multi-species model will increase the precision of the estimates of cue rate and EDR, but only up to a certain sample size. 
 	This is because there are many species for which NA-POPS has thousands to even tens of thousands of data points; given precision generally increases with sample size, after a certain sample size, a single-species vs multi-species estimate of cue rate and EDR should be similarly precise.
 	Put differently, we expect that species with a smaller sample size will have precision of cue rate and EDR estimates improved with a multi-species model.
-Finally, we expect that a multi-species model will be able to provide predictions of cue rate and EDR for species with little to no data, but should be able to provide better/more precise estimates for data-sparse species compared to data-absent species.
-Additionally, we expect to see a higher precision of estimates for species where there are more "similar" species.
-That is, species with more closely-related species modelled in the multi-species model will have more precise estimates of cue rate, and species with more species with similar traits will have more precise estimates of EDR.
+	\item\label{pred3} A multi-species model will have greater predictive capacity than a single-species model, particularly for species which are data-sparse.
+\end{enumerate}
+
+\par With the predictions above, we then expect to be able to generate estimates of cue rate and EDR for species with very little data, thus relaxing previous sample size requirements; additionally, we expect to be able to predict cue rate and EDR for species with no data by making use of the shrinkage effect offered by a multi-species model.
 
@@
 
+\subsection{Single Species Models}
+
+\par To directly compare the multi-species models described above to its single-species equivalent, we built two Bayesian single-species models in Stan.
+The single species models used the same likelihood statements as above for the count data, except rather than cue rate and EDR being estimated with shared information, they are estimated independently from each other.
+That is, $\log \phi \sim N(0,1) \forall \phi_{s}$, and $\log \tau \sim N(0,1) \forall \tau_s$, thus making it equivalent to the Null model in \citet{edwards_point_2023}, just estimated in a Bayesian framework rather than maximum likelihood.
+
 \subsection{Model Assessment}
 
-\par For both the Bayesian hierarchical QPAD Removal and Distance models described above, we compared the estimated cue rates and effective detection radii, respectively, to those produced within a single-species context in \citet{edwards_point_2023}.
-To do this, we modelled the difference as
+\par To test prediction \ref{pred1}, we ran the Bayesian hierarchical QPAD Removal and Distance models, as well as the single-species QPAD Removal and Distance models described above, on the set of species for which there was sufficient data in the original NA-POPS analysis.
+In this case, we ran both removal models for 316 species of birds, and we ran both distance models for 317 species of birds (Tricoloured Blackbird was included in the distance models but not the removal models).
+We then modelled the difference in estimated cue rates and EDRs as
 $$\Delta X \sim N(0,\sigma)$$
@@
 where $\Delta X$ is calculated as the mean multi-species estimate of cue rate or EDR minus the mean single-species estimate of cue rate of EDR.
-If our predictions are true, we would expect to see a distribution of modelled differences centred about 0, indicating no real difference in estimates of cue rate or EDR between single-species and multi-species models, for species for which we have sufficient data.
+If prediction \ref{pred1} is true, we would expect to see a distribution of modelled differences centred about 0, indicating no real difference in estimates of cue rate or EDR between single-species and multi-species models, for species for which we have sufficient data.
 
-\par We were also interested in determining if a multi-species model improved the precision around the estimates of cue rate or EDR. 
-To do this, we modelled the standard deviations about each estimate of cue rate or EDR from both the multi-species and single-species models from \citet{edwards_point_2023}, and regressed them against the log sample size per species.
+\par To test prediction \ref{pred2}, we modelled the standard deviations about each estimate of cue rate or EDR from both the multi-species and single-species models, and regressed them against the log sample size per species.
 That is
@@
 
-\par If our predictions are true, we would expect to see a higher value of $\alpha_1$ than for $\alpha_2$, corresponding to a higher baseline standard deviation in the single species models.
+\par If prediction \ref{pred2} is true, we would expect to see a higher value of $\alpha_1$ than for $\alpha_2$, corresponding to a higher baseline standard deviation in the single species models.
 We would also likely see a slightly smaller (i.e., more negative) value of $\beta_1$ than $\beta_2$ (which should also be negative), indicating that the standard deviation decreases with log sample size for both models, but should decrease at a slower rate with the multi-species model (which has a lower baseline standard deviation to begin with) than the single-species model.
+
+\par To test prediction \ref{pred3}, we ran a k-fold cross-validation with 5 folds, stratified by species.
+That is, for each cross-validation run, we held out 20\% of the data for each species for both the multi-species and single-species models.
+We then estimated multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR for each cross-validation run, and used these parameter estimates to calculate the posterior predictive density of each of the held out data points.
+We then calculated the difference in posterior predictive density for removal and distance models, by subtracting the pointwise posterior predictive density of the single-species model from the pointwise posterior predictive density of the multi-species model.
+Finally, we took the mean of the difference in pointwise posterior predictive density for each species to create a species-specific mean pointwise predictive density.
+With this, positive values of this mean would indicate that the multi-species model was preferred, and negative values would indicate that the single-species model was preferred.
+If prediction \ref{pred3} is true, then we would expect to see majority positive values of the specific-specific mean pointwise predictive density.
 
@@
 For species that did not have any data in the NA-POPS database, we generated a row of traits data for the species and modelled the species just as a parameter.
+We then ran the multi-species models described above using these additional data points (and hence an updated phylogenetic correlation structure for the removal model) to generate estimates of cue rate and EDR for these new species.
 

### 2023-10-18 article/article.tex
added introduction material to article, and corrected some areas of the methods

@@
 
+\par Detectability is an important metric for many conservation problems, but often goes unaccounted for, or detectability data just does not exist \citep{bennett_how_2023}.
+In conservation management situations, detectability can be used as a metric of sensitivity to evaluate when to monitor a system for a species, or when to act on current information \citep{canessa_when_2015, bennett_when_2018}.
+When estimating population sizes of an organism, detectability often comes into play as a way to convert observed counts into densities or estimates of true abundance.
+As such, accurate and precise estimates of detectability are important both for making accurate conservation decisions, and estimating accurate population sizes.
+
+\par The wealth of bird data that has been collected and is available in North America means that estimates of detectability for birds are relatively easy to come by \citep{bennett_how_2023}.
+The NA-POPS project is an effort to estimate detectability functions that take into account variability in detection through environmental and temporal covariates, for as many of North America's landbird species as possible \citep{edwards_point_2023}.
+To date, the project has estimated availability (the probability a bird sings or gives a cue) and perceptibility (the probability an observer sees or hears a cueing bird) for 319 species of birds, with an additional 19 species having estimates of availability or perceptibility (but not necessarily both) estimated.
+This project built off previous efforts by the Boreal Avian Modelling project \citep{cumming_toward_2010} to systematically estimate detectability in landbirds \citep{solymos_calibrating_2013, solymos_evaluating_2018}.
+Both the NA-POPS project and previous estimates by Boreal Avian Modelling project use the QPAD methodology, which is a flexible method to estimate availability and perceptibility given a harmonized dataset \citep{barker_ecological_2015, solymos_calibrating_2013}.
+
+\par Despite the efforts of the NA-POPS project, the 338 species of birds considered in the modelling exercise still only account for about 75\% of Partners in Flight's 2016 Landbird Conservation Plan \citep{edwards_point_2023, rosenberg_partners_2016}.
+This is because of sample size limitations that are recommended for removal and distance modelling, in that a modeller should have at least 75 data points for reasonable estimates \citep{buckland_introduction_2001}.
+Many of the species that fell short of that threshold in the NA-POPS database are systematically rare or undersampled species, that are either low in population size to begin with, or occur in rare habitats.
+Some examples of species not modelled include Kirtland's Warbler (SCIENTIFIC NAME) and Bicknell's Thrush (SCIENTIFIC NAME), both of which have low population sizes, and LeConte's Thrasher (SCIENTIFIC NAME) and Harris's Sparrow (SCIENTIFIC NAME), both of which have rare habitat associations  (PARNTERS IN FLIGHT WATCHLIST CITATION).
+Of course, despite not having good estimates of detectability availabile for these rare and undersampled species, these are the groups of species that need these estimates the most as their conservation status often depend on accurate population size estimates, and estimates of trend in those population sizes.
+
+\par Detectability in birds has been shown to at least be partly driven by variation in traits.
+\cite{johnston_species_2014} showed in a selection of European landbirds that body mass and habitat association had an effect on the detection distance, which influences the perceptibility.
+In this case, a larger body size tended to be associated with larger detection distances, which makes sense considering larger body sizes are associated with louder and lower-pitched songs in birds \citep{bowman_adaptive_1979, fletcher_acoustics_1999, ryan_role_1985}.
+Habitat associations tends to influence the attenuation of sound, whereby sound diminishes quicker in a closed-forested habitat compared to a more open habitat \citep{waide_tropical_1988, yip_sound_2017}.
+A study by \cite{solymos_phylogeny_2018} expanded on this effort by also considering the effects of species traits and phylogeny on both detection distance and cue rate, the latter of which influences the availability of birds.
+They found that cue rate is most influenced by phylogeny, and, to a lesser extent, whether a bird was a long-distance migrant or a resident species.
+They also found that while the best model that explained detection distance did include phylogeny, the strength of the phylogenetic relationship was reduced to effectively 0 after accounting for traits such as body mass, song pitch, habitat association, and migratory status.
+Both these studies and others (see CITATIONSSSSS) provide useful evidence that detectability in birds can be reasonably predicted by some combinations of phylogeny and traits.
+
+\par One recommendation for future research that come out of \cite{solymos_phylogeny_2018} was to try to use these phylogenetic and trait relationships in detectability to then predict components of detectability (i.e., availability through cue rate, and perceptibility through detection distance) in other birds.
+Indeed, by making use of shrinkage factors and partial pooling, such as what can be done in a Bayesian hierarchical models (CITATION), and by accounting phylogenetic relationships and the relative "relatedness" of different species, such as what can be done with Gaussian processes \citep{bernardo_regression_1998, mcelreath_continous_2020}, we can leverage the information provided by species with many data points to use to inform species with much fewer (included potentially zero) data points.
+Thus, the goal of this paper is to create a multi-species model of detectability that incorporates species traits and phylogeny, to then predict estimates of detectability for rare and undersampled species of North American landbirds.
+To do so, we developed a Bayesian hierarchical version of the QPAD Removal model initially derived in \cite{solymos_calibrating_2013}, that allows for information exchange between species based on phylogenetic distance and binned migratory status.
+Phylogenetic distance was coded as a correlation structure and used as a Gaussian random field in this model, and migratory status was coded similar to \cite{solymos_phylogeny_2018} where a species was considered either "migrant" or "resident.
+We also developed a Bayesian hierarchical version of the QPAD Distance model initially derived in \cite{solymos_calibrating_2013}, that allows for information exchange between species based only on the traits that came out as being significant in \cite{solymos_phylogeny_2018}.
+These traits include migratory status (coded the same as removal models), habitat association (coded as either "open" or "closed"), song pitch, and body mass.
+We also developed a Bayesian single species version of the QPAD Removal and Distance models (i.e., as opposed to the maximum likelihood models available in the \texttt{detect} R package; \cite{solymos_detect_2020}) to allow for direct comparison through cross validation.
+Finally, we ran these models with the full set of data points used in \cite{edwards_point_2023}, plus additional data for seven species that did not meet the initial sample size thresholds for NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird, and Spotted Owl.
+All seven of these bird species are listed on Partner's in Flight's watchlist for various threats (CITATION).
+
 \subsection{Predictions}
@@
 We start by introducing the Bayesian hierarchical QPAD removal and distance models, and then describe the input.
-
-\subsection{Data Collection}
-\subsubsection{Point Count Data}
-\par We used the same standardized database of point count data that was put together through the NA-POPS project \cite{edwards_point_2023}. 
-This consists of $712138$ point counts conducted across $292$ projects in Canada and the United States, $422514$ of which used removal sampling and $522820$ of which used distance sampling.
-
-\subsubsection{Phylogenetic Trees}
-\par We accessed 2000 pseudo-posterior trees from \citet{jetz_global_2012} that contains phylogeny for all species that contained sufficient data for removal modelling ($>75$ observations as per \cite{edwards_point_2023,solymos_evaluating_2018,buckland_introduction_2001}).
-We also requested that the phylogenies include the following bird species with insufficient data from NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird, and Spotted Owl.
-All seven of these bird species are listed on Partner's in Flight's watchlist for various threats (CITATION). 
-This resulted in 2000 pseudo-posterior trees that contained 323 species.
-
-\par We generated a phylogenetic correlation matrix using the same methods outlined in the Appendix of \citet{solymos_phylogeny_2018} that generate distances between species, and further processed using the R package "ape" \citet{paradis_ape_2019}.
-
-\subsubsection{Species Traits}
-\par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts of all the birds to model in this study (CITATION). 
-Body mass for each bird species was obtained form the Elton traits dataset (CITATION). 
-Where body mass was not available through the Elton traits dataset, the Birds of the World account was used instead.
+Finally, we describe the experiments conducted to assess model performance and model comparison, which then allow us to make predictions for the seven undersampled species of interest.
 
@@
 
-\par For any point count that employs removal and/or distance sampling, let $Y_{sijk}$ be the observed count of species $s$ during sampling event $i$, occuring in time band $j \in [1,J]$ and/or distance band $k \in [1,K]$.
-That is, if we have a point count where removal sampling or distance sampling (but not both) is not used, then we have that either $J = 1$ or $K = 1$, respectively. 
-The following two subsections detail the Bayesian hierarchical QPAD Removal and Distance Models.
+\par For any point count that employs removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002} and/or distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}, let $Y_{sijk}$ be the observed count of species $s$ during sampling event $i$, occuring in time band $j \in [1,J]$ and/or distance band $k \in [1,K]$.
+That is, if we have a point count where removal sampling or distance sampling (but not both) is not used, then we have that either $J = 1$ or $K = 1$, respectively; whereas, if we have a point count where both removal sampling and distance sampling is used, then we have that both $J,K \geq 2$.
+The case where $J = K = 1$ is where neither removal sampling nor distance sampling is used, and so these data points are not appropriate for the modelling techniques described here.
+
+\par The following two subsections detail the Bayesian hierarchical QPAD Removal and Distance Models.
 
@@
 \par For removal modelling, we are not interested in distance to the bird, and so we will sum counts from each sampling event $i$ over all distance bands. 
-Thus, we will be considering counts $Y_{sij.}$. 
+Thus, we will be considering counts $Y_{sij.} = \sum_{k=1}^{K}$. 
 From \citet{solymos_calibrating_2013}, we model $Y_{sij.}$ as
@@
 \par For sample $i$, $\Pi_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $J$ time bands; in essense, it is the mixing parameter.
-Let $t_ij$ be the maximum time for time band $j$ during sampling event $i$, and let $phi_s$ be the unknown cue rate for species $s$.
+Let $t_{ij}$ be the maximum time for time band $j$ during sampling event $i$, and let $\phi_s$ be the unknown cue rate for species $s$.
 We then have the following calculation of each component $\pi_{sij}$ of the mixing parameter $\Pi_{si.}$:
@@
 
-\par We are most interested in estimating $phi_s$, the cue rate for species $s$. 
-To allow for the partial pooling of estimates for each species to be based on a phylogenic relationship, we can use Gaussian processes, which are effectively allow for "continuous" factors. 
+\par We are most interested in estimating $\phi_s$, the cue rate for species $s$. 
+To allow for the partial pooling of estimates for each species to be based on a phylogenic relationship, we can use Gaussian processes, which effectively allow for "continuous" factors \citep{bernardo_regression_1998, mcelreath_continous_2020}.
 Let $\Phi$ be the vector of cue rates (of length $s$). 
@@
 \par The covariance function $\Sigma$ (denoted with subscript $[s \times s]$ to show dimensionality) determines the strength of the partial pooling, and can be any function that contains a measure of distance.
-In this case, we created a function based on phylogenetic distance $D_{[s \times s]}$, i.e.,
+In this case, we will create a function based on phylogenetic distance $D_{[s \times s]}$, i.e.,
 
@@
 \par For distance modelling, we are not interested in the time bin that the bird was recorded in, and so we will sum counts from each sampling event $i$ over all time bands. 
-Thus, we will be considering counts $Y_{si.k}$. 
+Thus, we will be considering counts $Y_{si.k} = \sum_{j=1}^{J}$. 
 From \citet{solymos_calibrating_2013}, we model $Y_{si.k}$ as
@@
 \par For sample $i$, $\Pi_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $K$ distance bins; in essense, it is the mixing parameter.
-Let $r_ik$ be the maximum distance for distance bin $k$ during sampling event $i$, and let $tau_s$ be the unknown effective detection radius for species $s$. 
+Let $r_{ik}$ be the maximum distance for distance bin $k$ during sampling event $i$, and let $\tau_s$ be the unknown effective detection radius for species $s$. 
 We then have the following calculation of each component $\pi_{sik}$ of the mixing parameter $\Pi_{si.}$:
@@
 		\dfrac{f(r_{i,k}, \tau_s) - f(r_{i,k-1}, \tau_s)}{f(r_{i,K}, \tau_s)} & \text{for } k > 1 \\
-		1 - \sum_{n = 2}^{J} \pi_{sin} & \text{for } k = 1
+		1 - \sum_{n = 2}^{K} \pi_{sin} & \text{for } k = 1
 	\end{cases}
@@
 
+\subsection{Data Collection}
+\subsubsection{Point Count Data}
+\par We used the same standardized database of point count data that was put together through the NA-POPS project \cite{edwards_point_2023}. 
+This consists of $712138$ point counts conducted across $292$ projects in Canada and the United States, $422514$ of which used removal sampling and $522820$ of which used distance sampling.
+
+\subsubsection{Phylogenetic Trees}
+\par We accessed 2000 pseudo-posterior trees from \citet{jetz_global_2012} that contains phylogeny for all species that contained sufficient data for removal modelling ($>75$ observations as per \cite{edwards_point_2023,solymos_evaluating_2018,buckland_introduction_2001}).
+We also requested that the phylogenies include the following bird species with insufficient data from NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird, and Spotted Owl.
+All seven of these bird species are listed on Partner's in Flight's watchlist for various threats (CITATION). 
+This resulted in 2000 pseudo-posterior trees that contained 323 species.
+
+\par We generated a phylogenetic correlation matrix $D$ using the same methods outlined in the Appendix of \citet{solymos_phylogeny_2018} that generate distances between species, and further processed using the R package "ape" \citet{paradis_ape_2019}.
+
+\subsubsection{Species Traits}
+\par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts of all the birds to model in this study (CITATION). 
+Body mass for each bird species was obtained form the Elton traits dataset (CITATION). 
+Where body mass was not available through the Elton traits dataset, the Birds of the World account was used instead.
+
 \subsection{Model Assessment}
@@
 where $\alpha$ is the model-specific intercept and $\beta$ is the model-specific slope.
+In both cases, $M=1$ corresponds to the single species model and $M=2$ corresponds to the multi-species model.
 
@@
 
+\subsection{Analysis}
+\par All analyses were performed in R version TO DO: CHECK (CITATION).
+All Bayesian models were coded in the Stan Probabilistic Language \cite{stan_development_team_stan_2019}, and models were run using the \texttt{cmdstanr} R package (CITATION) and visualized using the \texttt{bayesplot} R package (CITATION).
+Any manipulation of phylogenetic trees, including finding a consensus tree for visualization and generating variance-covariance matrices based on the 2000 bootstrapped trees, was done using the \texttt{ape} R package.
+
 \section{Results}

### 2023-10-18 article/article.tex
other small article updates

@@
 \usepackage{setspace}
-\onehalfspacing
+\doublespacing
 
@@
 \par All analyses were performed in R version TO DO: CHECK (CITATION).
+Manipulation of phylogenetic trees, including finding a consensus tree for visualization and generating variance-covariance matrices based on the 2000 bootstrapped trees, was done using the \texttt{ape} R package.
 All Bayesian models were coded in the Stan Probabilistic Language \cite{stan_development_team_stan_2019}, and models were run using the \texttt{cmdstanr} R package (CITATION) and visualized using the \texttt{bayesplot} R package (CITATION).
-Any manipulation of phylogenetic trees, including finding a consensus tree for visualization and generating variance-covariance matrices based on the 2000 bootstrapped trees, was done using the \texttt{ape} R package.
+Each of the single-species models and multi-species models running both the initial comparison set of species and the final predictive set of species, as well as the models created to assess performance, were run using 1000 warmup iterations and 2000 sampling iterations, each on 4 chains for a total of 8000 draws per model run.
+Cross-validation models were run with 250 warmup iterations and 500 sampling iterations, each on 4 chains for a total of 2000 draws per model run.
+All of the models made use of Stan's reduce\_sum functionality, which allows for within-chain parallelization as well as the usual across-chain parallelization (CITATION).
+This was particularly useful in our case as the multi-species removal model contained over 3.5 million data points and the multi-species distance model contained over 4.5 million data points, and so computational expense was a large concern.
+All models were run on one of two multi-core processing servers, one running Ubuntu 20.04.4 LTS and one running Ubuntu 20.04.6 LTS.
 

### 2023-10-18 src/scripts/distance/distance-multi-vs-single.R
move all distance, removal, and data prep scripts to subdirectory scripts

@@ -1,0 +1,88 @@
+####### Script Information ########################
+# Brandon P.M. Edwards
+# Multi-species QPAD Detectability
+# 07-run-distance-cv.R
+# Created October 2023
+# Last Updated October 2023
+
+####### Import Libraries and External Files #######
+
+library(cmdstanr)
+library(dplyr)
+
+source("src/functions/generate-distance-inits.R")
+source("src/functions/subset-distance-data.R")
+
+####### Load Data #################################
+
+load("data/generated/distance_stan_data.rda")
+cv_folds <- read.csv("data/generated/distance_cv_folds.csv")
+
+####### Set Constants #############################
+
+# Stan settings
+n_iter <- 2000
+n_warmup <- 1000
+n_chains <- 4
+refresh <- 10
+threads_per_chain <- 7
+
+####### Data Wrangling ############################
+
+#' First we must drop all species which are just not going to be involved
+#' in the cross validation at all. This will be the species for which we are
+#' generated predictions
+pred_drops <- c("LCTH", "LEPC", "HASP", "SPOW", "KIWA", "BITH")
+dis_data <- subset_distance_data(distance_stan_data = distance_stan_data,
+                                 sps = setdiff(unique(distance_stan_data$sp_list),
+                                               pred_drops))
+rm(distance_stan_data)
+
+ms_model <- cmdstan_model(stan_file = "models/distance_cp.stan",
+                          cpp_options = list(stan_threads = TRUE))
+
+ss_model <- cmdstan_model(stan_file = "models/distance_single.stan",
+                          cpp_options = list(stan_threads = TRUE))
+
+####### Run Full Models ###########################
+
+inits <- generate_distance_inits(n_chains = n_chains,
+                                 sp_list = setdiff(as.vector(dis_data$sp_all), pred_drops),
+                                 napops_skip = NULL,
+                                 param = "cp")
+
+# Multi species model
+stan_run_ms <- ms_model$sample(
+  data = dis_data,
+  iter_warmup = n_warmup,
+  iter_sampling = n_iter,
+  chains = n_chains,
+  parallel_chains = n_chains,
+  refresh = refresh,
+  threads_per_chain = threads_per_chain,
+  init = inits
+)
+stan_run_ms$save_object(file = paste0("output/model_runs/distance_ms.RDS"))
+
+# Single species model
[…]

### 2023-10-18 src/scripts/distance/3-distance-prior-check.R
updated names and ordering of scripts in distance subdirectory

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 05-distance-prior-check.R
+# 3-distance-prior-check.R
 # Created March 2023
 # Last Updated May 2023
 

### 2023-10-18 src/scripts/distance/4-distance-ms-vs-ss.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 07-run-distance-cv.R
+# 4-distance-ms-vs-ss.R
 # Created October 2023
 # Last Updated October 2023
 

### 2023-10-18 src/scripts/distance/6-distance-predictions.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 09-distance-model.R
+# 6-distance-predictions.R
 # Created October 2022
 # Last Updated October 2023
 

### 2023-10-18 src/scripts/distance/7-distance-cv-analysis.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 11-distance-cv-analysis.R
+# 7-distance-cv-analysis.R
 # Created October 2023
 # Last Updated October 2023
 ####### Import Libraries and External Files #######

### 2023-10-18 src/scripts/distance/8-distance-comparison-analysis.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 11-distance-analysis.R
+# 8-distance-comparison-analysis.R
 # Created May 2022
 # Last Updated October 2023
 ####### Import Libraries and External Files #######

### 2023-10-18 src/scripts/removal/5-removal-cv.R
updating naming and ordering of removal scripts

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 07-run-removal-cv.R
+# 5-removal-cv.R
 # Created October 2023
 # Last Updated October 2023
 

### 2023-10-18 src/scripts/removal/6-removal-predictions.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 07-removal-model.R
+# 6-removal-predictions.R
 # Created December 2022
 # Last Updated October 2023
 

### 2023-10-18 src/scripts/removal/7-removal-cv-analysis.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# removal-cv-analysis.R
+# 7-removal-cv-analysis.R
 # Created October 2023
 # Last Updated October 2023
 

### 2023-10-18 src/scripts/removal/8-removal-comparison-analysis.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 10-removal-analysis.R
+# 8-removal-comparison-analysis.R
 # Created April 2022
 # Last Updated October 2023
 

### 2023-10-18 src/scripts/distance/4-distance-ms-vs-ss.R
scripts for removal and distance multi species vs single species

@@ -16,7 +16,6 @@ source("src/functions/subset-distance-data.R")
 ####### Load Data #################################
 
 load("data/generated/distance_stan_data.rda")
-cv_folds <- read.csv("data/generated/distance_cv_folds.csv")
 
 ####### Set Constants #############################
 
@@ -83,6 +82,4 @@ stan_run_ss <- ss_model$sample(
   threads_per_chain = threads_per_chain,
   init = inits
 )
-stan_run_ss$save_object(file = paste0("output/model_runs/cv_distance/ss_fold_",
-                                      i,
-                                      ".RDS"))
+stan_run_ss$save_object(file = paste0("output/model_runs/distance_ss.RDS"))

### 2023-10-18 src/scripts/removal/4-removal-ms-vs-ss.R

@@ -1,0 +1,64 @@
+####### Script Information ########################
+# Brandon P.M. Edwards
+# Multi-species QPAD Detectability
+# 4-removal-ms-vs-ss.R
+# Created October 2023
+# Last Updated October 2023
+
+####### Import Libraries and External Files #######
+
+library(cmdstanr)
+library(dplyr)
+
+####### Load Data #################################
+
+load("data/generated/removal_stan_data_cv.rda")
+
+####### Set Constants #############################
+
+# Stan settings
+n_iter <- 2000
+n_warmup <- 1000
+n_chains <- 4
+refresh <- 10
+threads_per_chain <- 7
+
+####### Data Wrangling ############################
+
+ms_model <- cmdstan_model(stan_file = "models/removal.stan",
+                          cpp_options = list(stan_threads = TRUE))
+
+ss_model <- cmdstan_model(stan_file = "models/removal_single.stan",
+                          cpp_options = list(stan_threads = TRUE))
+
+####### Run Full Models ###########################
+
+# Multi species model
+stan_run_ms <- ms_model$sample(
+  data = removal_stan_data_cv,
+  iter_warmup = n_warmup,
+  iter_sampling = n_iter,
+  chains = n_chains,
+  parallel_chains = n_chains,
+  refresh = refresh,
+  threads_per_chain = threads_per_chain,
+  init = inits
+)
+stan_run_ms$save_object(file = paste0("output/model_runs/removal_ms.RDS"))
+
+# Single species model
+removal_stan_data_cv$phylo_corr <- NULL
+removal_stan_data_cv$m_mig_strat <- NULL
+removal_stan_data_cv$mig_strat <- NULL
+
+stan_run_ss <- ss_model$sample(
+  data = dis_data2,
+  iter_warmup = n_warmup,
+  iter_sampling = n_iter,
+  chains = n_chains,
+  parallel_chains = n_chains,
+  refresh = refresh,
+  threads_per_chain = threads_per_chain,
+  init = inits
+)
+stan_run_ss$save_object(file = paste0("output/model_runs/removal_ss.RDS"))

### 2023-10-18 src/scripts/removal/7-removal-cv-analysis.R
fixed small typo

@@ -131,7 +131,7 @@ preference <- data.frame(table(lppd_summary$Model_Preference))
 
 lppd_summary <- merge(lppd_summary, binomial[,c("Code", "Scientific_BT")],
                       by.x = "Species", by.y = "Code")
-lppd_summary$Scientific_BT <- gsub(" ", "_", lppd_summary)
+lppd_summary$Scientific_BT <- gsub(" ", "_", lppd_summary$Scientific_BT)
 lppd_summary <- merge(lppd_summary, traits[, c("Code", "Migrant")],
                       by.x = "Species", by.y = "Code")
 tree <- ape::consensus(phylo_tree)

### 2023-10-18 src/functions/dis-lik.R
allow tau to be a vector of same length as abund and bands

@@ -13,12 +13,12 @@ dis_lik <- function(abund = NULL,
     {
       if (k > 1)
       {
-        Pi[k] <- ((1 - exp(-(max_dist[i,k]^2 / (tau)^2))) - 
-                    (1 - exp(-(max_dist[i,k - 1]^2 / (tau)^2)))) / 
-          (1 - exp(-(max_dist[i,bands[i]]^2 / (tau)^2)))
+        Pi[k] <- ((1 - exp(-(max_dist[i,k]^2 / (tau[i])^2))) - 
+                    (1 - exp(-(max_dist[i,k - 1]^2 / (tau[i])^2)))) / 
+          (1 - exp(-(max_dist[i,bands[i]]^2 / (tau[i])^2)))
       }else{
-        Pi[k] = (1 - exp(-(max_dist[i,k]^2 / (tau)^2))) /
-          (1 - exp(-(max_dist[i,bands[i]]^2 / (tau)^2)))
+        Pi[k] = (1 - exp(-(max_dist[i,k]^2 / (tau[i])^2))) /
+          (1 - exp(-(max_dist[i,bands[i]]^2 / (tau[i])^2)))
       }
     }
     

### 2023-10-18 src/scripts/distance/7-distance-cv-analysis.R
update cross validation post hoc analysis for distance model to match current cross validation methods

@@ -21,6 +21,7 @@ source("src/functions/dis-lik.R")
 
 cv_folds <- read.csv("data/generated/distance_cv_folds.csv")
 load("data/generated/distance_stan_data.rda")
+traits <- read.csv("data/raw/traits.csv")
 
 ####### Data Wrangling ############################
 
@@ -33,58 +34,120 @@ dis_data <- subset_distance_data(distance_stan_data = distance_stan_data,
                                                pred_drops))
 rm(distance_stan_data)
 
-cv_folds$Index <- seq(1, nrow(cv_folds))
-cv_folds$EDR <- NA
+####### Main Code #################################
 
-for (i in 1:max(cv_folds$cv_fold))
+#' Create dataframe that will connect species, data point,
+#' and lppd
+lppd <- data.frame(Species = cv_folds$Species,
+                   CV_Fold = cv_folds$cv_fold,
+                   SS_LPPD = NA,
+                   MS_LPPD = NA,
+                   Difference = NA)
+
+for (f in 1:max(cv_folds$cv_fold))
 {
-  fold <- readRDS(paste0("output/model_runs/cv_distance/fold_",
-                         i,
-                         ".RDS"))
+  # Extract dataframe of tau from single species and multi species model
+  ss_mod <- readRDS(paste0("output/model_runs/cv_distance/ss_fold_",
+                           f,
+                           ".RDS"))
+  ms_mod <- readRDS(paste0("output/model_runs/cv_distance/ms_fold_",
+                           f,
+                           ".RDS"))
   
-  # Add the mean estimated EDR to the dataframe above
-  log_tau_df <- fold$summary("log_tau")
-  fold_indices <- which(cv_folds$cv_fold == i)
-  cv_folds[fold_indices, "EDR"] <- exp(log_tau_df[fold_indices, "mean"]) 
+  ss_tau_summary <- ss_mod$summary("log_tau")
+  ms_tau_summary <- ms_mod$summary("log_tau")
   
-  # Now calculate the median LPMF for each species
-  sp_fold <- cv_folds[which(cv_folds$cv_fold %in% i), c("Species", "Index")]
+  # Subset the data to just the test set
+  data_indices <- which(lppd$CV_Fold == f)
+  abund <- dis_data$abund_per_band[data_indices, ]
+  bands <- dis_data$bands_per_sample[data_indices]
+  max_dist <- dis_data$max_dist[data_indices, ] / 100
+  species <- dis_data$species[data_indices]
[…]

### 2023-10-19 article/article.tex
wrote rough preliminary results section

@@
 
-\par To directly compare the multi-species models described above to its single-species equivalent, we built two Bayesian single-species models in Stan.
+\par To directly compare the multi-species models described above to its single-species equivalent in a cross-validation (see below), we built two Bayesian single-species models in Stan.
 The single species models used the same likelihood statements as above for the count data, except rather than cue rate and EDR being estimated with shared information, they are estimated independently from each other.
@@
 
-\par To test prediction \ref{pred1}, we ran the Bayesian hierarchical QPAD Removal and Distance models, as well as the single-species QPAD Removal and Distance models described above, on the set of species for which there was sufficient data in the original NA-POPS analysis.
-In this case, we ran both removal models for 316 species of birds, and we ran both distance models for 317 species of birds (Tricoloured Blackbird was included in the distance models but not the removal models).
+\par To test prediction \ref{pred1}, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data in the original NA-POPS analysis.
+In this case, we ran both removal model for 316 species of birds, and we ran the distance model for 317 species of birds (Tricoloured Blackbird was included in the distance models but not the removal models).
+We accessed the single-species estimates of cue rate and EDR from the NA-POPS Null models using the \texttt{napops} R package (CITATION).
 We then modelled the difference in estimated cue rates and EDRs as
@@
 \par To test prediction \ref{pred3}, we ran a k-fold cross-validation with 5 folds, stratified by species.
-That is, for each cross-validation run, we held out 20\% of the data for each species for both the multi-species and single-species models.
+That is, for each cross-validation run, we held out 20\% of the data for each species for both the multi-species and Bayesian single-species models.
 We then estimated multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR for each cross-validation run, and used these parameter estimates to calculate the posterior predictive density of each of the held out data points.
@@
 \section{Results}
+\subsection{Removal Model Comparisons}
+
+\par We found very little evidence for an overall change in cue rate for a given species modelled by a single-species model vs a multi-species model (Figure \ref{fig:removal_1vs1}A).
+Seven out of the 316 species considered for removal modelling had a percent change in cue rate of more than 20\%.
+All seven of these species were among the lowest cue rates in both the single-species and multi-species context (Figure \ref{fig:removal_1vs1}B).
+
+\begin{figure}[h!]
 	\includegraphics{../output/plots/removal_1vs1.png}
+	\caption{TO DO}
+	\label{fig:removal_1vs1}
+\end{figure}
 
+\par We found much less support for a difference in cue rate between resident and migrant species than was found in \citet{solymos_evaluating_2018}. 
+Mean overall log cue rate in the multi-species model was TO DO; both estimated effects for resident and migrant species were effectively 0 (Figure \ref{fig:removal_params}).
+
+\begin{figure}[h!]
+	\includegraphics{../output/plots/removal_other_params.png}
+	\caption{TODO}
+	\label{fig:removal_params}
+\end{figure}
+
+\par We found some evidence that a multi-species removal model improved the precision of cue rate estimates compared to estimates from a single-species model.
+The intercept of standard deviations of cue rates estimated in a single species model was greater than the intercept of standard deviations of cue rates estimated in a multi-species model, with very little overlap in their 95\% credible intervals (Figure \ref{fig:removal_sd}A).
+This implies that the baseline standard deviation of cue rates is higher in a single species model than a multi species model for very low sample sizes.
+However, the slope of standard deviations of cue rates estimated in a single species model was farther from 0 (i.e., more negative) than the slope of standard deviations of cue rates estimated in a multi-species model, with very little overlap in their 95\% credible intervals (Figure \ref{fig:removal_sd}B).
+This implies that the decrease in standard deviation as sample size increases is lesser in a multi-species model than a single-species model.
+This indicates that after a certain sample size, it is likely that the difference in standard deviation between a single species and multi species model would be negligible.
+
+\begin{figure}[h!]
 	\includegraphics{../output/plots/removal_sd_plot.png}
+	\caption{TODO}
+	\label{fig:removal_sd}
+\end{figure}
 
+\par The cross validation revealed that single species model actually had better predictive capacity for more species than the multi-species model, but the difference in predictive capacity was overall very small.
+Mean overall difference in posterior predictive density was $-2.21 \times 10^{-5}$.
+Single-species models tended to have better predicitve capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:removal_cv}).
+There did not appear to be any patterns in model selection among phylogeny or traits (see Appendix SOMETHING for full phylogenetic tree).
+Six out of the 7 species that had a percent change in cue rate of more than 20\% had the multi-species model chosen as the better predictive model.
+Common Redpoll was the only species that had the single-species model chosen (but please see Discussion about this particular species).
+
+\begin{figure}[h!]
+	\includegraphics{../output/plots/cv_removal_diff_vs_n.png}
+	\caption{TODO}
+	\label{fig:removal_cv}
+\end{figure}
+
+\subsection{Distance Model Comparisons}
+
+\par We found some evidence for a change in modelled EDR between single species and multi-species distance models, in that the multi-species model tended to produce slightly smaller estimates of EDR (Figure \ref{fig:distance_1vs1}A).
+Still, the majority of the species had very little change; six out of the 317 species had a percent change in EDR of more than 20\% (Figure \ref{fig:distance_1vs1}B).
+
+\begin{figure}[h!]
+	\includegraphics{../output/plots/distance_1vs1.png}
+	\caption{TO DO}
+	\label{fig:distance_1vs1}
+\end{figure}
+
+\par We found little support for migration strategy having any effect on EDR.
+There is some evidence that a resident species may have a slightly higher EDR than a migrant species, but the 95\% credible intervals of both effects overlapped with the 95\% credible interval of the overall estimated intercept (Figure \ref{fig:distance_params}A).
+We did find similar support for change in EDR by body size and pitch, in that body size tended to have an increasing effect on EDR, and song pitch had a decreasing effect on EDR (Figure \ref{fig:distance_params}B).
+
+\begin{figure}[h!]
+	\includegraphics{../output/plots/distance_other_params.png}
+	\caption{TODO}
+	\label{fig:distance_params}
+\end{figure}
+
+\par We found little evidence that a multi-species model improves the precision of EDR compared to that of a single-species model.
+The 95\% credible intervals for both estimated slopes and intercepts overlapped with each other, indicating that the standard deviations are effectively the same for a single-species and multi-species estimate as sample size increases (Figure \ref{fig:distance_sd}).
+
+\begin{figure}[h!]
+	\includegraphics{../output/plots/distance_sd_plot.png}
+	\caption{TODO}
+	\label{fig:distance_sd}
+\end{figure}
+
+\par The cross validation revealed that the multi-species model tended to have slightly more support than the single-species model in terms of predicting EDR.
+However, similar to the removal model cross-validation results, the overall difference in posterior predictive density was very small, with a mean difference of $1.85 \times 10^{-4}$.
+Multi-species models tended to have better predicitve capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:distance_cv}).
+Interestingly, only two of the six species with a greater than 20\% change in EDR had the multi-species model chosen as the better predictive model (NOTE this could change as new results come in).
+TO DO: Any patterns among traits?
+
+\begin{figure}[h!]
+	\includegraphics{../output/plots/cv_distance_diff_vs_n.png}
+	\caption{TODO}
+	\label{fig:distance_cv}
+\end{figure}
+
+\subsection{Predictions of Cue Rate and EDR for Undersampled Species}
+
+\par We used the multi-species removal model and distance model to generate predictions of cue rate and EDR, respectively, for 7 species of undersampled birds from the original NA-POPS analysis (6 for distance modelling because Tricoloured Blackbird had sufficient distance sampling data).
+
+\par The multi-species removal model produced reasonable estimates of cue rates for all seven species of bird considered here (Figure \ref{fig:removal_predictions}).
+Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush and Kirtland's Warbler) had reasonable precision associated with the estimates.
+One exception to this was LeConte's Thrasher, where the precision was rather low.
+The species for which we had no data (i.e., Bicknell's Thrush and Kirtland's Warbler) had estimates of cue rate similar to that of their closely-related species, demonstrating the shrinkage effect of the multi-species model.
+
+\begin{figure}[h!]
 	\includegraphics{../output/plots/removal_predictions_plot.png}
+	\caption{TODO}
+	\label{fig:removal_predictions}
+\end{figure}
 
-\includegraphics{../output/plots/distance_1vs1.png}
+\par The multi-species distance model also produced reasonable estimates of EDR for all six species of landbird considered here (Figure \ref{fig:distance_predictions}).
+Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush) had fairly precise estimates, except for LeConte's Thrasher and Lesser Prairie Chicken where the estimated EDR spanned greater than 50 m in either direction.
+The species for which we had no data (i.e., Bicknell's Thrush) had an estimate of EDR similar to what would be expected by a bird of its size, song pitch, and habitat association, demonstrating the shrinkage effect of the multi-species model.
 
-\includegraphics{../output/plots/distance_sd_plot.png}
+\begin{figure}[h!]
+	\includegraphics{../output/plots/distance_predictions_plot.png}
+	\caption{TODO}
+	\label{fig:distance_predictions}
+\end{figure}
 
-\includegraphics{../output/plots/distance_predictions_plot.png}
+\section{Discussion}
+
+* results somewhat underwhelming for cross validation, but both models showed some support that a multi species model can improve precision of estimates (removal model in particular)
+
+*
+
+* migration strategy mattering less
+
+* talk about Common Redpoll removal modelling
+
+* talk about Bicknell's Thrush removal and distance modelling
 

### 2023-10-20 article/article.tex
wrote discussion for removal modelling

@@
 
-* results somewhat underwhelming for cross validation, but both models showed some support that a multi species model can improve precision of estimates (removal model in particular)
+\par In this paper, we developed a multi-species model of detectability by expanding upon the QPAD methodology developed by \citet{solymos_calibrating_2013}.
+We applied the multi-species QPAD removal model and multi-species QPAD distance model to the large dataset of point counts collected by the NA-POPS project \citep{edwards_point_2023}, and compared estimates of cue rate and EDR from a multi-species context to estimates of cue rate and EDR from the previous single species context.
+Given the results of these comparisons, we sought to predict these components of detectability for seven species of North American landbird that are rare and undersampled.
 
-*
+\subsection{Removal Modelling}
+\par Estimates of cue rate changed very little from species that were originally modelled by the single-species models in \citet{edwards_point_2023}, to the same ones modelled in the multi-species model here.
+The 5-fold cross validation tended to show either no preference for one model or another when sample size was large, and a preference (albeit an extremely small preference) for single-species models when sample size was smaller.
+When sample size was small, a multi-species model tended to produce more precise estimates of cue rate.
+Given the extremely small preference for single-species models under a lower sample size, it would appear that the bias-variance tradeoff is likely a welcome tradeoff here, because the decrease in variance is met by only a very slight increase in bias.
 
-* migration strategy mattering less
+\par There were seven species where the relative difference in cue rate between the single-species model and multi-species model was over 20\% (noting that this threshold was chosen arbitrarily).
+These species (Willow Ptarmagin, Rusty Blackbird, Red-breasted Sapsucker, Common Redpoll, Calliope Hummingbird, Allen's Hummingbird, and Bohemian Waxwing), mostly had fairly small sample sizes ($\leq 515$ for all species except Red-breasted Sapsucker [$n = 1854$] and Common Redpoll [$n = 2009$]), had the lowest cue rate in the single-species models, and had among the lowest cue rates in the multi-species model.
+For removal modelling, all those species except for Common Redpoll ended up having the multi-species model chosen as the more predictive model, indicating that sharing information among related species tended to improve their estimates of cue rate.
+Common Redpoll was likely an exception to this because the original estimate of cue rate in NA-POPS was a wild underestimate (cue rate of TO DO), likely because of a convergence issue in the \texttt{detect} package.
+It could be worth running Common Redpoll using the Bayesian single species model developed here to see if the estimate of cue rate is still extremely low, or if the vague but still somewhat informative prior helps to bring the estimated parameter into a more sensible space.
+It could also be worth investigating the merit of this type of modelling to estimate cue rate for species where the cue rate is extremely low.
+In our case, 41 out of the 79 species that had the lowest 25\% quantile of cue rate in the single-species model ended up having the multi-species model chosen as the preferred model.
+TO DO Re-read results of Martin-Schwarze.
 
-* talk about Common Redpoll removal modelling
+\par Although this was not a primary goal of our study, we found much less support than \citet{solymos_phylogeny_2018} for migration strategy having an effect on cue rate.
+Both the estimate effects for resident species and migratory species were effectively 0, with very little of the density above or below 0.025 of 0.
+One potential reason is that we had proportionally more resident species in our analysis (85 out of 316; 26.9\%) compared to \citet{solymos_phylogeny_2018} (23 out of 141; 16.3\%), and so the richer information provided by these additional species could have brought the effect down to 0.
+Additionally, while \citet{solymos_phylogeny_2018} only regressed estimates of cue rate against the predictor variables, we actually estimated cue rate within the context of the same predictors, and so the phylogenetic effect could have just overwhelmed the estimates, making the migratory strategy meaningless.
+
+\subsection{Distance Modelling}
+\par Estimates of EDR also changed very little from species that were originally modelled by the single-species models in \citet{edwards_point_2023}, to the same ones modelled in the multi-species model here.
+However, there was evidence for an average small downward shift in EDRs when modelled under a multi-species context.
+Still, the 5-fold cross validation tended to show little to no preference for one more or another when the sample size was large, and a slight (but still very small) preference for the multi-species model when the sample size was smaller.
+There was less evidence for the multi-species model being able to improve the precision of EDR estimates across sample sizes, indicating that a single-species model may mostly be sufficient for precise estimates.
+Under this context, it would appear that a multi-species model may slightly bring down estimates of EDR for species with a smaller sample size.
+
+\par Talk about species with a large relative difference
+
+\par Talk about parameters
+
+\subsection{Predictions of Cue Rate and EDR}
 

### 2023-10-20 src/scripts/removal/7-removal-cv-analysis.R
changed size of figures

@@ -175,12 +175,12 @@ print(lppd_difference_plot)
 dev.off()
 
 png(filename = "output/plots/cv_removal_diff_vs_n.png",
-    width = 10, height = 6, units = "in", res = 300)
+    width = 6, height = 4, units = "in", res = 300)
 print(lppd_diff_vs_n)
 dev.off()
 
-png(filename = "output/plots/cv_removal_tree.png",
-    width = 30, height = 60, units = "in", res = 300)
+pdf(file = "output/plots/cv_removal_tree.pdf",
+    width = 30, height = 60)
 print(model_pref_tree)
 dev.off()
 

### 2023-10-20 src/scripts/removal/8-removal-comparison-analysis.R

@@ -166,19 +166,17 @@ species_vars <- to_plot$variable
 ####### Output ####################################
 
 png("output/plots/removal_1vs1.png",
-    width = 6, height = 6, res = 600, units = "in")
-ggarrange(ggarrange(modelled_difference_plot, single_vs_multi_plot,
-                    ncol = 2,
-                    labels = c("A", "B")), difference_plot, nrow = 2, labels = c("", "C"))
+    width = 6, height = 3, res = 600, units = "in")
+ggarrange(modelled_difference_plot, single_vs_multi_plot, ncol = 2, labels = c("A", "B"))
 dev.off()
 
 png("output/plots/removal_sd_plot.png",
-    width = 6, height = 6, res = 600, units = "in")
+    width = 6, height = 4, res = 600, units = "in")
 ggarrange(sd_intercept_plot, sd_slope_plot, ncol = 2, labels = c("A", "B"))
 dev.off()
 
 png("output/plots/removal_predictions_plot.png",
-    width = 6, height = 6, res = 600, units = "in")
+    width = 6, height = 4, res = 600, units = "in")
 print(species_prediction_plot)
 dev.off()
 

### 2023-10-24 src/scripts/removal/8-removal-comparison-analysis.R
missed a spot

@@ -27,10 +27,10 @@ load("data/generated/removal_stan_data_pred.rda")
 rem_summary <- rem_model$summary("log_phi")
 
 # Add species names to these summaries
-rem_summary$Code <- removal_stan_data$sp_all
+rem_summary$Code <- removal_stan_data_pred$sp_all
 
 # Get data sample size for all species and add to summary
-species_n <- data.frame(table(removal_stan_data$species))
+species_n <- data.frame(table(removal_stan_data_pred$species))
 names(species_n) <- c("index", "N")
 rem_summary$index <- seq(1, nrow(rem_summary))
 rem_summary$N <- 0

### 2023-10-24 article/article.tex
writing some of distance discussion

@@
 However, there was evidence for an average small downward shift in EDRs when modelled under a multi-species context.
-Still, the 5-fold cross validation tended to show little to no preference for one more or another when the sample size was large, and a slight (but still very small) preference for the multi-species model when the sample size was smaller.
+Still, the 5-fold cross validation tended to show almost no preference for one model or another when the sample size was large, and a slight (but still very small) preference for the multi-species model when the sample size was small.
 There was less evidence for the multi-species model being able to improve the precision of EDR estimates across sample sizes, indicating that a single-species model may mostly be sufficient for precise estimates.
@@
 
-\par Talk about species with a large relative difference
+\par There were six species where the relative difference in EDR between the single-species model and multi-species model was over 20\%.
+Unlike removal modelling, these species (Sharp-tailed Grouse, Sedge Wren, Ring-necked Pheasant, Chimney Swift, Boat-tailed Grackle, and Baird's Sparrow) had little in common with each other in terms of sample size (range of 227 for Boat-tailed Grackle through 4466 for Baird's Sparrow) and EDR (ranging from 50.8 m to 226 m in single-species, and from 73.2 m to 159.1 m in the multi-species model).
+Additionally, there was mixed support among these species for single-species vs.
+multi-species: Chimney Swift and Boat-tailed Grackle had the multi-species model preferred, whereas the others had the single-species model preferred.
+It could be noted that both Chimney Swift and Boat-tailed Grackle showed an increase in estimated EDR from single-species to multi-species modelling and preferred the multi-species model, but Baird's Sparrow also showed an increase and preferred the single-species model.
 
-\par Talk about parameters
+\par Similar to removal modelling, we found less support than \citet{solymos_phylogeny_2018} for migration strategy having an effect on EDR, although the posterior distribution for both the resident and migrant species were much further away from zero, indicating at least some support.
+Habitat type, body mass, and song pitch, as expected, all had an overall effect on EDR, as was found in \citet{solymos_phylogeny_2018} and \cite{johnston_species_2014}.
 
@@
 
-* talk about Bicknell's Thrush removal and distance modelling
+\par We generated predictions of cue rate for seven species of landbirds that are listed on Partners in Flight's watchlist: Spotted Owl, Bicknell's Thrush, LeConte's Thrasher, Tricoloured Blackbird, Harris's Sparrow, Kirtland's Warbler, and Lesser Prairie-chicken.
+For all species except Bicknell's Thrush and Kirtland's Warbler, we had at least some samples, but fewer than the minimum threshold of 75 for NA-POPS.
+For these five species, we used the multi-species model as a way to relax the sample size requirements by allowing for the sharing of information among related species via phylogeny.
+For Bicknell's Thrush and Kirtland's Warbler, we had zero samples, and so we used the multi-species model purely to predict estimates of cue rate, again based off of information sharing from similar species.
+
+Overall, the estimates of cue rate fell roughly where we would expect based on related species.
+For example, Harris's Sparrow (n = 65) had an estimated cue rate of 0.30 cues per minute, with a 95\% credible interval of [0.21, 0.42].
 

### 2023-10-24 src/scripts/removal/9-removal-posthoc-analysis.Rmd
adding a script to do simple posthox analyses and output numbers needed in article

---
title: "9-removal-posthoc-analysis"
author: "Brandon Edwards"
date: "2023-10-24"
output: html_document
---

```{r setup, include=FALSE}
library(magrittr)
knitr::opts_chunk$set(echo = TRUE)
```
```{r read_data, echo = FALSE}
####### Read Data #################################

phi_df <- read.csv("../../../data/generated/phi.csv")
phi_diff <- read.csv("../../../data/generated/phi_differences.csv")
lppd_pointwise <- read.csv("../../../data/generated/lppd_removal.csv")
lppd_species <- read.csv("../../../data/generated/lppd_summary_removal.csv")
params <- read.csv("../../../data/generated/removal_params.csv")
traits <- read.csv("../../../data/raw/traits.csv")

ibp_codes <- read.csv("../../../data/raw/IBP-Alpha-Codes20.csv")
families <- read.csv("../../../data/raw/NACC_list_species.csv")

####### Set Constants #############################

sp_pred <- c("LCTH", "LEPC", "HASP", "TRBL", "SPOW", "KIWA", "BITH")
```

# Cross Validation Statistics

Table of Model Preference from Cross-validation:
```{r mod_pref, echo = FALSE}
print(data.frame(table(lppd_species$Model_Preference)))
```

Mean LPPD Difference overall: `r mean(lppd_pointwise$Difference)`

"Sample Size/Model Selection of Species with >20% difference in cue rate:
```{r n_mod_selection, echo = FALSE}
phi_diff_pred <- merge(phi_diff, lppd_species[, c("Species", "Mean_Difference", "Model_Preference", "Migrant")],
                       by = "Species")
print(phi_diff_pred[which(phi_diff_pred$Label != ""), ])
```

```{r, echo = FALSE}
l25 <- quantile(phi_diff$Single)[2]
```
25th quantile of cue rate in single-species model: `r l25`
```{r, echo = FALSE}
sp_l25 <- phi_diff_pred[which(phi_diff_pred$Single <= l25),]
```

Number of species with single-species cue rate at/below 25th quantile: `r nrow(sp_l25)`

Number of species at/below 25th quantile that selected multi-species: `r length(which(sp_l25$Model_Preference == "MS"))` (`r length(which(sp_l25$Model_Preference == "MS")) / nrow(sp_l25)`%)

# Parameter Statistics
```{r, echo = FALSE}
print(params[, c("variable", "mean", "median", "q5", "q95")])
```

```{r, echo = FALSE}
traits_red <- traits[which(traits$Code %in% phi_diff$Species), ]
```

Number of resident species: `r nrow(traits_red) - sum(traits_red$Migrant)`

Number of migrant species: `r sum(traits_red$Migrant)`

# Predictions
```{r, echo = FALSE}
fam_df <- data.frame(Species = phi_df$Code, Phi = exp(phi_df$mean)) %>%
  merge(ibp_codes[, c("SPEC", "COMMONNAME")],
        by.x = "Species", by.y = "SPEC") %>%
  merge(families[, c("common_name", "family")],
        by.x = "COMMONNAME", by.y = "common_name")

fam_df_sparrow <- fam_df[which(fam_df$family == "Passerellidae"), ]
fam_df_sparrow <- fam_df_sparrow[-which(fam_df_sparrow$Species == "HASP"), ]

fam_df_thrasher <- fam_df[which(fam_df$family == "Mimidae"), ]
fam_df_thrasher <- fam_df_thrasher[-which(fam_df_thrasher$Species == "LCTH"), ]

fam_df_warbler <- fam_df[which(fam_df$family == "Parulidae"), ]
fam_df_warbler <- fam_df_warbler[-which(fam_df_warbler$Species == "KIWA"), ]

fam_df_thrush <- fam_df[which(fam_df$family == "Turdidae"), ]
fam_df_thrush <- fam_df_thrush[-which(fam_df_thrush$Species == "BITH"), ]
```

Estimates of cue rate for predicted species:
```{r cr_pred, echo = FALSE}
phi_pred <- phi_df[which(phi_df$Code %in% sp_pred), ]
phi_pred$log_mean <- phi_pred$mean
phi_pred$mean <- exp(phi_pred$mean)
phi_pred$q5 <- exp(phi_pred$q5)
phi_pred$q95 <- exp(phi_pred$q95)
print(phi_pred[, c("Code", "mean", "q5", "q95", "N")])
```

Mean cue rate for sparrows (except HASP): `r mean(fam_df_sparrow$Phi)`, n = `r nrow(fam_df_sparrow)`

Mean cue rate for thrashers (except LCTH): `r mean(fam_df_thrasher$Phi)`, n = `r nrow(fam_df_thrasher)`

Mean cue rate for thrushes (except BITH): `r mean(fam_df_thrush$Phi)`, n = `r nrow(fam_df_thrush)`

Mean cue rate for warblers (except KIWA): `r mean(fam_df_warbler$Phi)`, n = `r nrow(fam_df_warbler)`

### 2023-10-24 src/scripts/removal/8-removal-comparison-analysis.R

@@ -100,6 +100,8 @@ diff_model_run <- diff_model$sample(
                                                    prob = 0.95) +
   xlab("Modelled Difference"))
 
+cr_differences <- to_plot
+
 ####### SD Comparison Plot ########################
 
 to_plot <- rem_summary[which(rem_summary$Code %in% napops_summary$Species), 
@@ -163,7 +165,13 @@ species_vars <- to_plot$variable
 (sigma_plot <- bayesplot::mcmc_areas(rem_model$draws(c("sigma")),
                                          prob = 0.95))
 
+params_summary <- rem_model$summary(c("intercept", "mu_mig_strat", "sigma"))
+
 ####### Output ####################################
+
+write.table(rem_summary, file = "data/generated/phi.csv", sep = ",", row.names = FALSE)
+write.table(cr_differences, file = "data/generated/phi_differences.csv", sep = ",", row.names = FALSE)
+write.table(params_summary, file = "data/generated/removal_params.csv", sep = ",", row.names = FALSE)
 
 png("output/plots/removal_1vs1.png",
     width = 6, height = 3, res = 600, units = "in")

### 2023-10-24 article/article.tex
update discussion with more removal modelling stuff and tlaking about predictions of cue rate

@@
 \par We found much less support for a difference in cue rate between resident and migrant species than was found in \citet{solymos_evaluating_2018}. 
-Mean overall log cue rate in the multi-species model was TO DO; both estimated effects for resident and migrant species were effectively 0 (Figure \ref{fig:removal_params}).
+Mean overall log cue rate in the multi-species model was -1.09 (95\% credible interval of [-1.24, -0.94]); both estimated effects for resident and migrant species were effectively 0 (Figure \ref{fig:removal_params}).
 
@@
 
-\par The cross validation revealed that single species model actually had better predictive capacity for more species than the multi-species model, but the difference in predictive capacity was overall very small.
+\par The cross validation revealed that single species model actually had better predictive capacity for more species (222 out of 316) than the multi-species model (94 out of 316), but the difference in predictive capacity was overall very small.
 Mean overall difference in posterior predictive density was $-2.21 \times 10^{-5}$.
@@
 
-\subsection{Predictions of Cue Rate and EDR}
+\subsection{Predictions of Cue Rate}
 
@@
 
-Overall, the estimates of cue rate fell roughly where we would expect based on related species.
-For example, Harris's Sparrow (n = 65) had an estimated cue rate of 0.30 cues per minute, with a 95\% credible interval of [0.21, 0.42].
+\par Overall, the estimates of cue rate fell roughly where we would expect based on related species, for species where we were simply relaxing the sample size requirements.
+For example, Harris's Sparrow (n = 65) had an estimated cue rate of 0.30 cues per minute, with a 95\% credible interval of [0.21, 0.42], compared to the overall mean cue rate of 0.37 cues per minute for all 32 other Passerellidae species in NA-POPS.
+Similarly, LeConte's Thrasher (n = 1) had an estimated cue rate of 0.32 cues per minute, with a 95\% credible interval of [0.18, 0.57].
+Though this was a much larger interval than Harris's Sparrow, it still falls within a reasonable distance from the overall mean cue rate of 0.36 cues per minute for all 6 other Mimidae species in NA-POPS.
+Of course, it is likely that the width of the credible intervals (i.e., the precision) of these relaxed sample-size species are going to depend both on the number of data points available for that particular species, the number of "similar" species, and the number of data points available for those similar species.
+
+\par For species where we were making predictions of cue rate based on no data (i.e., Bicknell's Thrush and Kirtland's Warbler), we found reasonable estimates but with fairly wide credible intervals.
+As expected, the predicted cue rate for Bicknell's Thrush (0.28 cues per minute) and the predicted cue rate for Kirtland's Warbler (0.29 cues per minute) fell very close to the overall mean cue rate of 0.27 for all 11 other Turdidae species and 0.31 for all 44 other Parulidae species in NA-POPS, respectively.
+That is, in the absence of any data, the estimated cue rate of Bicknell's Thrush and Kirtland's Warbler was shrunk toward cue rates of most similar species.
+However, the magnitude of contribution of data from related species appeared to play a role in the width of the credible intervals.
+Although both credible intervals were fairly wide, Bicknell's Thrush had a 95\% credible interval of [0.16, 0.48], representing an approximate 53\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [0.18, 0.48], representing an approximate 49\% change on either end.
+Although the difference in width of the intervals is small, this could be due to the greater number of other species contributing information to Kirtland's Warbler estimates than to Bicknell's Thrush.
+With that in mind, it would likely be important to consider how many "related" species are available with information (and how many data points do each of those related species have) when deciding on making a prediction of cue rate.
+
+\subsection{Predictions of Effective Detection Radius}
+
+\par We generated predictions of EDR for six species of landbirds that are listed in Partners in Flight's watchlist: Spotted Owl, Bicknell's Thrush, LeConte's Thrasher, Harris's Sparrow, Kirtland's Warbler, and Lesser Prairie-chicken.
+For all species except Bicknell's Thrush, we had at least some samples, but fewer than the minimum threshold of 75 for NA-POPS.
+For these five species, we used the multi-species model as a way to relax the sample size requirements by allowing for the sharing of information among related species via traits.
+For Bicknell's Thrush, we had zero samples, and so we used the multi-species model purely to predict estimates of EDR, again based off of information sharing from species with similar traits.
+
+\par Similar to multi-species removal modelling, estimates of EDR fell roughly where we would expect based on species of similar traits, for the species where we were relaxing samples sizes.
 

### 2023-10-25 src/scripts/distance/9-distance-posthoc-analysis.Rmd
added posthoc analysis of distance modelling stuff for paper

---
title: "9-distance-posthoc-analysis"
author: "Brandon Edwards"
date: "2023-10-24"
output: html_document
---

```{r setup, include=FALSE}
library(magrittr)
knitr::opts_chunk$set(echo = TRUE)
```
```{r read_data, echo = FALSE}
####### Read Data #################################

tau_df <- read.csv("../../../data/generated/tau.csv")
tau_diff <- read.csv("../../../data/generated/tau_differences.csv")
lppd_pointwise <- read.csv("../../../data/generated/lppd_distance.csv")
lppd_species <- read.csv("../../../data/generated/lppd_summary_distance.csv")
params <- read.csv("../../../data/generated/distance_params.csv")
traits <- read.csv("../../../data/raw/traits.csv")

ibp_codes <- read.csv("../../../data/raw/IBP-Alpha-Codes20.csv")
families <- read.csv("../../../data/raw/NACC_list_species.csv")

####### Set Constants #############################

sp_pred <- c("LCTH", "LEPC", "HASP", "SPOW", "KIWA", "BITH")
```

# Cross Validation Statistics

Table of Model Preference from Cross-validation:
```{r mod_pref, echo = FALSE}
print(data.frame(table(lppd_species$Model_Preference)))
```

Mean LPPD Difference overall: `r mean(lppd_pointwise$Difference)`

"Sample Size/Model Selection of Species with >20% difference in EDR:
```{r n_mod_selection, echo = FALSE}
tau_diff_pred <- merge(tau_diff, lppd_species[, c("Species", "Mean_Difference", "Model_Preference")],
                       by = "Species")
print(tau_diff_pred[which(tau_diff_pred$Label != ""), ])
```

```{r, echo = FALSE}
l25 <- quantile(tau_diff$Single)[2]
```
25th quantile of EDR in single-species model: `r l25`
```{r, echo = FALSE}
sp_l25 <- tau_diff_pred[which(tau_diff_pred$Single <= l25),]
```

Number of species with single-species EDR at/below 25th quantile: `r nrow(sp_l25)`

Number of species at/below 25th quantile that selected multi-species: `r length(which(sp_l25$Model_Preference == "MS"))` (`r length(which(sp_l25$Model_Preference == "MS")) / nrow(sp_l25)`%)

# Parameter Statistics
```{r, echo = FALSE}
print(params[, c("variable", "mean", "median", "q5", "q95")])
```

```{r, echo = FALSE}
traits_red <- traits[which(traits$Code %in% tau_diff$Species), ]
```

Number of resident species: `r nrow(traits_red) - sum(traits_red$Migrant)`

Number of migrant species: `r sum(traits_red$Migrant)`

# Predictions
```{r, echo = FALSE}
fam_df <- data.frame(Species = tau_df$Code, Tau = exp(tau_df$mean) * 100) %>%
  merge(ibp_codes[, c("SPEC", "COMMONNAME")],
        by.x = "Species", by.y = "SPEC") %>%
  merge(families[, c("common_name", "family")],
        by.x = "COMMONNAME", by.y = "common_name") %>%
  merge(traits[, c("Code", "Mass", "Pitch")],
        by.x = "Species", by.y = "Code")

fam_df_sparrow <- fam_df[which(fam_df$family == "Passerellidae"), ]
fam_df_sparrow <- fam_df_sparrow[-which(fam_df_sparrow$Species == "HASP"), ]

fam_df_thrasher <- fam_df[which(fam_df$family == "Mimidae"), ]
fam_df_thrasher <- fam_df_thrasher[-which(fam_df_thrasher$Species == "LCTH"), ]

fam_df_warbler <- fam_df[which(fam_df$family == "Parulidae"), ]
fam_df_warbler <- fam_df_warbler[-which(fam_df_warbler$Species == "KIWA"), ]

fam_df_thrush <- fam_df[which(fam_df$family == "Turdidae"), ]
fam_df_thrush <- fam_df_thrush[-which(fam_df_thrush$Species == "BITH"), ]
```

Estimates of EDR for predicted species:
```{r cr_pred, echo = FALSE}
tau_pred <- tau_df[which(tau_df$Code %in% sp_pred), ]
tau_pred <- merge(tau_pred, traits[, c("Code", "Mass", "Pitch")],
                  by = "Code")
tau_pred$log_mean <- tau_pred$mean
tau_pred$mean <- exp(tau_pred$mean) * 100
tau_pred$q5 <- exp(tau_pred$q5) * 100
tau_pred$q95 <- exp(tau_pred$q95) * 100
print(tau_pred[, c("Code", "mean", "q5", "q95", "N", "Mass", "Pitch")])
```

Mean EDR for sparrows (except HASP): `r mean(fam_df_sparrow$Tau)`, 
Song Pitch: `r mean(fam_df_sparrow$Pitch)`, 
Mass: `r mean(fam_df_sparrow$Mass)`, 
n = `r nrow(fam_df_sparrow)`

Mean EDR for thrashers (except LCTH): `r mean(fam_df_thrasher$Tau)`, n = `r nrow(fam_df_thrasher)`

Mean EDR for thrushes (except BITH): `r mean(fam_df_thrush$Tau)`, n = `r nrow(fam_df_thrush)`

Mean EDR for warblers (except KIWA): `r mean(fam_df_warbler$Tau)`, n = `r nrow(fam_df_warbler)`

### 2023-10-25 src/scripts/distance/8-distance-comparison-analysis.R

@@ -98,6 +98,8 @@ diff_model_run <- diff_model$sample(
                                                    prob = 0.95) +
     xlab("Modelled Difference"))
 
+edr_differences <- to_plot
+
 ####### SD Comparison Plot ########################
 
 to_plot <- dis_summary[which(dis_summary$Code %in% napops_summary$Species), 
@@ -166,7 +168,15 @@ species_vars <- to_plot$variable
 (sigma_plot <- bayesplot::mcmc_areas(dis_model$draws(c("sigma")),
                                      prob = 0.95))
 
+params_summary <- dis_model$summary(c("intercept", "mu_mig_strat",
+                                      "mu_habitat", "beta_mass", "beta_pitch",
+                                      "sigma"))
+
 ####### Output ####################################
+
+write.table(dis_summary, file = "data/generated/tau.csv", sep = ",", row.names = FALSE)
+write.table(edr_differences, file = "data/generated/tau_differences.csv", sep = ",", row.names = FALSE)
+write.table(params_summary, file = "data/generated/distance_params.csv", sep = ",", row.names = FALSE)
 
 png("output/plots/distance_1vs1.png",
     width = 6, height = 3, res = 600, units = "in")

### 2023-10-25 article/article.tex
points of what to write next
added more posthoc analysis discussion to paper

@@
 \par Similar to multi-species removal modelling, estimates of EDR fell roughly where we would expect based on species of similar traits, for the species where we were relaxing samples sizes.
+For example, Harris's Sparrow (n = 13) had an estimated mean EDR of 65.6 m, with a 95\% credible interval of [53.1, 82.5].
+This range does fall below the mean EDR of 85.3 m for the other sparrows in NA-POPS, but Harris's Sparrow song pitch of approximately 3.8 Hz falls above the average song pitch of 3.0 Hz for the other sparrows in NA-POPS, which could contribute to the smaller EDR.
+On the other hand, LeConte's Thrasher (n = 1) had an estimated EDR of 134.2 m, with a 95\% credible interval of [81.8, 230.5], which falls roughly in line with the mean EDR of 113.2 m of the six other thrashers in NA-POPS.
+
+\subsection{Model Performance}
+
+* removal modelling was always okay
+
+* distance model struggled significantly
+
+\subsection{Recommendations for Multi-species Predictions of Detectability}
 

### 2023-10-25 article/article.tex
added discussion piece comparing detectabilities to other studies

@@
 
-\subsection{Predictions of Cue Rate}
+\subsection{Relaxation of Sample Size}
 
-\par We generated predictions of cue rate for seven species of landbirds that are listed on Partners in Flight's watchlist: Spotted Owl, Bicknell's Thrush, LeConte's Thrasher, Tricoloured Blackbird, Harris's Sparrow, Kirtland's Warbler, and Lesser Prairie-chicken.
-For all species except Bicknell's Thrush and Kirtland's Warbler, we had at least some samples, but fewer than the minimum threshold of 75 for NA-POPS.
-For these five species, we used the multi-species model as a way to relax the sample size requirements by allowing for the sharing of information among related species via phylogeny.
-For Bicknell's Thrush and Kirtland's Warbler, we had zero samples, and so we used the multi-species model purely to predict estimates of cue rate, again based off of information sharing from similar species.
+\par For five out of the seven species where we generated new predictions of detectability in this study, we were interested in seeing how a multi-species QPAD removal model could estimate cue rate with few data, and how a multi-species QPAD distance model could estimate EDR with few data.
 
@@
 Though this was a much larger interval than Harris's Sparrow, it still falls within a reasonable distance from the overall mean cue rate of 0.36 cues per minute for all 6 other Mimidae species in NA-POPS.
-Of course, it is likely that the width of the credible intervals (i.e., the precision) of these relaxed sample-size species are going to depend both on the number of data points available for that particular species, the number of "similar" species, and the number of data points available for those similar species.
-
-\par For species where we were making predictions of cue rate based on no data (i.e., Bicknell's Thrush and Kirtland's Warbler), we found reasonable estimates but with fairly wide credible intervals.
-As expected, the predicted cue rate for Bicknell's Thrush (0.28 cues per minute) and the predicted cue rate for Kirtland's Warbler (0.29 cues per minute) fell very close to the overall mean cue rate of 0.27 for all 11 other Turdidae species and 0.31 for all 44 other Parulidae species in NA-POPS, respectively.
-That is, in the absence of any data, the estimated cue rate of Bicknell's Thrush and Kirtland's Warbler was shrunk toward cue rates of most similar species.
-However, the magnitude of contribution of data from related species appeared to play a role in the width of the credible intervals.
-Although both credible intervals were fairly wide, Bicknell's Thrush had a 95\% credible interval of [0.16, 0.48], representing an approximate 53\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [0.18, 0.48], representing an approximate 49\% change on either end.
-Although the difference in width of the intervals is small, this could be due to the greater number of other species contributing information to Kirtland's Warbler estimates than to Bicknell's Thrush.
-With that in mind, it would likely be important to consider how many "related" species are available with information (and how many data points do each of those related species have) when deciding on making a prediction of cue rate.
-
-\subsection{Predictions of Effective Detection Radius}
-
-\par We generated predictions of EDR for six species of landbirds that are listed in Partners in Flight's watchlist: Spotted Owl, Bicknell's Thrush, LeConte's Thrasher, Harris's Sparrow, Kirtland's Warbler, and Lesser Prairie-chicken.
-For all species except Bicknell's Thrush, we had at least some samples, but fewer than the minimum threshold of 75 for NA-POPS.
-For these five species, we used the multi-species model as a way to relax the sample size requirements by allowing for the sharing of information among related species via traits.
-For Bicknell's Thrush, we had zero samples, and so we used the multi-species model purely to predict estimates of EDR, again based off of information sharing from species with similar traits.
 
@@
 On the other hand, LeConte's Thrasher (n = 1) had an estimated EDR of 134.2 m, with a 95\% credible interval of [81.8, 230.5], which falls roughly in line with the mean EDR of 113.2 m of the six other thrashers in NA-POPS.
+
+\par Of course, it is likely that the width of the credible intervals (i.e., the precision) of these relaxed sample-size species are going to depend both on the number of data points available for that particular species, the number of "similar" species, and the number of data points available for those similar species.
+However, species that do not have many other similar species from which to borrow information still have the benefit of using information from the priors that we set in the model, which is a large advantage to modelling detectability in a Bayesian context.
+Additionally, it is important to note that we did not account for other sources of variation at the survey level, such as when the survey was conducted (i.e., time of day, time of year), or where the survey was conducted (e.g., in a forest, by a roadside).
+It is well known that the ability to detect a species can be very dependent on the time of day or year that an observer is surveying, as well as the environmental conditions of the survey \citep{edwards_point_2023}.
+Thus, some of the variation in these estimates of cue rate and EDR could be due to not accounting for survey-level variation that affects detectability, and so it is important that future iterations of this model incorporate both species-level effects on detectability \textit{and} survey-level effects on detectability.
+
+\subsection{Predicting Detectability for Species with No Data}
+
+\par We generated predictions of cue rate and EDR for both Bicknell's Thrush and Kirtland's Warbler.
+For both species, we had zero removal sampling data, and so cue rates were solely based off information sharing among phylogeny (and to a much lesser extent, migration status).
+For Kirtland's Warbler, we did have one data point for distance sampling, so predictions of EDR for Kirtland's Warbler here is not entirely based off zero data points.
+Bicknell's Thrush, however, had zero data points, and so EDRs for this species were based solely off information sharing among other species with similar traits.
+
+\par As expected, the predicted cue rate for Bicknell's Thrush (0.28 cues per minute) and the predicted cue rate for Kirtland's Warbler (0.29 cues per minute) fell very close to the overall mean cue rate of 0.27 for all 11 other Turdidae species and 0.31 for all 44 other Parulidae species in NA-POPS, respectively.
+That is, in the absence of any data, the estimated cue rate of Bicknell's Thrush and Kirtland's Warbler was shrunk toward cue rates of most similar species.
+
+\par For distance modelling, the predicted EDR for Bicknell's Thrush (73.1 m) and the predicted EDR for Kirtland's Warbler (75.0 m) fell reasonably close to the overall mean EDR of 92.7 m for all other 11 Turdidae species and 61.2 m for all 44 other Parulidae species in NA-POPS, respectively.
+That is, in the absence of any data (or with very little data in the case of Kirtland's Warbler), the estimated EDR of Bicknell's Thrush and Kirtland's Warbler was s
+
+\par In both the removal modelling and distance modelling, the magnitude of contribution of data from related species appeared to play a role in the width of the credible intervals.
+For removal modelling, although both credible intervals were fairly wide, Bicknell's Thrush had a 95\% credible interval of [0.16, 0.48], representing an approximate 53\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [0.18, 0.48], representing an approximate 49\% change on either end.
+For distance modelling, both credible intervals were also fairly wide, but Bicknell's Thrush had a 95\% credible interval of [40.4, 135.0], representing an approximate 58\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [46.6, 124.8], representing an approximate 47\% change on either end.
+Although the difference in width of the intervals is small, this could be due to the greater number of other species contributing information to Kirtland's Warbler estimates than to Bicknell's Thrush.
+In the case of distance modelling, the one data point contributing to Kirtland's Warbler's EDR could have also played a part.
+With that in mind, it would likely be important to consider how many "related" species are available with information (and how many data points do each of those related species have) when deciding on making a prediction of cue rate.
+
+\subsection{Comparisons of Predicted Detectability to Other Studies}
+
+\par Using the predictions of cue rate and EDR for Bicknell's Thrush and Kirtland's Warbler, we can calculate availability and perceptibility, and hence detectability, for a given survey type, and compare to previous estimates of detectability obtained from other studies.
+As a reminder, following derivations in \citet{solymos_calibrating_2013}, we note that detectability is the product of availability and perceptibility.
+Availability $p(t)$ is defined as the probability of a bird giving a cue during a survey run for $t$ minutes, and is calculated as
+$p(t) = 1 - e^{-t\phi}$
+where $\phi$ is the cue rate of the species.
+Perceptibility $q(r)$ is defined as the probability of an observer hearing or seeing a bird, given the bird gives a cue, with a survey radius of $r$ metres, and is calculated as
+$q(r) = \dfrac{\tau^2 \left\{1 - e^{-\dfrac{r^2}{\tau^2}}\right\}}{r^2}$
+where $\tau$ is the EDR of the species.
+
+\par For Bicknell's Thrush, using the equations above, the predicted cue rate of 0.28, predicted EDR of 73.1 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.34.
+We can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respectively credible intervals) to get a range of detectabilities from 0.09 through 0.70.
+A study by \citet{aubry_bicknells_2018} noted that they had estimated detection probabilities $\geq$ 0.74 and $\leq$ 0.88, and so even our highest potential value of detectability falls short of their estimated range.
+However, we note that \citet{aubry_bicknells_2018} are using data that specifically target Bicknell's Thrush in their range, and they are using surveys conducted at peak Bicknell's Thrush breeding time, and so it is likely that detectability will be high in that specific case.
+
+\par For Kirtland's Warbler, using the equations above, the predicted cue rate of 0.29, predicted EDR of 75.0 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.36.
+Similar to our Bicknell's Thrush example, we can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respective credible intervals) to obtain a range of detectabilities from 0.13 through 0.67.
+A study by \citet{van_dyke_comparative_2022} found esitmated detection probabilities for Kirtland's Warbler that ranged from effectively 0, through to a detection probability of 0.73.
+In that study, they compared estimates of detectability among different jack pine stands and red pine stands, where the detectability was highest in jack pine stands than the red pine stands.
+
+Both these examples highlight the need for future iterations of this multi-species model to consider environmental and temporal covariates that affect peaks in detectability through the year and for different environmental conditions.
+This is especially true for the types of species we are interested in generating predictions for, i.e., rare and/or undersampled species.
+In general, rare and undersampled species will likely prefer rare habitats or be confined to a very small range (as is the case with Bicknell's Thrush and Kirtland's Warbler above).
+Thus, while we can generate reasonable estimates of the components of detectability (i.e., availability through cue rate and perceptibility through EDR) based on information borrowed from similar species (either through phylogeny or traits), this model as it stands may fail to glean more specific information that takes into account detectability peaks throughout the year, or detectability peaks given specific habitats.
+With this in mind, it may be beneficial to include temporal covariates at the survey level, similar to what was done in \citet{solymos_calibrating_2013, solymos_evaluating_2018, edwards_point_2023}, so that further information about cue rate may be shared among similar species giving cues at similar times of day or year.
+Similarly, it may be beneficial to include environmental covariates at the survey level, similar to what was done in \citet{solymos_calibrating_2013, edwards_point_2023}, so that further information about EDR may be shared among similar species being surveyed at similar habitats.
+This could also provide an opportunity to disentangle species x survey interactions in detectability.
+
+\par The examples above additionally highlight an excellent opportunity to make use of a powerful feature of Bayesian models, that being the inclusion of expert knowledge as prior information.
+For example, if a Bicknell's Thrush expert knows that there is something particular about how often they sing or give a cue compared to other thrushes, or how disproportionately quiet or loud they are comapred to species with similar traits, this can be coded directly into the prior distribution.
+Similarly, this may also provide a unique opportunity to make use of traditional ecological knowledge to further inform predictions of detectability.
+For example, in the Anishinaabe language, bird taxonomy is based both on physical features, but also behaviours such as flight, calls, morning rituals, and weather (CITATION Joe Pitawanakwat presentation).
+This taxonomy is based off observations of bird behaviour that has been passed down through generations, and may vary slightly (or heavily) from taxonomy based on genomics.
+In fact, it could be argued that a taxonomy based on how humans observe birds may actually be more relevant for detectability, and so a future opportunity could be to adopt a two-eyed seeing approach \cite{reid_twoeyed_2021} to predicting detectability in rare birds.
 

### 2023-10-25 src/scripts/distance/9-distance-posthoc-analysis.Rmd
added discussion piece comparing detectabilities to other studies

@@
 Mean EDR for warblers (except KIWA): `r mean(fam_df_warbler$Tau)`, n = `r nrow(fam_df_warbler)`
+
+# Detectability Estimates
+```{r detectability}
+# Set constants based off predictions here and in the cue rate output
+bith_cr <- c(0.28, 0.16, 0.48)
+bith_edr <- c(73.1, 40.4, 135.0)
+kiwa_cr <- c(0.29, 0.18, 0.48)
+kiwa_edr <- c(75, 46.6, 124.8)
+t = 5
+r = 100
+
+bith_avail <- 1 - exp(-t * bith_cr)
+bith_percept <- (bith_edr^2 * (1 - exp(-r^2/bith_edr^2))) / (r^2)
+bith_detect <- bith_avail * bith_percept
+print(bith_detect)
+
+kiwa_avail <- 1 - exp(-t * kiwa_cr)
+kiwa_percept <- (kiwa_edr^2 * (1 - exp(-r^2/kiwa_edr^2))) / (r^2)
+kiwa_detect <- kiwa_avail * kiwa_percept
+print(kiwa_detect)
+```

### 2023-10-25 article/article.tex
added model performance subsection

@@
 
-* removal modelling was always okay
+\par The multi-species QPAD Distance model had several issues related to initialization, sampling, and convergence throughout this study.
+For many iterations of the model development, Stan's generation of initial values proved difficult.
+For all unconstrained parameters, Stan draws initial values from a uniform(-2,2) distribution.
+On the scale that we modelled detection distances (i.e., log detection distance in hundreds of metres), these are reasonable values.
+At the low end, an initial value of -2 would correspond to a detection distance of $e^{-2} \times 100 = 13.5$ m, which is low but somewhat plausible for a small-bodied, high-pitched species such as a hummingbird.
+At the high end, an initial value of 2 would correspond to a detection distance of $e^{2} \times 100 = 739$ m, which is high but still somewhat plausible for large-bodied, low-pitched species like a grouse or prairie-chickens.
+However, where Stan seemed to struggle is when these randomly assigned values for detection distances were effectively impossible for a large set of species.
+That is to say, if the values were initialized such that hummingbirds had a detection distance of 500 m or more, and grouse or prairie-chickens had a detection distance of 15 m or less, this would result in an impossible likelihood, forcing Stan to reject the initial values.
+Our fix was to set initial values to be jittered mean detection distances originally estimated in \citet{edwards_point_2023}; for species without estimates, we used the mean EDR for all species.
+This approach seemed to work as the initial values were no longer rejected, but we caution that this could very well be a "bandaid solution" to an underlying issue with the model itself.
 
-* distance model struggled significantly
+\par The distance models also seemed to struggle with sampling the posterior distribution, in that over a quarter of the iterations resulted in a divergent transition (CITATION for divergent transitions).
+This is likely due to hierarchical nature of the model and the parameterization choices that we made.
+\citet{betancourt_hierarchical_2020} discuss at length some issues that arise when sampling from hierarchical models in cases where the hierarchical groups are data rich, where a "centred parameterization" should be used, or data poor, where a "non-centred parameterization" should be used.
+We experimented with using only a non-centred parameterization which resulted in extremely low effective sample sizes (as low as 4 for 8000 iterations), large R-hat values, and took an unreasonable amount of time to finish sampling (approximately 2.5 weeks).
+The model presented in this paper is a centred parameterization, which produced much more reasonable effective sample sizes and R hat values, finished within a day and a half, but still suffered from degeneracies such as hierarchical funneling as described by \citet{betancourt_hierarchical_2020}.
+In our case, because we are modelling on a hierarchy that contain many data points for some species x trait groups and fewer data points for other species x trait groups, our divergent transitions may be resolved if we were to implement our model using a mixture of centred and non-centred parametrization as described by \citet{betancourt_hierarchical_2020}.
 
-\subsection{Recommendations for Multi-species Predictions of Detectability}
+\par For removal modelling, we experience no issues related to initialization, sampling, or convergence. 
+The model finished within approximately 24 hours, and all parameters had R-hat values of 1, or within 0.001 of it.
+We note that even though we used a Gaussian process to estimate cue rate, we did not have to estimate the covariance matrix used for the Gaussian process, because it was effectively a known parameter provided by the phylogenetic relationship.
+This would account for the relatively quick performance of this model given the number of parameters to estimate and the amount of data used; had the covariance matrix needed to be estimated, this model would have likely taken much longer to complete.
+
+\subsection{Conservation Implications}
 

### 2023-10-25 article/article.tex
fixing up a few citations and areas where i had to come back to in introduciton and methods, some formatting

@@
 
-\par Detectability is an important metric for many conservation problems, but often goes unaccounted for, or detectability data just does not exist \citep{bennett_how_2023}.
+\par Detectability is an important metric for many conservation problems, but often goes unaccounted for in studies, or detectability data just does not exist \citep{bennett_how_2023}.
 In conservation management situations, detectability can be used as a metric of sensitivity to evaluate when to monitor a system for a species, or when to act on current information \citep{canessa_when_2015, bennett_when_2018}.
-When estimating population sizes of an organism, detectability often comes into play as a way to convert observed counts into densities or estimates of true abundance.
+When estimating population sizes of an organism, detectability often comes into play as a way to convert observed counts into densities or estimates of true abundance \citep{solymos_calibrating_2013, johnson_defense_2008}.
 As such, accurate and precise estimates of detectability are important both for making accurate conservation decisions, and estimating accurate population sizes.
@@
 The NA-POPS project is an effort to estimate detectability functions that take into account variability in detection through environmental and temporal covariates, for as many of North America's landbird species as possible \citep{edwards_point_2023}.
-To date, the project has estimated availability (the probability a bird sings or gives a cue) and perceptibility (the probability an observer sees or hears a cueing bird) for 319 species of birds, with an additional 19 species having estimates of availability or perceptibility (but not necessarily both) estimated.
+To date, the project has estimated availability (the probability a bird sings or gives a cue) and perceptibility (the probability an observer sees or hears a cueing bird) for 319 species of birds, with an additional 19 species having estimates of availability or perceptibility (but not necessarily both).
 This project built off previous efforts by the Boreal Avian Modelling project \citep{cumming_toward_2010} to systematically estimate detectability in landbirds \citep{solymos_calibrating_2013, solymos_evaluating_2018}.
-Both the NA-POPS project and previous estimates by Boreal Avian Modelling project use the QPAD methodology, which is a flexible method to estimate availability and perceptibility given a harmonized dataset \citep{barker_ecological_2015, solymos_calibrating_2013}.
+Both the NA-POPS project and previous estimates by Boreal Avian Modelling project use the QPAD methodology \citep{solymos_calibrating_2013}, which is a flexible method to estimate availability and perceptibility given a harmonized dataset \citep{barker_ecological_2015}.
 
 \par Despite the efforts of the NA-POPS project, the 338 species of birds considered in the modelling exercise still only account for about 75\% of Partners in Flight's 2016 Landbird Conservation Plan \citep{edwards_point_2023, rosenberg_partners_2016}.
-This is because of sample size limitations that are recommended for removal and distance modelling, in that a modeller should have at least 75 data points for reasonable estimates \citep{buckland_introduction_2001}.
+This is because of sample size limitations that are recommended for removal and distance modelling, in that a modeller should have at least 75 data points for reasonable estimates \citep{buckland_introduction_2001, solymos_calibrating_2013}.
 Many of the species that fell short of that threshold in the NA-POPS database are systematically rare or undersampled species, that are either low in population size to begin with, or occur in rare habitats.
-Some examples of species not modelled include Kirtland's Warbler (SCIENTIFIC NAME) and Bicknell's Thrush (SCIENTIFIC NAME), both of which have low population sizes, and LeConte's Thrasher (SCIENTIFIC NAME) and Harris's Sparrow (SCIENTIFIC NAME), both of which have rare habitat associations  (PARNTERS IN FLIGHT WATCHLIST CITATION).
+Some examples of species not modelled include Kirtland's Warbler (\textit{Setophaga kirtlandii}) and Bicknell's Thrush (\textit{Catharus bicknelli}), both of which have low population sizes, and LeConte's Thrasher (\textit{Toxostoma lecontei}) and Harris's Sparrow (\textit{Zonotrichia querula}), both of which have rare habitat associations \citep{will_handbook_2020}.
 Of course, despite not having good estimates of detectability availabile for these rare and undersampled species, these are the groups of species that need these estimates the most as their conservation status often depend on accurate population size estimates, and estimates of trend in those population sizes.
@@
 They also found that while the best model that explained detection distance did include phylogeny, the strength of the phylogenetic relationship was reduced to effectively 0 after accounting for traits such as body mass, song pitch, habitat association, and migratory status.
-Both these studies and others (see CITATIONSSSSS) provide useful evidence that detectability in birds can be reasonably predicted by some combinations of phylogeny and traits.
 
-\par One recommendation for future research that come out of \cite{solymos_phylogeny_2018} was to try to use these phylogenetic and trait relationships in detectability to then predict components of detectability (i.e., availability through cue rate, and perceptibility through detection distance) in other birds.
-Indeed, by making use of shrinkage factors and partial pooling, such as what can be done in a Bayesian hierarchical models (CITATION), and by accounting phylogenetic relationships and the relative "relatedness" of different species, such as what can be done with Gaussian processes \citep{bernardo_regression_1998, mcelreath_continous_2020}, we can leverage the information provided by species with many data points to use to inform species with much fewer (included potentially zero) data points.
-Thus, the goal of this paper is to create a multi-species model of detectability that incorporates species traits and phylogeny, to then predict estimates of detectability for rare and undersampled species of North American landbirds.
+\par One recommendation for future research that came out of \cite{solymos_phylogeny_2018} was to try to use these phylogenetic and trait relationships in detectability to then predict components of detectability (i.e., availability through cue rate, and perceptibility through detection distance) in other birds.
+Indeed, by making use of shrinkage factors and partial pooling, such as what can be done in a Bayesian hierarchical models \citep{gelman_what_2021}, and by accounting phylogenetic relationships and the relative "relatedness" of different species, such as what can be done with Gaussian processes \citep{bernardo_regression_1998, mcelreath_continous_2020}, we can leverage the information provided by species with many data points to use to inform species with much fewer (included potentially zero) data points.
+
+\par The goal of this paper is to create a multi-species model of detectability that incorporates species traits and phylogeny, to both allow for the relaxation of the sample size requirements for estimating detectability, and to predict estimates of detectability for bird species without any data.
 To do so, we developed a Bayesian hierarchical version of the QPAD Removal model initially derived in \cite{solymos_calibrating_2013}, that allows for information exchange between species based on phylogenetic distance and binned migratory status.
-Phylogenetic distance was coded as a correlation structure and used as a Gaussian random field in this model, and migratory status was coded similar to \cite{solymos_phylogeny_2018} where a species was considered either "migrant" or "resident.
+Phylogenetic distance was coded as a correlation structure and used as a Gaussian random field in this model, and migratory status was coded similar to \cite{solymos_phylogeny_2018} where a species was considered either "migrant" or "resident".
 We also developed a Bayesian hierarchical version of the QPAD Distance model initially derived in \cite{solymos_calibrating_2013}, that allows for information exchange between species based only on the traits that came out as being significant in \cite{solymos_phylogeny_2018}.
@@
 We also developed a Bayesian single species version of the QPAD Removal and Distance models (i.e., as opposed to the maximum likelihood models available in the \texttt{detect} R package; \cite{solymos_detect_2020}) to allow for direct comparison through cross validation.
-Finally, we ran these models with the full set of data points used in \cite{edwards_point_2023}, plus additional data for seven species that did not meet the initial sample size thresholds for NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird, and Spotted Owl.
-All seven of these bird species are listed on Partner's in Flight's watchlist for various threats (CITATION).
+Finally, we ran these models with the full set of data points used in \cite{edwards_point_2023}, plus additional data for seven species that did not meet the initial sample size thresholds for NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird (\textit{Agelaius tricolor}), and Spotted Owl (\textit{Strix occidentalis}).
+All seven of these bird species are listed on Partner's in Flight's watchlist for various threats \citep{will_handbook_2020}.
 
@@
 
-\par For removal modelling, we are not interested in distance to the bird, and so we will sum counts from each sampling event $i$ over all distance bands. 
+\par For removal modelling, we are only interested in the time bin in which the bird was recorded, and not the distance, and so we will sum counts from each sampling event $i$ over all distance bands.
 Thus, we will be considering counts $Y_{sij.} = \sum_{k=1}^{K}$. 
 From \citet{solymos_calibrating_2013}, we model $Y_{sij.}$ as
-$$Y_{sij.} \sim multinomial\left(Y_{si..}, \Pi_{si.}\right).$$
+$$Y_{sij.} \sim multinomial\left(Y_{si..}, \mathbf{\Pi}_{si.}\right).$$
 
-\par For sample $i$, $\Pi_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $J$ time bands; in essense, it is the mixing parameter.
+\par For sample $i$, $\mathbf{\Pi}_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $J$ time bands; in essense, it is the mixing parameter.
 Let $t_{ij}$ be the maximum time for time band $j$ during sampling event $i$, and let $\phi_s$ be the unknown cue rate for species $s$.
-We then have the following calculation of each component $\pi_{sij}$ of the mixing parameter $\Pi_{si.}$:
+We then have the following calculation of each component $\pi_{sij}$ of the mixing parameter $\mathbf{\Pi}_{si.}$:
 
@@
 To allow for the partial pooling of estimates for each species to be based on a phylogenic relationship, we can use Gaussian processes, which effectively allow for "continuous" factors \citep{bernardo_regression_1998, mcelreath_continous_2020}.
-Let $\Phi$ be the vector of cue rates (of length $s$). 
+Let $\mathbf{\Phi}$ be the vector of cue rates (of length $s$). 
 We then have:
 
-$$\log \Phi \sim MVN\left( \mu_{[s]}, \Sigma_{[s \times s]} \right).$$
+$$\log \mathbf{\Phi} \sim MVN\left( \mathbf{\mu}_{[s]}, \mathbf{\Sigma}_{[s \times s]} \right).$$
 
-\par The covariance function $\Sigma$ (denoted with subscript $[s \times s]$ to show dimensionality) determines the strength of the partial pooling, and can be any function that contains a measure of distance.
-In this case, we will create a function based on phylogenetic distance $D_{[s \times s]}$, i.e.,
+\par The covariance function $\mathbf{\Sigma}$ (denoted with subscript $[s \times s]$ to show dimensionality) determines the strength of the partial pooling, and can be any function that contains a measure of distance.
+In this case, we will create a function based on phylogenetic distance $\mathbf{D}_{[s \times s]}$, i.e.,
 
-$$\Sigma_{[s \times s]} = D_{[s \times s]}\lambda\sigma.$$
+$$\mathbf{\Sigma}_{[s \times s]} = \mathbf{D}_{[s \times s]}\lambda\sigma.$$
 
-\par Here, $\lambda$ is Pagal's Lambda (CITATION), which determines the strength of the phylogenetic relationship. 
+\par Here, $\lambda$ is Pagal's Lambda \citep{pagel_inferring_1999}, which determines the strength of the phylogenetic relationship. 
 We set $\lambda = 0.76$, consistent with previous results \cite{solymos_phylogeny_2018}. 
@@
 
-\par Finally, the vector of mean cue rate $\mu_{[s]}$ is just a function of the species migratory strategy $\alpha_M$, where the migratory strategy $M$ could either be Resident or Migrant. 
+\par Finally, the vector of mean cue rate $\mathbf{\mu}_{[s]}$ is just a function of the species migratory strategy $\alpha_M$, where the migratory strategy $M$ could either be Resident or Migrant. 
 We set priors on both migratory strategies to be $\alpha_M \sim N(-1, 0.01)$, because on the log scale we would expect cue rate to be negative.
-
-\par Results of simulated priors can be found in Appendix TO DO.
 
@@
 
-\par For distance modelling, we are not interested in the time bin that the bird was recorded in, and so we will sum counts from each sampling event $i$ over all time bands. 
+\par For distance modelling, we are only interested in the distance to the cueing bird, and not the time bin that the bird was recorded in, and so we will sum counts from each sampling event $i$ over all time bands.
 Thus, we will be considering counts $Y_{si.k} = \sum_{j=1}^{J}$. 
 From \citet{solymos_calibrating_2013}, we model $Y_{si.k}$ as
-$$Y_{si.k} \sim multinomial\left(Y_{si..}, \Pi_{si.}\right).$$
+$$Y_{si.k} \sim multinomial\left(Y_{si..}, \mathbf{\Pi}_{si.}\right).$$
 
-\par For sample $i$, $\Pi_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $K$ distance bins; in essense, it is the mixing parameter.
+\par For sample $i$, $\mathbf{\Pi}_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $K$ distance bins; in essense, it is the mixing parameter.
 Let $r_{ik}$ be the maximum distance for distance bin $k$ during sampling event $i$, and let $\tau_s$ be the unknown effective detection radius for species $s$. 
-We then have the following calculation of each component $\pi_{sik}$ of the mixing parameter $\Pi_{si.}$:
-
-%\begin{equation*}
-%	\pi_{sik} = 
-%	\begin{cases}
-%		\dfrac{\left( 1 - \exp\left\{ -\dfrac{r_{i,k}^2}{\exp\left\{\tau_s^2\right\}} \right\} \right) - \left( 1 - \exp\left\{ -\dfrac{r_{i,k-1}^2}{\exp\left\{\tau_s^2\right\}} \right\} \right)}
-%		{ 1 - \exp\left\{ -\dfrac{r_{i,K}^2}{\exp\left\{\tau_s^2\right\}} \right\} } & \text{for } k > 1 \\
-%		1 - \sum_{n = 2}^{J} \pi_{sin} & \text{for } k = 1
-%	\end{cases}
-%\end{equation*}
+We then have the following calculation of each component $\pi_{sik}$ of the mixing parameter $\mathbf{\Pi}_{si.}$:
 
@@
 
-\par Results of simulated priors can be found in Appendix TO DO.
-
 \subsection{Single Species Models}
@@
 The single species models used the same likelihood statements as above for the count data, except rather than cue rate and EDR being estimated with shared information, they are estimated independently from each other.
-That is, $\log \phi \sim N(0,1) \forall \phi_{s}$, and $\log \tau \sim N(0,1) \forall \tau_s$, thus making it equivalent to the Null model in \citet{edwards_point_2023}, just estimated in a Bayesian framework rather than maximum likelihood.
+That is, $\log \phi_s \sim N(0,1) \forall s$, and $\log \tau_s \sim N(0,1) \forall s$, thus making it equivalent to the Null model in \citet{edwards_point_2023}, just estimated in a Bayesian framework rather than maximum likelihood.
 
@@
 This consists of $712138$ point counts conducted across $292$ projects in Canada and the United States, $422514$ of which used removal sampling and $522820$ of which used distance sampling.
+All of these point counts were standardized using practices developed by the Boreal Avian Modelling project \citep{barker_ecological_2015}.
 
@@
 We also requested that the phylogenies include the following bird species with insufficient data from NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird, and Spotted Owl.
-All seven of these bird species are listed on Partner's in Flight's watchlist for various threats (CITATION). 
+All seven of these bird species are listed on Partner's in Flight's watchlist for various threats \citep{will_handbook_2020}. 
 This resulted in 2000 pseudo-posterior trees that contained 323 species.
 
-\par We generated a phylogenetic correlation matrix $D$ using the same methods outlined in the Appendix of \citet{solymos_phylogeny_2018} that generate distances between species, and further processed using the R package "ape" \citet{paradis_ape_2019}.
+\par We generated a phylogenetic correlation matrix $\mathbf{D}$ using the same methods outlined in the Appendix of \citet{solymos_phylogeny_2018} that generate distances between species, and further processed using the R package "ape" \citet{paradis_ape_2019}.
 
 \subsubsection{Species Traits}
-\par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts of all the birds to model in this study (CITATION). 
-Body mass for each bird species was obtained form the Elton traits dataset (CITATION). 
+\par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts of all the birds to model in this study \citep{billerman_birds_2022}. 
+Body mass for each bird species was obtained form the Elton traits dataset \citep{wilman_eltontraits_2014}. 
 Where body mass was not available through the Elton traits dataset, the Birds of the World account was used instead.
@@
 In this case, we ran both removal model for 316 species of birds, and we ran the distance model for 317 species of birds (Tricoloured Blackbird was included in the distance models but not the removal models).
-We accessed the single-species estimates of cue rate and EDR from the NA-POPS Null models using the \texttt{napops} R package (CITATION).
+We accessed the single-species estimates of cue rate and EDR from the NA-POPS Null models using the \texttt{napops} R package \citep{edwards_napops_2023, edwards_point_2023}.
 We then modelled the difference in estimated cue rates and EDRs as
@@
 We then estimated multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR for each cross-validation run, and used these parameter estimates to calculate the posterior predictive density of each of the held out data points.
-We then calculated the difference in posterior predictive density for removal and distance models, by subtracting the pointwise posterior predictive density of the single-species model from the pointwise posterior predictive density of the multi-species model.
-Finally, we took the mean of the difference in pointwise posterior predictive density for each species to create a species-specific mean pointwise predictive density.
-With this, positive values of this mean would indicate that the multi-species model was preferred, and negative values would indicate that the single-species model was preferred.
-If prediction \ref{pred3} is true, then we would expect to see majority positive values of the specific-specific mean pointwise predictive density.
+We then calculated the difference in posterior predictive density for removal and distance models, by subtracting the pointwise posterior predictive density of the single-species model from the pointwise posterior predictive density of the multi-species model (i.e., multi-species minus single-species).
+Finally, we took the mean of the difference in pointwise posterior predictive density for each species to create a species-specific mean difference in pointwise predictive density.
+With this, positive values would indicate that the multi-species model was preferred, and negative values would indicate that the single-species model was preferred.
+If prediction \ref{pred3} is true, then we would expect to see majority positive values of the species-specific mean pointwise predictive density.
 
@@
 \subsection{Analysis}
-\par All analyses were performed in R version TO DO: CHECK (CITATION).
-Manipulation of phylogenetic trees, including finding a consensus tree for visualization and generating variance-covariance matrices based on the 2000 bootstrapped trees, was done using the \texttt{ape} R package.
-All Bayesian models were coded in the Stan Probabilistic Language \cite{stan_development_team_stan_2019}, and models were run using the \texttt{cmdstanr} R package (CITATION) and visualized using the \texttt{bayesplot} R package (CITATION).
+\par All analyses were performed in R version 4.2.2 \citep{r_core_team_r_2022}.
+Manipulation of phylogenetic trees, including finding a consensus tree for visualization and generating variance-covariance matrices based on the 2000 bootstrapped trees, was done using the \texttt{ape} R package \citep{paradis_ape_2019}.
+All Bayesian models were coded in the Stan Probabilistic Language \cite{stan_development_team_stan_2019}, and models were run using the \texttt{cmdstanr} R package \citep{gabry_cmdstanr_2022} and visualized using the \texttt{bayesplot} R package \citep{gabry_visualization_2019}.
 Each of the single-species models and multi-species models running both the initial comparison set of species and the final predictive set of species, as well as the models created to assess performance, were run using 1000 warmup iterations and 2000 sampling iterations, each on 4 chains for a total of 8000 draws per model run.
 Cross-validation models were run with 250 warmup iterations and 500 sampling iterations, each on 4 chains for a total of 2000 draws per model run.
-All of the models made use of Stan's reduce\_sum functionality, which allows for within-chain parallelization as well as the usual across-chain parallelization (CITATION).
+All of the models made use of Stan's reduce\_sum functionality, which allows for within-chain parallelization as well as the usual across-chain parallelization.
 This was particularly useful in our case as the multi-species removal model contained over 3.5 million data points and the multi-species distance model contained over 4.5 million data points, and so computational expense was a large concern.

### 2023-10-26 article/article.tex
added figure captions

@@
 Each of the single-species models and multi-species models running both the initial comparison set of species and the final predictive set of species, as well as the models created to assess performance, were run using 1000 warmup iterations and 2000 sampling iterations, each on 4 chains for a total of 8000 draws per model run.
+To mitigate issues surrounding initial values being rejected in the distance model (see Discussion), we set the initial values of EDR to be the mean EDR from NA-POPS estimates; for species where EDR estimates did not exist in NA-POPS, we used the overall mean EDR as initial value.
 Cross-validation models were run with 250 warmup iterations and 500 sampling iterations, each on 4 chains for a total of 2000 draws per model run.
 All of the models made use of Stan's reduce\_sum functionality, which allows for within-chain parallelization as well as the usual across-chain parallelization.
-This was particularly useful in our case as the multi-species removal model contained over 3.5 million data points and the multi-species distance model contained over 4.5 million data points, and so computational expense was a large concern.
+This was particularly useful in our case as the multi-species removal model contained over 3 million data points and the multi-species distance model contained over 4 million data points, and so computational expense was a large concern.
 All models were run on one of two multi-core processing servers, one running Ubuntu 20.04.4 LTS and one running Ubuntu 20.04.6 LTS.
@@
 Seven out of the 316 species considered for removal modelling had a percent change in cue rate of more than 20\%.
+These species were Willow Ptarmigan (\textit{Lagopus lagopus}), Rusty Blackbird (\textit{Euphagus carolinus}), Red-bellied Sapsucker (\textit{Sphyrapicus ruber}), Common Redpoll (\textit{Acanthis flammea}), Calliope Hummingbird (\textit{Selasphorus calliope}), Bohemian Waxwing (\textit{Bombycilla garrulus}), and Allen's Hummingbird (\textit{Selasphorus sasin}).
 All seven of these species were among the lowest cue rates in both the single-species and multi-species context (Figure \ref{fig:removal_1vs1}B).
@@
 	\includegraphics{../output/plots/removal_1vs1.png}
-	\caption{TO DO}
+	\caption{Modelled difference with 95\% credible interval (shaded area) between cue rates estimated from the multi-species model developed here, minus the single-species model in NA-POPS \citep{edwards_point_2023} (A); and 1-to-1 plot comparing cue rates estimated from the single-species model vs the multi-species model for each species (B).}
 	\label{fig:removal_1vs1}
@@
 	\includegraphics{../output/plots/removal_other_params.png}
-	\caption{TODO}
+	\caption{Posterior distributions with 95\% credible intervals (shaded area) of parameters estimated in the multi-species removal model, with overall mean log cue rate as the intercept (A), migratory strategy as departures from the intercept based on whether a species was a resident (mu\_mig\_strat[1]) or migrant (mu\_mig\_strat[B]) (B), and extra variance term after accounting for the phylogenetic variation (C).}
 	\label{fig:removal_params}
@@
 	\includegraphics{../output/plots/removal_sd_plot.png}
-	\caption{TODO}
+	\caption{Posterior distributions with 95\% credible intervals (shaded areas) of (A) modelled intercepts of standard deviation for the single-species removal model (alpha[1]) and multi-species removal model (alpha[2]), and (B) modelled slopes of standard deviation with increasing sample size for the single-species removal model (beta[1]) and multi-species removal model (beta[2]).}
 	\label{fig:removal_sd}
@@
 Single-species models tended to have better predicitve capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:removal_cv}).
-There did not appear to be any patterns in model selection among phylogeny or traits (see Appendix SOMETHING for full phylogenetic tree).
+There did not appear to be any patterns in model selection among phylogeny or traits (see Supplemental Figure 1 for full phylogenetic tree).
 Six out of the 7 species that had a percent change in cue rate of more than 20\% had the multi-species model chosen as the better predictive model.
@@
 	\includegraphics{../output/plots/cv_removal_diff_vs_n.png}
-	\caption{TODO}
+	\caption{Mean difference in pointwise posterior predictive density.
+Points are mean difference by species.
+Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species model preference, whereas negative differences (coloured in blue for clarity) reprepresent a single-species model preference.
+Regression line shows the relationship between mean difference and species sample size.}
 	\label{fig:removal_cv}
@@
 Still, the majority of the species had very little change; six out of the 317 species had a percent change in EDR of more than 20\% (Figure \ref{fig:distance_1vs1}B).
+These species were Sharp-tailed Grouse (\textit{Tympanuchus phasianellus}), Sedge Wren (\textit{Cistothorus stellaris}), Ring-necked Pheasant (\textit{Phasianus colchicus}), Chimney Swift (\textit{Chaetura pelagica}), Boat-tailed Grackle (\textit{Quiscalus major}), and Baird's Sparrow (\textit{Ammodramus bairdii}).
 
@@
 	\includegraphics{../output/plots/distance_1vs1.png}
-	\caption{TO DO}
+	\caption{Modelled difference with 95\% credible interval (shaded area) between EDR estimated from the multi-species model developed here, minus the single-species model in NA-POPS \citep{edwards_point_2023} (A); and 1-to-1 plot comparing EDR estimated from the single-species model vs the multi-species model for each species (B).}
 	\label{fig:distance_1vs1}
@@
 	\includegraphics{../output/plots/distance_other_params.png}
-	\caption{TODO}
+	\caption{Posterior distributions with 95\% credible intervals (shaded area) of parameters estimated in the multi-species distance model, with (A) overall mean log EDR as the intercept, with departures from the intercept modelled as migratory strategy (resident: mu\_mig\_strat[1], migrant: mu\_mig\_strat[B]), (B) slope parameters of changes in EDR given change in mass and pitch, and (C) extra variance term after accounting for traits.}
 	\label{fig:distance_params}
@@
 	\includegraphics{../output/plots/distance_sd_plot.png}
-	\caption{TODO}
+	\caption{Posterior distributions with 95\% credible intervals (shaded areas) of (A) modelled intercepts of standard deviation for the single-species distance model (alpha[1]) and multi-species distance model (alpha[2]), and (B) modelled slopes of standard deviation with increasing sample size for the single-species distance model (beta[1]) and multi-species distance model (beta[2]).}
 	\label{fig:distance_sd}
@@
 Multi-species models tended to have better predicitve capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:distance_cv}).
-Interestingly, only two of the six species with a greater than 20\% change in EDR had the multi-species model chosen as the better predictive model (NOTE this could change as new results come in).
-TO DO: Any patterns among traits?
-
+Interestingly, only two (Chimney Swift and Boat-tailed Grackle) of the six species with a greater than 20\% change in EDR had the multi-species model chosen as the better predictive model.
 \begin{figure}[h!]
 	\includegraphics{../output/plots/cv_distance_diff_vs_n.png}
-	\caption{TODO}
+	\caption{Mean difference in pointwise posterior predictive density.
+Points are mean difference by species.
+Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species distance model preference, whereas negative differences (coloured in blue for clarity) represent a single-species distance model preference.
+Regression line shows the relationship between mean difference and species sample size.}
 	\label{fig:distance_cv}
@@
 	\includegraphics{../output/plots/removal_predictions_plot.png}
-	\caption{TODO}
+	\caption{Cue rates (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for seven species of rare and undersampled North American landbirds.
+Species are 4 letter banding codes: SPOW = Spotted Owl (n = 7), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), TRBL = Tricoloured Blackbird (n = 74), HASP = Harris's Sparrow (n = 65), KIWA = Kirtland's Warbler (n = 0), and LEPC = Lesser Prairie-chicken (n = 45).}
 	\label{fig:removal_predictions}
@@
 	\includegraphics{../output/plots/distance_predictions_plot.png}
-	\caption{TODO}
+	\caption{Effective detection radius (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for six species of rare and undersampled North American landbirds.
+Species are 4 letter banding codes: SPOW = Spotted Owl (n = 28), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), HASP = Harris's Sparrow (n = 13), KIWA = Kirtland's Warbler (n = 1), and LEPC = Lesser Prairie-chicken (n = 45).}
 	\label{fig:distance_predictions}
@@
 We applied the multi-species QPAD removal model and multi-species QPAD distance model to the large dataset of point counts collected by the NA-POPS project \citep{edwards_point_2023}, and compared estimates of cue rate and EDR from a multi-species context to estimates of cue rate and EDR from the previous single species context.
-Given the results of these comparisons, we sought to predict these components of detectability for seven species of North American landbird that are rare and undersampled.
+Given the results of these comparisons, we predicted these components of detectability for seven species of North American landbird that are rare and undersampled.
 
@@
 \par Estimates of cue rate changed very little from species that were originally modelled by the single-species models in \citet{edwards_point_2023}, to the same ones modelled in the multi-species model here.
-The 5-fold cross validation tended to show either no preference for one model or another when sample size was large, and a preference (albeit an extremely small preference) for single-species models when sample size was smaller.
+The 5-fold cross validation tended to show no preference for one model or another when sample size was large, and a preference (albeit an extremely small preference) for single-species models when sample size was smaller.
 When sample size was small, a multi-species model tended to produce more precise estimates of cue rate.
@@
 For removal modelling, all those species except for Common Redpoll ended up having the multi-species model chosen as the more predictive model, indicating that sharing information among related species tended to improve their estimates of cue rate.
-Common Redpoll was likely an exception to this because the original estimate of cue rate in NA-POPS was a wild underestimate (cue rate of TO DO), likely because of a convergence issue in the \texttt{detect} package.
+Common Redpoll was likely an exception to this because the original estimate of cue rate in NA-POPS was a wild underestimate (cue rate of $5.9\times 10^{-9}$), likely because of a convergence issue in the \texttt{detect} package.
 It could be worth running Common Redpoll using the Bayesian single species model developed here to see if the estimate of cue rate is still extremely low, or if the vague but still somewhat informative prior helps to bring the estimated parameter into a more sensible space.

### 2023-10-26 article/article.tex
moved model performance to the top of the discussion section

@@
 
+\subsection{Model Performance}
+
+\par The multi-species QPAD Distance model had several issues related to initialization, sampling, and convergence throughout this study.
+For many iterations of the model development, Stan's generation of initial values proved difficult.
+For all unconstrained parameters, Stan draws initial values from a uniform(-2,2) distribution.
+On the scale that we modelled detection distances (i.e., log detection distance in hundreds of metres), these are reasonable values.
+At the low end, an initial value of -2 would correspond to a detection distance of $e^{-2} \times 100 = 13.5$ m, which is low but somewhat plausible for a small-bodied, high-pitched species such as a hummingbird.
+At the high end, an initial value of 2 would correspond to a detection distance of $e^{2} \times 100 = 739$ m, which is high but still somewhat plausible for large-bodied, low-pitched species like a grouse or prairie-chickens.
+However, where Stan seemed to struggle is when these randomly assigned values for detection distances were effectively impossible for a large set of species.
+That is to say, if the values were initialized such that hummingbirds had a detection distance of 500 m or more, and grouse or prairie-chickens had a detection distance of 15 m or less, this would result in an impossible likelihood, forcing Stan to reject the initial values.
+Our fix was to set initial values to be jittered mean detection distances originally estimated in \citet{edwards_point_2023}; for species without estimates, we used the mean EDR for all species.
+This approach seemed to work as the initial values were no longer rejected, but we caution that this could very well be a "bandaid solution" to an underlying issue with the model itself.
+
+\par The distance models also seemed to struggle with sampling the posterior distribution, in that over a quarter of the iterations resulted in a divergent transition (CITATION for divergent transitions).
+This is likely due to hierarchical nature of the model and the parameterization choices that we made.
+\citet{betancourt_hierarchical_2020} discuss at length some issues that arise when sampling from hierarchical models in cases where the hierarchical groups are data rich, where a "centred parameterization" should be used, or data poor, where a "non-centred parameterization" should be used.
+We experimented with using only a non-centred parameterization which resulted in extremely low effective sample sizes (as low as 4 for 8000 iterations), large R-hat values, and took an unreasonable amount of time to finish sampling (approximately 2.5 weeks).
+The model presented in this paper is a centred parameterization, which produced much more reasonable effective sample sizes and R hat values, finished within a day and a half, but still suffered from degeneracies such as hierarchical funneling as described by \citet{betancourt_hierarchical_2020}.
+In our case, because we are modelling on a hierarchy that contain many data points for some species x trait groups and fewer data points for other species x trait groups, our divergent transitions may be resolved if we were to implement our model using a mixture of centred and non-centred parametrization as described by \citet{betancourt_hierarchical_2020}.
+
+\par For removal modelling, we experience no issues related to initialization, sampling, or convergence. 
+The model finished within approximately 24 hours, and all parameters had R-hat values of 1, or within 0.001 of it.
+We note that even though we used a Gaussian process to estimate cue rate, we did not have to estimate the covariance matrix used for the Gaussian process, because it was effectively a known parameter provided by the phylogenetic relationship.
+This would account for the relatively quick performance of this model given the number of parameters to estimate and the amount of data used; had the covariance matrix needed to be estimated, this model would have likely taken much longer to complete.
+
 \subsection{Removal Modelling}
@@
 \subsection{Distance Modelling}
-\par Estimates of EDR also changed very little from species that were originally modelled by the single-species models in \citet{edwards_point_2023}, to the same ones modelled in the multi-species model here.
+\par Estimates of EDR also changed very little, albeit a non-zero amount, from species that were originally modelled by the single-species models in \citet{edwards_point_2023}, to the same ones modelled in the multi-species model here.
 However, there was evidence for an average small downward shift in EDRs when modelled under a multi-species context.
@@
 
-\par Similar to multi-species removal modelling, estimates of EDR fell roughly where we would expect based on species of similar traits, for the species where we were relaxing samples sizes.
+\par Estimates of EDR also fell roughly where we would expect based on species of similar traits, for the species where we were relaxing samples sizes.
 For example, Harris's Sparrow (n = 13) had an estimated mean EDR of 65.6 m, with a 95\% credible interval of [53.1, 82.5].
@@
 \par Using the predictions of cue rate and EDR for Bicknell's Thrush and Kirtland's Warbler, we can calculate availability and perceptibility, and hence detectability, for a given survey type, and compare to previous estimates of detectability obtained from other studies.
-As a reminder, following derivations in \citet{solymos_calibrating_2013}, we note that detectability is the product of availability and perceptibility.
+As a reminder, following derivations in \citet{solymos_calibrating_2013}, detectability is the product of availability and perceptibility.
 Availability $p(t)$ is defined as the probability of a bird giving a cue during a survey run for $t$ minutes, and is calculated as
@@
 We can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respectively credible intervals) to get a range of detectabilities from 0.09 through 0.70.
-A study by \citet{aubry_bicknells_2018} noted that they had estimated detection probabilities $\geq$ 0.74 and $\leq$ 0.88, and so even our highest potential value of detectability falls short of their estimated range.
+\citet{aubry_bicknells_2018}, however, estimate detection probabilities of $\geq$ 0.74 and $\leq$ 0.88, and so even our highest potential value of detectability falls short of their estimated range.
 However, we note that \citet{aubry_bicknells_2018} are using data that specifically target Bicknell's Thrush in their range, and they are using surveys conducted at peak Bicknell's Thrush breeding time, and so it is likely that detectability will be high in that specific case.
@@
 
-Both these examples highlight the need for future iterations of this multi-species model to consider environmental and temporal covariates that affect peaks in detectability through the year and for different environmental conditions.
+\subsection{Improvements to this model}
+
+Both examples in the previous section highlight the need for future iterations of this multi-species model to consider environmental and temporal covariates that affect peaks in detectability through the year and for different environmental conditions.
 This is especially true for the types of species we are interested in generating predictions for, i.e., rare and/or undersampled species.
-In general, rare and undersampled species will likely prefer rare habitats or be confined to a very small range (as is the case with Bicknell's Thrush and Kirtland's Warbler above).
+In general, rare species will prefer rare habitats, or will at least be confined to a very small range (as is the case with Bicknell's Thrush and Kirtland's Warbler above).
 Thus, while we can generate reasonable estimates of the components of detectability (i.e., availability through cue rate and perceptibility through EDR) based on information borrowed from similar species (either through phylogeny or traits), this model as it stands may fail to glean more specific information that takes into account detectability peaks throughout the year, or detectability peaks given specific habitats.
@@
 Similarly, this may also provide a unique opportunity to make use of traditional ecological knowledge to further inform predictions of detectability.
-For example, in the Anishinaabe language, bird taxonomy is based both on physical features, but also behaviours such as flight, calls, morning rituals, and weather (CITATION Joe Pitawanakwat presentation).
+For example, in the Anishinaabe language, bird taxonomy is based both on physical features, but also behaviours such as flight, calls, morning rituals, and weather \citep{pitawanakwat_evening_2022}.
 This taxonomy is based off observations of bird behaviour that has been passed down through generations, and may vary slightly (or heavily) from taxonomy based on genomics.
 In fact, it could be argued that a taxonomy based on how humans observe birds may actually be more relevant for detectability, and so a future opportunity could be to adopt a two-eyed seeing approach \cite{reid_twoeyed_2021} to predicting detectability in rare birds.
-
-\subsection{Model Performance}
-
-\par The multi-species QPAD Distance model had several issues related to initialization, sampling, and convergence throughout this study.
-For many iterations of the model development, Stan's generation of initial values proved difficult.
-For all unconstrained parameters, Stan draws initial values from a uniform(-2,2) distribution.
-On the scale that we modelled detection distances (i.e., log detection distance in hundreds of metres), these are reasonable values.
-At the low end, an initial value of -2 would correspond to a detection distance of $e^{-2} \times 100 = 13.5$ m, which is low but somewhat plausible for a small-bodied, high-pitched species such as a hummingbird.
-At the high end, an initial value of 2 would correspond to a detection distance of $e^{2} \times 100 = 739$ m, which is high but still somewhat plausible for large-bodied, low-pitched species like a grouse or prairie-chickens.
-However, where Stan seemed to struggle is when these randomly assigned values for detection distances were effectively impossible for a large set of species.
-That is to say, if the values were initialized such that hummingbirds had a detection distance of 500 m or more, and grouse or prairie-chickens had a detection distance of 15 m or less, this would result in an impossible likelihood, forcing Stan to reject the initial values.
-Our fix was to set initial values to be jittered mean detection distances originally estimated in \citet{edwards_point_2023}; for species without estimates, we used the mean EDR for all species.
-This approach seemed to work as the initial values were no longer rejected, but we caution that this could very well be a "bandaid solution" to an underlying issue with the model itself.
-
-\par The distance models also seemed to struggle with sampling the posterior distribution, in that over a quarter of the iterations resulted in a divergent transition (CITATION for divergent transitions).
-This is likely due to hierarchical nature of the model and the parameterization choices that we made.
-\citet{betancourt_hierarchical_2020} discuss at length some issues that arise when sampling from hierarchical models in cases where the hierarchical groups are data rich, where a "centred parameterization" should be used, or data poor, where a "non-centred parameterization" should be used.
-We experimented with using only a non-centred parameterization which resulted in extremely low effective sample sizes (as low as 4 for 8000 iterations), large R-hat values, and took an unreasonable amount of time to finish sampling (approximately 2.5 weeks).
-The model presented in this paper is a centred parameterization, which produced much more reasonable effective sample sizes and R hat values, finished within a day and a half, but still suffered from degeneracies such as hierarchical funneling as described by \citet{betancourt_hierarchical_2020}.
-In our case, because we are modelling on a hierarchy that contain many data points for some species x trait groups and fewer data points for other species x trait groups, our divergent transitions may be resolved if we were to implement our model using a mixture of centred and non-centred parametrization as described by \citet{betancourt_hierarchical_2020}.
-
-\par For removal modelling, we experience no issues related to initialization, sampling, or convergence. 
-The model finished within approximately 24 hours, and all parameters had R-hat values of 1, or within 0.001 of it.
-We note that even though we used a Gaussian process to estimate cue rate, we did not have to estimate the covariance matrix used for the Gaussian process, because it was effectively a known parameter provided by the phylogenetic relationship.
-This would account for the relatively quick performance of this model given the number of parameters to estimate and the amount of data used; had the covariance matrix needed to be estimated, this model would have likely taken much longer to complete.
 

### 2023-10-26 article/article.tex
finished up discussion section, some things still to do before sending around for draft

@@
 
-\par For distance modelling, the predicted EDR for Bicknell's Thrush (73.1 m) and the predicted EDR for Kirtland's Warbler (75.0 m) fell reasonably close to the overall mean EDR of 92.7 m for all other 11 Turdidae species and 61.2 m for all 44 other Parulidae species in NA-POPS, respectively.
+\par TO DO For distance modelling, the predicted EDR for Bicknell's Thrush (73.1 m) and the predicted EDR for Kirtland's Warbler (75.0 m) fell reasonably close to the overall mean EDR of 92.7 m for all other 11 Turdidae species and 61.2 m for all 44 other Parulidae species in NA-POPS, respectively.
 That is, in the absence of any data (or with very little data in the case of Kirtland's Warbler), the estimated EDR of Bicknell's Thrush and Kirtland's Warbler was s
@@
 \subsection{Conservation Implications}
+\par Detectability often goes unaccounted for in conservation problems, often because the estimates of detection probability simply do not exist, or they are extremely uncertain \citep{bennett_how_2023}.
+This model provides an excellent opportunity for detection probabilities to be estimated with higher precision for several species of landbirds in North America, and also provides an opportunity to predict detection probabilities for species which have very little (or even no) data.
+This means that conservation problems that rely on estimates of detectability such as Value of Information analyses \citep{canessa_when_2015, bennett_when_2018} or prioritizations \citep{hanson_prioritizr_2023} can look to make more accurate decisions.
+For organizations such as Partners in Flight, Boreal Avian Modelling project (\citep{cumming_toward_2010}), and others that rely on detectability information for bird population size estimates, these improved estimates (and new estimates and predictions) provide a way forward for continuing to produce accurate estimates of population size that are based on data-driven detectability estimates.
+
+\par The model presented here can be used to fill out the remaining missing species in the NA-POPS database, in order to estimate detectability for every species of North American landbird.
+By making use of phylogeny and trait information, bird species with fewer than 75 data points can have estimates generated based on the few data points available, and information available from other species.
+As shown here, birds with no data points can have detectability predicted that falls closely in line with similar species.
+We caution that birds that are to have detectability predicted should be noted as such on any communications, such that organizations know that the estimate is simply a "best guess" based on other similar species.
+
+\par Finally, we see this model as a proof-of-concept for use with other taxa.
+Obviously, given the vast amount of bird data available, it was easy enough to obtain estimates for enough data-rich species such that the few data-poor species could have enough information to borrow.
+However, one benefit that we have shown here is that Bayesian models can still come up with reasonable predictions if there are informed enough priors on which to fall back.
+That is to say, for a taxon such as butterflies where there are some species with data and others with very little data, but there is overall very little data in general (CITE Lewthwaite paper(s)), with reasonably-informed priors, a researcher may still be able to glean reasonable estimates of detectability with the little data available.
+We suggest that this type of model where information about detectability is shared among species be tested with taxon that generally have fewer data collected.
 

### 2023-10-27 article/article.tex
final touches for draft 1

@@
 Some examples of species not modelled include Kirtland's Warbler (\textit{Setophaga kirtlandii}) and Bicknell's Thrush (\textit{Catharus bicknelli}), both of which have low population sizes, and LeConte's Thrasher (\textit{Toxostoma lecontei}) and Harris's Sparrow (\textit{Zonotrichia querula}), both of which have rare habitat associations \citep{will_handbook_2020}.
-Of course, despite not having good estimates of detectability availabile for these rare and undersampled species, these are the groups of species that need these estimates the most as their conservation status often depend on accurate population size estimates, and estimates of trend in those population sizes.
+Of course, despite not having good estimates of detectability available for these rare and undersampled species, these are the groups of species that need these estimates the most as their conservation status often depend on accurate population size estimates, and estimates of trend in those population sizes.
 
@@
 We also developed a Bayesian single species version of the QPAD Removal and Distance models (i.e., as opposed to the maximum likelihood models available in the \texttt{detect} R package; \cite{solymos_detect_2020}) to allow for direct comparison through cross validation.
-Finally, we ran these models with the full set of data points used in \cite{edwards_point_2023}, plus additional data for seven species that did not meet the initial sample size thresholds for NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird (\textit{Agelaius tricolor}), and Spotted Owl (\textit{Strix occidentalis}).
+Finally, we ran these models with the full set of data points used in \cite{edwards_point_2023}, plus additional data for seven species that did not meet the initial sample size thresholds for NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Harris's Sparrow, Kirtland's Warbler, Tricoloured Blackbird (\textit{Agelaius tricolor}), and Spotted Owl (\textit{Strix occidentalis}).
 All seven of these bird species are listed on Partner's in Flight's watchlist for various threats \citep{will_handbook_2020}.
@@
 
-\par For any point count that employs removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002} and/or distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}, let $Y_{sijk}$ be the observed count of species $s$ during sampling event $i$, occuring in time band $j \in [1,J]$ and/or distance band $k \in [1,K]$.
+\par For any point count that employs removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002} and/or distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}, let $Y_{sijk}$ be the observed count of species $s$ during sampling event $i$, occurring in time band $j \in [1,J]$ and/or distance band $k \in [1,K]$.
 That is, if we have a point count where removal sampling or distance sampling (but not both) is not used, then we have that either $J = 1$ or $K = 1$, respectively; whereas, if we have a point count where both removal sampling and distance sampling is used, then we have that both $J,K \geq 2$.
@@
 
-\par For sample $i$, $\mathbf{\Pi}_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $J$ time bands; in essense, it is the mixing parameter.
+\par For sample $i$, $\mathbf{\Pi}_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $J$ time bands; in essence, it is the mixing parameter.
 Let $t_{ij}$ be the maximum time for time band $j$ during sampling event $i$, and let $\phi_s$ be the unknown cue rate for species $s$.
@@
 
-\par For sample $i$, $\mathbf{\Pi}_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $K$ distance bins; in essense, it is the mixing parameter.
+\par For sample $i$, $\mathbf{\Pi}_{si.}$ is the corresponding probability vector that determines how the total number of species $s$ observed at sample $i$ (i.e., $Y_{si..}$) are to be distributed across the $K$ distance bins; in essence, it is the mixing parameter.
 Let $r_{ik}$ be the maximum distance for distance bin $k$ during sampling event $i$, and let $\tau_s$ be the unknown effective detection radius for species $s$. 
@@
 \par We accessed 2000 pseudo-posterior trees from \citet{jetz_global_2012} that contains phylogeny for all species that contained sufficient data for removal modelling ($>75$ observations as per \cite{edwards_point_2023,solymos_evaluating_2018,buckland_introduction_2001}).
-We also requested that the phylogenies include the following bird species with insufficient data from NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird, and Spotted Owl.
+We also requested that the phylogenies include the following bird species with insufficient data from NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricoloured Blackbird, and Spotted Owl.
 All seven of these bird species are listed on Partner's in Flight's watchlist for various threats \citep{will_handbook_2020}. 
@@
 Mean overall difference in posterior predictive density was $-2.21 \times 10^{-5}$.
-Single-species models tended to have better predicitve capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:removal_cv}).
+Single-species models tended to have better predictive capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:removal_cv}).
 There did not appear to be any patterns in model selection among phylogeny or traits (see Supplemental Figure 1 for full phylogenetic tree).
@@
 Points are mean difference by species.
-Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species model preference, whereas negative differences (coloured in blue for clarity) reprepresent a single-species model preference.
+Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species model preference, whereas negative differences (coloured in blue for clarity) represent a single-species model preference.
 Regression line shows the relationship between mean difference and species sample size.}
@@
 However, similar to the removal model cross-validation results, the overall difference in posterior predictive density was very small, with a mean difference of $1.85 \times 10^{-4}$.
-Multi-species models tended to have better predicitve capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:distance_cv}).
+Multi-species models tended to have better predictive capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:distance_cv}).
 Interestingly, only two (Chimney Swift and Boat-tailed Grackle) of the six species with a greater than 20\% change in EDR had the multi-species model chosen as the better predictive model.
@@
 
-\par The distance models also seemed to struggle with sampling the posterior distribution, in that over a quarter of the iterations resulted in a divergent transition (CITATION for divergent transitions).
+\par The distance models also seemed to struggle with sampling the posterior distribution, in that over a quarter of the iterations resulted in a divergent transition \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
 This is likely due to hierarchical nature of the model and the parameterization choices that we made.
@@
 We experimented with using only a non-centred parameterization which resulted in extremely low effective sample sizes (as low as 4 for 8000 iterations), large R-hat values, and took an unreasonable amount of time to finish sampling (approximately 2.5 weeks).
-The model presented in this paper is a centred parameterization, which produced much more reasonable effective sample sizes and R hat values, finished within a day and a half, but still suffered from degeneracies such as hierarchical funneling as described by \citet{betancourt_hierarchical_2020}.
+The model presented in this paper is a centred parameterization, which produced much more reasonable effective sample sizes and R hat values, finished within a day and a half, but still suffered from degeneracies such as hierarchical funnelling as described by \citet{betancourt_hierarchical_2020}.
 In our case, because we are modelling on a hierarchy that contain many data points for some species x trait groups and fewer data points for other species x trait groups, our divergent transitions may be resolved if we were to implement our model using a mixture of centred and non-centred parametrization as described by \citet{betancourt_hierarchical_2020}.
@@
 When sample size was small, a multi-species model tended to produce more precise estimates of cue rate.
-Given the extremely small preference for single-species models under a lower sample size, it would appear that the bias-variance tradeoff is likely a welcome tradeoff here, because the decrease in variance is met by only a very slight increase in bias.
+Given the extremely small preference for single-species models under a lower sample size, it would appear that the bias-variance trade-off is likely a welcome trade-off here, because the decrease in variance is met by only a very slight increase in bias.
 
@@
 In our case, 41 out of the 79 species that had the lowest 25\% quantile of cue rate in the single-species model ended up having the multi-species model chosen as the preferred model.
-TO DO Re-read results of Martin-Schwarze.
 
 \par Although this was not a primary goal of our study, we found much less support than \citet{solymos_phylogeny_2018} for migration strategy having an effect on cue rate.
-Both the estimate effects for resident species and migratory species were effectively 0, with very little of the density above or below 0.025 of 0.
+Both the estimate effects for resident species and migratory species were effectively 0, with very little of the density above or below 0.025 from 0.
 One potential reason is that we had proportionally more resident species in our analysis (85 out of 316; 26.9\%) compared to \citet{solymos_phylogeny_2018} (23 out of 141; 16.3\%), and so the richer information provided by these additional species could have brought the effect down to 0.
@@
 
-\par TO DO For distance modelling, the predicted EDR for Bicknell's Thrush (73.1 m) and the predicted EDR for Kirtland's Warbler (75.0 m) fell reasonably close to the overall mean EDR of 92.7 m for all other 11 Turdidae species and 61.2 m for all 44 other Parulidae species in NA-POPS, respectively.
-That is, in the absence of any data (or with very little data in the case of Kirtland's Warbler), the estimated EDR of Bicknell's Thrush and Kirtland's Warbler was s
+\par For distance modelling, the predicted EDR for Bicknell's Thrush (73.1 m) and the predicted EDR for Kirtland's Warbler (75.0 m) fell reasonably close to the overall mean EDR of 92.7 m for all other 11 Turdidae species and 61.2 m for all 44 other Parulidae species in NA-POPS, respectively.
+That is, in the absence of any data (or with very little data in the case of Kirtland's Warbler), the estimated EDR of Bicknell's Thrush and Kirtland's Warbler was similar to those of other Thrushes and Warblers, respectively, but with departures from overall mean based on specific traits.
 
@@
 Similar to our Bicknell's Thrush example, we can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respective credible intervals) to obtain a range of detectabilities from 0.13 through 0.67.
-A study by \citet{van_dyke_comparative_2022} found esitmated detection probabilities for Kirtland's Warbler that ranged from effectively 0, through to a detection probability of 0.73.
+A study by \citet{van_dyke_comparative_2022} found estimated detection probabilities for Kirtland's Warbler that ranged from effectively 0, through to a detection probability of 0.73.
 In that study, they compared estimates of detectability among different jack pine stands and red pine stands, where the detectability was highest in jack pine stands than the red pine stands.
@@
 \par The examples above additionally highlight an excellent opportunity to make use of a powerful feature of Bayesian models, that being the inclusion of expert knowledge as prior information.
-For example, if a Bicknell's Thrush expert knows that there is something particular about how often they sing or give a cue compared to other thrushes, or how disproportionately quiet or loud they are comapred to species with similar traits, this can be coded directly into the prior distribution.
+For example, if a Bicknell's Thrush expert knows that there is something particular about how often they sing or give a cue compared to other thrushes, or how disproportionately quiet or loud they are compared to species with similar traits, this can be coded directly into the prior distribution.
 Similarly, this may also provide a unique opportunity to make use of traditional ecological knowledge to further inform predictions of detectability.
@@
 However, one benefit that we have shown here is that Bayesian models can still come up with reasonable predictions if there are informed enough priors on which to fall back.
-That is to say, for a taxon such as butterflies where there are some species with data and others with very little data, but there is overall very little data in general (CITE Lewthwaite paper(s)), with reasonably-informed priors, a researcher may still be able to glean reasonable estimates of detectability with the little data available.
+That is to say, for a taxon such as butterflies where there are some species with data and others with very little data, but there is overall very little data in general \citep{lewthwaite_geographical_2022} with reasonably-informed priors, a researcher may still be able to glean reasonable estimates of detectability with the little data available.
 We suggest that this type of model where information about detectability is shared among species be tested with taxon that generally have fewer data collected.

### 2023-11-30 article/article.tex
revisions of introduction and methods

@@
 
-\par Detectability is an important metric for many conservation problems, but often goes unaccounted for in studies, or detectability data just does not exist \citep{bennett_how_2023}.
+\par Accurate and precise estimates of detectability are important both for making efficient conservation decisions, and estimating accurate population sizes.
 In conservation management situations, detectability can be used as a metric of sensitivity to evaluate when to monitor a system for a species, or when to act on current information \citep{canessa_when_2015, bennett_when_2018}.
 When estimating population sizes of an organism, detectability often comes into play as a way to convert observed counts into densities or estimates of true abundance \citep{solymos_calibrating_2013, johnson_defense_2008}.
-As such, accurate and precise estimates of detectability are important both for making accurate conservation decisions, and estimating accurate population sizes.
-
-\par The wealth of bird data that has been collected and is available in North America means that estimates of detectability for birds are relatively easy to come by \citep{bennett_how_2023}.
-The NA-POPS project is an effort to estimate detectability functions that take into account variability in detection through environmental and temporal covariates, for as many of North America's landbird species as possible \citep{edwards_point_2023}.
+However, despite its importance, detectability often goes unaccounted for in studies, or detectability data just does not exist \citep{bennett_how_2023}.
+The wealth of openly available bird data in North America means that estimates of detectability for birds are relatively easy to come by, in comparison to most other taxonomic groups \citep{bennett_how_2023}.
+The NA-POPS project is an effort to estimate detectability functions that take into account variability in detection through environmental and temporal covariates, for as many of North America’s landbird species as possible \citep{edwards_point_2023}.
 To date, the project has estimated availability (the probability a bird sings or gives a cue) and perceptibility (the probability an observer sees or hears a cueing bird) for 319 species of birds, with an additional 19 species having estimates of availability or perceptibility (but not necessarily both).
@@
 
-\par Despite the efforts of the NA-POPS project, the 338 species of birds considered in the modelling exercise still only account for about 75\% of Partners in Flight's 2016 Landbird Conservation Plan \citep{edwards_point_2023, rosenberg_partners_2016}.
+\par Despite the efforts of the NA-POPS project, approximately 25\% of the species in the Partners in Flight’s 2016 Landbird Conservation Plan \citep{rosenberg_partners_2016} have insufficient data to directly estimate detectability \citep{edwards_point_2023}.
 This is because of sample size limitations that are recommended for removal and distance modelling, in that a modeller should have at least 75 data points for reasonable estimates \citep{buckland_introduction_2001, solymos_calibrating_2013}.
-Many of the species that fell short of that threshold in the NA-POPS database are systematically rare or undersampled species, that are either low in population size to begin with, or occur in rare habitats.
-Some examples of species not modelled include Kirtland's Warbler (\textit{Setophaga kirtlandii}) and Bicknell's Thrush (\textit{Catharus bicknelli}), both of which have low population sizes, and LeConte's Thrasher (\textit{Toxostoma lecontei}) and Harris's Sparrow (\textit{Zonotrichia querula}), both of which have rare habitat associations \citep{will_handbook_2020}.
-Of course, despite not having good estimates of detectability available for these rare and undersampled species, these are the groups of species that need these estimates the most as their conservation status often depend on accurate population size estimates, and estimates of trend in those population sizes.
+Many of the species that fell short of that threshold in the NA-POPS database are rare or undersampled species, that are either have low population sizes, or occur in rare or inaccessible habitats.
+Some examples of species not modelled include Kirtland’s Warbler (\textit{Setophaga kirtlandii}) and Bicknell’s Thrush (\textit{Catharus bicknelli}), both of which have low population sizes, and LeConte’s Thrasher (\textit{Toxostoma lecontei}) and Harris’s Sparrow (\textit{Zonotrichia querula}), both of which have rare habitat associations \citep{will_handbook_2020}.
+Such species are often conservation concerns, and thus detectability information for them would be particularly important.
 
-\par Detectability in birds has been shown to at least be partly driven by variation in traits.
-\cite{johnston_species_2014} showed in a selection of European landbirds that body mass and habitat association had an effect on the detection distance, which influences the perceptibility.
+\par Detectability in birds has been shown to be at least partly driven by variation in traits.
+\cite{johnston_species_2014} showed in a selection of European landbirds that body mass and habitat association had an effect on detection distance, which influences perceptibility
 In this case, a larger body size tended to be associated with larger detection distances, which makes sense considering larger body sizes are associated with louder and lower-pitched songs in birds \citep{bowman_adaptive_1979, fletcher_acoustics_1999, ryan_role_1985}.
-Habitat associations tends to influence the attenuation of sound, whereby sound diminishes quicker in a closed-forested habitat compared to a more open habitat \citep{waide_tropical_1988, yip_sound_2017}.
+Habitat associations tend to influence the attenuation of sound, whereby sound diminishes quicker in a closed-forested habitat than more open habitat \citep{waide_tropical_1988, yip_sound_2017}.
 A study by \cite{solymos_phylogeny_2018} expanded on this effort by also considering the effects of species traits and phylogeny on both detection distance and cue rate, the latter of which influences the availability of birds.
-They found that cue rate is most influenced by phylogeny, and, to a lesser extent, whether a bird was a long-distance migrant or a resident species.
-They also found that while the best model that explained detection distance did include phylogeny, the strength of the phylogenetic relationship was reduced to effectively 0 after accounting for traits such as body mass, song pitch, habitat association, and migratory status.
+They found that cue rate is most related to phylogeny, and, to a lesser extent, whether a bird was a long-distance migrant or a resident species.
+They also found detection distance was related to species’ traits such as body mass, song pitch, habitat association, and migratory status and that after accounting for these traits, phylogeny explained very little.
 
-\par One recommendation for future research that came out of \cite{solymos_phylogeny_2018} was to try to use these phylogenetic and trait relationships in detectability to then predict components of detectability (i.e., availability through cue rate, and perceptibility through detection distance) in other birds.
-Indeed, by making use of shrinkage factors and partial pooling, such as what can be done in a Bayesian hierarchical models \citep{gelman_what_2021}, and by accounting phylogenetic relationships and the relative "relatedness" of different species, such as what can be done with Gaussian processes \citep{bernardo_regression_1998, mcelreath_continous_2020}, we can leverage the information provided by species with many data points to use to inform species with much fewer (included potentially zero) data points.
+\par One recommendation for future research that came out of \cite{solymos_phylogeny_2018} was to use these phylogenetic and trait relationships with detectability to predict components of detectability (i.e., availability through cue rate, and perceptibility through detection distance) in other birds.
+Indeed, by making use of shrinkage factors and partial pooling, as can be done in Bayesian hierarchical models \citep{gelman_what_2021}, and by accounting for phylogenetic relationships and the relative "relatedness" of different species, as can be done with Gaussian processes \citep{bernardo_regression_1998, mcelreath_continous_2020}, we can leverage the information provided by species with many data to inform estimates for species with much fewer (included potentially zero) data points.
 
-\par The goal of this paper is to create a multi-species model of detectability that incorporates species traits and phylogeny, to both allow for the relaxation of the sample size requirements for estimating detectability, and to predict estimates of detectability for bird species without any data.
+\par The goal of this paper is to create a multi-species model of detectability that incorporates species traits and phylogeny, to both allow for the relaxation of the sample size requirements for estimating detectability, and to derive estimates of detectability for bird species without any data.
 To do so, we developed a Bayesian hierarchical version of the QPAD Removal model initially derived in \cite{solymos_calibrating_2013}, that allows for information exchange between species based on phylogenetic distance and binned migratory status.
@@
 Finally, we ran these models with the full set of data points used in \cite{edwards_point_2023}, plus additional data for seven species that did not meet the initial sample size thresholds for NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Harris's Sparrow, Kirtland's Warbler, Tricoloured Blackbird (\textit{Agelaius tricolor}), and Spotted Owl (\textit{Strix occidentalis}).
-All seven of these bird species are listed on Partner's in Flight's watchlist for various threats \citep{will_handbook_2020}.
+All seven of these bird species are listed on Partners in Flight's watchlist for various threats \citep{will_handbook_2020}.
 
@@
 	This is because the species that were modelled in \citet{edwards_point_2023} had a sufficient number of data points to stand on their own, and so adding in species-level effects should have little effect on estimates.
-	\item\label{pred2} A multi-species model will increase the precision of the estimates of cue rate and EDR, but only up to a certain sample size. 
-	This is because there are many species for which NA-POPS has thousands to even tens of thousands of data points; given precision generally increases with sample size, after a certain sample size, a single-species vs multi-species estimate of cue rate and EDR should be similarly precise.
-	Put differently, we expect that species with a smaller sample size will have precision of cue rate and EDR estimates improved with a multi-species model.
+	\item\label{pred2} Improvement in precision will be related to sample size, such that precision of estimates will improve the most for species with smaller sample size, while precision of estimates will improve the least for species with larger sample sizes.
 	\item\label{pred3} A multi-species model will have greater predictive capacity than a single-species model, particularly for species which are data-sparse.
@@
 \par For any point count that employs removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002} and/or distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}, let $Y_{sijk}$ be the observed count of species $s$ during sampling event $i$, occurring in time band $j \in [1,J]$ and/or distance band $k \in [1,K]$.
-That is, if we have a point count where removal sampling or distance sampling (but not both) is not used, then we have that either $J = 1$ or $K = 1$, respectively; whereas, if we have a point count where both removal sampling and distance sampling is used, then we have that both $J,K \geq 2$.
+That is, if we have a point count where removal sampling or distance sampling (but not both) is not used, then either $J = 1$ or $K = 1$, respectively; whereas, if we have a point count where both removal sampling and distance sampling is used, then both $J,K \geq 2$.
 The case where $J = K = 1$ is where neither removal sampling nor distance sampling is used, and so these data points are not appropriate for the modelling techniques described here.
-
-\par The following two subsections detail the Bayesian hierarchical QPAD Removal and Distance Models.
 
@@
 
-\par Here, $\lambda$ is Pagal's Lambda \citep{pagel_inferring_1999}, which determines the strength of the phylogenetic relationship. 
+\par Here, $\lambda$ is Pagel's Lambda \citep{pagel_inferring_1999}, which determines the strength of the phylogenetic relationship. 
 We set $\lambda = 0.76$, consistent with previous results \cite{solymos_phylogeny_2018}. 
-We also have the extra variance term $\sigma \sim exp(5)$.
+We also estimated the extra variance term and used an exponential prior $\sigma \sim exp(5)$.
 
-\par Finally, the vector of mean cue rate $\mathbf{\mu}_{[s]}$ is just a function of the species migratory strategy $\alpha_M$, where the migratory strategy $M$ could either be Resident or Migrant. 
+\par Finally, the vector of mean cue rate $\mathbf{\mu}_{[s]}$ is a function of the species migratory strategy $\alpha_M$, where the migratory strategy $M$ could either be Resident or Migrant. 
 We set priors on both migratory strategies to be $\alpha_M \sim N(-1, 0.01)$, because on the log scale we would expect cue rate to be negative.
@@
 \subsubsection{Point Count Data}
-\par We used the same standardized database of point count data that was put together through the NA-POPS project \cite{edwards_point_2023}. 
+\par We used the same standardized database of point count data that was assembled through the NA-POPS project \cite{edwards_point_2023}. 
 This consists of $712138$ point counts conducted across $292$ projects in Canada and the United States, $422514$ of which used removal sampling and $522820$ of which used distance sampling.
-All of these point counts were standardized using practices developed by the Boreal Avian Modelling project \citep{barker_ecological_2015}.
+All of these point counts were recorded in the database in a standard way, using practices developed by the Boreal Avian Modelling project \citep{barker_ecological_2015}.
 
 \subsubsection{Phylogenetic Trees}
-\par We accessed 2000 pseudo-posterior trees from \citet{jetz_global_2012} that contains phylogeny for all species that contained sufficient data for removal modelling ($>75$ observations as per \cite{edwards_point_2023,solymos_evaluating_2018,buckland_introduction_2001}).
+\par We accessed 2000 pseudo-posterior trees from \citet{jetz_global_2012} for all species with sufficient data for removal modelling ($>75$ observations as per \cite{edwards_point_2023,solymos_evaluating_2018,buckland_introduction_2001}).
 We also requested that the phylogenies include the following bird species with insufficient data from NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricoloured Blackbird, and Spotted Owl.
@@
 \par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts of all the birds to model in this study \citep{billerman_birds_2022}. 
-Body mass for each bird species was obtained form the Elton traits dataset \citep{wilman_eltontraits_2014}. 
-Where body mass was not available through the Elton traits dataset, the Birds of the World account was used instead.
+Body mass for each bird species was obtained from the Elton traits dataset \citep{wilman_eltontraits_2014}. 
+Where body mass was not available through the Elton traits dataset, the Birds of the World account \citep{billerman_birds_2022} was used instead.
 
@@
 \par To test prediction \ref{pred1}, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data in the original NA-POPS analysis.
-In this case, we ran both removal model for 316 species of birds, and we ran the distance model for 317 species of birds (Tricoloured Blackbird was included in the distance models but not the removal models).
+In this case, we ran both models for 316 species of birds.
+Rufous Hummingbird (\textit{Selasphorus rufus}) was dropped from the distance modelling due to issues with divergent transitions \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
+Tricoloured Blackbird had sufficient distance data, and so was included in the distance modelling, but not the removal modelling.
 We accessed the single-species estimates of cue rate and EDR from the NA-POPS Null models using the \texttt{napops} R package \citep{edwards_napops_2023, edwards_point_2023}.
@@
 That is, for each cross-validation run, we held out 20\% of the data for each species for both the multi-species and Bayesian single-species models.
-We then estimated multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR for each cross-validation run, and used these parameter estimates to calculate the posterior predictive density of each of the held out data points.
+We then derived multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR for each cross-validation run, and used these parameter estimates to calculate the posterior predictive density of each of the held out data points.
 We then calculated the difference in posterior predictive density for removal and distance models, by subtracting the pointwise posterior predictive density of the single-species model from the pointwise posterior predictive density of the multi-species model (i.e., multi-species minus single-species).
@@
 This is different from \citet{edwards_point_2023, solymos_calibrating_2013, solymos_evaluating_2018}, where any species with less than 75 data points were removed from the analysis.
-For species that did not have any data in the NA-POPS database, we generated a row of traits data for the species and modelled the species just as a parameter.
+For species that did not have any data in the NA-POPS database, we generated a row of traits data for the species and estimated its cue rate and EDR as parameters in the model.
 We then ran the multi-species models described above using these additional data points (and hence an updated phylogenetic correlation structure for the removal model) to generate estimates of cue rate and EDR for these new species.
@@
 To mitigate issues surrounding initial values being rejected in the distance model (see Discussion), we set the initial values of EDR to be the mean EDR from NA-POPS estimates; for species where EDR estimates did not exist in NA-POPS, we used the overall mean EDR as initial value.
-Cross-validation models were run with 250 warmup iterations and 500 sampling iterations, each on 4 chains for a total of 2000 draws per model run.
+Cross-validation models were run with 1000 warmup iterations and 500 sampling iterations, each on 4 chains for a total of 2000 draws per model run.
 All of the models made use of Stan's reduce\_sum functionality, which allows for within-chain parallelization as well as the usual across-chain parallelization.

### 2023-12-06 src/scripts/removal/8-removal-comparison-analysis.R
convering other parameters plot into dot and error bar

@@ -3,7 +3,7 @@
 # Multi-species QPAD Detectability
 # 8-removal-comparison-analysis.R
 # Created April 2022
-# Last Updated October 2023
+# Last Updated December 2023
 
 ####### Import Libraries and External Files #######
 
@@ -13,6 +13,8 @@ library(napops)
 library(plyr)
 library(ggplot2)
 library(ggpubr)
+library(ggrepel)
+library(patchwork)
 theme_set(theme_pubclean())
 bayesplot::color_scheme_set("red")
 
@@ -67,7 +69,7 @@ for (i in 1:nrow(to_plot))
   geom_abline(slope = 1) +
   xlim(0,1) +
   ylim(0,1) +
-  geom_text(aes(label = Label)) +
+  geom_text_repel(box.padding = 0.5, max.overlaps = Inf, aes(label = Label)) +
   xlab("Cue Rate (Single Species)") + 
     ylab("Cue Rate (Multi Species)") +
   NULL)
@@ -101,6 +103,25 @@ diff_model_run <- diff_model$sample(
   xlab("Modelled Difference"))
 
 cr_differences <- to_plot
+
+####### Plotting Other Parameters #################
+
+(intercept_plot <- bayesplot::mcmc_areas(rem_model$draws(c("intercept")),
+                                         prob = 0.95))
+
+(mig_strat_plot <- bayesplot::mcmc_areas(rem_model$draws(c("mu_mig_strat[1]", "mu_mig_strat[2]")),
+                                         prob = 0.95))
+
+(sigma_plot <- bayesplot::mcmc_areas(rem_model$draws(c("sigma")),
+                                     prob = 0.95))
+
+(other_params_plot <- bayesplot::mcmc_intervals(rem_model$draws(c("intercept",
+                                                                     "mu_mig_strat[1]", 
+                                                                     "mu_mig_strat[2]",
+                                                                     "sigma"))) +
+    scale_y_discrete(labels = c("Intercept", "Resident", "Migrant", "Sigma")))
+
+params_summary <- rem_model$summary(c("intercept", "mu_mig_strat", "sigma"))
 
 ####### SD Comparison Plot ########################
 
@@ -154,28 +175,19 @@ species_vars <- to_plot$variable
   scale_y_discrete(labels = (to_plot$Code)) +
   coord_flip())
 
-####### Plotting Other Parameters #################
-
-(intercept_plot <- bayesplot::mcmc_areas(rem_model$draws(c("intercept")),
-                                         prob = 0.95))
-
[…]

### 2023-12-06 src/scripts/post-hoc/1-1vs1-figure.R
adding separate post-hoc subdirectory for scripts related to generating figures. will eventually move all figure-generating scripts here

@@ -1,0 +1,180 @@
+####### Script Information ########################
+# Brandon P.M. Edwards
+# Multi-species QPAD Detectability
+# posthoc/1-1vs1-figure.R
+# Created December 2023
+# Last Updated December 2023
+
+####### Import Libraries and External Files #######
+
+library(cmdstanr)
+library(bayesplot)
+library(napops)
+library(plyr)
+library(ggplot2)
+library(ggpubr)
+library(ggrepel)
+library(patchwork)
+theme_set(theme_pubclean())
+bayesplot::color_scheme_set("red")
+
+####### Read Data #################################
+
+rem_model <- readRDS("output/model_runs/removal_predictions.RDS")
+load("data/generated/removal_stan_data_pred.rda")
+dis_model <- readRDS("output/model_runs/distance_predictions.RDS")
+load("data/generated/distance_stan_data.rda")
+
+####### Removal Model #############################
+
+# Extract log_phi summary statistics from full Stan model runs
+rem_summary <- rem_model$summary("log_phi")
+
+# Add species names to these summaries
+rem_summary$Code <- removal_stan_data_pred$sp_all
+
+# Get data sample size for all species and add to summary
+species_n <- data.frame(table(removal_stan_data_pred$species))
+names(species_n) <- c("index", "N")
+rem_summary$index <- seq(1, nrow(rem_summary))
+rem_summary$N <- 0
+for (i in 1:nrow(species_n))
+{
+  rem_summary[which(rem_summary$index == species_n$index[i]), "N"] <-
+    species_n$N[i]
+}
+
+# Get original single-species NA-POPS estimates
+napops_summary <- napops::coef_removal(species = rem_summary$Code, model = 1)
+
+####### 1-to-1 Plot ###############################
+
+to_plot <- merge(rem_summary[,c("Code", "mean", "N")], napops_summary[, c("Species", "Intercept")],
+                 by.x = "Code", by.y = "Species")
+names(to_plot) <- c("Species", "Multi", "N", "Single")
+to_plot$Label <- ""
+to_plot$Multi <- exp(to_plot$Multi)
+to_plot$Single <- exp(to_plot$Single)
+to_plot$diff <- to_plot$Multi - to_plot$Single
+
+# Check for a 20% relative difference in cue rates
+for (i in 1:nrow(to_plot))
+{
+  if ((abs(to_plot$diff[i]) / to_plot$Single[i]) > 0.20)
+  {
+    to_plot$Label[i] <- to_plot$Species[i]
+  }
+}
+
[…]

### 2023-12-06 src/scripts/post-hoc/2-sd-figure.R

@@ -1,0 +1,121 @@
+####### Script Information ########################
+# Brandon P.M. Edwards
+# Multi-species QPAD Detectability
+# posthoc/2-sd-figure.R
+# Created December 2023
+# Last Updated December 2023
+
+####### Import Libraries and External Files #######
+
+library(cmdstanr)
+library(bayesplot)
+library(napops)
+library(plyr)
+library(ggplot2)
+library(ggpubr)
+library(ggrepel)
+library(patchwork)
+theme_set(theme_pubclean())
+bayesplot::color_scheme_set("red")
+
+####### Read Data #################################
+
+rem_model <- readRDS("output/model_runs/removal_predictions.RDS")
+load("data/generated/removal_stan_data_pred.rda")
+dis_model <- readRDS("output/model_runs/distance_predictions.RDS")
+load("data/generated/distance_stan_data.rda")
+
+####### Removal Model #############################
+
+# Extract log_phi summary statistics from full Stan model runs
+rem_summary <- rem_model$summary("log_phi")
+
+# Add species names to these summaries
+rem_summary$Code <- removal_stan_data_pred$sp_all
+
+# Get data sample size for all species and add to summary
+species_n <- data.frame(table(removal_stan_data_pred$species))
+names(species_n) <- c("index", "N")
+rem_summary$index <- seq(1, nrow(rem_summary))
+rem_summary$N <- 0
+for (i in 1:nrow(species_n))
+{
+  rem_summary[which(rem_summary$index == species_n$index[i]), "N"] <-
+    species_n$N[i]
+}
+
+# Get original single-species NA-POPS estimates
+napops_summary <- napops::coef_removal(species = rem_summary$Code, model = 1)
+
+to_plot <- rem_summary[which(rem_summary$Code %in% napops_summary$Species), 
+                       c("Code", "N", "sd")]
+names(to_plot) <- c("Code", "N", "SD_Multi")
+to_plot$SD_Single <- NA
+# Get original NA-POPS SDs
+for (i in 1:nrow(to_plot))
+{
+  to_plot$SD_Single[i] <- sqrt(napops::get_vcv(species = to_plot$Code[i],
+                                               model_type = "rem",
+                                               model_num = 1))
+}
[…]

### 2023-12-06 src/scripts/post-hoc/2-sd-figure.R
reverting back to simple cmparison plot of SD vs sample size for both models, combined into one figure

@@ -49,7 +49,7 @@ napops_summary <- napops::coef_removal(species = rem_summary$Code, model = 1)
 
 to_plot <- rem_summary[which(rem_summary$Code %in% napops_summary$Species), 
                        c("Code", "N", "sd")]
-names(to_plot) <- c("Code", "N", "SD_Multi")
+names(to_plot) <- c("Code", "N", "Multi")
 to_plot$SD_Single <- NA
 # Get original NA-POPS SDs
 for (i in 1:nrow(to_plot))
@@ -59,63 +59,63 @@ for (i in 1:nrow(to_plot))
                                                model_num = 1))
 }
 to_plot_long <- reshape2::melt(to_plot, id = c("Code", "N"), variable_name = "Model")
+(sd_comp_removal <- ggplot(data = to_plot_long, 
+                       aes(x = log(N), y = (value),
+                           group = variable, color = variable)) + 
+  geom_smooth() +
+  xlab("log(Sample Size)") +
+  ylab("Standard Deviation") +
+    theme(legend.position = "none") +
+  NULL)
 
-sd_model_data <- list(n_samples = nrow(to_plot_long),
-                      stdev = to_plot_long$value,
-                      N = to_plot_long$N,
-                      model_type = factor(to_plot_long$variable,
-                                          levels = c("SD_Single", "SD_Multi")))
+####### Distance Model #############################
 
-sd_model <- cmdstan_model(stan_file = "models/sd_model.stan")
+# Extract log_phi summary statistics from full Stan model runs
+dis_summary <- dis_model$summary("log_tau")
 
-sd_model_run <- sd_model$sample(
-  data = sd_model_data,
-  iter_warmup = 1000,
-  iter_sampling = 2000,
-  chains = 4,
-  parallel_chains = 4,
-  refresh = 10
-)
+# Add species names to these summaries
+dis_summary$Code <- distance_stan_data$sp_all
 
-n_vals <- seq(min(to_plot_long$N), max(to_plot_long$N))
-
-ss_draws <- as.matrix(sd_model_run$draws(c("alpha[1]", "beta[1]"), format = "df"))[, 1:2]
-ss_pred <- matrix(data = NA, nrow = length(n_vals), ncol = nrow(ss_draws))
-for (i in 1:nrow(ss_draws))
+# Get data sample size for all species and add to summary
[…]

### 2023-12-07 src/scripts/post-hoc/3-cv-figure.R
WIP for cross validation

@@ -1,0 +1,292 @@
+####### Script Information ########################
+# Brandon P.M. Edwards
+# Multi-species QPAD Detectability
+# posthoc/3-cv-figure.R
+# Created December 2023
+# Last Updated December 2023
+
+####### Import Libraries and External Files #######
+
+library(cmdstanr)
+library(bayesplot)
+library(ggplot2)
+library(ggpubr)
+library(ape)
+library(ggtree)
+theme_set(theme_pubclean())
+bayesplot::color_scheme_set("red")
+
+source("src/functions/rem-lik.R")
+source("src/functions/subset-distance-data.R")
+source("src/functions/dis-lik.R")
+
+####### Read Data #################################
+
+load("data/generated/removal_stan_data_cv.rda")
+load("data/generated/distance_stan_data.rda")
+cv_folds_rem <- read.csv("data/generated/removal_cv_folds.csv")
+cv_folds_dis <- read.csv("data/generated/distance_cv_folds.csv")
+phylo_tree <- ape::read.nexus(file = "data/raw/all_species.nex")
+binomial <- read.csv("data/generated/binomial_names.csv")
+traits <- read.csv("data/raw/traits.csv")
+
+####### Removal Model #############################
+
+#' Create dataframe that will connect species, data point,
+#' and lppd
+lppd <- data.frame(Species = cv_folds_rem$Species,
+                   CV_Fold = cv_folds_rem$cv_fold,
+                   SS_LPPD = NA,
+                   MS_LPPD = NA,
+                   Difference = NA)
+
+for (f in 1:max(cv_folds_rem$cv_fold))
+{
+  # Extract dataframe of phi from single species and multi species model
+  ss_mod <- readRDS(paste0("output/model_runs/cv_removal/ss_fold_",
+                           f,
+                           ".RDS"))
+  ms_mod <- readRDS(paste0("output/model_runs/cv_removal/ms_fold_",
+                           f,
+                           ".RDS"))
+  
+  ss_phi_summary <- ss_mod$summary("log_phi")
+  ms_phi_summary <- ms_mod$summary("log_phi")
+  
+  # Subset the data to just the test set
+  data_indices <- which(lppd$CV_Fold == f)
+  abund <- removal_stan_data_cv$abund_per_band[data_indices, ]
+  bands <- removal_stan_data_cv$bands_per_sample[data_indices]
[…]

### 2023-12-07 src/scripts/post-hoc/3-cv-figure.R
cross validation figures script

@@ -248,13 +248,13 @@ lppd_summary[which((lppd_summary$Mean_Difference - lppd_summary$StdErr > 0) &
 
 preference <- data.frame(table(lppd_summary$Model_Preference))
 
-(lppd_difference_plot <- ggplot(data = lppd_summary, aes(x = Species, y = Mean_Difference)) +
+(lppd_difference_plot_dis <- ggplot(data = lppd_summary, aes(x = Species, y = Mean_Difference)) +
     geom_errorbar(aes(ymin = Mean_Difference - StdErr, ymax = Mean_Difference + StdErr)) +
     theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
     #coord_flip() +
     NULL)
 
-(lppd_diff_vs_n <- ggplot(data = lppd_summary, aes(x = N, y = Mean_Difference)) +
+(lppd_diff_vs_n_dis <- ggplot(data = lppd_summary, aes(x = N, y = Mean_Difference)) +
     geom_point(aes(color = Model_Preference)) +
     stat_smooth(method = "lm", 
                 formula = y ~ x, 
@@ -262,31 +262,14 @@ preference <- data.frame(table(lppd_summary$Model_Preference))
     NULL
 )
 
-
-
-
-
-
-
-
-
-
 ####### Output ####################################
 
-write.table(lppd, file = "data/generated/lppd_removal.csv", sep = ",", row.names = FALSE)
-write.table(lppd_summary, file = "data/generated/lppd_summary_removal.csv", sep = ",", row.names = FALSE)
+write.table(lppd_removal, file = "data/generated/lppd_removal.csv", sep = ",", row.names = FALSE)
+write.table(lppd_summary_removal, file = "data/generated/lppd_summary_removal.csv", sep = ",", row.names = FALSE)
+write.table(lppd, file = "data/generated/lppd_distance.csv", sep = ",", row.names = FALSE)
+write.table(lppd_summary, file = "data/generated/lppd_summary_distance.csv", sep = ",", row.names = FALSE)
 
-png(filename = "output/plots/cv_removal.png",
-    width = 20, height = 6, units = "in", res = 300)
-print(lppd_difference_plot)
-dev.off()
-
-png(filename = "output/plots/cv_removal_diff_vs_n.png",
-    width = 6, height = 4, units = "in", res = 300)
-print(lppd_diff_vs_n)
-dev.off()
-
[…]

### 2023-12-07 src/scripts/post-hoc/1-1vs1-figure.R
output csv files of phi, tau, and differences

@@ -100,6 +100,8 @@ removal_plot <- single_vs_multi_plot + inset_element(modelled_difference_plot,
                                                      right = 1, 
                                                      top = 0.5)
 
+cr_differences <- to_plot
+
 ####### Distance Model ############################
 
 # Extract log_phi summary statistics from full Stan model runs
@@ -172,7 +174,15 @@ distance_plot <- single_vs_multi_plot + inset_element(modelled_difference_plot,
                                                      right = 1, 
                                                      top = 0.5)
 
+edr_differences <- to_plot
+
 ####### Output ####################################
+
+write.table(rem_summary, file = "data/generated/phi.csv", sep = ",", row.names = FALSE)
+write.table(dis_summary, file = "data/generated/tau.csv", sep = ",", row.names = FALSE)
+
+write.table(cr_differences, file = "data/generated/phi_differences.csv", sep = ",", row.names = FALSE)
+write.table(edr_differences, file = "data/generated/tau_differences.csv", sep = ",", row.names = FALSE)
 
 tiff("output/plots/1vs1.tiff",
      width = 6, height = 6, res = 600, units = "in")

### 2023-12-07 src/scripts/post-hoc/4-predictions-figure.R
script to generate predictions plot figure

@@ -1,0 +1,101 @@
+####### Script Information ########################
+# Brandon P.M. Edwards
+# Multi-species QPAD Detectability
+# posthoc/4-predictions-figure.R
+# Created December 2023
+# Last Updated December 2023
+
+####### Import Libraries and External Files #######
+
+library(cmdstanr)
+library(bayesplot)
+library(napops)
+library(plyr)
+library(ggplot2)
+library(ggpubr)
+library(ggrepel)
+library(patchwork)
+theme_set(theme_pubclean())
+bayesplot::color_scheme_set("red")
+
+####### Read Data #################################
+
+rem_model <- readRDS("output/model_runs/removal_predictions.RDS")
+load("data/generated/removal_stan_data_pred.rda")
+dis_model <- readRDS("output/model_runs/distance_predictions.RDS")
+load("data/generated/distance_stan_data.rda")
+
+####### Removal Model #############################
+
+# Extract log_phi summary statistics from full Stan model runs
+rem_summary <- rem_model$summary("log_phi")
+
+# Add species names to these summaries
+rem_summary$Code <- removal_stan_data_pred$sp_all
+
+# Get data sample size for all species and add to summary
+species_n <- data.frame(table(removal_stan_data_pred$species))
+names(species_n) <- c("index", "N")
+rem_summary$index <- seq(1, nrow(rem_summary))
+rem_summary$N <- 0
+for (i in 1:nrow(species_n))
+{
+  rem_summary[which(rem_summary$index == species_n$index[i]), "N"] <-
+    species_n$N[i]
+}
+
+# Get original single-species NA-POPS estimates
+napops_summary <- napops::coef_removal(species = rem_summary$Code, model = 1)
+
+sp <- c("LCTH", "LEPC", "HASP", "TRBL", "SPOW", "KIWA", "BITH")
+
+to_plot <- rem_summary[which(rem_summary$Code %in% sp), ]
+
+species_vars <- to_plot$variable
+
+(removal_plot <- bayesplot::mcmc_intervals(exp(rem_model$draws(species_vars))) + 
+    xlab("Predicted Cue Rate") + 
+    ylab("Species") +
+    scale_y_discrete(labels = (to_plot$Code)) +
+    coord_flip())
+
+####### Distance Model ############################
+
+# Extract log_phi summary statistics from full Stan model runs
+dis_summary <- dis_model$summary("log_tau")
+
[…]

### 2023-12-07 src/scripts/post-hoc/5-params-figure.R
script to generate parameter estimate figure

@@ -1,0 +1,63 @@
+####### Script Information ########################
+# Brandon P.M. Edwards
+# Multi-species QPAD Detectability
+# posthoc/5-params-figure.R
+# Created December 2023
+# Last Updated December 2023
+
+####### Import Libraries and External Files #######
+
+library(cmdstanr)
+library(bayesplot)
+library(napops)
+library(plyr)
+library(ggplot2)
+library(ggpubr)
+library(ggrepel)
+library(patchwork)
+theme_set(theme_pubclean())
+bayesplot::color_scheme_set("red")
+
+####### Read Data #################################
+
+rem_model <- readRDS("output/model_runs/removal_predictions.RDS")
+dis_model <- readRDS("output/model_runs/distance_predictions.RDS")
+
+####### Removal Model #############################
+
+(removal_plot <- bayesplot::mcmc_intervals(rem_model$draws(c("sigma",
+                                                             "mu_mig_strat[2]", 
+                                                             "mu_mig_strat[1]",
+                                                             "intercept"))) +
+    scale_y_discrete(labels = c("Sigma", "Migrant", "Resident", "Intercept")))
+
+params_removal <- rem_model$summary(c("intercept", "mu_mig_strat", "sigma"))
+
+####### Distance Model ############################
+
+(distance_plot <- bayesplot::mcmc_intervals(dis_model$draws(c("sigma",
+                                                              "beta_pitch",
+                                                              "beta_mass", 
+                                                              "mu_habitat[2]",
+                                                              "mu_habitat[1]",
+                                                              "mu_mig_strat[2]",
+                                                              "mu_mig_strat[1]",
+                                                              "intercept"))) +
+   scale_y_discrete(labels = c("Sigma", "Pitch Slope","Mass Slope","Closed Habitat",
+                               "Open Habitat","Migrant","Resident","Intercept")))
+
+params_distance <- dis_model$summary(c("intercept", "mu_mig_strat",
+                                      "mu_habitat", "beta_mass", "beta_pitch",
+                                      "sigma"))
+
+####### Output ####################################
+
+write.table(params_removal, file = "data/generated/removal_params.csv", sep = ",", row.names = FALSE)
+write.table(params_distance, file = "data/generated/distance_params.csv", sep = ",", row.names = FALSE)
+
+tiff(filename = "output/plots/parameters_plot.tiff",
+     width = 6, height = 6, units = "in", res = 600)
+ggarrange(removal_plot, distance_plot, nrow = 2, labels = c("A", "B"))
+dev.off()
+
+

### 2023-12-07 src/scripts/post-hoc/3-cv-figure.R
chagne to tiff output

@@ -269,7 +269,7 @@ write.table(lppd_summary_removal, file = "data/generated/lppd_summary_removal.cs
 write.table(lppd, file = "data/generated/lppd_distance.csv", sep = ",", row.names = FALSE)
 write.table(lppd_summary, file = "data/generated/lppd_summary_distance.csv", sep = ",", row.names = FALSE)
 
-tiff(filename = "output/plots/cv_plot.png",
+tiff(filename = "output/plots/cv_plot.tiff",
     width = 6, height = 6, units = "in", res = 600)
 ggarrange(lppd_diff_vs_n_rem, lppd_diff_vs_n_dis, nrow = 2, labels = c("A", "B"))
 dev.off()

### 2023-12-07 src/scripts/distance/7-distance-cv-analysis.R
remove unecessary files

@@ -1,153 +1,0 @@
-####### Script Information ########################
-# Brandon P.M. Edwards
-# Multi-species QPAD Detectability
-# 7-distance-cv-analysis.R
-# Created October 2023
-# Last Updated October 2023
-####### Import Libraries and External Files #######
-
-library(cmdstanr)
-library(bayesplot)
-library(plyr)
-library(ggplot2)
-library(ggpubr)
-theme_set(theme_pubclean())
-bayesplot::color_scheme_set("red")
-
-source("src/functions/subset-distance-data.R")
-source("src/functions/dis-lik.R")
-
-####### Read Data #################################
-
-cv_folds <- read.csv("data/generated/distance_cv_folds.csv")
-load("data/generated/distance_stan_data.rda")
-traits <- read.csv("data/raw/traits.csv")
-
-####### Data Wrangling ############################
-
-#' First we must drop all species which are just not going to be involved
-#' in the cross validation at all. This will be the species for which we are
-#' generated predictions
-pred_drops <- c("LCTH", "LEPC", "HASP", "SPOW", "KIWA", "BITH")
-dis_data <- subset_distance_data(distance_stan_data = distance_stan_data,
-                                 sps = setdiff(unique(distance_stan_data$sp_list),
-                                               pred_drops))
-rm(distance_stan_data)
-
-####### Main Code #################################
-
-#' Create dataframe that will connect species, data point,
-#' and lppd
-lppd <- data.frame(Species = cv_folds$Species,
-                   CV_Fold = cv_folds$cv_fold,
-                   SS_LPPD = NA,
-                   MS_LPPD = NA,
-                   Difference = NA)
-
-for (f in 1:max(cv_folds$cv_fold))
-{
-  # Extract dataframe of tau from single species and multi species model
-  ss_mod <- readRDS(paste0("output/model_runs/cv_distance/ss_fold_",
-                           f,
-                           ".RDS"))
-  ms_mod <- readRDS(paste0("output/model_runs/cv_distance/ms_fold_",
-                           f,
-                           ".RDS"))
-  
-  ss_tau_summary <- ss_mod$summary("log_tau")
-  ms_tau_summary <- ms_mod$summary("log_tau")
-  
-  # Subset the data to just the test set
[…]

### 2023-12-07 src/scripts/distance/8-distance-comparison-analysis.R

@@ -1,201 +1,0 @@
-####### Script Information ########################
-# Brandon P.M. Edwards
-# Multi-species QPAD Detectability
-# 8-distance-comparison-analysis.R
-# Created May 2022
-# Last Updated October 2023
-####### Import Libraries and External Files #######
-
-library(cmdstanr)
-library(bayesplot)
-library(napops)
-library(plyr)
-library(ggplot2)
-library(ggpubr)
-theme_set(theme_pubclean())
-bayesplot::color_scheme_set("red")
-
-####### Read Data #################################
-
-dis_model <- readRDS("output/model_runs/distance_predictions.RDS")
-load("data/generated/distance_stan_data.rda")
-
-####### Data Wrangling ############################
-
-# Extract log_phi summary statistics from full Stan model runs
-dis_summary <- dis_model$summary("log_tau")
-
-# Add species names to these summaries
-dis_summary$Code <- distance_stan_data$sp_all
-
-# Get data sample size for all species and add to summary
-species_n <- data.frame(table(distance_stan_data$species))
-names(species_n) <- c("index", "N")
-dis_summary$index <- seq(1, nrow(dis_summary))
-dis_summary$N <- 0
-for (i in 1:nrow(species_n))
-{
-  dis_summary[which(dis_summary$index == species_n$index[i]), "N"] <-
-    species_n$N[i]
-}
-
-# Get original single-species NA-POPS estimates
-napops_summary <- napops::coef_distance(species = dis_summary$Code, model = 1)
-
-####### 1-to-1 Plot ###############################
-
-to_plot <- merge(dis_summary[,c("Code", "mean")], napops_summary[, c("Species", "Intercept")],
-                 by.x = "Code", by.y = "Species")
-names(to_plot) <- c("Species", "Multi", "Single")
-to_plot$Label <- ""
-to_plot$Multi <- exp(to_plot$Multi) * 100
-to_plot$Single <- exp(to_plot$Single)
-to_plot$diff <- to_plot$Multi - to_plot$Single
-
-# Check for a 20% relative difference in EDR
-for (i in 1:nrow(to_plot))
-{
-  if ((abs(to_plot$diff[i]) / to_plot$Single[i]) > 0.20)
-  {
-    to_plot$Label[i] <- to_plot$Species[i]
-  }
-}
-
-(single_vs_multi_plot <- ggplot(data = to_plot, mapping = aes(x = Single, y = Multi)) +
[…]

### 2023-12-07 src/scripts/removal/7-removal-cv-analysis.R

@@ -1,186 +1,0 @@
-####### Script Information ########################
-# Brandon P.M. Edwards
-# Multi-species QPAD Detectability
-# 7-removal-cv-analysis.R
-# Created October 2023
-# Last Updated October 2023
-
-####### Import Libraries and External Files #######
-
-library(cmdstanr)
-library(bayesplot)
-library(ggplot2)
-library(ggpubr)
-library(ape)
-library(ggtree)
-theme_set(theme_pubclean())
-bayesplot::color_scheme_set("red")
-
-source("src/functions/rem-lik.R")
-
-####### Read Data #################################
-
-load("data/generated/removal_stan_data_cv.rda")
-cv_folds <- read.csv("data/generated/removal_cv_folds.csv")
-phylo_tree <- ape::read.nexus(file = "data/raw/all_species.nex")
-binomial <- read.csv("data/generated/binomial_names.csv")
-traits <- read.csv("data/raw/traits.csv")
-
-####### Main Code #################################
-
-#' Create dataframe that will connect species, data point,
-#' and lppd
-lppd <- data.frame(Species = cv_folds$Species,
-                   CV_Fold = cv_folds$cv_fold,
-                   SS_LPPD = NA,
-                   MS_LPPD = NA,
-                   Difference = NA)
-
-for (f in 1:max(cv_folds$cv_fold))
-{
-  # Extract dataframe of phi from single species and multi species model
-  ss_mod <- readRDS(paste0("output/model_runs/cv_removal/ss_fold_",
-                    f,
-                    ".RDS"))
-  ms_mod <- readRDS(paste0("output/model_runs/cv_removal/ms_fold_",
-                           f,
-                           ".RDS"))
-  
-  ss_phi_summary <- ss_mod$summary("log_phi")
-  ms_phi_summary <- ms_mod$summary("log_phi")
-  
-  # Subset the data to just the test set
-  data_indices <- which(lppd$CV_Fold == f)
-  abund <- removal_stan_data_cv$abund_per_band[data_indices, ]
-  bands <- removal_stan_data_cv$bands_per_sample[data_indices]
-  max_time <- removal_stan_data_cv$max_time[data_indices, ]
-  species <- removal_stan_data_cv$species[data_indices]
-  max_intervals <- removal_stan_data_cv$max_intervals
-
-  phi_ss <- exp(ss_phi_summary$mean[species])
[…]

### 2023-12-07 src/scripts/removal/8-removal-comparison-analysis.R

@@ -1,207 +1,0 @@
-####### Script Information ########################
-# Brandon P.M. Edwards
-# Multi-species QPAD Detectability
-# 8-removal-comparison-analysis.R
-# Created April 2022
-# Last Updated December 2023
-
-####### Import Libraries and External Files #######
-
-library(cmdstanr)
-library(bayesplot)
-library(napops)
-library(plyr)
-library(ggplot2)
-library(ggpubr)
-library(ggrepel)
-library(patchwork)
-theme_set(theme_pubclean())
-bayesplot::color_scheme_set("red")
-
-####### Read Data #################################
-
-rem_model <- readRDS("output/model_runs/removal_predictions.RDS")
-load("data/generated/removal_stan_data_pred.rda")
-
-####### Data Wrangling ############################
-
-# Extract log_phi summary statistics from full Stan model runs
-rem_summary <- rem_model$summary("log_phi")
-
-# Add species names to these summaries
-rem_summary$Code <- removal_stan_data_pred$sp_all
-
-# Get data sample size for all species and add to summary
-species_n <- data.frame(table(removal_stan_data_pred$species))
-names(species_n) <- c("index", "N")
-rem_summary$index <- seq(1, nrow(rem_summary))
-rem_summary$N <- 0
-for (i in 1:nrow(species_n))
-{
-  rem_summary[which(rem_summary$index == species_n$index[i]), "N"] <-
-    species_n$N[i]
-}
-
-# Get original single-species NA-POPS estimates
-napops_summary <- napops::coef_removal(species = rem_summary$Code, model = 1)
-
-####### 1-to-1 Plot ###############################
-
-to_plot <- merge(rem_summary[,c("Code", "mean", "N")], napops_summary[, c("Species", "Intercept")],
-                 by.x = "Code", by.y = "Species")
-names(to_plot) <- c("Species", "Multi", "N", "Single")
-to_plot$Label <- ""
-to_plot$Multi <- exp(to_plot$Multi)
-to_plot$Single <- exp(to_plot$Single)
-to_plot$diff <- to_plot$Multi - to_plot$Single
-
-# Check for a 20% relative difference in cue rates
-for (i in 1:nrow(to_plot))
-{
-  if ((abs(to_plot$diff[i]) / to_plot$Single[i]) > 0.20)
-  {
-    to_plot$Label[i] <- to_plot$Species[i]
-  }
-}
-
-(single_vs_multi_plot <- ggplot(data = to_plot, mapping = aes(x = Single, y = Multi)) +
[…]

### 2023-12-07 src/functions/dis-lik.R
use log predictive density rather than predictive density. also changed aesthetics on the plots

@@ -24,7 +24,7 @@ dis_lik <- function(abund = NULL,
     
     Pi[bands[i]] = 1 - sum(Pi)
     
-    loglik[i] <- dmultinom(x = abund[i,1:bands[i]], prob = Pi)
+    loglik[i] <- dmultinom(x = abund[i,1:bands[i]], prob = Pi, log = TRUE)
   }
   
   return(loglik)

### 2023-12-07 src/functions/rem-lik.R

@@ -17,7 +17,7 @@ rem_lik <- function(abund = NULL,
     }
     Pi[1] = 1 - sum(Pi)
     
-    lp[i] <- dmultinom(x = abund[i,1:bands[i]], prob = Pi)
+    lp[i] <- dmultinom(x = abund[i,1:bands[i]], prob = Pi, log = TRUE)
   }
   
   return(lp)

### 2023-12-07 src/scripts/post-hoc/3-cv-figure.R

@@ -90,6 +90,25 @@ for (f in 1:max(cv_folds_rem$cv_fold))
 
 lppd$Difference <- lppd$MS_LPPD - lppd$SS_LPPD
 
+# cv_model <- cmdstan_model(stan_file = "models/cv_difference.stan")
+# cv_model_data <- list(N = nrow(lppd),
+#                       n_species = length(unique(lppd$Species)),
+#                       difference = lppd$Difference,
+#                       species = removal_stan_data_cv$species)
+# 
+# cv_model_run <- cv_model$sample(
+#   data = cv_model_data,
+#   iter_warmup = 1000,
+#   iter_sampling = 500,
+#   chains = 4,
+#   parallel_chains = 4,
+#   refresh = 10
+# )
+# 
+# saveRDS(cv_model_run, file = "output/model_runs/cv_removal/cv_model_removal.RDS")
+# 
+# cv_model_summary <- cv_model_run$summary(c("nu", "overall_difference", "species_difference", "sigma"))
+
 lppd_summary <- data.frame(Species = unique(lppd$Species),
                            Mean_Difference = NA,
                            StdErr = NA,
@@ -125,13 +144,10 @@ preference <- data.frame(table(lppd_summary$Model_Preference))
     #coord_flip() +
     NULL)
 
-(lppd_diff_vs_n_rem <- ggplot(data = lppd_summary, aes(x = N, y = Mean_Difference)) +
+(lppd_diff_vs_n_rem <- ggplot(data = lppd_summary, aes(x = log(N), y = Mean_Difference)) +
     geom_point(aes(color = Model_Preference)) +
-    stat_smooth(method = "lm", 
-                formula = y ~ x, 
-                geom = "smooth") +  
-    NULL
-)
+    geom_hline(yintercept = 0) +
+    NULL)
 
 lppd_summary <- merge(lppd_summary, binomial[,c("Code", "Scientific_BT")],
                       by.x = "Species", by.y = "Code")
@@ -254,11 +270,9 @@ preference <- data.frame(table(lppd_summary$Model_Preference))
     #coord_flip() +
     NULL)
 
-(lppd_diff_vs_n_dis <- ggplot(data = lppd_summary, aes(x = N, y = Mean_Difference)) +
+(lppd_diff_vs_n_dis <- ggplot(data = lppd_summary, aes(x = log(N), y = Mean_Difference)) +
     geom_point(aes(color = Model_Preference)) +
[…]

### 2023-12-07 article/article.tex
small updates to the article, and output both tiff and png formats for figures

@@
 \section{Results}
-\subsection{Removal Model Comparisons}
+\subsection{Parameter Estimates}
 
-\par We found very little evidence for an overall change in cue rate for a given species modelled by a single-species model vs a multi-species model (Figure \ref{fig:removal_1vs1}A).
+\par We found much less support for a difference in cue rate between resident and migrant species than was found in \citet{solymos_evaluating_2018}. 
+Mean overall log cue rate in the multi-species model was -1.09 (95\% credible interval of [-1.24, -0.94]); both estimated effects for resident and migrant species were effectively 0 (Figure \ref{fig:params}A).
+We also found little support for migration strategy having any effect on EDR.
+There is some evidence that a resident species may have a slightly higher EDR than a migrant species, but the 95\% credible intervals of both effects overlapped with the 95\% credible interval of the overall estimated intercept (Figure \ref{fig:params}B).
+We did find similar support for change in EDR by body size and pitch, in that body size tended to have an increasing effect on EDR, and song pitch had a decreasing effect on EDR (Figure \ref{fig:params}B).
+
+\begin{figure}[h!]
+	\includegraphics{../output/plots/parameters_plot.png}
+	\caption{Posterior distributions with 95\% credible intervals (shaded area) of parameters estimated in the multi-species removal model, with overall mean log cue rate as the intercept (A), migratory strategy as departures from the intercept based on whether a species was a resident (mu\_mig\_strat[1]) or migrant (mu\_mig\_strat[B]) (B), and extra variance term after accounting for the phylogenetic variation (C).
+Posterior distributions with 95\% credible intervals (shaded area) of parameters estimated in the multi-species distance model, with (A) overall mean log EDR as the intercept, with departures from the intercept modelled as migratory strategy (resident: mu\_mig\_strat[1], migrant: mu\_mig\_strat[B]), (B) slope parameters of changes in EDR given change in mass and pitch, and (C) extra variance term after accounting for traits.}
+	\label{fig:params}
+\end{figure}
+
+\subsection{Comparison of Estimates between Single Species and Multi Species}
+
+\par We found very little evidence for an overall change in cue rate for a given species modelled by a single-species model vs a multi-species model (Figure \ref{fig:1vs1}A).
 Seven out of the 316 species considered for removal modelling had a percent change in cue rate of more than 20\%.
 These species were Willow Ptarmigan (\textit{Lagopus lagopus}), Rusty Blackbird (\textit{Euphagus carolinus}), Red-bellied Sapsucker (\textit{Sphyrapicus ruber}), Common Redpoll (\textit{Acanthis flammea}), Calliope Hummingbird (\textit{Selasphorus calliope}), Bohemian Waxwing (\textit{Bombycilla garrulus}), and Allen's Hummingbird (\textit{Selasphorus sasin}).
-All seven of these species were among the lowest cue rates in both the single-species and multi-species context (Figure \ref{fig:removal_1vs1}B).
+All seven of these species were among the lowest cue rates in both the single-species and multi-species context (Figure \ref{fig:1vs1}A).
+
+\par We found little evidence for a change in modelled EDR between single species and multi-species distance models (Figure \ref{fig:1vs1}B). 
+(NEEDS UPDATING) Seven out of the 317 species had a percent change in EDR of more than 20\% (Figure \ref{fig:1vs1}B).
+These species were Sharp-tailed Grouse (\textit{Tympanuchus phasianellus}), Sedge Wren (\textit{Cistothorus stellaris}), Ring-necked Pheasant (\textit{Phasianus colchicus}), Chimney Swift (\textit{Chaetura pelagica}), Boat-tailed Grackle (\textit{Quiscalus major}), and Baird's Sparrow (\textit{Ammodramus bairdii}).
 
 \begin{figure}[h!]
-	\includegraphics{../output/plots/removal_1vs1.png}
-	\caption{Modelled difference with 95\% credible interval (shaded area) between cue rates estimated from the multi-species model developed here, minus the single-species model in NA-POPS \citep{edwards_point_2023} (A); and 1-to-1 plot comparing cue rates estimated from the single-species model vs the multi-species model for each species (B).}
-	\label{fig:removal_1vs1}
+	\includegraphics{../output/plots/1vs1.png}
+	\caption{Modelled difference with 95\% credible interval (shaded area) between cue rates estimated from the multi-species model developed here, minus the single-species model in NA-POPS \citep{edwards_point_2023} (A); and 1-to-1 plot comparing cue rates estimated from the single-species model vs the multi-species model for each species (B).
+Modelled difference with 95\% credible interval (shaded area) between EDR estimated from the multi-species model developed here, minus the single-species model in NA-POPS \citep{edwards_point_2023} (A); and 1-to-1 plot comparing EDR estimated from the single-species model vs the multi-species model for each species (B).}
+	\label{fig:1vs1}
 \end{figure}
 
-\par We found much less support for a difference in cue rate between resident and migrant species than was found in \citet{solymos_evaluating_2018}. 
-Mean overall log cue rate in the multi-species model was -1.09 (95\% credible interval of [-1.24, -0.94]); both estimated effects for resident and migrant species were effectively 0 (Figure \ref{fig:removal_params}).
+\subsection{Comparison of Precision between Single Species and Multi Species}
 
-\begin{figure}[h!]
-	\includegraphics{../output/plots/removal_other_params.png}
-	\caption{Posterior distributions with 95\% credible intervals (shaded area) of parameters estimated in the multi-species removal model, with overall mean log cue rate as the intercept (A), migratory strategy as departures from the intercept based on whether a species was a resident (mu\_mig\_strat[1]) or migrant (mu\_mig\_strat[B]) (B), and extra variance term after accounting for the phylogenetic variation (C).}
-	\label{fig:removal_params}
-\end{figure}
-
-\par We found some evidence that a multi-species removal model improved the precision of cue rate estimates compared to estimates from a single-species model.
-The intercept of standard deviations of cue rates estimated in a single species model was greater than the intercept of standard deviations of cue rates estimated in a multi-species model, with very little overlap in their 95\% credible intervals (Figure \ref{fig:removal_sd}A).
+\par (NEEDS UPDATING) We found some evidence that a multi-species removal model improved the precision of cue rate estimates compared to estimates from a single-species model.
+The intercept of standard deviations of cue rates estimated in a single species model was greater than the intercept of standard deviations of cue rates estimated in a multi-species model, with very little overlap in their 95\% credible intervals (Figure \ref{fig:sd}A).
 This implies that the baseline standard deviation of cue rates is higher in a single species model than a multi species model for very low sample sizes.
-However, the slope of standard deviations of cue rates estimated in a single species model was farther from 0 (i.e., more negative) than the slope of standard deviations of cue rates estimated in a multi-species model, with very little overlap in their 95\% credible intervals (Figure \ref{fig:removal_sd}B).
+However, the slope of standard deviations of cue rates estimated in a single species model was farther from 0 (i.e., more negative) than the slope of standard deviations of cue rates estimated in a multi-species model, with very little overlap in their 95\% credible intervals (Figure \ref{fig:sd}A).
 This implies that the decrease in standard deviation as sample size increases is lesser in a multi-species model than a single-species model.
@@
 
+\par (NEEDS UPDATING) We found little evidence that a multi-species model improves the precision of EDR compared to that of a single-species model.
+The 95\% credible intervals for both estimated slopes and intercepts overlapped with each other, indicating that the standard deviations are effectively the same for a single-species and multi-species estimate as sample size increases (Figure \ref{fig:sd}A).
+
 \begin{figure}[h!]
-	\includegraphics{../output/plots/removal_sd_plot.png}
-	\caption{Posterior distributions with 95\% credible intervals (shaded areas) of (A) modelled intercepts of standard deviation for the single-species removal model (alpha[1]) and multi-species removal model (alpha[2]), and (B) modelled slopes of standard deviation with increasing sample size for the single-species removal model (beta[1]) and multi-species removal model (beta[2]).}
-	\label{fig:removal_sd}
+	\includegraphics{../output/plots/sd_comp.png}
+	\caption{Posterior distributions with 95\% credible intervals (shaded areas) of (A) modelled intercepts of standard deviation for the single-species removal model (alpha[1]) and multi-species removal model (alpha[2]), and (B) modelled slopes of standard deviation with increasing sample size for the single-species removal model (beta[1]) and multi-species removal model (beta[2]).
+Posterior distributions with 95\% credible intervals (shaded areas) of (A) modelled intercepts of standard deviation for the single-species distance model (alpha[1]) and multi-species distance model (alpha[2]), and (B) modelled slopes of standard deviation with increasing sample size for the single-species distance model (beta[1]) and multi-species distance model (beta[2]).}
+	\label{fig:sd}
 \end{figure}
+
+\subsection{Comparison of Predictive Accuracy between Single Species and Multi Species}
 
@@
 Mean overall difference in posterior predictive density was $-2.21 \times 10^{-5}$.
-Single-species models tended to have better predictive capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:removal_cv}).
+Single-species models tended to have better predictive capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:cv}A).
 There did not appear to be any patterns in model selection among phylogeny or traits (see Supplemental Figure 1 for full phylogenetic tree).
@@
 
+\par The cross validation revealed that the multi-species model tended to have slightly more support than the single-species model in terms of predicting EDR.
+However, similar to the removal model cross-validation results, the overall difference in posterior predictive density was very small, with a mean difference of $1.85 \times 10^{-4}$.
+Multi-species models tended to have better predictive capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:cv}B).
+Interestingly, only two (Chimney Swift and Boat-tailed Grackle) of the six species with a greater than 20\% change in EDR had the multi-species model chosen as the better predictive model.
+
 \begin{figure}[h!]
-	\includegraphics{../output/plots/cv_removal_diff_vs_n.png}
+	\includegraphics{../output/plots/cv_plot.png}
 	\caption{Mean difference in pointwise posterior predictive density.
@@
 Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species model preference, whereas negative differences (coloured in blue for clarity) represent a single-species model preference.
-Regression line shows the relationship between mean difference and species sample size.}
-	\label{fig:removal_cv}
-\end{figure}
-
-\subsection{Distance Model Comparisons}
-
-\par We found some evidence for a change in modelled EDR between single species and multi-species distance models, in that the multi-species model tended to produce slightly smaller estimates of EDR (Figure \ref{fig:distance_1vs1}A).
-Still, the majority of the species had very little change; six out of the 317 species had a percent change in EDR of more than 20\% (Figure \ref{fig:distance_1vs1}B).
-These species were Sharp-tailed Grouse (\textit{Tympanuchus phasianellus}), Sedge Wren (\textit{Cistothorus stellaris}), Ring-necked Pheasant (\textit{Phasianus colchicus}), Chimney Swift (\textit{Chaetura pelagica}), Boat-tailed Grackle (\textit{Quiscalus major}), and Baird's Sparrow (\textit{Ammodramus bairdii}).
-
-\begin{figure}[h!]
-	\includegraphics{../output/plots/distance_1vs1.png}
-	\caption{Modelled difference with 95\% credible interval (shaded area) between EDR estimated from the multi-species model developed here, minus the single-species model in NA-POPS \citep{edwards_point_2023} (A); and 1-to-1 plot comparing EDR estimated from the single-species model vs the multi-species model for each species (B).}
-	\label{fig:distance_1vs1}
-\end{figure}
-
-\par We found little support for migration strategy having any effect on EDR.
-There is some evidence that a resident species may have a slightly higher EDR than a migrant species, but the 95\% credible intervals of both effects overlapped with the 95\% credible interval of the overall estimated intercept (Figure \ref{fig:distance_params}A).
-We did find similar support for change in EDR by body size and pitch, in that body size tended to have an increasing effect on EDR, and song pitch had a decreasing effect on EDR (Figure \ref{fig:distance_params}B).
-
-\begin{figure}[h!]
-	\includegraphics{../output/plots/distance_other_params.png}
-	\caption{Posterior distributions with 95\% credible intervals (shaded area) of parameters estimated in the multi-species distance model, with (A) overall mean log EDR as the intercept, with departures from the intercept modelled as migratory strategy (resident: mu\_mig\_strat[1], migrant: mu\_mig\_strat[B]), (B) slope parameters of changes in EDR given change in mass and pitch, and (C) extra variance term after accounting for traits.}
-	\label{fig:distance_params}
-\end{figure}
-
-\par We found little evidence that a multi-species model improves the precision of EDR compared to that of a single-species model.
-The 95\% credible intervals for both estimated slopes and intercepts overlapped with each other, indicating that the standard deviations are effectively the same for a single-species and multi-species estimate as sample size increases (Figure \ref{fig:distance_sd}).
-
-\begin{figure}[h!]
-	\includegraphics{../output/plots/distance_sd_plot.png}
-	\caption{Posterior distributions with 95\% credible intervals (shaded areas) of (A) modelled intercepts of standard deviation for the single-species distance model (alpha[1]) and multi-species distance model (alpha[2]), and (B) modelled slopes of standard deviation with increasing sample size for the single-species distance model (beta[1]) and multi-species distance model (beta[2]).}
-	\label{fig:distance_sd}
-\end{figure}
-
-\par The cross validation revealed that the multi-species model tended to have slightly more support than the single-species model in terms of predicting EDR.
-However, similar to the removal model cross-validation results, the overall difference in posterior predictive density was very small, with a mean difference of $1.85 \times 10^{-4}$.
-Multi-species models tended to have better predictive capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:distance_cv}).
-Interestingly, only two (Chimney Swift and Boat-tailed Grackle) of the six species with a greater than 20\% change in EDR had the multi-species model chosen as the better predictive model.
-\begin{figure}[h!]
-	\includegraphics{../output/plots/cv_distance_diff_vs_n.png}
-	\caption{Mean difference in pointwise posterior predictive density.
+Regression line shows the relationship between mean difference and species sample size.
+Mean difference in pointwise posterior predictive density.
 Points are mean difference by species.
@@
 Regression line shows the relationship between mean difference and species sample size.}
-	\label{fig:distance_cv}
+	\label{fig:cv}
 \end{figure}
@@
 
-\par The multi-species removal model produced reasonable estimates of cue rates for all seven species of bird considered here (Figure \ref{fig:removal_predictions}).
+\par The multi-species removal model produced reasonable estimates of cue rates for all seven species of bird considered here (Figure \ref{fig:predictions}A).
 Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush and Kirtland's Warbler) had reasonable precision associated with the estimates.
@@
 
-\begin{figure}[h!]
-	\includegraphics{../output/plots/removal_predictions_plot.png}
-	\caption{Cue rates (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for seven species of rare and undersampled North American landbirds.
-Species are 4 letter banding codes: SPOW = Spotted Owl (n = 7), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), TRBL = Tricoloured Blackbird (n = 74), HASP = Harris's Sparrow (n = 65), KIWA = Kirtland's Warbler (n = 0), and LEPC = Lesser Prairie-chicken (n = 45).}
-	\label{fig:removal_predictions}
-\end{figure}
-
-\par The multi-species distance model also produced reasonable estimates of EDR for all six species of landbird considered here (Figure \ref{fig:distance_predictions}).
+\par The multi-species distance model also produced reasonable estimates of EDR for all six species of landbird considered here (Figure \ref{fig:predictions}B).
 Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush) had fairly precise estimates, except for LeConte's Thrasher and Lesser Prairie Chicken where the estimated EDR spanned greater than 50 m in either direction.
@@
 \begin{figure}[h!]
-	\includegraphics{../output/plots/distance_predictions_plot.png}
-	\caption{Effective detection radius (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for six species of rare and undersampled North American landbirds.
+	\includegraphics{../output/plots/predictions_figure.png}
+	\caption{Cue rates (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for seven species of rare and undersampled North American landbirds.
+Species are 4 letter banding codes: SPOW = Spotted Owl (n = 7), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), TRBL = Tricoloured Blackbird (n = 74), HASP = Harris's Sparrow (n = 65), KIWA = Kirtland's Warbler (n = 0), and LEPC = Lesser Prairie-chicken (n = 45).
+Effective detection radius (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for six species of rare and undersampled North American landbirds.
 Species are 4 letter banding codes: SPOW = Spotted Owl (n = 28), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), HASP = Harris's Sparrow (n = 13), KIWA = Kirtland's Warbler (n = 1), and LEPC = Lesser Prairie-chicken (n = 45).}
-	\label{fig:distance_predictions}
+	\label{fig:predictions}
 \end{figure}

### 2023-12-07 src/scripts/post-hoc/3-cv-figure.R

@@ -287,3 +287,8 @@ tiff(filename = "output/plots/cv_plot.tiff",
     width = 6, height = 6, units = "in", res = 600)
 ggarrange(lppd_diff_vs_n_rem, lppd_diff_vs_n_dis, nrow = 2, labels = c("A", "B"))
 dev.off()
+
+png(filename = "output/plots/cv_plot.png",
+     width = 6, height = 6, units = "in", res = 600)
+ggarrange(lppd_diff_vs_n_rem, lppd_diff_vs_n_dis, nrow = 2, labels = c("A", "B"))
+dev.off()

### 2023-12-07 article/article.tex
got rid of specific predictions subseciton and just turned it into a paragraph

@@
 
-\subsection{Predictions}
-\begin{enumerate}
-	\item\label{pred1} For both the Bayesian hierarchical QPAD removal model and Bayesian hierarchical QPAD distance model, estimates of cue rate and EDR, respectively, will change very little from those estimated from the original NA-POPS database \cite{edwards_point_2023}.
+\par We predicted that for both the Bayesian hierarchical QPAD removal model and Bayesian hierarchical QPAD distance model, estimates of cue rate and EDR, respectively, will change very little from those estimated from the original NA-POPS database \cite{edwards_point_2023}.
 This is because the species that were modelled in \citet{edwards_point_2023} had a sufficient number of data points to stand on their own, and so adding in species-level effects should have little effect on estimates.
-	\item\label{pred2} Improvement in precision will be related to sample size, such that precision of estimates will improve the most for species with smaller sample size, while precision of estimates will improve the least for species with larger sample sizes.
-	\item\label{pred3} A multi-species model will have greater predictive capacity than a single-species model, particularly for species which are data-sparse.
-\end{enumerate}
-
-\par With the predictions above, we then expect to be able to generate estimates of cue rate and EDR for species with very little data, thus relaxing previous sample size requirements; additionally, we expect to be able to predict cue rate and EDR for species with no data by making use of the shrinkage effect offered by a multi-species model.
+Furthermore, we predicted that improvement in precision from a multi-species model will be related to sample size, such that precision of estimates will improve the most for species with smaller sample size, while precision of estimates will improve the least for species with larger sample sizes.
+Finally, we predicted that a multi-species model will have greater predictive capacity than a single-species model, particularly for species which are data-sparse.
+With the predictions above, we then expect to be able to generate estimates of cue rate and EDR for species with very little data, thus relaxing previous sample size requirements; additionally, we expect to be able to predict cue rate and EDR for species with no data by making use of the shrinkage effect offered by a multi-species model.
 

### 2023-12-07 article/article.tex
updated methods, results, and discussion of article based on j and a suggestions. still waiting for distance CV to finish

@@
 
-\par To test prediction \ref{pred1}, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data in the original NA-POPS analysis.
+\par To test our prediction that there would be little change between estimates of species already modelled in NA-POPS, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data in the original NA-POPS analysis.
 In this case, we ran both models for 316 species of birds.
@@
 where $\Delta X$ is calculated as the mean multi-species estimate of cue rate or EDR minus the mean single-species estimate of cue rate of EDR.
-If prediction \ref{pred1} is true, we would expect to see a distribution of modelled differences centred about 0, indicating no real difference in estimates of cue rate or EDR between single-species and multi-species models, for species for which we have sufficient data.
+If our prediction is true, we would expect to see a distribution of modelled differences centred about 0, indicating no real difference in estimates of cue rate or EDR between single-species and multi-species models, for species for which we have sufficient data.
 
-\par To test prediction \ref{pred2}, we modelled the standard deviations about each estimate of cue rate or EDR from both the multi-species and single-species models, and regressed them against the log sample size per species.
-That is
-$$SD \sim N(\alpha_M + \beta_M\times \log(N), \sigma)$$
-$$\alpha, \beta \sim N(0,1)$$
-$$\sigma \sim exp(1)$$
+\par To test our prediction about improved precision, we simply plotted the species-specific estimates of standard deviations of cue rate and EDR, respectively, against the log sample size, for each of the single-species and multi-species model.
+We then fit a smooth LOESS curve through each of the model-specific points.
+If our prediction is true, we would expect to see a plot showing a higher standard deviation in single-species models than for multi-species models, but only up until a certain sample size when they should both be similar.
 
-where $\alpha$ is the model-specific intercept and $\beta$ is the model-specific slope.
-In both cases, $M=1$ corresponds to the single species model and $M=2$ corresponds to the multi-species model.
-
-\par If prediction \ref{pred2} is true, we would expect to see a higher value of $\alpha_1$ than for $\alpha_2$, corresponding to a higher baseline standard deviation in the single species models.
-We would also likely see a slightly smaller (i.e., more negative) value of $\beta_1$ than $\beta_2$ (which should also be negative), indicating that the standard deviation decreases with log sample size for both models, but should decrease at a slower rate with the multi-species model (which has a lower baseline standard deviation to begin with) than the single-species model.
-
-\par To test prediction \ref{pred3}, we ran a k-fold cross-validation with 5 folds, stratified by species.
+\par To test our prediction about improved accuracy, we ran a k-fold cross-validation with 5 folds, stratified by species.
 That is, for each cross-validation run, we held out 20\% of the data for each species for both the multi-species and Bayesian single-species models. 
-We then derived multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR for each cross-validation run, and used these parameter estimates to calculate the posterior predictive density of each of the held out data points.
+We then derived multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR, for each cross-validation run, and used these parameter estimates to calculate the estimated pointwise log posterior density of each of the held out data points.
 We then calculated the difference in posterior predictive density for removal and distance models, by subtracting the pointwise posterior predictive density of the single-species model from the pointwise posterior predictive density of the multi-species model (i.e., multi-species minus single-species).
-Finally, we took the mean of the difference in pointwise posterior predictive density for each species to create a species-specific mean difference in pointwise predictive density.
-With this, positive values would indicate that the multi-species model was preferred, and negative values would indicate that the single-species model was preferred.
-If prediction \ref{pred3} is true, then we would expect to see majority positive values of the species-specific mean pointwise predictive density.
+Finally, we took the mean and standard error of the difference in pointwise log posterior density for each species to create a species-specific mean and standard error of difference in pointwise posterior density.
+With this, positive intervals (i.e., where mean $\pm$ standard error is all positive) would indicate that the multi-species model was preferred, and negative intervals would indicate that the single-species model was preferred.
+If our prediction is true, then we would expect to see majority positive values of the species-specific mean pointwise posterior density.
 
@@
 \par To predict estimates of cue rate or EDR from the multi-species models, we adopted two possible approaches.
-If the NA-POPS database contained some point-count data for the species of interest, we simply added these data into the model.
+If the NA-POPS database contained at least some point-count data for the species of interest, we simply added these data into the model.
 This is different from \citet{edwards_point_2023, solymos_calibrating_2013, solymos_evaluating_2018}, where any species with less than 75 data points were removed from the analysis.
@@
 All Bayesian models were coded in the Stan Probabilistic Language \cite{stan_development_team_stan_2019}, and models were run using the \texttt{cmdstanr} R package \citep{gabry_cmdstanr_2022} and visualized using the \texttt{bayesplot} R package \citep{gabry_visualization_2019}.
-Each of the single-species models and multi-species models running both the initial comparison set of species and the final predictive set of species, as well as the models created to assess performance, were run using 1000 warmup iterations and 2000 sampling iterations, each on 4 chains for a total of 8000 draws per model run.
+Each of the single-species models and multi-species models, as well as the models created to assess performance, were run using 1000 warmup iterations and 2000 sampling iterations, each on 4 chains for a total of 8000 draws per model run.
 To mitigate issues surrounding initial values being rejected in the distance model (see Discussion), we set the initial values of EDR to be the mean EDR from NA-POPS estimates; for species where EDR estimates did not exist in NA-POPS, we used the overall mean EDR as initial value.
@@
 
-\par (NEEDS UPDATING) We found some evidence that a multi-species removal model improved the precision of cue rate estimates compared to estimates from a single-species model.
-The intercept of standard deviations of cue rates estimated in a single species model was greater than the intercept of standard deviations of cue rates estimated in a multi-species model, with very little overlap in their 95\% credible intervals (Figure \ref{fig:sd}A).
-This implies that the baseline standard deviation of cue rates is higher in a single species model than a multi species model for very low sample sizes.
-However, the slope of standard deviations of cue rates estimated in a single species model was farther from 0 (i.e., more negative) than the slope of standard deviations of cue rates estimated in a multi-species model, with very little overlap in their 95\% credible intervals (Figure \ref{fig:sd}A).
-This implies that the decrease in standard deviation as sample size increases is lesser in a multi-species model than a single-species model.
-This indicates that after a certain sample size, it is likely that the difference in standard deviation between a single species and multi species model would be negligible.
+\par We found some evidence that a multi-species removal model improved the precision of cue rate estimates compared to estimates from a single-species model (Figure \ref{fig:sd}A).
+This was particularly evident for species with sample sizes less than approximately $\exp\left\{7\right\} \approx 1097$, where the estimated LOESS curve for single-species standard deviation was always higher than the estimated LOESS curve for multi-species standard deviations.
+For sample sizes greater than 1097, standard deviations appeared similar in both the single-species and multi-species model, indicating no real improvement in precision.
 
-\par (NEEDS UPDATING) We found little evidence that a multi-species model improves the precision of EDR compared to that of a single-species model.
-The 95\% credible intervals for both estimated slopes and intercepts overlapped with each other, indicating that the standard deviations are effectively the same for a single-species and multi-species estimate as sample size increases (Figure \ref{fig:sd}A).
+\par We found little evidence that a multi-species model improves the precision of EDR compared to that of a single-species model (Figure \ref{fig:sd}B).
+Both estimated LOESS curves for single-species standard deviations and multi-species standard deviations were roughly equal across all sample sizes, indicating no real improvement (or decrease) in precision.
 
@@
 
-\par The cross validation revealed that single species model actually had better predictive capacity for more species (222 out of 316) than the multi-species model (94 out of 316), but the difference in predictive capacity was overall very small.
-Mean overall difference in posterior predictive density was $-2.21 \times 10^{-5}$.
-Single-species models tended to have better predictive capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:cv}A).
+\par The cross validation revealed that multi-species model had better predictive capacity for more species (200 out of 316) than the single-species model (99 out of 316).
+Seventeen out of 316 species showed no better predictive capacity with either model.
+Mean overall difference in log predictive density was $4.78 \times 10^{-6}$.
+As expected, multi-species models tended to have better predictive capacity when sample size was low, with a negligible difference as sample size increased (Figure \ref{fig:cv}A).
 There did not appear to be any patterns in model selection among phylogeny or traits (see Supplemental Figure 1 for full phylogenetic tree).
-Six out of the 7 species that had a percent change in cue rate of more than 20\% had the multi-species model chosen as the better predictive model.
-Common Redpoll was the only species that had the single-species model chosen (but please see Discussion about this particular species).
 
-\par The cross validation revealed that the multi-species model tended to have slightly more support than the single-species model in terms of predicting EDR.
+\par (NEEDS UPDATING) The cross validation revealed that the multi-species model tended to have slightly more support than the single-species model in terms of predicting EDR.
 However, similar to the removal model cross-validation results, the overall difference in posterior predictive density was very small, with a mean difference of $1.85 \times 10^{-4}$.
@@
 
-\par The multi-species removal model produced reasonable estimates of cue rates for all seven species of bird considered here (Figure \ref{fig:predictions}A).
+\par The multi-species removal model produced reasonable estimates of cue rates for all seven species considered here (Figure \ref{fig:predictions}A).
 Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush and Kirtland's Warbler) had reasonable precision associated with the estimates.
@@
 
-\par The multi-species distance model also produced reasonable estimates of EDR for all six species of landbird considered here (Figure \ref{fig:predictions}B).
+\par The multi-species distance model also produced reasonable estimates of EDR for all six species considered here (Figure \ref{fig:predictions}B).
 Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush) had fairly precise estimates, except for LeConte's Thrasher and Lesser Prairie Chicken where the estimated EDR spanned greater than 50 m in either direction.
@@
 
+\par In both cases of QPAD removal modelling and QPAD distance modelling, we were able to generate estimates of detectability for species with no data, and precise estimates of detectability for data-sparse species.
+Overall, the shrinkage effect offered by the multi-species modelling approach allowed for more accurate estimates of cue rate and EDR to be derived in species with smaller sample sizes, without affected estimates in species with large sample sizes.
+The best model for predictive accuracy tended to depend on the species, but overall differences in the estimates were small enough that using a multi-species model should not produce extremely inaccurate estimates.
+
+\par Although this was not a primary goal of our study, when considering the species-level covariates, we found much less support than \citet{solymos_phylogeny_2018} for migration strategy having an effect on cue rate and EDR.
+Both the estimated effects for resident species and migratory species were effectively 0 in the removal model, with very little of the density above or below 0.025 from 0.
+These estimates varied slightly more in the distance model, but the 95\% credible interval still captured 0.
+One potential reason is that we had proportionally more resident species in our analysis (85 out of 316; 26.9\%) compared to \citet{solymos_phylogeny_2018} (23 out of 141; 16.3\%), and so the richer information provided by these additional species could have brought the effect down to 0.
+Habitat type, body mass, and song pitch, as expected, all had an overall effect on EDR, as was found in \citet{solymos_phylogeny_2018} and \cite{johnston_species_2014}.
+
+\subsection{Predicting Detectability for Data-sparse Species}
+
+\par For five out of the seven species for which we generated new predictions of detectability in this study, we were interested in seeing how a multi-species QPAD removal model could estimate cue rate with few data, and how a multi-species QPAD distance model could estimate EDR with few data.
+
+\par Overall, the estimates of cue rate fell roughly where we would expect based on related species, for species for which we were simply relaxing the sample size requirements.
+For example, Harris's Sparrow (n = 65) had an estimated cue rate of 0.30 cues per minute, with a 95\% credible interval of [0.21, 0.42], compared to the overall mean cue rate of 0.37 cues per minute for all 32 other Passerellidae species in NA-POPS.
+Similarly, LeConte's Thrasher (n = 1) had an estimated cue rate of 0.32 cues per minute, with a 95\% credible interval of [0.18, 0.57].
+Though this was a much larger interval than Harris's Sparrow, it still falls within a reasonable distance from the overall mean cue rate of 0.36 cues per minute for all 6 other Mimidae species in NA-POPS.
+
+\par Estimates of EDR also fell roughly where we would expect based on species of similar traits, for the species for which we were relaxing samples sizes.
+For example, Harris's Sparrow (n = 13) had an estimated mean EDR of 65.6 m, with a 95\% credible interval of [53.1, 82.5].
+This range does fall below the mean EDR of 85.3 m for the other sparrows in NA-POPS, but Harris's Sparrow song pitch of approximately 3.8 Hz falls above the average song pitch of 3.0 Hz for the other sparrows in NA-POPS, which could contribute to the smaller EDR.
+On the other hand, LeConte's Thrasher (n = 1) had an estimated EDR of 134.2 m, with a 95\% credible interval of [81.8, 230.5], which falls roughly in line with the mean EDR of 113.2 m of the six other thrashers in NA-POPS.
+
+\par It is likely that the width of the credible intervals (i.e., the precision) of these relaxed sample-size species will depend on the number of data points available for that particular species, the number of "similar" species, and the number of data points available for those similar species.
+Additionally, it is important to note that we did not account for other sources of variation at the survey level, such as when the survey was conducted (i.e., time of day, time of year), or where the survey was conducted (e.g., in a forest, by a roadside).
+It is well known that the ability to detect a species can be very dependent on the time of day or year that an observer is surveying, as well as the environmental conditions of the survey \citep{edwards_point_2023}.
+Thus, some of the variation in these estimates of cue rate and EDR could be due to not accounting for survey-level variation that affects detectability, and so it is important that future iterations of this model incorporate both species-level effects on detectability \textit{and} survey-level effects on detectability.
+
+\subsection{Predicting Detectability for Species with No Data}
+
+\par We generated predictions of cue rate and EDR for both Bicknell’s Thrush and Kirtland’s Warbler.
+For both species, we had zero removal sampling data, and so cue rates were solely based on information sharing among phylogeny (and to a much lesser extent, migration status). 
+For Kirtland’s Warbler, we did have one data point for distance sampling, so predictions of EDR for Kirtland’s Warbler here are not entirely based on zero data points.
+Bicknell’s Thrush, however, had zero data points, and so EDRs for this species were based solely on information sharing among other species with similar traits.
+
+\par As expected, the predicted cue rate for Bicknell’s Thrush (0.28 cues per minute) and the predicted cue rate for Kirtland’s Warbler (0.29 cues per minute) fell very close to the overall mean cue rate of 0.27 for all 11 other Turdidae species and 0.31 for all 44 other Parulidae species in NA-POPS, respectively.
+That is, in the absence of any data, the estimated cue rate of Bicknell’s Thrush and Kirtland’s Warbler was shrunk toward cue rates of most similar species.
+
+\par For distance modelling, the predicted EDR for Bicknell's Thrush (73.6 m) and the predicted EDR for Kirtland's Warbler (75.3 m) fell reasonably close to the overall mean EDR of 92.7 m for all other 11 Turdidae species and 61.2 m for all 44 other Parulidae species in NA-POPS, respectively.
+hat is, in the absence of any data (or with very little data in the case of Kirtland’s Warbler), the estimated EDR of Bicknell’s Thrush and Kirtland’s Warbler was similar to those of other Thrushes and Warblers, respectively, but with departures from overall mean based on specific traits.
+
+\par In both the removal modelling and distance modelling, the magnitude of contribution of data from related species appeared to play a role in the width of the credible intervals. 
+For removal modelling, although both credible intervals were fairly wide, Bicknell’s Thrush had a 95\% credible interval of [0.16, 0.48], representing an approximate 53\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [0.18, 0.48], representing an approximate 49\% change on either end.
+For distance modelling, both credible intervals were also fairly wide, but Bicknell's Thrush had a 95\% credible interval of [40.5, 134.4], representing an approximate 58\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [45.9, 125.6], representing an approximate 47\% change on either end.
+Although the difference in width of the intervals is small, this could be due to the greater number of other species contributing information to Kirtland's Warbler estimates than to Bicknell's Thrush.
+In the case of distance modelling, the one data point contributing to Kirtland's Warbler's EDR could have also played a part.
+With that in mind, it would likely be important to consider how many "related" species are available with information (and how many data points do each of those related species have) when deciding on making a prediction of cue rate.
+
+\subsection{Comparisons of Predicted Detectability to Other Studies}
+
+\par Using the predictions of cue rate and EDR for Bicknell’s Thrush and Kirtland’s Warbler, we can calculate availability and perceptibility, and hence detectability, for a given survey type, and compare to previous estimates of detectability obtained from other studies.
+As a reminder, following derivations in \citet{solymos_calibrating_2013}, detectability is the product of availability and perceptibility.
+Availability $p(t)$ is defined as the probability of a bird giving a cue during a survey run for $t$ minutes, and is calculated as
+$p(t) = 1 - e^{-t\phi}$
+where $\phi$ is the cue rate of the species.
+Perceptibility $q(r)$ is defined as the probability of an observer hearing or seeing a bird, given the bird gives a cue, with a survey radius of $r$ metres, and is calculated as
+$q(r) = \dfrac{\tau^2 \left\{1 - e^{-\dfrac{r^2}{\tau^2}}\right\}}{r^2}$
+where $\tau$ is the EDR of the species.
+
+\par For Bicknell's Thrush, using the equations above, the predicted cue rate of 0.28, predicted EDR of 73.3 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.34.
+We can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respectively credible intervals) to get a range of detectabilities from 0.09 through 0.70.
+\citet{aubry_bicknells_2018}, however, estimate detection probabilities of $\geq$ 0.74 and $\leq$ 0.88, and so even our highest potential value of detectability falls short of their estimated range.
+However, we note that \citet{aubry_bicknells_2018} are using data that specifically target Bicknell's Thrush in their range, and they are using surveys conducted at peak Bicknell's Thrush breeding time, and so it is likely that detectability will be high in that specific case.
+
+\par For Kirtland's Warbler, using the equations above, the predicted cue rate of 0.29, predicted EDR of 75.3 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.36.
+Similar to our Bicknell's Thrush example, we can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respective credible intervals) to obtain a range of detectabilities from 0.12 through 0.67.
+A study by \citet{van_dyke_comparative_2022} found estimated detection probabilities for Kirtland’s Warbler that ranged from effectively 0, through to a detection probability of 0.73
+In that study, they compared estimates of detectability among different jack pine stands and red pine stands, where the detectability was higher in jack pine stands than the red pine stands.
+
 \subsection{Model Performance}
@@
 \par The multi-species QPAD Distance model had several issues related to initialization, sampling, and convergence throughout this study.
-For many iterations of the model development, Stan's generation of initial values proved difficult.
+For many iterations of the model development, generation of initial values using Stan proved difficult.
 For all unconstrained parameters, Stan draws initial values from a uniform(-2,2) distribution.
@@
 \par The distance models also seemed to struggle with sampling the posterior distribution, in that over a quarter of the iterations resulted in a divergent transition \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
-This is likely due to hierarchical nature of the model and the parameterization choices that we made.
-\citet{betancourt_hierarchical_2020} discuss at length some issues that arise when sampling from hierarchical models in cases where the hierarchical groups are data rich, where a "centred parameterization" should be used, or data poor, where a "non-centred parameterization" should be used.
-We experimented with using only a non-centred parameterization which resulted in extremely low effective sample sizes (as low as 4 for 8000 iterations), large R-hat values, and took an unreasonable amount of time to finish sampling (approximately 2.5 weeks).
-The model presented in this paper is a centred parameterization, which produced much more reasonable effective sample sizes and R hat values, finished within a day and a half, but still suffered from degeneracies such as hierarchical funnelling as described by \citet{betancourt_hierarchical_2020}.
-In our case, because we are modelling on a hierarchy that contain many data points for some species x trait groups and fewer data points for other species x trait groups, our divergent transitions may be resolved if we were to implement our model using a mixture of centred and non-centred parametrization as described by \citet{betancourt_hierarchical_2020}.
+However, after several iterations of this model, we identified that the samplers were not appropriately sampling Rufous Hummingbird's EDR, in that there appeared to be a self-imposed barrier in being able to sample EDRs less than 25 m.
+Once we removed Rufous Hummingbird from the analysis, all divergent transitions vanished.
+We demonstrate some of the odd sampling of the posterior for Rufous Hummingbird in Supplemental Materials B.
+One potential explanation is that Rufous Hummingbird had the smallest EDR of all the species in NA-POPS, and so there could be an issue with numerical stability of sampling very low values, due to the exponentiation of a quotient in the distance modelling likelihood.
+Future studies should seek to test the limits of this distance model, particularly in a Bayesian setting, using simulated data with very low EDRs.
 
@@
 
-\subsection{Removal Modelling}
-\par Estimates of cue rate changed very little from species that were originally modelled by the single-species models in \citet{edwards_point_2023}, to the same ones modelled in the multi-species model here.
-The 5-fold cross validation tended to show no preference for one model or another when sample size was large, and a preference (albeit an extremely small preference) for single-species models when sample size was smaller.
-When sample size was small, a multi-species model tended to produce more precise estimates of cue rate.
-Given the extremely small preference for single-species models under a lower sample size, it would appear that the bias-variance trade-off is likely a welcome trade-off here, because the decrease in variance is met by only a very slight increase in bias.
-
-\par There were seven species where the relative difference in cue rate between the single-species model and multi-species model was over 20\% (noting that this threshold was chosen arbitrarily).
-These species (Willow Ptarmagin, Rusty Blackbird, Red-breasted Sapsucker, Common Redpoll, Calliope Hummingbird, Allen's Hummingbird, and Bohemian Waxwing), mostly had fairly small sample sizes ($\leq 515$ for all species except Red-breasted Sapsucker [$n = 1854$] and Common Redpoll [$n = 2009$]), had the lowest cue rate in the single-species models, and had among the lowest cue rates in the multi-species model.
-For removal modelling, all those species except for Common Redpoll ended up having the multi-species model chosen as the more predictive model, indicating that sharing information among related species tended to improve their estimates of cue rate.
-Common Redpoll was likely an exception to this because the original estimate of cue rate in NA-POPS was a wild underestimate (cue rate of $5.9\times 10^{-9}$), likely because of a convergence issue in the \texttt{detect} package.
-It could be worth running Common Redpoll using the Bayesian single species model developed here to see if the estimate of cue rate is still extremely low, or if the vague but still somewhat informative prior helps to bring the estimated parameter into a more sensible space.
-It could also be worth investigating the merit of this type of modelling to estimate cue rate for species where the cue rate is extremely low.
-In our case, 41 out of the 79 species that had the lowest 25\% quantile of cue rate in the single-species model ended up having the multi-species model chosen as the preferred model.
-
-\par Although this was not a primary goal of our study, we found much less support than \citet{solymos_phylogeny_2018} for migration strategy having an effect on cue rate.
-Both the estimate effects for resident species and migratory species were effectively 0, with very little of the density above or below 0.025 from 0.
-One potential reason is that we had proportionally more resident species in our analysis (85 out of 316; 26.9\%) compared to \citet{solymos_phylogeny_2018} (23 out of 141; 16.3\%), and so the richer information provided by these additional species could have brought the effect down to 0.
-Additionally, while \citet{solymos_phylogeny_2018} only regressed estimates of cue rate against the predictor variables, we actually estimated cue rate within the context of the same predictors, and so the phylogenetic effect could have just overwhelmed the estimates, making the migratory strategy meaningless.
-
-\subsection{Distance Modelling}
-\par Estimates of EDR also changed very little, albeit a non-zero amount, from species that were originally modelled by the single-species models in \citet{edwards_point_2023}, to the same ones modelled in the multi-species model here.
-However, there was evidence for an average small downward shift in EDRs when modelled under a multi-species context.
-Still, the 5-fold cross validation tended to show almost no preference for one model or another when the sample size was large, and a slight (but still very small) preference for the multi-species model when the sample size was small.
-There was less evidence for the multi-species model being able to improve the precision of EDR estimates across sample sizes, indicating that a single-species model may mostly be sufficient for precise estimates.
-Under this context, it would appear that a multi-species model may slightly bring down estimates of EDR for species with a smaller sample size.
-
-\par There were six species where the relative difference in EDR between the single-species model and multi-species model was over 20\%. 
-Unlike removal modelling, these species (Sharp-tailed Grouse, Sedge Wren, Ring-necked Pheasant, Chimney Swift, Boat-tailed Grackle, and Baird's Sparrow) had little in common with each other in terms of sample size (range of 227 for Boat-tailed Grackle through 4466 for Baird's Sparrow) and EDR (ranging from 50.8 m to 226 m in single-species, and from 73.2 m to 159.1 m in the multi-species model).
-Additionally, there was mixed support among these species for single-species vs.
-multi-species: Chimney Swift and Boat-tailed Grackle had the multi-species model preferred, whereas the others had the single-species model preferred.
-It could be noted that both Chimney Swift and Boat-tailed Grackle showed an increase in estimated EDR from single-species to multi-species modelling and preferred the multi-species model, but Baird's Sparrow also showed an increase and preferred the single-species model.
-
-\par Similar to removal modelling, we found less support than \citet{solymos_phylogeny_2018} for migration strategy having an effect on EDR, although the posterior distribution for both the resident and migrant species were much further away from zero, indicating at least some support.
-Habitat type, body mass, and song pitch, as expected, all had an overall effect on EDR, as was found in \citet{solymos_phylogeny_2018} and \cite{johnston_species_2014}.
-
-\subsection{Relaxation of Sample Size}
-
-\par For five out of the seven species where we generated new predictions of detectability in this study, we were interested in seeing how a multi-species QPAD removal model could estimate cue rate with few data, and how a multi-species QPAD distance model could estimate EDR with few data.
-
-\par Overall, the estimates of cue rate fell roughly where we would expect based on related species, for species where we were simply relaxing the sample size requirements.
-For example, Harris's Sparrow (n = 65) had an estimated cue rate of 0.30 cues per minute, with a 95\% credible interval of [0.21, 0.42], compared to the overall mean cue rate of 0.37 cues per minute for all 32 other Passerellidae species in NA-POPS.
-Similarly, LeConte's Thrasher (n = 1) had an estimated cue rate of 0.32 cues per minute, with a 95\% credible interval of [0.18, 0.57].
-Though this was a much larger interval than Harris's Sparrow, it still falls within a reasonable distance from the overall mean cue rate of 0.36 cues per minute for all 6 other Mimidae species in NA-POPS.
-
-\par Estimates of EDR also fell roughly where we would expect based on species of similar traits, for the species where we were relaxing samples sizes.
-For example, Harris's Sparrow (n = 13) had an estimated mean EDR of 65.6 m, with a 95\% credible interval of [53.1, 82.5].
-This range does fall below the mean EDR of 85.3 m for the other sparrows in NA-POPS, but Harris's Sparrow song pitch of approximately 3.8 Hz falls above the average song pitch of 3.0 Hz for the other sparrows in NA-POPS, which could contribute to the smaller EDR.
-On the other hand, LeConte's Thrasher (n = 1) had an estimated EDR of 134.2 m, with a 95\% credible interval of [81.8, 230.5], which falls roughly in line with the mean EDR of 113.2 m of the six other thrashers in NA-POPS.
-
-\par Of course, it is likely that the width of the credible intervals (i.e., the precision) of these relaxed sample-size species are going to depend both on the number of data points available for that particular species, the number of "similar" species, and the number of data points available for those similar species.
-However, species that do not have many other similar species from which to borrow information still have the benefit of using information from the priors that we set in the model, which is a large advantage to modelling detectability in a Bayesian context.
-Additionally, it is important to note that we did not account for other sources of variation at the survey level, such as when the survey was conducted (i.e., time of day, time of year), or where the survey was conducted (e.g., in a forest, by a roadside).
-It is well known that the ability to detect a species can be very dependent on the time of day or year that an observer is surveying, as well as the environmental conditions of the survey \citep{edwards_point_2023}.
-Thus, some of the variation in these estimates of cue rate and EDR could be due to not accounting for survey-level variation that affects detectability, and so it is important that future iterations of this model incorporate both species-level effects on detectability \textit{and} survey-level effects on detectability.
-
-\subsection{Predicting Detectability for Species with No Data}
-
-\par We generated predictions of cue rate and EDR for both Bicknell's Thrush and Kirtland's Warbler.
-For both species, we had zero removal sampling data, and so cue rates were solely based off information sharing among phylogeny (and to a much lesser extent, migration status).
-For Kirtland's Warbler, we did have one data point for distance sampling, so predictions of EDR for Kirtland's Warbler here is not entirely based off zero data points.
-Bicknell's Thrush, however, had zero data points, and so EDRs for this species were based solely off information sharing among other species with similar traits.
-
-\par As expected, the predicted cue rate for Bicknell's Thrush (0.28 cues per minute) and the predicted cue rate for Kirtland's Warbler (0.29 cues per minute) fell very close to the overall mean cue rate of 0.27 for all 11 other Turdidae species and 0.31 for all 44 other Parulidae species in NA-POPS, respectively.
-That is, in the absence of any data, the estimated cue rate of Bicknell's Thrush and Kirtland's Warbler was shrunk toward cue rates of most similar species.
-
-\par For distance modelling, the predicted EDR for Bicknell's Thrush (73.1 m) and the predicted EDR for Kirtland's Warbler (75.0 m) fell reasonably close to the overall mean EDR of 92.7 m for all other 11 Turdidae species and 61.2 m for all 44 other Parulidae species in NA-POPS, respectively.
-That is, in the absence of any data (or with very little data in the case of Kirtland's Warbler), the estimated EDR of Bicknell's Thrush and Kirtland's Warbler was similar to those of other Thrushes and Warblers, respectively, but with departures from overall mean based on specific traits.
-
-\par In both the removal modelling and distance modelling, the magnitude of contribution of data from related species appeared to play a role in the width of the credible intervals.
-For removal modelling, although both credible intervals were fairly wide, Bicknell's Thrush had a 95\% credible interval of [0.16, 0.48], representing an approximate 53\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [0.18, 0.48], representing an approximate 49\% change on either end.
-For distance modelling, both credible intervals were also fairly wide, but Bicknell's Thrush had a 95\% credible interval of [40.4, 135.0], representing an approximate 58\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [46.6, 124.8], representing an approximate 47\% change on either end.
-Although the difference in width of the intervals is small, this could be due to the greater number of other species contributing information to Kirtland's Warbler estimates than to Bicknell's Thrush.
-In the case of distance modelling, the one data point contributing to Kirtland's Warbler's EDR could have also played a part.
-With that in mind, it would likely be important to consider how many "related" species are available with information (and how many data points do each of those related species have) when deciding on making a prediction of cue rate.
-
-\subsection{Comparisons of Predicted Detectability to Other Studies}
-
-\par Using the predictions of cue rate and EDR for Bicknell's Thrush and Kirtland's Warbler, we can calculate availability and perceptibility, and hence detectability, for a given survey type, and compare to previous estimates of detectability obtained from other studies.
-As a reminder, following derivations in \citet{solymos_calibrating_2013}, detectability is the product of availability and perceptibility.
-Availability $p(t)$ is defined as the probability of a bird giving a cue during a survey run for $t$ minutes, and is calculated as
-$p(t) = 1 - e^{-t\phi}$
-where $\phi$ is the cue rate of the species.
-Perceptibility $q(r)$ is defined as the probability of an observer hearing or seeing a bird, given the bird gives a cue, with a survey radius of $r$ metres, and is calculated as
-$q(r) = \dfrac{\tau^2 \left\{1 - e^{-\dfrac{r^2}{\tau^2}}\right\}}{r^2}$
-where $\tau$ is the EDR of the species.
-
-\par For Bicknell's Thrush, using the equations above, the predicted cue rate of 0.28, predicted EDR of 73.1 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.34.
-We can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respectively credible intervals) to get a range of detectabilities from 0.09 through 0.70.
-\citet{aubry_bicknells_2018}, however, estimate detection probabilities of $\geq$ 0.74 and $\leq$ 0.88, and so even our highest potential value of detectability falls short of their estimated range.
-However, we note that \citet{aubry_bicknells_2018} are using data that specifically target Bicknell's Thrush in their range, and they are using surveys conducted at peak Bicknell's Thrush breeding time, and so it is likely that detectability will be high in that specific case.
-
-\par For Kirtland's Warbler, using the equations above, the predicted cue rate of 0.29, predicted EDR of 75.0 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.36.
-Similar to our Bicknell's Thrush example, we can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respective credible intervals) to obtain a range of detectabilities from 0.13 through 0.67.
-A study by \citet{van_dyke_comparative_2022} found estimated detection probabilities for Kirtland's Warbler that ranged from effectively 0, through to a detection probability of 0.73.
-In that study, they compared estimates of detectability among different jack pine stands and red pine stands, where the detectability was highest in jack pine stands than the red pine stands.
-
 \subsection{Improvements to this model}
@@
 This is especially true for the types of species we are interested in generating predictions for, i.e., rare and/or undersampled species.
-In general, rare species will prefer rare habitats, or will at least be confined to a very small range (as is the case with Bicknell's Thrush and Kirtland's Warbler above).
+In general, rare species will prefer rare habitats, or will at least be confined to a very small range (as is the case with Bicknell’s Thrush and Kirtland’s Warbler above). 
 Thus, while we can generate reasonable estimates of the components of detectability (i.e., availability through cue rate and perceptibility through EDR) based on information borrowed from similar species (either through phylogeny or traits), this model as it stands may fail to glean more specific information that takes into account detectability peaks throughout the year, or detectability peaks given specific habitats.
@@
 Similarly, it may be beneficial to include environmental covariates at the survey level, similar to what was done in \citet{solymos_calibrating_2013, edwards_point_2023}, so that further information about EDR may be shared among similar species being surveyed at similar habitats.
-This could also provide an opportunity to disentangle species x survey interactions in detectability.
+This could also provide an opportunity to disentangle species by survey interactions in detectability.
 
-\par The examples above additionally highlight an excellent opportunity to make use of a powerful feature of Bayesian models, that being the inclusion of expert knowledge as prior information.
-For example, if a Bicknell's Thrush expert knows that there is something particular about how often they sing or give a cue compared to other thrushes, or how disproportionately quiet or loud they are compared to species with similar traits, this can be coded directly into the prior distribution.
-Similarly, this may also provide a unique opportunity to make use of traditional ecological knowledge to further inform predictions of detectability.
+\par The examples above additionally highlight an excellent opportunity to use a powerful feature of Bayesian models, which is the inclusion of expert knowledge as prior information. 
+For example, if a Bicknell’s Thrush expert knows that there is something particular about how often they sing or give a cue compared to other thrushes, or how disproportionately quiet or loud they are compared to species with similar traits, this can be coded directly into the prior distribution.
+Similarly, this may also provide a unique opportunity to use traditional ecological knowledge to further inform predictions of detectability.
 For example, in the Anishinaabe language, bird taxonomy is based both on physical features, but also behaviours such as flight, calls, morning rituals, and weather \citep{pitawanakwat_evening_2022}.
-This taxonomy is based off observations of bird behaviour that has been passed down through generations, and may vary slightly (or heavily) from taxonomy based on genomics.
-In fact, it could be argued that a taxonomy based on how humans observe birds may actually be more relevant for detectability, and so a future opportunity could be to adopt a two-eyed seeing approach \cite{reid_twoeyed_2021} to predicting detectability in rare birds.
+This taxonomy is based on observations of bird behaviour that has been passed down through generations, and may vary from taxonomy based on genomics.
+In fact, it could be argued that a taxonomy based on how humans observe birds may actually be more relevant for detectability, and so a future opportunity could be to adopt a two-eyed seeing approach \cite{reid_twoeyed_2021} using both Indigenous and western scientific knowledge, to predict detectability in rare birds.
 
 \subsection{Conservation Implications}
-\par Detectability often goes unaccounted for in conservation problems, often because the estimates of detection probability simply do not exist, or they are extremely uncertain \citep{bennett_how_2023}.
+\par Detectability frequently goes unaccounted for in conservation problems, often because the estimates of detection probability simply do not exist, or they are extremely uncertain \citep{bennett_how_2023}.
 This model provides an excellent opportunity for detection probabilities to be estimated with higher precision for several species of landbirds in North America, and also provides an opportunity to predict detection probabilities for species which have very little (or even no) data.
-This means that conservation problems that rely on estimates of detectability such as Value of Information analyses \citep{canessa_when_2015, bennett_when_2018} or prioritizations \citep{hanson_prioritizr_2023} can look to make more accurate decisions.
+This means that conservation problems that rely on estimates of detectability such as Value of Information analyses \citep{canessa_when_2015, bennett_when_2018} or prioritizations \citep{hanson_prioritizr_2023} can make more accurate decisions.
 For organizations such as Partners in Flight, Boreal Avian Modelling project (\citep{cumming_toward_2010}), and others that rely on detectability information for bird population size estimates, these improved estimates (and new estimates and predictions) provide a way forward for continuing to produce accurate estimates of population size that are based on data-driven detectability estimates.
@@
 As shown here, birds with no data points can have detectability predicted that falls closely in line with similar species.
-We caution that birds that are to have detectability predicted should be noted as such on any communications, such that organizations know that the estimate is simply a "best guess" based on other similar species.
+However, we caution that detectability predictions based on models using similar species should be noted as such, so that organizations know that the estimate is simply a "best guess" based on other similar species.
 
 \par Finally, we see this model as a proof-of-concept for use with other taxa.
-Obviously, given the vast amount of bird data available, it was easy enough to obtain estimates for enough data-rich species such that the few data-poor species could have enough information to borrow.
+Obviously, given the vast amount of bird data available, it was relatively easy to obtain estimates for enough data-rich species such that the few data-poor species could have enough information to borrow.
 However, one benefit that we have shown here is that Bayesian models can still come up with reasonable predictions if there are informed enough priors on which to fall back.
-That is to say, for a taxon such as butterflies where there are some species with data and others with very little data, but there is overall very little data in general \citep{lewthwaite_geographical_2022} with reasonably-informed priors, a researcher may still be able to glean reasonable estimates of detectability with the little data available.
+For taxa such as butterflies for which some species have reasonable data and most have very little data \citep{lewthwaite_geographical_2022}, a researcher may still be able to glean reasonable estimates of detectability with the little data available.
 We suggest that this type of model where information about detectability is shared among species be tested with taxon that generally have fewer data collected.

### 2023-12-07 src/scripts/distance/9-distance-posthoc-analysis.Rmd
updated methods, results, and discussion of article based on j and a suggestions. still waiting for distance CV to finish

@@
 bith_cr <- c(0.28, 0.16, 0.48)
-bith_edr <- c(73.1, 40.4, 135.0)
+bith_edr <- c(73.6, 40.5, 134.4)
 kiwa_cr <- c(0.29, 0.18, 0.48)
-kiwa_edr <- c(75, 46.6, 124.8)
+kiwa_edr <- c(75.3, 45.9, 125.6)
 t = 5

### 2024-01-04 article/article.tex
final touches on draft 2 including figure captions and text

@@
 	\and
-	Solymos, Peter\\
+	Gahbauer, Marcel\\
 	\and
-	Stralberg, Diana\\
+	Grinde, Alexis\\
+	\and
+	Hope, David\\
+	\and
+	Knight, Elly\\
 	\and
@@
 	\and
-	Gahbauer, Marcel\\
-	\and
-	Grinde, Alexis\\
-	\and
-	Hope, David\\
+	Solymos, Peter\\
 	\and
@@
 
+	Species-specific detection probability is an important metric in conservation, but is often unaccounted for or does not exist on a species level.
+Despite having several estimates of detection probabilities for North American birds, there are still several species of rare and undersampled birds that do not have sufficient data to estimate detection probabilities.
+Here, we take advantage of similarities in detection probabilities across phylogeny to estimate detection probabilities in seven species of rare North American birds, using a hierarchical Bayesian implementation of previously-developed removal and distance models.
+For two species, Bicknell's Thrush and Kirtland's Warbler, we demonstrate the multi-species' model ability to predict detection probabilities given little to no data, and compare these predictions to previous estimates found in the literature.
+We also demonstrate the multi-species models' ability to provide more precise estimates of detectability for species where there is sufficient data.
+We recommend that this multi-species model be combined with survey-level covariates to then be used to augment previous estimates of detection probabilities in the NA-POPS database.
+
 \end{abstract}
@@
 However, despite its importance, detectability often goes unaccounted for in studies, or detectability data just does not exist \citep{bennett_how_2023}.
-The wealth of openly available bird data in North America means that estimates of detectability for birds are relatively easy to come by, in comparison to most other taxonomic groups \citep{bennett_how_2023}.
+
+\par The wealth of openly available bird data in North America means that estimates of detectability for birds are relatively easy to come by, in comparison to most other taxonomic groups \citep{bennett_how_2023}.
 The NA-POPS project is an effort to estimate detectability functions that take into account variability in detection through environmental and temporal covariates, for as many of North America’s landbird species as possible \citep{edwards_point_2023}.
@@
 \par Detectability in birds has been shown to be at least partly driven by variation in traits.
-\cite{johnston_species_2014} showed in a selection of European landbirds that body mass and habitat association had an effect on detection distance, which influences perceptibility
+\cite{johnston_species_2014} showed in a selection of European landbirds that body mass and habitat association had an effect on detection distance, which influences perceptibility.
 In this case, a larger body size tended to be associated with larger detection distances, which makes sense considering larger body sizes are associated with louder and lower-pitched songs in birds \citep{bowman_adaptive_1979, fletcher_acoustics_1999, ryan_role_1985}.
@@
 If the NA-POPS database contained at least some point-count data for the species of interest, we simply added these data into the model.
-This is different from \citet{edwards_point_2023, solymos_calibrating_2013, solymos_evaluating_2018}, where any species with less than 75 data points were removed from the analysis.
+This is different from \citet{edwards_point_2023}, \citet{solymos_calibrating_2013}, and \citet{solymos_evaluating_2018}, where any species with less than 75 data points were removed from the analysis.
 For species that did not have any data in the NA-POPS database, we generated a row of traits data for the species and estimated its cue rate and EDR as parameters in the model.
@@
 
+\par All code is available open-source at https://github.com/BrandonEdwards/multispecies-qpad-detectability and will be archived with a DOI upon acceptance of this paper.
+
 \section{Results}
@@
 We did find similar support for change in EDR by body size and pitch, in that body size tended to have an increasing effect on EDR, and song pitch had a decreasing effect on EDR (Figure \ref{fig:params}B).
-
-\begin{figure}[h!]
-	\includegraphics{../output/plots/parameters_plot.png}
-	\caption{Posterior distributions with 95\% credible intervals (shaded area) of parameters estimated in the multi-species removal model, with overall mean log cue rate as the intercept (A), migratory strategy as departures from the intercept based on whether a species was a resident (mu\_mig\_strat[1]) or migrant (mu\_mig\_strat[B]) (B), and extra variance term after accounting for the phylogenetic variation (C).
-Posterior distributions with 95\% credible intervals (shaded area) of parameters estimated in the multi-species distance model, with (A) overall mean log EDR as the intercept, with departures from the intercept modelled as migratory strategy (resident: mu\_mig\_strat[1], migrant: mu\_mig\_strat[B]), (B) slope parameters of changes in EDR given change in mass and pitch, and (C) extra variance term after accounting for traits.}
-	\label{fig:params}
-\end{figure}
 
@@
 (NEEDS UPDATING) Seven out of the 317 species had a percent change in EDR of more than 20\% (Figure \ref{fig:1vs1}B).
-These species were Sharp-tailed Grouse (\textit{Tympanuchus phasianellus}), Sedge Wren (\textit{Cistothorus stellaris}), Ring-necked Pheasant (\textit{Phasianus colchicus}), Chimney Swift (\textit{Chaetura pelagica}), Boat-tailed Grackle (\textit{Quiscalus major}), and Baird's Sparrow (\textit{Ammodramus bairdii}).
-
-\begin{figure}[h!]
-	\includegraphics{../output/plots/1vs1.png}
-	\caption{Modelled difference with 95\% credible interval (shaded area) between cue rates estimated from the multi-species model developed here, minus the single-species model in NA-POPS \citep{edwards_point_2023} (A); and 1-to-1 plot comparing cue rates estimated from the single-species model vs the multi-species model for each species (B).
-Modelled difference with 95\% credible interval (shaded area) between EDR estimated from the multi-species model developed here, minus the single-species model in NA-POPS \citep{edwards_point_2023} (A); and 1-to-1 plot comparing EDR estimated from the single-species model vs the multi-species model for each species (B).}
-	\label{fig:1vs1}
-\end{figure}
+These species were Sharp-tailed Grouse (\textit{Tympanuchus phasianellus}), Sedge Wren (\textit{Cistothorus stellaris}), Ring-necked Pheasant (\textit{Phasianus colchicus}), Chimney Swift (\textit{Chaetura pelagica}), Chihuahuan Raven (\textit{Corvus cryptoleucus}), Boat-tailed Grackle (\textit{Quiscalus major}), and Baird's Sparrow (\textit{Ammodramus bairdii}).
 
@@
 
-\begin{figure}[h!]
-	\includegraphics{../output/plots/sd_comp.png}
-	\caption{Posterior distributions with 95\% credible intervals (shaded areas) of (A) modelled intercepts of standard deviation for the single-species removal model (alpha[1]) and multi-species removal model (alpha[2]), and (B) modelled slopes of standard deviation with increasing sample size for the single-species removal model (beta[1]) and multi-species removal model (beta[2]).
-Posterior distributions with 95\% credible intervals (shaded areas) of (A) modelled intercepts of standard deviation for the single-species distance model (alpha[1]) and multi-species distance model (alpha[2]), and (B) modelled slopes of standard deviation with increasing sample size for the single-species distance model (beta[1]) and multi-species distance model (beta[2]).}
-	\label{fig:sd}
-\end{figure}
-
 \subsection{Comparison of Predictive Accuracy between Single Species and Multi Species}
 
-\par The cross validation revealed that multi-species model had better predictive capacity for more species (200 out of 316) than the single-species model (99 out of 316).
+\par The cross validation revealed that the multi-species removal model had better predictive capacity for more species (200 out of 316) than the single-species removal models (99 out of 316).
 Seventeen out of 316 species showed no better predictive capacity with either model.
-Mean overall difference in log predictive density was $4.78 \times 10^{-6}$.
+Mean overall difference in log predictive density was $4.78 \times 10^{-6}$, indicating a small overall preference for the multi-species model.
 As expected, multi-species models tended to have better predictive capacity when sample size was low, with a negligible difference as sample size increased (Figure \ref{fig:cv}A).
 There did not appear to be any patterns in model selection among phylogeny or traits (see Supplemental Figure 1 for full phylogenetic tree).
+Among the seven species with greater than 20\% change in cue rate, three species (Allen's Hummingbird, Bohemian Waxwing, and Rusty Blackbird) had the multi-species model chosen as the better model, three species (Common Redpoll, Red-bellied Sapsucker, and Willow Ptarmigan) had the single-species model chosen as the better model, and one species (Calliope Hummingbird) had neither model chosen as the better model.
 
-\par (NEEDS UPDATING) The cross validation revealed that the multi-species model tended to have slightly more support than the single-species model in terms of predicting EDR.
+\par The cross validation revealed that the multi-species distance model also tended to have better predictive capacity for more species (156 out of 316) than the single-species distance models (140 out of 316), but to a lesser extent than the removal models.
+Twenty out of the 316 species showed no better predictive capacity with either model.
+Mean overall difference in log predictive density was $7.66 \times 10^{-10}$, indicating a small overall preference for the multi-species model.
 However, similar to the removal model cross-validation results, the overall difference in posterior predictive density was very small, with a mean difference of $1.85 \times 10^{-4}$.
-Multi-species models tended to have better predictive capacity when sample size was low (again noting a very small difference), and a negligible difference as sample size increased (Figure \ref{fig:cv}B).
-Interestingly, only two (Chimney Swift and Boat-tailed Grackle) of the six species with a greater than 20\% change in EDR had the multi-species model chosen as the better predictive model.
-
-\begin{figure}[h!]
-	\includegraphics{../output/plots/cv_plot.png}
-	\caption{Mean difference in pointwise posterior predictive density.
-Points are mean difference by species.
-Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species model preference, whereas negative differences (coloured in blue for clarity) represent a single-species model preference.
-Regression line shows the relationship between mean difference and species sample size.
-Mean difference in pointwise posterior predictive density.
-Points are mean difference by species.
-Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species distance model preference, whereas negative differences (coloured in blue for clarity) represent a single-species distance model preference.
-Regression line shows the relationship between mean difference and species sample size.}
-	\label{fig:cv}
-\end{figure}
+Similar to removal models, multi-species distance models tended to have better predictive capacity when sample size was low, with a negligible difference as sample size increased (Figure \ref{fig:cv}B).
+Among the seven species with greater than 20\% change in EDR, three species (Baird's Sparrow, Sedge Wren, and Sharp-tailed Grouse) had the multi-species model chosen as the better model, and four species (Boat-tailed Grackle, Chihuahuan Raven, Chimney Swift, and Ring-necked Pheasant) has the single-species model chosen as the better model.
 
@@
 The species for which we had no data (i.e., Bicknell's Thrush) had an estimate of EDR similar to what would be expected by a bird of its size, song pitch, and habitat association, demonstrating the shrinkage effect of the multi-species model.
-
-\begin{figure}[h!]
-	\includegraphics{../output/plots/predictions_figure.png}
-	\caption{Cue rates (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for seven species of rare and undersampled North American landbirds.
-Species are 4 letter banding codes: SPOW = Spotted Owl (n = 7), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), TRBL = Tricoloured Blackbird (n = 74), HASP = Harris's Sparrow (n = 65), KIWA = Kirtland's Warbler (n = 0), and LEPC = Lesser Prairie-chicken (n = 45).
-Effective detection radius (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for six species of rare and undersampled North American landbirds.
-Species are 4 letter banding codes: SPOW = Spotted Owl (n = 28), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), HASP = Harris's Sparrow (n = 13), KIWA = Kirtland's Warbler (n = 1), and LEPC = Lesser Prairie-chicken (n = 45).}
-	\label{fig:predictions}
-\end{figure}
 
@@
 For example, Harris's Sparrow (n = 13) had an estimated mean EDR of 65.6 m, with a 95\% credible interval of [53.1, 82.5].
-This range does fall below the mean EDR of 85.3 m for the other sparrows in NA-POPS, but Harris's Sparrow song pitch of approximately 3.8 Hz falls above the average song pitch of 3.0 Hz for the other sparrows in NA-POPS, which could contribute to the smaller EDR.
+This range does fall below the mean EDR of 85.3 m for the other sparrows in NA-POPS, but Harris's Sparrow song pitch of approximately 3.8 Hz falls above the average song pitch of 3.0 Hz for the other sparrows in NA-POPS, which would contribute to the smaller EDR.
 On the other hand, LeConte's Thrasher (n = 1) had an estimated EDR of 134.2 m, with a 95\% credible interval of [81.8, 230.5], which falls roughly in line with the mean EDR of 113.2 m of the six other thrashers in NA-POPS.
@@
 \par For distance modelling, the predicted EDR for Bicknell's Thrush (73.6 m) and the predicted EDR for Kirtland's Warbler (75.3 m) fell reasonably close to the overall mean EDR of 92.7 m for all other 11 Turdidae species and 61.2 m for all 44 other Parulidae species in NA-POPS, respectively.
-hat is, in the absence of any data (or with very little data in the case of Kirtland’s Warbler), the estimated EDR of Bicknell’s Thrush and Kirtland’s Warbler was similar to those of other Thrushes and Warblers, respectively, but with departures from overall mean based on specific traits.
+That is, in the absence of any data (or with very little data in the case of Kirtland’s Warbler), the estimated EDR of Bicknell’s Thrush and Kirtland’s Warbler was similar to those of other Thrushes and Warblers, respectively, but with departures from overall mean based on specific traits.
 
@@
 
-\par For removal modelling, we experience no issues related to initialization, sampling, or convergence. 
+\par For removal modelling, we experienced no issues related to initialization, sampling, or convergence. 
 The model finished within approximately 24 hours, and all parameters had R-hat values of 1, or within 0.001 of it.
@@
 Similarly, this may also provide a unique opportunity to use traditional ecological knowledge to further inform predictions of detectability.
-For example, in the Anishinaabe language, bird taxonomy is based both on physical features, but also behaviours such as flight, calls, morning rituals, and weather \citep{pitawanakwat_evening_2022}.
+For example, in the Anishinaabe language, bird taxonomy is based not only on physical features, but also behaviours such as flight, calls, morning rituals, and weather \citep{pitawanakwat_evening_2022}.
 This taxonomy is based on observations of bird behaviour that has been passed down through generations, and may vary from taxonomy based on genomics.
@@
 This means that conservation problems that rely on estimates of detectability such as Value of Information analyses \citep{canessa_when_2015, bennett_when_2018} or prioritizations \citep{hanson_prioritizr_2023} can make more accurate decisions.
-For organizations such as Partners in Flight, Boreal Avian Modelling project (\citep{cumming_toward_2010}), and others that rely on detectability information for bird population size estimates, these improved estimates (and new estimates and predictions) provide a way forward for continuing to produce accurate estimates of population size that are based on data-driven detectability estimates.
+For organizations such as Partners in Flight, Boreal Avian Modelling project \citep{cumming_toward_2010}, and others that rely on detectability information for bird population size estimates, these improved estimates (and new estimates and predictions) provide a way forward for continuing to produce accurate estimates of population size that are based on data-driven detectability estimates.
 
-\par The model presented here can be used to fill out the remaining missing species in the NA-POPS database, in order to estimate detectability for every species of North American landbird.
+\par The models presented here can be used to fill out the remaining missing species in the NA-POPS database, in order to estimate detectability for every species of North American landbird.
 By making use of phylogeny and trait information, bird species with fewer than 75 data points can have estimates generated based on the few data points available, and information available from other species.
@@
 
+\section{Figures}
+
+\begin{figure}[h]
+	\includegraphics{../output/plots/parameters_plot.png}
+	\caption{Posterior distributions with mean (points), 50\% credible intervals (bolded lines), and 95\% credible intervals (thin lines) of parameters estimated in the multi-species removal model (A) and the multi-species distance model (B).}
+	\label{fig:params}
+\end{figure}
+
+\begin{figure}[h]
+	\includegraphics{../output/plots/1vs1.png}
+	\caption{1-to-1 plot comparing cue rates (A) and EDRs (B) estimated from the single-species model vs the multi-species model for each species.
+Labelled species are those which had a greater than 20\% difference in cue rate or EDR between the single-species and multi-species models.
+Insets are modelled differences with 95\% credible interval (shaded area) between cue rates (A) and EDRs (B) estimated from the multi-species model developed here, minus the single-species model in NA-POPS \citep{edwards_point_2023}.}
+	\label{fig:1vs1}
+\end{figure}
+
+\begin{figure}[h]
+	\includegraphics{../output/plots/sd_comp.png}
+	\caption{LOESS curves of standard deviations of cue rate (A) and EDR (B) for species modelled by a single-species model (blue lines) and multi-species model (red lines).}
+	\label{fig:sd}
+\end{figure}
+
+\begin{figure}[h]
+	\includegraphics{../output/plots/cv_plot.png}
+	\caption{Mean difference in pointwise posterior predictive density for cue rate (A) and EDR (B) vs.
+log species sample size.
+Points are mean difference by species.
+Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species model preference, whereas negative differences (coloured in blue for clarity) represent a single-species model preference.
+Grey points imply no differences.
+}
+	\label{fig:cv}
+\end{figure}
+
+\begin{figure}[h]
+	\includegraphics{../output/plots/predictions_figure.png}
+	\caption{(A) Cue rates (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for seven species of rare and undersampled North American landbirds.
+Species are 4 letter banding codes: SPOW = Spotted Owl (n = 7), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), TRBL = Tricoloured Blackbird (n = 74), HASP = Harris's Sparrow (n = 65), KIWA = Kirtland's Warbler (n = 0), and LEPC = Lesser Prairie-chicken (n = 45).
+Black points are cue rate estimates of other similar species with the same migration strategy, with the opacity of the point indicating the strength of the relationship.
+(B) Effective detection radius (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for six species of rare and undersampled North American landbirds.
+Species are 4 letter banding codes: SPOW = Spotted Owl (n = 28), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), HASP = Harris's Sparrow (n = 13), KIWA = Kirtland's Warbler (n = 1), and LEPC = Lesser Prairie-chicken (n = 45).
+Black points are EDR estimates of species with the same migration strategy and habitat preference, and with a mass and pitch within 40\% in either direction.}
+	\label{fig:predictions}
+\end{figure}
+
 \end{document}

### 2024-02-09 article/article.tex
some introduction fixes

@@
 
-\par Accurate and precise estimates of detectability are important both for making efficient conservation decisions, and estimating accurate population sizes.
-In conservation management situations, detectability can be used as a metric of sensitivity to evaluate when to monitor a system for a species, or when to act on current information \citep{canessa_when_2015, bennett_when_2018}.
-When estimating population sizes of an organism, detectability often comes into play as a way to convert observed counts into densities or estimates of true abundance \citep{solymos_calibrating_2013, johnson_defense_2008}.
-However, despite its importance, detectability often goes unaccounted for in studies, or detectability data just does not exist \citep{bennett_how_2023}.
+\par Accurate and precise estimates of detection probability (or detectability) are important both for estimating accurate population sizes, as well as for making effective conservation decisions.
+In conservation management, prior knowledge of detectability can be used as a metric of sensitivity to evaluate when and how to monitor a system for a species, or when to act on current information \citep{canessa_when_2015, bennett_when_2018}.
+When estimating population size, detectability often comes into play as a way to convert observed counts into densities or estimates of true abundance \citep{solymos_calibrating_2013, johnson_defense_2008}.
+Despite its importance, detectability often goes unaccounted for in studies; and for many species, the ancillary data needed to estimate detectability does not exist \citep{bennett_how_2023}.
 
-\par The wealth of openly available bird data in North America means that estimates of detectability for birds are relatively easy to come by, in comparison to most other taxonomic groups \citep{bennett_how_2023}.
-The NA-POPS project is an effort to estimate detectability functions that take into account variability in detection through environmental and temporal covariates, for as many of North America’s landbird species as possible \citep{edwards_point_2023}.
-To date, the project has estimated availability (the probability a bird sings or gives a cue) and perceptibility (the probability an observer sees or hears a cueing bird) for 319 species of birds, with an additional 19 species having estimates of availability or perceptibility (but not necessarily both).
-This project built off previous efforts by the Boreal Avian Modelling project \citep{cumming_toward_2010} to systematically estimate detectability in landbirds \citep{solymos_calibrating_2013, solymos_evaluating_2018}.
-Both the NA-POPS project and previous estimates by Boreal Avian Modelling project use the QPAD methodology \citep{solymos_calibrating_2013}, which is a flexible method to estimate availability and perceptibility given a harmonized dataset \citep{barker_ecological_2015}.
+\par The wealth of openly available bird observation data in North America means that estimates of detectability for birds are relatively easy to estimate, in comparison to most other taxonomic groups \citep{bennett_how_2023}.
+The NA-POPS project is an effort to estimate detection probabilities that account for variability in detection processes through environmental and temporal covariates for as many of North America’s landbird species as possible \citep{edwards_point_2023}.
+To date, the project has estimated the two components of detectability--availability (the probability that a bird sings or gives a cue during the survey period) and perceptibility (the probability that an observer sees or hears a cueing bird)--for 319 species of birds, with an additional 19 species having estimates of either availability or perceptibility.
+This project expanded previous efforts by the Boreal Avian Modelling project \citep{cumming_toward_2010} to systematically estimate detectability in landbirds \citep{solymos_calibrating_2013, solymos_evaluating_2018}.
+Both the NA-POPS project and previous estimates by Boreal Avian Modelling project use QPAD \citep{solymos_calibrating_2013}, which is a flexible method to estimate availability and perceptibility given a harmonized dataset with the necessary survey-specific data (i.e., distance to bird from observer and time to first detection; \citet{barker_ecological_2015}.
 

### 2024-04-02 src/scripts/removal/5-removal-cv.R
up number of iterations for cross validation

@@ -17,7 +17,7 @@ cv_folds <- read.csv("data/generated/removal_cv_folds.csv")
 ####### Set Constants #############################
 
 # Stan settings
-n_iter <- 500
+n_iter <- 2000
 n_warmup <- 1000
 n_chains <- 4
 refresh <- 10

### 2024-04-04 article/article.tex
updates to introduction

@@
 	\and
-	Knight, Elly\\
+	Knight, Elly C.\\
 	\and
-	Michel, Nicole\\
+	Michel, Nicole L.\\
 	\and
@@
 
-	Species-specific detection probability is an important metric in conservation, but is often unaccounted for or does not exist on a species level.
-Despite having several estimates of detection probabilities for North American birds, there are still several species of rare and undersampled birds that do not have sufficient data to estimate detection probabilities.
-Here, we take advantage of similarities in detection probabilities across phylogeny to estimate detection probabilities in seven species of rare North American birds, using a hierarchical Bayesian implementation of previously-developed removal and distance models.
-For two species, Bicknell's Thrush and Kirtland's Warbler, we demonstrate the multi-species' model ability to predict detection probabilities given little to no data, and compare these predictions to previous estimates found in the literature.
-We also demonstrate the multi-species models' ability to provide more precise estimates of detectability for species where there is sufficient data.
-We recommend that this multi-species model be combined with survey-level covariates to then be used to augment previous estimates of detection probabilities in the NA-POPS database.
+	Accurate assessments of population status and trends rely on accounting for probability of detection, making it an important metric in conservation, yet it is often unaccounted for or unavailable on a species level.
+	Although detection probability methods and estimates exist for many North American birds, probability estimates are most accurate for common, easy to detect species. 
+	There is therefore a need for methodology that provides detection estimates for species that are rare or undersampled, which generally do not have sufficient data for accurate estimates. 
+	Here, we take advantage of similarities in detection probabilities among phylogenetically related species and species with similar traits to estimate detection probabilities in seven species of rare North American birds, using a hierarchical Bayesian multi-species removal and distance models.
+	For two species, Bicknell’s Thrush and Kirtland’s Warbler, we demonstrate our multi-species model’s ability to predict detection probabilities given little to no observation data, and compare these predictions to previous estimates found in the literature.
+	We also demonstrate the multi-species models’ ability to provide more precise estimates of detectability for species where there is sufficient data. 
+	We recommend that this multi-species model be combined with survey-level covariates to augment previous estimates of detection probabilities in the NA-POPS database of detectability offsets for North American birds.
 
@@
 \par Despite the efforts of the NA-POPS project, approximately 25\% of the species in the Partners in Flight’s 2016 Landbird Conservation Plan \citep{rosenberg_partners_2016} have insufficient data to directly estimate detectability \citep{edwards_point_2023}.
-This is because of sample size limitations that are recommended for removal and distance modelling, in that a modeller should have at least 75 data points for reasonable estimates \citep{buckland_introduction_2001, solymos_calibrating_2013}.
+These species fall short of the threshold of 75 distinct observations recommended for removal and distance modelling \citep{buckland_introduction_2001, solymos_calibrating_2013}. 
 Many of the species that fell short of that threshold in the NA-POPS database are rare or undersampled species, that are either have low population sizes, or occur in rare or inaccessible habitats.
-Some examples of species not modelled include Kirtland’s Warbler (\textit{Setophaga kirtlandii}) and Bicknell’s Thrush (\textit{Catharus bicknelli}), both of which have low population sizes, and LeConte’s Thrasher (\textit{Toxostoma lecontei}) and Harris’s Sparrow (\textit{Zonotrichia querula}), both of which have rare habitat associations \citep{will_handbook_2020}.
-Such species are often conservation concerns, and thus detectability information for them would be particularly important.
+%Some examples of species not modelled include Kirtland’s Warbler (\textit{Setophaga kirtlandii}) and Bicknell’s Thrush (\textit{Catharus bicknelli}), both of which have low population sizes, and LeConte’s Thrasher (\textit{Toxostoma lecontei}) and Harris’s Sparrow (\textit{Zonotrichia querula}), both of which have rare habitat associations \citep{will_handbook_2020}.
+Species with small population sizes and rare habitats are also often considered vulnerable, and thus detectability information for them would be particularly important to inform their conservation.
 
-\par Detectability in birds has been shown to be at least partly driven by variation in traits.
-\cite{johnston_species_2014} showed in a selection of European landbirds that body mass and habitat association had an effect on detection distance, which influences perceptibility.
-In this case, a larger body size tended to be associated with larger detection distances, which makes sense considering larger body sizes are associated with louder and lower-pitched songs in birds \citep{bowman_adaptive_1979, fletcher_acoustics_1999, ryan_role_1985}.
-Habitat associations tend to influence the attenuation of sound, whereby sound diminishes quicker in a closed-forested habitat than more open habitat \citep{waide_tropical_1988, yip_sound_2017}.
-A study by \cite{solymos_phylogeny_2018} expanded on this effort by also considering the effects of species traits and phylogeny on both detection distance and cue rate, the latter of which influences the availability of birds.
-They found that cue rate is most related to phylogeny, and, to a lesser extent, whether a bird was a long-distance migrant or a resident species.
-They also found detection distance was related to species’ traits such as body mass, song pitch, habitat association, and migratory status and that after accounting for these traits, phylogeny explained very little.
+\par Detectability in birds has been shown to be at least partially driven by variation in species traits.
+\cite{johnston_species_2014} showed that detection distance - and consequently perceptibility - varied with body mass and habitat association in a selection of European landbirds.
+In this case, larger body size tended to be associated with longer detection distances, which makes sense considering larger body sizes are associated with louder and lower-pitched songs in birds \citep{bowman_adaptive_1979, fletcher_acoustics_1999, ryan_role_1985}.
+Moreover, vegetation structure influences the attenuation of sound, whereby sound diminishes quicker in a closed-forested habitat than more open habitats, thus leading to variation in detectability among species associated with different habitats \citep{waide_tropical_1988, yip_sound_2017}.
+\citet{solymos_phylogeny_2018} considered the effects of species traits and phylogeny on both detection distance and cue rate, the latter of which influences the availability of birds for detection.
+They found that cue rate is most related to phylogeny, and, to a lesser extent, migratory strategy (i.e., whether a bird was a long-distance migrant or a resident species).
+They also found that detection distance was related to species’ traits such as body mass, song pitch, habitat association, and that after accounting for these traits, phylogeny explained very little.
 
-\par One recommendation for future research that came out of \cite{solymos_phylogeny_2018} was to use these phylogenetic and trait relationships with detectability to predict components of detectability (i.e., availability through cue rate, and perceptibility through detection distance) in other birds.
-Indeed, by making use of shrinkage factors and partial pooling, as can be done in Bayesian hierarchical models \citep{gelman_what_2021}, and by accounting for phylogenetic relationships and the relative "relatedness" of different species, as can be done with Gaussian processes \citep{bernardo_regression_1998, mcelreath_continous_2020}, we can leverage the information provided by species with many data to inform estimates for species with much fewer (included potentially zero) data points.
+\par \citet{solymos_phylogeny_2018} recommended using these phylogenetic and trait relationships to predict components of detectability (i.e., availability through cue rate, and perceptibility through detection distance) more broadly.
+By leveraging the information provided by species with many data it may be possible to inform estimates for species with much fewer (including potentially zero) data points.
+This could be done in two ways: by making use of shrinkage factors and partial pooling, as done in Bayesian hierarchical models (Gelman and Vehtari 2021), and by accounting for phylogenetic relationships and the relative "relatedness" of different species, as done with Gaussian processes (Bernardo et al.
+1998; McElreath 2020).
 
 \par The goal of this paper is to create a multi-species model of detectability that incorporates species traits and phylogeny, to both allow for the relaxation of the sample size requirements for estimating detectability, and to derive estimates of detectability for bird species without any data.
-To do so, we developed a Bayesian hierarchical version of the QPAD Removal model initially derived in \cite{solymos_calibrating_2013}, that allows for information exchange between species based on phylogenetic distance and binned migratory status.
-Phylogenetic distance was coded as a correlation structure and used as a Gaussian random field in this model, and migratory status was coded similar to \cite{solymos_phylogeny_2018} where a species was considered either "migrant" or "resident".
-We also developed a Bayesian hierarchical version of the QPAD Distance model initially derived in \cite{solymos_calibrating_2013}, that allows for information exchange between species based only on the traits that came out as being significant in  \cite{solymos_phylogeny_2018}.
+To do so, we developed a Bayesian hierarchical version of the QPAD removal model initially derived in \citet{solymos_calibrating_2013}, that allows for information exchange between species based on phylogenetic distance and binned migratory status.
+Phylogenetic distance was coded as a correlation structure and used as a Gaussian random field in this model, and migratory status was coded similarly to \citet{solymos_phylogeny_2018} in which a species was considered either "migrant" or "resident".
+We also developed a Bayesian hierarchical version of the QPAD distance model initially derived in \citet{solymos_calibrating_2013} that allows for information exchange among species based only on the traits shown to be significant by \citet{solymos_phylogeny_2018}.
 These traits include migratory status (coded the same as removal models), habitat association (coded as either "open" or "closed"), song pitch, and body mass. 
-We also developed a Bayesian single species version of the QPAD Removal and Distance models (i.e., as opposed to the maximum likelihood models available in the \texttt{detect} R package; \cite{solymos_detect_2020}) to allow for direct comparison through cross validation.
-Finally, we ran these models with the full set of data points used in \cite{edwards_point_2023}, plus additional data for seven species that did not meet the initial sample size thresholds for NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Harris's Sparrow, Kirtland's Warbler, Tricoloured Blackbird (\textit{Agelaius tricolor}), and Spotted Owl (\textit{Strix occidentalis}).
-All seven of these bird species are listed on Partners in Flight's watchlist for various threats \citep{will_handbook_2020}.
+We also developed a Bayesian single species version of the QPAD removal and distance models (i.e., as opposed to the maximum likelihood models available in the \texttt{detect} R package; \citet{solymos_detect_2020} to allow for direct comparison through cross validation.
+Finally, we ran these models with the full set of data points used in \citet{edwards_point_2023}, plus additional data for seven species that did not meet the initial sample size thresholds for NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird (\textit{Agelaius tricolor}), and Spotted Owl (\textit{Strix occidentalis}).
+All seven of these bird species are listed on Partners in Flight’s watchlist, indicating their conservation importance \citep{will_handbook_2020}.
 
-\par We predicted that for both the Bayesian hierarchical QPAD removal model and Bayesian hierarchical QPAD distance model, estimates of cue rate and EDR, respectively, will change very little from those estimated from the original NA-POPS database \cite{edwards_point_2023}.
+\par We predicted that for both the Bayesian hierarchical QPAD removal model and Bayesian hierarchical QPAD distance model, estimates of cue rate and EDR, respectively, will differ very little from those estimated from the original NA-POPS database \cite{edwards_point_2023}.
 This is because the species that were modelled in \citet{edwards_point_2023} had a sufficient number of data points to stand on their own, and so adding in species-level effects should have little effect on estimates.

### 2024-04-10 article/article.tex

@@
 
-\par For any point count that employs removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002} and/or distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}, let $Y_{sijk}$ be the observed count of species $s$ during sampling event $i$, occurring in time band $j \in [1,J]$ and/or distance band $k \in [1,K]$.
-That is, if we have a point count where removal sampling or distance sampling (but not both) is not used, then either $J = 1$ or $K = 1$, respectively; whereas, if we have a point count where both removal sampling and distance sampling is used, then both $J,K \geq 2$.
-The case where $J = K = 1$ is where neither removal sampling nor distance sampling is used, and so these data points are not appropriate for the modelling techniques described here.
+\par For any point count that employed removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002} and/or distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}, let $Y_{sijk}$ be the observed count of species $s$ during sampling event $i$, occurring in time band $j \in [1,J]$ and/or distance band $k \in [1,K]$.
+That is, if we have a point count where removal sampling or distance sampling (but not both) was not used, then either $J = 1$ or $K = 1$, respectively; whereas, if we have a point count where both removal sampling and distance sampling was used, then both $J,K \geq 2$.
+The case where $J = K = 1$ is where neither removal sampling nor distance sampling was used, and so these data points are not appropriate for the modelling techniques described here.
 
@@
 
-\par For removal modelling, we are only interested in the time bin in which the bird was recorded, and not the distance, and so we will sum counts from each sampling event $i$ over all distance bands.
+\par For removal modelling, we were only interested in the time bin in which the bird was recorded, and not the distance, and so we will sum counts from each sampling event $i$ over all distance bands.
 Thus, we will be considering counts $Y_{sij.} = \sum_{k=1}^{K}$. 
@@
 \par We are most interested in estimating $\phi_s$, the cue rate for species $s$. 
-To allow for the partial pooling of estimates for each species to be based on a phylogenic relationship, we can use Gaussian processes, which effectively allow for "continuous" factors \citep{bernardo_regression_1998, mcelreath_continous_2020}.
+To allow for the partial pooling of estimates for each species to be based on a phylogenetic relationship, we can use Gaussian processes, which effectively allow for "continuous" factors \citep{bernardo_regression_1998, mcelreath_continous_2020}.
 Let $\mathbf{\Phi}$ be the vector of cue rates (of length $s$). 
@@
 
-\par Finally, the vector of mean cue rate $\mathbf{\mu}_{[s]}$ is a function of the species migratory strategy $\alpha_M$, where the migratory strategy $M$ could either be Resident or Migrant. 
+\par Finally, the vector of mean cue rate $\mathbf{\mu}_{[s]}$ is a function of the species migratory strategy $\alpha_M$, where the migratory strategy $M$ could either be resident or migrant. 
 We set priors on both migratory strategies to be $\alpha_M \sim N(-1, 0.01)$, because on the log scale we would expect cue rate to be negative.
@@
 
-\par In this model, mean effective detection radius for species $s$ (i.e., $\mu_s$) is a function of the following traits as described in \cite{solymos_phylogeny_2018}: an overall intercept term $\alpha_0$; migratory strategy $\alpha_M$, where $M$ can be Migratory or Resident; habitat preference $\alpha_H$, where $H$ can be Open or Closed habitat; log body size $\beta_b$; and log song pitch $\beta_p$.
+\par In this model, mean effective detection radius for species $s$ (i.e., $\mu_s$) is a function of the following traits as described in \cite{solymos_phylogeny_2018}: an overall intercept term $\alpha_0$; migratory strategy $\alpha_M$, where $M$ can be migratory or resident; habitat preference $\alpha_H$, where $H$ can be open or closed habitat; log body size $\beta_b$; and log song pitch $\beta_p$.
 That is, we have
@@
 \par We used the same standardized database of point count data that was assembled through the NA-POPS project \cite{edwards_point_2023}. 
-This consists of $712138$ point counts conducted across $292$ projects in Canada and the United States, $422514$ of which used removal sampling and $522820$ of which used distance sampling.
+This consists of $712,138$ point counts conducted across $292$ projects in Canada and the United States, $422,514$ of which used removal sampling and $522,820$ of which used distance sampling.
 All of these point counts were recorded in the database in a standard way, using practices developed by the Boreal Avian Modelling project \citep{barker_ecological_2015}.
@@
 \par We accessed 2000 pseudo-posterior trees from \citet{jetz_global_2012} for all species with sufficient data for removal modelling ($>75$ observations as per \cite{edwards_point_2023,solymos_evaluating_2018,buckland_introduction_2001}).
-We also requested that the phylogenies include the following bird species with insufficient data from NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricoloured Blackbird, and Spotted Owl.
+We also specified that the phylogenies include the following bird species with insufficient data from NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird, and Spotted Owl.
 All seven of these bird species are listed on Partner's in Flight's watchlist for various threats \citep{will_handbook_2020}. 

### 2024-04-10 src/scripts/post-hoc/4-species-cv-figure.R
removal model species-level CV posthoc

@@ -1,0 +1,65 @@
+####### Script Information ########################
+# Brandon P.M. Edwards
+# Multi-species QPAD Detectability
+# posthoc/4-species-cv-figure.R
+# Created April 2024
+# Last Updated April 2024
+
+####### Import Libraries and External Files #######
+
+library(cmdstanr)
+library(ggplot2)
+library(ggpubr)
+theme_set(theme_pubclean())
+
+####### Set Constants #############################
+
+k <- 10
+
+####### Read Data #################################
+
+load("data/generated/removal_stan_data_cv.rda")
+removal_all <- readRDS("output/model_runs/removal_ms.RDS")
+removal_folds <- vector(mode = "list", length = k)
+for (i in 1:k)
+{
+  removal_folds[[i]] <- readRDS(paste0("output/model_runs/cv_removal/species/fold_", i, ".RDS"))
+}
+rem_fold_nums <- read.csv("data/generated/removal_cv_folds_species.csv")
+
+####### Main Code #################################
+
+phi_all <- removal_all$summary("log_phi")
+phi_all$Species <- removal_stan_data_cv$sp_all
+
+rem_cv_df <- rem_fold_nums[!duplicated(rem_fold_nums$Species), ]
+rem_cv_df <- merge(rem_cv_df, phi_all[, c("mean", "Species")],
+                   by = "Species")
+rem_cv_df$Predicted <- NA
+names(rem_cv_df)[3] <- "True"
+
+for (i in 1:k)
+{
+  phi_fold <- removal_folds[[i]]$summary("log_phi")
+  phi_fold$Species <- removal_stan_data_cv$sp_all
+  
+  sps <- rem_cv_df[which(rem_cv_df$cv_fold == i), "Species"]
+  
+  for (sp in sps)
+  {
+    rem_cv_df[which(rem_cv_df$Species == sp), "Predicted"] <- phi_fold[which(phi_fold$Species == sp), "mean"]
+  }
+}
+
+rem_cv_df$True_Exp <- exp(rem_cv_df$True)
+rem_cv_df$Predicted_Exp <- exp(rem_cv_df$Predicted)
+rem_cv_df$Difference <- rem_cv_df$True_Exp - rem_cv_df$Predicted_Exp
+rem_cv_df$Abs_Diff <- abs(rem_cv_df$Difference)
+
+(removal_cv_plot <- ggplot(data = rem_cv_df, aes(x = True_Exp, y = Predicted_Exp)) +
[…]

### 2024-04-12 src/scripts/post-hoc/4-species-cv-figure.R
output removal species cv plot

@@ -56,10 +56,15 @@ rem_cv_df$Predicted_Exp <- exp(rem_cv_df$Predicted)
 rem_cv_df$Difference <- rem_cv_df$True_Exp - rem_cv_df$Predicted_Exp
 rem_cv_df$Abs_Diff <- abs(rem_cv_df$Difference)
 
-(removal_cv_plot <- ggplot(data = rem_cv_df, aes(x = True_Exp, y = Predicted_Exp)) +
+removal_cv_plot <- ggplot(data = rem_cv_df, aes(x = True_Exp, y = Predicted_Exp)) +
   geom_point() +
   geom_abline(slope = 1, intercept = 0) +
   xlim(0,0.8) + ylim(0,0.8) +
-  NULL)
+  NULL
 
 ####### Output ####################################
+
+png("output/plots/species_cv_plot.png",
+    width = 6, height = 6, res = 300, units = "in")
+print(removal_cv_plot)
+dev.off()

### 2024-05-01 src/scripts/distance/6-distance-predictions.R
udpates to distance prediction script

@@ -3,19 +3,17 @@
 # Multi-species QPAD Detectability
 # 6-distance-predictions.R
 # Created October 2022
-# Last Updated October 2023
+# Last Updated April 2024
 
 ####### Import Libraries and External Files #######
 
 library(cmdstanr)
-library(dplyr)
 
 source("src/functions/generate-distance-inits.R")
-source("src/functions/subset-distance-data.R")
 
 ####### Load Data #################################
 
-load("data/generated/distance_stan_data.rda")
+load("data/generated/distance_stan_data_pred.rda")
 
 ####### Set Constants #############################
 
@@ -24,26 +22,24 @@ n_iter <- 2000
 n_warmup <- 1000
 n_chains <- 4
 refresh <- 10
-threads_per_chain <- 7
+threads_per_chain <- 4
 
-distance_stan_data$grainsize <- 1
+distance_stan_data_pred$grainsize <- 1
 
 inits <- generate_distance_inits(n_chains = n_chains,
                                  napops_skip = c("BITH", "HASP", "KIWA", "LCTH", "LEPC", "SPOW"),
-                                 sp_list = distance_stan_data$sp_all,
-                                 param = "cp",
-                                 species_cp = distance_stan_data$species_cp,
-                                 species_ncp = distance_stan_data$species_ncp)
+                                 sp_list = distance_stan_data_pred$sp_all,
+                                 param = "cp")
 
-distance_stan_data$sp_list <- NULL
-distance_stan_data$sp_all <- NULL
+distance_stan_data_pred$sp_list <- NULL
+distance_stan_data_pred$sp_all <- NULL
 
 # Scale the maximum distances for computational ease
-distance_stan_data$max_dist <- distance_stan_data$max_dist / 100
+distance_stan_data_pred$max_dist <- distance_stan_data_pred$max_dist / 100
 
 # get rid of centre/scale attributes for modelling
-distance_stan_data$pitch <- distance_stan_data$pitch[,1]
-distance_stan_data$mass <- distance_stan_data$mass[,1]
+distance_stan_data_pred$pitch <- distance_stan_data_pred$pitch[,1]
+distance_stan_data_pred$mass <- distance_stan_data_pred$mass[,1]
 
 ####### Run Model #################################
 
@@ -51,14 +47,13 @@ model_file <- cmdstan_model(stan_file = "models/distance_cp.stan",
                             cpp_options = list(stan_threads = TRUE))
 
[…]

### 2024-05-02 article/article.tex

@@
 \par The goal of this paper is to create a multi-species model of detectability that incorporates species traits and phylogeny, to both allow for the relaxation of the sample size requirements for estimating detectability, and to derive estimates of detectability for bird species without any data.
-To do so, we developed a Bayesian hierarchical version of the QPAD removal model initially derived in \citet{solymos_calibrating_2013}, that allows for information exchange between species based on phylogenetic distance and binned migratory status.
-Phylogenetic distance was coded as a correlation structure and used as a Gaussian random field in this model, and migratory status was coded similarly to \citet{solymos_phylogeny_2018} in which a species was considered either "migrant" or "resident".
-We also developed a Bayesian hierarchical version of the QPAD distance model initially derived in \citet{solymos_calibrating_2013} that allows for information exchange among species based only on the traits shown to be significant by \citet{solymos_phylogeny_2018}.
-These traits include migratory status (coded the same as removal models), habitat association (coded as either "open" or "closed"), song pitch, and body mass. 
-We also developed a Bayesian single species version of the QPAD removal and distance models (i.e., as opposed to the maximum likelihood models available in the \texttt{detect} R package; \citet{solymos_detect_2020} to allow for direct comparison through cross validation.
-Finally, we ran these models with the full set of data points used in \citet{edwards_point_2023}, plus additional data for seven species that did not meet the initial sample size thresholds for NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird (\textit{Agelaius tricolor}), and Spotted Owl (\textit{Strix occidentalis}).
+To achieve this goal, we wanted to satisfy the following objectives and predictions:
+\begin{enumerate}
+	\item Compare and contrast estimates of cue rate and EDR produced from a Bayesian hierarchical QPAD removal model and distance model, respectively, to estimates produced from single-species model which does not share information.
+	We predicted that for both the Bayesian hierarchical QPAD removal model and Bayesian hierarchical QPAD distance model, estimates of cue rate and EDR, respectively, will differ very little from those estimated from a single-species model where no information is shared \cite{edwards_point_2023}, for species where we have sufficient sample size.
+	\item Compare precision in estimates of cue rate and EDR produced from a Bayesian hierarchical QPAD removal model and distance model, respectively, to estimates produced from single-species model which does not share information.
+	We predicted that estimates of cue rate and EDR will be more precise when produced by a Bayesian hierarchical QPAD model than with a single-species QPAD model, but this increased precision will be most evident with smaller sample sizes.
+	\item Assess the capacity for a multi-species model to produces estimates of cue rate or EDR for species that are data sparse.
+	We predicted that a multi-species model will be able to produce better fitting estimates of cue rate or EDR than a single species model, particularly for data sparse species.
+	\item Assess the capacity for a multi-species model to make out-of-sample predictions of cue rate or EDR for species which have no data.
+	We predicted that a multi-species model will be able to produce estimates of cue rate or EDR that fall within a reasonable range of estimates from related species.
+\end{enumerate}
+
+Given these comparisons, we provide a case study of predicting detectability components for seven species for which we had little to no data: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird (\textit{Agelaius tricolor}), and Spotted Owl (\textit{Strix occidentalis}).
 All seven of these bird species are listed on Partners in Flight’s watchlist, indicating their conservation importance \citep{will_handbook_2020}.
-
-\par We predicted that for both the Bayesian hierarchical QPAD removal model and Bayesian hierarchical QPAD distance model, estimates of cue rate and EDR, respectively, will differ very little from those estimated from the original NA-POPS database \cite{edwards_point_2023}.
-This is because the species that were modelled in \citet{edwards_point_2023} had a sufficient number of data points to stand on their own, and so adding in species-level effects should have little effect on estimates.
-Furthermore, we predicted that improvement in precision from a multi-species model will be related to sample size, such that precision of estimates will improve the most for species with smaller sample size, while precision of estimates will improve the least for species with larger sample sizes.
-Finally, we predicted that a multi-species model will have greater predictive capacity than a single-species model, particularly for species which are data-sparse.
-With the predictions above, we then expect to be able to generate estimates of cue rate and EDR for species with very little data, thus relaxing previous sample size requirements; additionally, we expect to be able to predict cue rate and EDR for species with no data by making use of the shrinkage effect offered by a multi-species model.
 
@@
 Finally, we describe the experiments conducted to assess model performance and model comparison, which then allow us to make predictions for the seven undersampled species of interest.
+
+\subsection{Data Collection}
+\subsubsection{Point Count Data}
+\par We used the same standardized database of point count data that was assembled through the NA-POPS project \cite{edwards_point_2023}. 
+This consists of $712,138$ point counts conducted across $292$ projects in Canada and the United States, $422,514$ of which used removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002} and $522,820$ of which used distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}.
+All of these point counts were recorded in the database in a standard way, using practices developed by the Boreal Avian Modelling project \citep{barker_ecological_2015}.
+
+\subsubsection{Phylogenetic Trees}
+\par We accessed 2000 pseudo-posterior trees from \citet{jetz_global_2012} for all species with sufficient data for removal modelling ($>75$ observations as per \cite{edwards_point_2023,solymos_evaluating_2018,buckland_introduction_2001}).
+We also requested that the phylogenies include the following bird species with insufficient removal and/or distance data: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird, and Spotted Owl.
+All seven of these bird species are listed on Partner's in Flight's watchlist for various threats \citep{will_handbook_2020}. 
+This resulted in 2000 pseudo-posterior trees that contained 323 species.
+
+\par We generated a phylogenetic correlation matrix $\mathbf{D}$ using the same methods outlined in the Appendix of \citet{solymos_phylogeny_2018} that generate distances between species, and further processed using the R package "ape" \citet{paradis_ape_2019}.
+
+\subsubsection{Species Traits}
+\par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts of all the birds to model in this study \citep{billerman_birds_2022}. 
+Birds were classified as "closed habitat" if their primary habitat was forests, and "open habitat" if their primary habitat was grasslands or shrublands.
+Any bird species that has specific areas of their range that is only for breeding (i.e., not year-round throughout their range) was classified as "migratory".
+Body mass for each bird species was obtained from the Elton traits dataset \citep{wilman_eltontraits_2014}. 
+Where body mass was not available through the Elton traits dataset, the Birds of the World account \citep{billerman_birds_2022} was used instead.
 
@@
 
-\par To directly compare the multi-species models described above to its single-species equivalent in a cross-validation (see below), we built two Bayesian single-species models in Stan.
+\par To directly compare the multi-species models described above to its single-species equivalent (see below), we built two Bayesian single-species models in Stan.
 The single species models used the same likelihood statements as above for the count data, except rather than cue rate and EDR being estimated with shared information, they are estimated independently from each other.
@@
 
-\subsection{Data Collection}
-\subsubsection{Point Count Data}
-\par We used the same standardized database of point count data that was assembled through the NA-POPS project \cite{edwards_point_2023}. 
-This consists of $712,138$ point counts conducted across $292$ projects in Canada and the United States, $422,514$ of which used removal sampling and $522,820$ of which used distance sampling.
-All of these point counts were recorded in the database in a standard way, using practices developed by the Boreal Avian Modelling project \citep{barker_ecological_2015}.
-
-\subsubsection{Phylogenetic Trees}
-\par We accessed 2000 pseudo-posterior trees from \citet{jetz_global_2012} for all species with sufficient data for removal modelling ($>75$ observations as per \cite{edwards_point_2023,solymos_evaluating_2018,buckland_introduction_2001}).
-We also specified that the phylogenies include the following bird species with insufficient data from NA-POPS: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird, and Spotted Owl.
-All seven of these bird species are listed on Partner's in Flight's watchlist for various threats \citep{will_handbook_2020}. 
-This resulted in 2000 pseudo-posterior trees that contained 323 species.
-
-\par We generated a phylogenetic correlation matrix $\mathbf{D}$ using the same methods outlined in the Appendix of \citet{solymos_phylogeny_2018} that generate distances between species, and further processed using the R package "ape" \citet{paradis_ape_2019}.
-
-\subsubsection{Species Traits}
-\par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts of all the birds to model in this study \citep{billerman_birds_2022}. 
-Body mass for each bird species was obtained from the Elton traits dataset \citep{wilman_eltontraits_2014}. 
-Where body mass was not available through the Elton traits dataset, the Birds of the World account \citep{billerman_birds_2022} was used instead.
-
 \subsection{Model Assessment}
-
-\par To test our prediction that there would be little change between estimates of species already modelled in NA-POPS, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data in the original NA-POPS analysis.
+\subsubsection{Objective 1: Single-Species vs. Multi-Species Estimates}
+\par To test our prediction that there would be little change between estimates of species with sufficient sizes, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data.
 In this case, we ran both models for 316 species of birds.
 Rufous Hummingbird (\textit{Selasphorus rufus}) was dropped from the distance modelling due to issues with divergent transitions \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
-Tricoloured Blackbird had sufficient distance data, and so was included in the distance modelling, but not the removal modelling.
-We accessed the single-species estimates of cue rate and EDR from the NA-POPS Null models using the \texttt{napops} R package \citep{edwards_napops_2023, edwards_point_2023}.
-We then modelled the difference in estimated cue rates and EDRs as
+Tricolored Blackbird was only included in the distance model because there was sufficient distance sampling data.
+We also ran the single species removal and distance models for the same suite of birds.
+We then subtracted the estimate of the single-species cue rate or EDR from the estimate of the multi-species cue rate or EDR, and modelled the difference in estimated cue rates and EDRs $\Delta X$ as
 $$\Delta X \sim N(0,\sigma)$$
@@
 
-where $\Delta X$ is calculated as the mean multi-species estimate of cue rate or EDR minus the mean single-species estimate of cue rate of EDR.
-If our prediction is true, we would expect to see a distribution of modelled differences centred about 0, indicating no real difference in estimates of cue rate or EDR between single-species and multi-species models, for species for which we have sufficient data.
+If our prediction is true, we would expect to see a distribution of modelled differences centred about 0, indicating no difference in estimates of cue rate or EDR between single-species and multi-species models, for species for which we have sufficient data.
 
-\par To test our prediction about improved precision, we simply plotted the species-specific estimates of standard deviations of cue rate and EDR, respectively, against the log sample size, for each of the single-species and multi-species model.
+\subsubsection{Objective 2: Comparison of Precision}
+\par To test our prediction about improved precision, we plotted the species-specific estimates of standard deviations of cue rate and EDR, respectively, against the log sample size, for each of the single-species and multi-species model.
 We then fit a smooth LOESS curve through each of the model-specific points.
-If our prediction is true, we would expect to see a plot showing a higher standard deviation in single-species models than for multi-species models, but only up until a certain sample size when they should both be similar.
+If our prediction is true, we would expect to see a plot showing a higher standard deviation in single-species models than for multi-species models, until sample size is such that they would both be similar.
+
+\subsubsection{Objective 3: Assessment of Estimates for Data-sparse Species}
 
 \par To test our prediction about improved accuracy, we ran a k-fold cross-validation with 5 folds, stratified by species.
-That is, for each cross-validation run, we held out 20\% of the data for each species for both the multi-species and Bayesian single-species models. 
-We then derived multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR, for each cross-validation run, and used these parameter estimates to calculate the estimated pointwise log posterior density of each of the held out data points.
-We then calculated the difference in posterior predictive density for removal and distance models, by subtracting the pointwise posterior predictive density of the single-species model from the pointwise posterior predictive density of the multi-species model (i.e., multi-species minus single-species).
-Finally, we took the mean and standard error of the difference in pointwise log posterior density for each species to create a species-specific mean and standard error of difference in pointwise posterior density.
+That is, for each of the five cross-validation run, we held out 20\% of the data for each species for both the multi-species and single-species models. 
+We then derived multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR, for each cross-validation run, and used these parameter estimates to calculate the estimated log pointwise posterior density (LPPD) of each of the held out data points.
+We then calculated the difference in LPPD for removal and distance models, by subtracting the LPPD of the single-species model from the LPPD of the multi-species model (i.e., multi-species minus single-species).
+Finally, we took the mean and standard error of the difference in LPPD for each species to create a species-specific mean and standard error of difference in LPPD.
 With this, positive intervals (i.e., where mean $\pm$ standard error is all positive) would indicate that the multi-species model was preferred, and negative intervals would indicate that the single-species model was preferred.
-If our prediction is true, then we would expect to see majority positive values of the species-specific mean pointwise posterior density.
+If our prediction is true, then we would expect to see majority positive values of the species-specific mean LPPD.
 
-\subsection{Predicting Detection Probabilities}
-\par To predict estimates of cue rate or EDR from the multi-species models, we adopted two possible approaches.
-If the NA-POPS database contained at least some point-count data for the species of interest, we simply added these data into the model.
-This is different from \citet{edwards_point_2023}, \citet{solymos_calibrating_2013}, and \citet{solymos_evaluating_2018}, where any species with less than 75 data points were removed from the analysis.
-For species that did not have any data in the NA-POPS database, we generated a row of traits data for the species and estimated its cue rate and EDR as parameters in the model.
-We then ran the multi-species models described above using these additional data points (and hence an updated phylogenetic correlation structure for the removal model) to generate estimates of cue rate and EDR for these new species.
+\subsubsection{Objective 4: Assessment of Out-of-sample Predictions}
+\par To test our prediction that a multi-species model would be able to predict cue rates or EDRs for species with no data, we ran a 10-fold leave-species-out cross validation.
+That is, for each of the ten cross-validation runs, we held out all data for 10\% of the species for both the multi-species and single-species models.
+To determine which species would be left out in each fold of the removal model, we first split the phylogenetic tree into 10 parts using the "cutree" function in base R, which preserves relatedness in the 10 parts of the split tree.
+Then, for each of the parts of the tree, we randomly assigned a number between 1 and 10 (drawn from a uniform distribution) to all migratory then all resident species.
+This ensured that for each fold, there would be a balance of related and unrelated species left out, as well as a relatively equal balance of migratory and resident species left out.
+To determine which species would be left out in each fold of the distance model, we randomly assigned a number between 1 and 10 (drawn from a uniform distribution) to each species within a migratory resident $\times$ habitat preference, which ensured that there would be a good balance of species left out for each group of species.
+by either phylogeny and migratory status (for removal modelling) or by migratory status and habitat preference (for distance modelling).
+Because Stan allows draws from a posterior distribution even with no data, we drew values from the posterior distribution of all of the held out species for each cross-validation fold, and used mean and standard deviation of those draws as the predicted cue rate or EDR.
+We then modelled the difference in those predicted cue rates or EDRs to the "true" cue rates or EDRs obtained from the multi-species models in Objective 1.
+
+\subsection{Case Study: Predicting Detection Probabilities for Rare and Undersampled Species}
+\par We ran the multi-species removal and distance models again, this time with the minimal data available for seven species of undersampled landbirds, and generated estimates of cue rate and EDR for these seven species.
+For tricolored blackbird, we had sufficient data points for distance modelling for Objectives 1 - 4, and so we only generated estimates of cue rate here.
+For two of the species, Kirtland's Warbler and Bicknell's Thrush, we used the predicted cue rate and EDR to calculate availability and perceptibility, and hence detectability, for a given survey type, and compared these metrics to previous estimates of detectability obtained from other studies.
+As a reminder, following derivations in \citet{solymos_calibrating_2013}, detectability is the product of availability and perceptibility.
+Availability $p(t)$ is defined as the probability of a bird giving a cue during a survey run for $t$ minutes, and is calculated as
+$p(t) = 1 - e^{-t\phi}$
+where $\phi$ is the cue rate of the species.
+Perceptibility $q(r)$ is defined as the probability of an observer hearing or seeing a bird, given the bird gives a cue, with a survey radius of $r$ metres, and is calculated as
+$q(r) = \dfrac{\tau^2 \left\{1 - e^{-\dfrac{r^2}{\tau^2}}\right\}}{r^2}$
+where $\tau$ is the EDR of the species.
+For Kirtland's Warbler, we compared our estimates of detectability to those obtained in \citet{van_dyke_comparative_2022} which compared estimates of Kirtland's Warbler detectability among different jack pine stands and red pine stands.
+For Bicknell's Thrush, we compared our estimates of detectability to those obtained in \citet{aubry_bicknells_2018} which estimated detectability in a small portion of Bicknell's Thrush's breeding range.
 
@@
 Each of the single-species models and multi-species models, as well as the models created to assess performance, were run using 1000 warmup iterations and 2000 sampling iterations, each on 4 chains for a total of 8000 draws per model run.
-To mitigate issues surrounding initial values being rejected in the distance model (see Discussion), we set the initial values of EDR to be the mean EDR from NA-POPS estimates; for species where EDR estimates did not exist in NA-POPS, we used the overall mean EDR as initial value.
-Cross-validation models were run with 1000 warmup iterations and 500 sampling iterations, each on 4 chains for a total of 2000 draws per model run.
+To mitigate issues surrounding initial values being rejected in the distance model (see Discussion), we set the initial values of EDR to be the mean EDR from NA-POPS \citep{edwards_point_2023} estimates; for species where EDR estimates did not exist in NA-POPS, we used the overall mean EDR as initial value.
 All of the models made use of Stan's reduce\_sum functionality, which allows for within-chain parallelization as well as the usual across-chain parallelization.
@@
 \section{Results}
-\subsection{Parameter Estimates}
-
-\par We found much less support for a difference in cue rate between resident and migrant species than was found in \citet{solymos_evaluating_2018}. 
-Mean overall log cue rate in the multi-species model was -1.09 (95\% credible interval of [-1.24, -0.94]); both estimated effects for resident and migrant species were effectively 0 (Figure \ref{fig:params}A).
-We also found little support for migration strategy having any effect on EDR.
-There is some evidence that a resident species may have a slightly higher EDR than a migrant species, but the 95\% credible intervals of both effects overlapped with the 95\% credible interval of the overall estimated intercept (Figure \ref{fig:params}B).
-We did find similar support for change in EDR by body size and pitch, in that body size tended to have an increasing effect on EDR, and song pitch had a decreasing effect on EDR (Figure \ref{fig:params}B).
 
@@
 
-\par We used the multi-species removal model and distance model to generate predictions of cue rate and EDR, respectively, for 7 species of undersampled birds from the original NA-POPS analysis (6 for distance modelling because Tricoloured Blackbird had sufficient distance sampling data).
+\par We used the multi-species removal model and distance model to generate predictions of cue rate and EDR, respectively, for seven species of undersampled birds from the original NA-POPS analysis (6 for distance modelling because Tricoloured Blackbird had sufficient distance sampling data).
 
@@
 The best model for predictive accuracy tended to depend on the species, but overall differences in the estimates were small enough that using a multi-species model should not produce extremely inaccurate estimates.
-
-\par Although this was not a primary goal of our study, when considering the species-level covariates, we found much less support than \citet{solymos_phylogeny_2018} for migration strategy having an effect on cue rate and EDR.
-Both the estimated effects for resident species and migratory species were effectively 0 in the removal model, with very little of the density above or below 0.025 from 0.
-These estimates varied slightly more in the distance model, but the 95\% credible interval still captured 0.
-One potential reason is that we had proportionally more resident species in our analysis (85 out of 316; 26.9\%) compared to \citet{solymos_phylogeny_2018} (23 out of 141; 16.3\%), and so the richer information provided by these additional species could have brought the effect down to 0.
-Habitat type, body mass, and song pitch, as expected, all had an overall effect on EDR, as was found in \citet{solymos_phylogeny_2018} and \cite{johnston_species_2014}.
 
@@
 \par Using the predictions of cue rate and EDR for Bicknell’s Thrush and Kirtland’s Warbler, we can calculate availability and perceptibility, and hence detectability, for a given survey type, and compare to previous estimates of detectability obtained from other studies.
-As a reminder, following derivations in \citet{solymos_calibrating_2013}, detectability is the product of availability and perceptibility.
-Availability $p(t)$ is defined as the probability of a bird giving a cue during a survey run for $t$ minutes, and is calculated as
-$p(t) = 1 - e^{-t\phi}$
-where $\phi$ is the cue rate of the species.
-Perceptibility $q(r)$ is defined as the probability of an observer hearing or seeing a bird, given the bird gives a cue, with a survey radius of $r$ metres, and is calculated as
-$q(r) = \dfrac{\tau^2 \left\{1 - e^{-\dfrac{r^2}{\tau^2}}\right\}}{r^2}$
-where $\tau$ is the EDR of the species.
 
@@
 \begin{figure}[h]
-	\includegraphics{../output/plots/parameters_plot.png}
-	\caption{Posterior distributions with mean (points), 50\% credible intervals (bolded lines), and 95\% credible intervals (thin lines) of parameters estimated in the multi-species removal model (A) and the multi-species distance model (B).}
-	\label{fig:params}
-\end{figure}
-
-\begin{figure}[h]
 	\includegraphics{../output/plots/1vs1.png}

### 2024-05-02 article/article.tex
udpates to article and posthoc analysis

@@
 
-\subsection{Comparison of Estimates between Single Species and Multi Species}
+\subsection{Single-Species vs. Multi-Species Estimates}
 
-\par We found very little evidence for an overall change in cue rate for a given species modelled by a single-species model vs a multi-species model (Figure \ref{fig:1vs1}A).
-Seven out of the 316 species considered for removal modelling had a percent change in cue rate of more than 20\%.
-These species were Willow Ptarmigan (\textit{Lagopus lagopus}), Rusty Blackbird (\textit{Euphagus carolinus}), Red-bellied Sapsucker (\textit{Sphyrapicus ruber}), Common Redpoll (\textit{Acanthis flammea}), Calliope Hummingbird (\textit{Selasphorus calliope}), Bohemian Waxwing (\textit{Bombycilla garrulus}), and Allen's Hummingbird (\textit{Selasphorus sasin}).
-All seven of these species were among the lowest cue rates in both the single-species and multi-species context (Figure \ref{fig:1vs1}A).
+\par We found slight evidence for an overall change in cue rate for a given species modelled by a single-species model vs a multi-species model.
+The mean modelled difference in cue rate was $-1.57 \times 10^{-3}$, indicating that estimates of cue rate from a multi-species model may be slightly less than estimates of cue rate from a single-species model.
+Ten out of the 316 species considered for removal modelling had a percent change in cue rate of more than 10\%.
+These species were: Willow Ptarmigan (\textit{Lagopus lagopus}), Rusty Blackbird (\textit{Euphagus carolinus}), Red-bellied Sapsucker (\textit{Sphyrapicus ruber}), Prairie Falcon (\textit{Falco mexicanus}), Nuttall's Woodpecker (\textit{Picoides nuttallii}), Northern Goshawk (\textit{Accipiter atricapillus}), Ferruginous Hawk (\textit{Buteo regalis}), Common Redpoll (\textit{Acanthis flammea}), Calliope Hummingbird (\textit{Selasphorus calliope}), and Bohemian Waxwing (\textit{Bombycilla garrulus})(Figure \ref{fig:1vs1}A).
 
-\par We found little evidence for a change in modelled EDR between single species and multi-species distance models (Figure \ref{fig:1vs1}B). 
-(NEEDS UPDATING) Seven out of the 317 species had a percent change in EDR of more than 20\% (Figure \ref{fig:1vs1}B).
-These species were Sharp-tailed Grouse (\textit{Tympanuchus phasianellus}), Sedge Wren (\textit{Cistothorus stellaris}), Ring-necked Pheasant (\textit{Phasianus colchicus}), Chimney Swift (\textit{Chaetura pelagica}), Chihuahuan Raven (\textit{Corvus cryptoleucus}), Boat-tailed Grackle (\textit{Quiscalus major}), and Baird's Sparrow (\textit{Ammodramus bairdii}).
+\par We found little evidence for a change in modelled EDR between single species and multi-species distance models. 
+The mean modelled difference in EDR was 0.97, but the 90\% credible interval overlapped zero.
+Two out of the 315 species had a percent change in EDR of more than 10\%.
+These species were Greater Prairie-chicken (\textit{Tympanuchus cupido}) and Chihuahuan Raven (\textit{Corvus cryptoleucus}) (Figure \ref{fig:1vs1}B).
 
@@
 
-\par We found some evidence that a multi-species removal model improved the precision of cue rate estimates compared to estimates from a single-species model (Figure \ref{fig:sd}A).
-This was particularly evident for species with sample sizes less than approximately $\exp\left\{7\right\} \approx 1097$, where the estimated LOESS curve for single-species standard deviation was always higher than the estimated LOESS curve for multi-species standard deviations.
-For sample sizes greater than 1097, standard deviations appeared similar in both the single-species and multi-species model, indicating no real improvement in precision.
-
-\par We found little evidence that a multi-species model improves the precision of EDR compared to that of a single-species model (Figure \ref{fig:sd}B).
-Both estimated LOESS curves for single-species standard deviations and multi-species standard deviations were roughly equal across all sample sizes, indicating no real improvement (or decrease) in precision.
+\par For both the removal and distance models, we found little evidence that a multi-species model improved the precision of estimates of cue rate or EDR, respectively.
+For removal modelling, the LOESS curves show some difference in sample sizes less than approximately $\exp\left\{7\right\} \approx 1097$, where the estimated LOESS curve for single-species standard deviation was slightly higher than the estimated LOESS curve for multi-species standard deviations (Figure \ref{fig:sd}A).
+For distance modelling, the curves were relatively equivalent throughout the range of sample sizes (Figure \ref{fig:sd}B).
 

### 2024-05-02 src/scripts/post-hoc/1-1vs1-figure.R

@@ -3,7 +3,7 @@
 # Multi-species QPAD Detectability
 # posthoc/1-1vs1-figure.R
 # Created December 2023
-# Last Updated April 2024
+# Last Updated May 2024
 
 ####### Import Libraries and External Files #######
 
@@ -18,8 +18,6 @@ library(patchwork)
 theme_set(theme_pubclean())
 bayesplot::color_scheme_set("red")
 
-source("src/functions/subset-distance-data.R")
-
 ####### Read Data #################################
 
 rem_ms <- readRDS("output/model_runs/removal_ms.RDS")
@@ -28,7 +26,7 @@ load("data/generated/removal_stan_data_cv.rda")
 
 dis_ms <- readRDS("output/model_runs/distance_ms.RDS")
 dis_ss <- readRDS("output/model_runs/distance_ss.RDS")
-load("data/generated/distance_stan_data.rda")
+load("data/generated/distance_stan_data_cv.rda")
 
 ####### Removal Model #############################
 
@@ -67,7 +65,7 @@ to_plot$Multi <- exp(to_plot$Multi)
 to_plot$Single <- exp(to_plot$Single)
 to_plot$diff <- to_plot$Multi - to_plot$Single
 
-# Check for a 20% relative difference in cue rates
+# Check for a 10% relative difference in cue rates
 for (i in 1:nrow(to_plot))
 {
   if ((abs(to_plot$diff[i]) / to_plot$Single[i]) > 0.10)
@@ -111,18 +109,16 @@ removal_plot <- single_vs_multi_plot + inset_element(modelled_difference_plot,
                                                      top = 0.5)
 
 cr_differences <- to_plot
+cr_diff_model <- diff_model_run$summary()
 
 ####### Distance Model ############################
 
-pred_drops <- c("LCTH", "LEPC", "HASP", "SPOW", "KIWA", "BITH")
-distance_stan_data_cv <- distance_stan_data
-
 # Extract log_tau summary statistics from full Stan model runs and add species Codes
 dis_ms_summary <- dis_ms$summary("log_tau")
-dis_ms_summary$Code <- setdiff(distance_stan_data_cv$sp_all, pred_drops)
+dis_ms_summary$Code <- distance_stan_data_cv$sp_all
 
 dis_ss_summary <- dis_ss$summary("log_tau")
-dis_ss_summary$Code <- setdiff(distance_stan_data_cv$sp_all, pred_drops)
[…]

### 2024-05-02 src/scripts/post-hoc/2-sd-figure.R

@@ -23,8 +23,9 @@ rem_ms <- readRDS("output/model_runs/removal_ms.RDS")
 rem_ss <- readRDS("output/model_runs/removal_ss.RDS")
 load("data/generated/removal_stan_data_cv.rda")
 
-dis_model <- readRDS("output/model_runs/distance_predictions.RDS")
-load("data/generated/distance_stan_data.rda")
+dis_ms <- readRDS("output/model_runs/distance_ms.RDS")
+dis_ss <- readRDS("output/model_runs/distance_ss.RDS")
+load("data/generated/distance_stan_data_cv.rda")
 
 ####### Removal Model #############################
 
@@ -66,50 +67,49 @@ to_plot <- data.frame(Species = rep(removal_stan_data_cv$sp_all, 2),
   geom_smooth() +
   xlab("log(Sample Size)") +
   ylab("Standard Deviation") +
-  #theme(legend.position = "none") +
+  theme(legend.position = "none") +
   NULL)
 
 ####### Distance Model #############################
 
-# Extract log_phi summary statistics from full Stan model runs
-dis_summary <- dis_model$summary("log_tau")
+# Extract log_tau summary statistics from full Stan model runs
+dis_summary_ms <- dis_ms$summary("log_tau")
+dis_summary_ss <- dis_ss$summary("log_tau")
 
-# Add species names to these summaries
-dis_summary$Code <- distance_stan_data$sp_all
+# Add the species names
+dis_summary_ms$Code <- distance_stan_data_cv$sp_all
+dis_summary_ss$Code <- distance_stan_data_cv$sp_all
 
 # Get data sample size for all species and add to summary
-species_n <- data.frame(table(distance_stan_data$species))
+species_n <- data.frame(table(distance_stan_data_cv$species))
 names(species_n) <- c("index", "N")
-dis_summary$index <- seq(1, nrow(dis_summary))
-dis_summary$N <- 0
+dis_summary_ms$index <- seq(1, nrow(dis_summary_ms))
+dis_summary_ss$index <- seq(1, nrow(dis_summary_ss))
+
+dis_summary_ms$N <- 0
+dis_summary_ss$N <- 0
+
 for (i in 1:nrow(species_n))
 {
-  dis_summary[which(dis_summary$index == species_n$index[i]), "N"] <-
+  dis_summary_ms[which(dis_summary_ms$index == species_n$index[i]), "N"] <-
+    species_n$N[i]
+  
[…]

### 2024-05-06 src/scripts/distance/5-distance-cv.R
cahnges to distance cv script + ms fold 1 distance

@@ -3,19 +3,17 @@
 # Multi-species QPAD Detectability
 # 5-distance-cv.R
 # Created October 2023
-# Last Updated March 2024
+# Last Updated May 2024
 
 ####### Import Libraries and External Files #######
 
 library(cmdstanr)
-library(dplyr)
 
 source("src/functions/generate-distance-inits.R")
-source("src/functions/subset-distance-data.R")
 
 ####### Load Data #################################
 
-load("data/generated/distance_stan_data.rda")
+load("data/generated/distance_stan_data_cv.rda")
 cv_folds <- read.csv("data/generated/distance_cv_folds.csv")
 
 ####### Set Constants #############################
@@ -25,18 +23,10 @@ n_iter <- 2000
 n_warmup <- 1000
 n_chains <- 4
 refresh <- 10
-threads_per_chain <- 7
+threads_per_chain <- 3
+dis_data <- distance_stan_data_cv # lazy
 
 ####### Data Wrangling ############################
-
-#' First we must drop all species which are just not going to be involved
-#' in the cross validation at all. This will be the species for which we are
-#' generated predictions
-pred_drops <- c("LCTH", "LEPC", "HASP", "SPOW", "KIWA", "BITH")
-dis_data <- subset_distance_data(distance_stan_data = distance_stan_data,
-                                 sps = setdiff(unique(distance_stan_data$sp_list),
-                                               pred_drops))
-rm(distance_stan_data)
 
 ms_model <- cmdstan_model(stan_file = "models/distance_cp.stan",
                           cpp_options = list(stan_threads = TRUE))
@@ -61,8 +51,8 @@ for (i in 1:max(cv_folds$cv_fold))
                        mig_strat = dis_data$mig_strat,
                        n_habitat = dis_data$n_habitat,
                        habitat = dis_data$habitat,
-                       mass = dis_data$mass,
-                       pitch = dis_data$pitch,
+                       mass = dis_data$mass[,1],
+                       pitch = dis_data$pitch[,1],
                        sp_all = dis_data$sp_all)
   
   # Check to see if we can drop any training columns (i.e. if bands per sample is smaller)
@@ -79,7 +69,7 @@ for (i in 1:max(cv_folds$cv_fold))
   stan_cv_data$max_dist <- stan_cv_data$max_dist / 100
   
   inits <- generate_distance_inits(n_chains = n_chains,
-                                   sp_list = setdiff(as.vector(stan_cv_data$sp_all), pred_drops),
[…]

### 2024-05-06 article/article.tex
updates to posthoc scripts and article

@@
 
-\subsection{Comparison of Predictive Accuracy between Single Species and Multi Species}
+\subsection{Assessment of Model Preference for Data-sparse Species}
 
-\par The cross validation revealed that the multi-species removal model had better predictive capacity for more species (200 out of 316) than the single-species removal models (99 out of 316).
-Seventeen out of 316 species showed no better predictive capacity with either model.
-Mean overall difference in log predictive density was $4.78 \times 10^{-6}$, indicating a small overall preference for the multi-species model.
+\par The k-fold cross validation revealed that the multi-species removal model had better predictive performance for more species (200 out of 316) than the single-species removal model (94 out of 316).
+Twenty-two out of 316 species showed no preference between either model.
+Mean overall difference in species-wise mean LPPD was $2.65 \times 10^{-4}$, indicating a small overall preference for the multi-species model.
 As expected, multi-species models tended to have better predictive capacity when sample size was low, with a negligible difference as sample size increased (Figure \ref{fig:cv}A).
-There did not appear to be any patterns in model selection among phylogeny or traits (see Supplemental Figure 1 for full phylogenetic tree).
-Among the seven species with greater than 20\% change in cue rate, three species (Allen's Hummingbird, Bohemian Waxwing, and Rusty Blackbird) had the multi-species model chosen as the better model, three species (Common Redpoll, Red-bellied Sapsucker, and Willow Ptarmigan) had the single-species model chosen as the better model, and one species (Calliope Hummingbird) had neither model chosen as the better model.
+Among the ten species with greater than 10\% change in cue rate, three species (Bohemian Waxwing, Northern Goshawk, and Rusty Blackbird) had the multi-species model chosen as the better model, whereas the remaining seven species had the single-species model chosen as the better model.
 
-\par The cross validation revealed that the multi-species distance model also tended to have better predictive capacity for more species (156 out of 316) than the single-species distance models (140 out of 316), but to a lesser extent than the removal models.
+\par NEEDS UPDATING The cross validation revealed that the multi-species distance model also tended to have better predictive capacity for more species (156 out of 316) than the single-species distance models (140 out of 316), but to a lesser extent than the removal models.
 Twenty out of the 316 species showed no better predictive capacity with either model.
@@
 Among the seven species with greater than 20\% change in EDR, three species (Baird's Sparrow, Sedge Wren, and Sharp-tailed Grouse) had the multi-species model chosen as the better model, and four species (Boat-tailed Grackle, Chihuahuan Raven, Chimney Swift, and Ring-necked Pheasant) has the single-species model chosen as the better model.
+
+\subsection{Assessment of Out-of-sample Predictions}
 

### 2024-05-06 src/scripts/post-hoc/3-cv-figure.R

@@ -3,7 +3,7 @@
 # Multi-species QPAD Detectability
 # posthoc/3-cv-figure.R
 # Created December 2023
-# Last Updated April 2024
+# Last Updated May 2024
 
 ####### Import Libraries and External Files #######
 
@@ -17,13 +17,12 @@ theme_set(theme_pubclean())
 bayesplot::color_scheme_set("red")
 
 source("src/functions/rem-lik.R")
-source("src/functions/subset-distance-data.R")
 source("src/functions/dis-lik.R")
 
 ####### Read Data #################################
 
 load("data/generated/removal_stan_data_cv.rda")
-load("data/generated/distance_stan_data.rda")
+load("data/generated/distance_stan_data_cv.rda")
 cv_folds_rem <- read.csv("data/generated/removal_cv_folds.csv")
 cv_folds_dis <- read.csv("data/generated/distance_cv_folds.csv")
 phylo_tree <- ape::read.nexus(file = "data/raw/all_species.nex")
@@ -40,8 +39,8 @@ lppd <- data.frame(Species = cv_folds_rem$Species,
                    MS_LPPD = NA,
                    Difference = NA)
 
-for (f in 1:max(cv_folds_rem$cv_fold))
-{
+for (f in 1:max(cv_folds_rem$cv_fold)){
+  
   # Extract dataframe of phi from single species and multi species model
   ss_mod <- readRDS(paste0("output/model_runs/cv_removal/ms-vs-ss/ss_fold_",
                            f,

### 2024-05-06 src/scripts/post-hoc/4-species-cv-figure.R

@@ -3,7 +3,7 @@
 # Multi-species QPAD Detectability
 # posthoc/4-species-cv-figure.R
 # Created April 2024
-# Last Updated April 2024
+# Last Updated May 2024
 
 ####### Import Libraries and External Files #######
 
@@ -11,8 +11,6 @@ library(cmdstanr)
 library(ggplot2)
 library(ggpubr)
 theme_set(theme_pubclean())
-
-source("src/functions/subset-distance-data.R")
 
 ####### Set Constants #############################
 
@@ -29,7 +27,7 @@ for (i in 1:k)
 }
 rem_fold_nums <- read.csv("data/generated/removal_cv_folds_species.csv")
 
-load("data/generated/distance_stan_data.rda")
+load("data/generated/distance_stan_data_cv.rda")
 distance_all <- readRDS("output/model_runs/distance_ms.RDS")
 distance_folds <- vector(mode = "list", length = k)
 for (i in 1:k)
@@ -66,6 +64,7 @@ rem_cv_df$True_Exp <- exp(rem_cv_df$True)
 rem_cv_df$Predicted_Exp <- exp(rem_cv_df$Predicted)
 rem_cv_df$Difference <- rem_cv_df$True_Exp - rem_cv_df$Predicted_Exp
 rem_cv_df$Abs_Diff <- abs(rem_cv_df$Difference)
+rem_cv_df$Percent_Diff <- rem_cv_df$Difference / rem_cv_df$True_Exp
 
 removal_cv_plot <- ggplot(data = rem_cv_df, aes(x = True_Exp, y = Predicted_Exp)) +
   geom_point() +

### 2024-06-24 src/scripts/post-hoc/3-cv-figure.R
changing name of distance data

@@ -148,15 +148,6 @@ lppd_summary_removal <- lppd_summary
 
 ####### Distance Model ############################
 
-#' First we must drop all species which are just not going to be involved
-#' in the cross validation at all. This will be the species for which we are
-#' generated predictions
-pred_drops <- c("LCTH", "LEPC", "HASP", "SPOW", "KIWA", "BITH")
-dis_data <- subset_distance_data(distance_stan_data = distance_stan_data,
-                                 sps = setdiff(unique(distance_stan_data$sp_list),
-                                               pred_drops))
-rm(distance_stan_data)
-
 #' Create dataframe that will connect species, data point,
 #' and lppd
 lppd <- data.frame(Species = cv_folds_dis$Species,
@@ -180,11 +171,11 @@ for (f in 1:max(cv_folds_dis$cv_fold))
   
   # Subset the data to just the test set
   data_indices <- which(lppd$CV_Fold == f)
-  abund <- dis_data$abund_per_band[data_indices, ]
-  bands <- dis_data$bands_per_sample[data_indices]
-  max_dist <- dis_data$max_dist[data_indices, ] / 100
-  species <- dis_data$species[data_indices]
-  max_intervals <- dis_data$max_intervals
+  abund <- distance_stan_data_cv$abund_per_band[data_indices, ]
+  bands <- distance_stan_data_cv$bands_per_sample[data_indices]
+  max_dist <- distance_stan_data_cv$max_dist[data_indices, ] / 100
+  species <- distance_stan_data_cv$species[data_indices]
+  max_intervals <- distance_stan_data_cv$max_intervals
   
   tau_ss <- exp(ss_tau_summary$mean[species])
   tau_ms <- exp(ms_tau_summary$mean[species])
@@ -220,7 +211,7 @@ lppd_summary <- data.frame(Species = unique(lppd$Species),
                            StdErr = NA,
                            Model_Preference = NA)
 
-sp_sample_size <- data.frame(table(dis_data$sp_list))
+sp_sample_size <- data.frame(table(distance_stan_data_cv$sp_list))
 names(sp_sample_size) <- c("Species", "N")
 lppd_summary <- dplyr::left_join(lppd_summary, sp_sample_size, by = "Species")
 

### 2024-06-26 article/article.tex
renaming of posthoc scripts and final touches on figures

@@
 To determine which species would be left out in each fold of the distance model, we randomly assigned a number between 1 and 10 (drawn from a uniform distribution) to each species within a migratory resident $\times$ habitat preference, which ensured that there would be a good balance of species left out for each group of species.
-by either phylogeny and migratory status (for removal modelling) or by migratory status and habitat preference (for distance modelling).
-Because Stan allows draws from a posterior distribution even with no data, we drew values from the posterior distribution of all of the held out species for each cross-validation fold, and used mean and standard deviation of those draws as the predicted cue rate or EDR.
+Because Stan allows draws from a posterior distribution even with no data, we drew values from the posterior distribution of all the held out species for each cross-validation fold, and used mean and standard deviation of those draws as the predicted cue rate or EDR.
 We then modelled the difference in those predicted cue rates or EDRs to the "true" cue rates or EDRs obtained from the multi-species models in Objective 1.
+If our prediction is true, then we would expect to have predicted cue rates and EDRs be similar to true cue rates and EDRs obtained from modelling with data.
 
@@
 \par We found slight evidence for an overall change in cue rate for a given species modelled by a single-species model vs a multi-species model.
-The mean modelled difference in cue rate was $-1.57 \times 10^{-3}$, indicating that estimates of cue rate from a multi-species model may be slightly less than estimates of cue rate from a single-species model.
+The mean modelled difference in cue rate was $-1.56 \times 10^{-3}$, indicating that estimates of cue rate from a multi-species model may be slightly less than estimates of cue rate from a single-species model.
 Ten out of the 316 species considered for removal modelling had a percent change in cue rate of more than 10\%.
@@
 
-\par NEEDS UPDATING The cross validation revealed that the multi-species distance model also tended to have better predictive capacity for more species (156 out of 316) than the single-species distance models (140 out of 316), but to a lesser extent than the removal models.
-Twenty out of the 316 species showed no better predictive capacity with either model.
-Mean overall difference in log predictive density was $7.66 \times 10^{-10}$, indicating a small overall preference for the multi-species model.
-However, similar to the removal model cross-validation results, the overall difference in posterior predictive density was very small, with a mean difference of $1.85 \times 10^{-4}$.
+\par The k-fold cross validation revealed that the multi-species distance model also tended to have better predictive capacity for more species (166 out of 315) than the single-species distance models (130 out of 315), but to a lesser extent than the removal models.
+Nineteen out of the 316 species showed no better predictive capacity with either model.
+Mean overall difference in species-wise mean LPPD was $2.75 \times 10^{-5}$, indicating a small overall preference for the multi-species model.
 Similar to removal models, multi-species distance models tended to have better predictive capacity when sample size was low, with a negligible difference as sample size increased (Figure \ref{fig:cv}B).
-Among the seven species with greater than 20\% change in EDR, three species (Baird's Sparrow, Sedge Wren, and Sharp-tailed Grouse) had the multi-species model chosen as the better model, and four species (Boat-tailed Grackle, Chihuahuan Raven, Chimney Swift, and Ring-necked Pheasant) has the single-species model chosen as the better model.
+Among the two species with greater than 10\% change in EDR, Chihuahuan Raven had the single-species model chosen as the preferred model, and Greater Prairie-chicken had no preference.
 

### 2024-06-26 src/scripts/post-hoc/1-1vs1-figure.R

@@ -14,7 +14,7 @@ library(plyr)
 library(ggplot2)
 library(ggpubr)
 library(ggrepel)
-library(patchwork)
+#library(patchwork)
 theme_set(theme_pubclean())
 bayesplot::color_scheme_set("red")
 
@@ -54,8 +54,6 @@ for (i in 1:nrow(species_n))
     species_n$N[i]
 }
 
-####### 1-to-1 Plot ###############################
-
 to_plot <- data.frame(Species = rem_ms_summary$Code,
                       Single = rem_ss_summary$mean,
                       Multi = rem_ms_summary$mean,
@@ -74,20 +72,10 @@ for (i in 1:nrow(to_plot))
   }
 }
 
-(single_vs_multi_plot <- ggplot(data = to_plot, mapping = aes(x = Single, y = Multi)) +
-    geom_point() +
-    geom_abline(slope = 1) +
-    xlim(0,1) +
-    ylim(0,1) +
-    geom_text_repel(box.padding = 0.5, max.overlaps = Inf, aes(label = Label)) +
-    xlab("Cue Rate (Single Species)") + 
-    ylab("Cue Rate (Multi Species)") +
-    NULL)
-
 diff_model_data <- list(n_samples = nrow(to_plot),
-                        difference = to_plot$diff,
-                        N = to_plot$N)
-diff_model <- cmdstan_model(stan_file = "models/difference.stan")
+                        single = to_plot$Single,
+                        multi = to_plot$Multi)
+diff_model <- cmdstan_model(stan_file = "models/single-vs-multi.stan")
 
 diff_model_run <- diff_model$sample(
   data = diff_model_data,
@@ -98,15 +86,21 @@ diff_model_run <- diff_model$sample(
   refresh = 10
 )
 
-(modelled_difference_plot <- bayesplot::mcmc_areas(diff_model_run$draws(c("mu")),
-                                                   prob = 0.95) +
-    xlab("Modelled Difference"))
+rem_diff_model_draws <- diff_model_run$draws(format = "df")
 
-removal_plot <- single_vs_multi_plot + inset_element(modelled_difference_plot,
-                                                     left = 0.6, 
-                                                     bottom = 0, 
-                                                     right = 1, 
-                                                     top = 0.5)
+(removal_plot <- ggplot(data = to_plot, mapping = aes(x = Single, y = Multi)) +
[…]

### 2024-06-26 src/scripts/post-hoc/3-kfold-cv-figure.R.R

@@ -1,9 +1,9 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# posthoc/3-cv-figure.R
+# posthoc/3-kfold-cv-figure.R
 # Created December 2023
-# Last Updated May 2024
+# Last Updated June 2024
 
 ####### Import Libraries and External Files #######
 
@@ -127,6 +127,9 @@ preference <- data.frame(table(lppd_summary$Model_Preference))
 (lppd_diff_vs_n_rem <- ggplot(data = lppd_summary, aes(x = log(N), y = Mean_Difference)) +
     geom_point(aes(color = Model_Preference)) +
     geom_hline(yintercept = 0) +
+    xlab("Log(Sample Size)") +
+    ylab("Mean LPPD Difference") +
+    theme(legend.position = "none") +
     NULL)
 
 lppd_summary <- merge(lppd_summary, binomial[,c("Code", "Scientific_BT")],
@@ -244,6 +247,9 @@ preference <- data.frame(table(lppd_summary$Model_Preference))
 (lppd_diff_vs_n_dis <- ggplot(data = lppd_summary, aes(x = log(N), y = Mean_Difference)) +
     geom_point(aes(color = Model_Preference)) +
     geom_hline(yintercept = 0) +
+    xlab("Log(Sample Size)") +
+    ylab("Mean LPPD Difference") +
+    theme(legend.position = "none") +
     NULL
 )
 
@@ -254,12 +260,13 @@ write.table(lppd_summary_removal, file = "data/generated/lppd_summary_removal.cs
 write.table(lppd, file = "data/generated/lppd_distance.csv", sep = ",", row.names = FALSE)
 write.table(lppd_summary, file = "data/generated/lppd_summary_distance.csv", sep = ",", row.names = FALSE)
 
-tiff(filename = "output/plots/cv_plot.tiff",
+tiff(filename = "output/plots/kfold_cv_plot.tiff",
     width = 6, height = 6, units = "in", res = 600)
 ggarrange(lppd_diff_vs_n_rem, lppd_diff_vs_n_dis, nrow = 2, labels = c("A", "B"))
 dev.off()
 
-png(filename = "output/plots/cv_plot.png",
+png(filename = "output/plots/kfold_cv_plot.png",
      width = 6, height = 6, units = "in", res = 600)
 ggarrange(lppd_diff_vs_n_rem, lppd_diff_vs_n_dis, nrow = 2, labels = c("A", "B"))
 dev.off()
+

### 2024-06-26 src/scripts/post-hoc/4-species-cv-figure.R

@@ -3,7 +3,7 @@
 # Multi-species QPAD Detectability
 # posthoc/4-species-cv-figure.R
 # Created April 2024
-# Last Updated May 2024
+# Last Updated June 2024
 
 ####### Import Libraries and External Files #######
 
@@ -66,10 +66,29 @@ rem_cv_df$Difference <- rem_cv_df$True_Exp - rem_cv_df$Predicted_Exp
 rem_cv_df$Abs_Diff <- abs(rem_cv_df$Difference)
 rem_cv_df$Percent_Diff <- rem_cv_df$Difference / rem_cv_df$True_Exp
 
+rem_model_data <- list(n_samples = nrow(rem_cv_df),
+                       true_values = rem_cv_df$True_Exp,
+                       predicted_values = rem_cv_df$Predicted_Exp)
+
+rem_cv_model <- cmdstan_model(stan_file = "models/predicted-vs-true.stan")
+rem_cv_model_run <- rem_cv_model$sample(data = rem_model_data,
+                                          iter_warmup = 1000,
+                                          iter_sampling = 2000,
+                                          chains = 4,
+                                          parallel_chains = 4,
+                                          refresh = 10)
+rem_cv_model_draws <- rem_cv_model_run$draws(format = "df")
+
 removal_cv_plot <- ggplot(data = rem_cv_df, aes(x = True_Exp, y = Predicted_Exp)) +
+  geom_abline(intercept = rem_cv_model_draws$intercept, slope = rem_cv_model_draws$slope,
+              color = "grey", alpha = 0.1) +
+  geom_abline(intercept = mean(rem_cv_model_draws$intercept), slope = mean(rem_cv_model_draws$slope),
+              color = "black") +
+  geom_abline(slope = 1, color = "red", linetype = 2) +
   geom_point() +
-  geom_abline(slope = 1, intercept = 0) +
   xlim(0,0.8) + ylim(0,0.8) +
+  xlab("Cue Rate (Full Data)") +
+  ylab("Cue Rate (No Data)") +
   NULL
 
 ####### Distance Model #################################
@@ -101,16 +120,39 @@ dis_cv_df$Predicted_Exp <- exp(dis_cv_df$Predicted) * 100
 dis_cv_df$Difference <- dis_cv_df$True_Exp - dis_cv_df$Predicted_Exp
 dis_cv_df$Abs_Diff <- abs(dis_cv_df$Difference)
 
-dis_cv_df <- dis_cv_df[-which(is.na(dis_cv_df$Predicted)),]
[…]

### 2024-06-26 src/scripts/post-hoc/5-predictions-figure.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# posthoc/4-predictions-figure.R
+# posthoc/5-predictions-figure.R
 # Created December 2023
 # Last Updated April 2024
 
@@ -14,7 +14,6 @@ library(ggplot2)
 library(ggpubr)
 library(ggrepel)
 library(gginnards)
-library(patchwork)
 theme_set(theme_pubclean())
 bayesplot::color_scheme_set("red")
 
@@ -94,6 +93,7 @@ attributes(rem_draws)$dimnames$variable <- to_plot$Code
 (removal_plot <- bayesplot::mcmc_intervals(rem_draws) + 
     xlab("Predicted Cue Rate") + 
     ylab("Species") +
+    xlim(0,1)+
     coord_flip())
 
 removal_plot$layers <- c(geom_point(data = other_sp_df, aes(x = mean, y = Species, alpha = Alpha), size = 0.5, position=position_jitter(height=0.2)),
@@ -160,6 +160,7 @@ attributes(dis_draws)$dimnames$variable <- to_plot$Code
 (distance_plot <- bayesplot::mcmc_intervals(dis_draws) + 
     xlab("Predicted EDR") + 
     ylab("Species") +
+    xlim(0,600) +
     coord_flip())
 
 distance_plot$layers <- c(geom_point(data = other_sp_df, aes(x = mean, y = Species), size = 0.5, position=position_jitter(height=0.4)),

### 2024-06-26 src/scripts/removal/4-removal-prior-check.R
clean up models directory and removal scripts

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 3-removal-prior-check.R
+# 4-removal-prior-check.R
 # Created May 2023
 # Last Updated May 2023
 

### 2024-06-26 src/scripts/removal/5-removal-ms-vs-ss.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 4-removal-ms-vs-ss.R
+# 5-removal-ms-vs-ss.R
 # Created October 2023
 # Last Updated March 2024
 

### 2024-06-26 src/scripts/removal/6-removal-kfold-cv.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 5-removal-cv.R
+# 6-removal-kfold-cv.R
 # Created October 2023
 # Last Updated March 2024
 

### 2024-06-26 src/scripts/removal/7-removal-species-cv.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 5-removal-species-cv.R
+# 7-removal-species-cv.R
 # Created March 2024
 # Last Updated March 2024
 

### 2024-06-26 src/scripts/removal/8-removal-predictions.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 6-removal-predictions.R
+# 8-removal-predictions.R
 # Created December 2022
 # Last Updated April 2024
 

### 2024-06-26 src/scripts/distance/4-distance-prior-check.R
cleaning up distance scripts

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 3-distance-prior-check.R
+# 4-distance-prior-check.R
 # Created March 2023
 # Last Updated May 2023
 

### 2024-06-26 src/scripts/distance/5-distance-ms-vs-ss.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 4-distance-ms-vs-ss.R
+# 5-distance-ms-vs-ss.R
 # Created October 2023
 # Last Updated April 2024
 

### 2024-06-26 src/scripts/distance/6-distance-kfold-cv.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 5-distance-cv.R
+# 6-distance-kfold-cv.R
 # Created October 2023
 # Last Updated May 2024
 

### 2024-06-26 src/scripts/distance/7-distance-species-cv.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 5-distance-species-cv.R
+# 7-distance-species-cv.R
 # Created March 2024
 # Last Updated May 2024
 

### 2024-06-26 src/scripts/distance/8-distance-predictions.R

@@ -1,7 +1,7 @@
 ####### Script Information ########################
 # Brandon P.M. Edwards
 # Multi-species QPAD Detectability
-# 6-distance-predictions.R
+# 8-distance-predictions.R
 # Created October 2022
 # Last Updated April 2024
 

### 2024-06-27 src/scripts/post-hoc/1-1vs1-figure.R
adding figure that highlights conservative estimates of EDR

@@ -3,7 +3,7 @@
 # Multi-species QPAD Detectability
 # posthoc/1-1vs1-figure.R
 # Created December 2023
-# Last Updated May 2024
+# Last Updated June 2024
 
 ####### Import Libraries and External Files #######
 

### 2024-06-27 src/scripts/post-hoc/4-species-cv-figure.R

@@ -36,6 +36,8 @@ for (i in 1:k)
 }
 dis_fold_nums <- read.csv("data/generated/distance_cv_folds_species.csv")
 
+traits <- read.csv("data/raw/traits.csv")
+
 ####### Removal Model #################################
 
 phi_all <- removal_all$summary("log_phi")
@@ -78,14 +80,15 @@ rem_cv_model_run <- rem_cv_model$sample(data = rem_model_data,
                                           parallel_chains = 4,
                                           refresh = 10)
 rem_cv_model_draws <- rem_cv_model_run$draws(format = "df")
+rem_cv_model_summary <- rem_cv_model_run$summary()
 
 removal_cv_plot <- ggplot(data = rem_cv_df, aes(x = True_Exp, y = Predicted_Exp)) +
   geom_abline(intercept = rem_cv_model_draws$intercept, slope = rem_cv_model_draws$slope,
               color = "grey", alpha = 0.1) +
   geom_abline(intercept = mean(rem_cv_model_draws$intercept), slope = mean(rem_cv_model_draws$slope),
               color = "black") +
-  geom_abline(slope = 1, color = "red", linetype = 2) +
   geom_point() +
+  geom_abline(slope = 1, color = "red", linetype = 2) +
   xlim(0,0.8) + ylim(0,0.8) +
   xlab("Cue Rate (Full Data)") +
   ylab("Cue Rate (No Data)") +
@@ -132,20 +135,48 @@ dis_cv_model_run <- dis_cv_model$sample(data = dis_model_data,
                                         parallel_chains = 4,
                                         refresh = 10)
 dis_cv_model_draws <- dis_cv_model_run$draws(format = "df")
+dis_cv_model_summary <- dis_cv_model_run$summary()
 
 distance_cv_plot <- ggplot(data = dis_cv_df, aes(x = True_Exp, y = Predicted_Exp)) +
   geom_abline(intercept = dis_cv_model_draws$intercept, slope = dis_cv_model_draws$slope,
               color = "grey", alpha = 0.1) +
   geom_abline(intercept = mean(dis_cv_model_draws$intercept), slope = mean(dis_cv_model_draws$slope),
               color = "black") +
-  geom_abline(slope = 1, color = "red", linetype = 2) +
   geom_point() +
+  geom_abline(slope = 1, color = "red", linetype = 2) +
   xlim(0,500) + ylim(0,500) +
[…]

### 2024-06-27 article/article.tex
Final touches to draft 3 prior to sending out to coauthors

@@
 	Here, we take advantage of similarities in detection probabilities among phylogenetically related species and species with similar traits to estimate detection probabilities in seven species of rare North American birds, using a hierarchical Bayesian multi-species removal and distance models.
-	For two species, Bicknell’s Thrush and Kirtland’s Warbler, we demonstrate our multi-species model’s ability to predict detection probabilities given little to no observation data, and compare these predictions to previous estimates found in the literature.
-	We also demonstrate the multi-species models’ ability to provide more precise estimates of detectability for species where there is sufficient data. 
-	We recommend that this multi-species model be combined with survey-level covariates to augment previous estimates of detection probabilities in the NA-POPS database of detectability offsets for North American birds.
+	We first show that a multi-species model of detectability can improve precision of detectability estimates for species with low sample size, and even with species up to 1000 data points.
+	However, when there is no data available for a species, the multi-species models presented here tend to be overly conservative in predicting detectability.
+	Thus, we recommend using these models to 1) provide more precise estimates of detectability for species where there are sufficient data, 2) generate estimates of detectability for species for which we have sparse (but non-zero) data, and 3) augment previous estimates of detection probabilities in the NA-POPS database of detectability offsets for North American birds by combining this model with survey-level covariates.
 
@@
 When estimating population size, detectability often comes into play as a way to convert observed counts into densities or estimates of true abundance \citep{solymos_calibrating_2013, johnson_defense_2008}.
-Despite its importance, detectability often goes unaccounted for in studies; and for many species, the ancillary data needed to estimate detectability does not exist \citep{bennett_how_2023}.
+Despite its importance, detectability often goes unaccounted for in studies; and for many species, the ancillary data needed to estimate detectability does not exist \citep{bennett_how_2024}.
 
-\par The wealth of openly available bird observation data in North America means that estimates of detectability for birds are relatively easy to estimate, in comparison to most other taxonomic groups \citep{bennett_how_2023}.
+\par The wealth of openly available bird observation data in North America means that estimates of detectability for birds are relatively easy to estimate, in comparison to most other taxonomic groups \citep{bennett_how_2024}.
 The NA-POPS project is an effort to estimate detection probabilities that account for variability in detection processes through environmental and temporal covariates for as many of North America’s landbird species as possible \citep{edwards_point_2023}.
@@
 This project expanded previous efforts by the Boreal Avian Modelling project \citep{cumming_toward_2010} to systematically estimate detectability in landbirds \citep{solymos_calibrating_2013, solymos_evaluating_2018}.
-Both the NA-POPS project and previous estimates by Boreal Avian Modelling project use QPAD \citep{solymos_calibrating_2013}, which is a flexible method to estimate availability and perceptibility given a harmonized dataset with the necessary survey-specific data (i.e., distance to bird from observer and time to first detection; \citet{barker_ecological_2015}.
+Both the NA-POPS project and previous estimates by Boreal Avian Modelling project use QPAD \citep{solymos_calibrating_2013}, which is a flexible method to estimate availability and perceptibility given a harmonized dataset with the necessary survey-specific data (i.e., distance to bird from observer and time to first detection; \citep{barker_ecological_2015}).
 
@@
 These species fall short of the threshold of 75 distinct observations recommended for removal and distance modelling \citep{buckland_introduction_2001, solymos_calibrating_2013}. 
-Many of the species that fell short of that threshold in the NA-POPS database are rare or undersampled species, that are either have low population sizes, or occur in rare or inaccessible habitats.
+Many of the species that fell short of that threshold in the NA-POPS database are rare or undersampled species, that either have low population sizes, or occur in rare or inaccessible habitats.
 %Some examples of species not modelled include Kirtland’s Warbler (\textit{Setophaga kirtlandii}) and Bicknell’s Thrush (\textit{Catharus bicknelli}), both of which have low population sizes, and LeConte’s Thrasher (\textit{Toxostoma lecontei}) and Harris’s Sparrow (\textit{Zonotrichia querula}), both of which have rare habitat associations \citep{will_handbook_2020}.
@@
 \par The following subsections outline the modelling techniques used in this study as well as the data that were collected. 
-We start by introducing the Bayesian hierarchical QPAD removal and distance models, and then describe the input.
+We start by describing the data that will be used in the models, and then introduce the Bayesian hierarchical QPAD removal and distance models.
 Finally, we describe the experiments conducted to assess model performance and model comparison, which then allow us to make predictions for the seven undersampled species of interest.
@@
 
-\par We generated a phylogenetic correlation matrix $\mathbf{D}$ using the same methods outlined in the Appendix of \citet{solymos_phylogeny_2018} that generate distances between species, and further processed using the R package "ape" \citet{paradis_ape_2019}.
+\par We generated a phylogenetic correlation matrix $\mathbf{C}$ using the same methods outlined in the Appendix of \citet{solymos_phylogeny_2018} that generate distances between species, and further processed using the R package "ape" \citet{paradis_ape_2019}.
 
@@
 \par The covariance function $\mathbf{\Sigma}$ (denoted with subscript $[s \times s]$ to show dimensionality) determines the strength of the partial pooling, and can be any function that contains a measure of distance.
-In this case, we will create a function based on phylogenetic distance $\mathbf{D}_{[s \times s]}$, i.e.,
+In this case, we will create a function based on phylogenetic distance $\mathbf{C}_{[s \times s]}$, i.e.,
 
-$$\mathbf{\Sigma}_{[s \times s]} = \mathbf{D}_{[s \times s]}\lambda\sigma.$$
+$$\mathbf{\Sigma}_{[s \times s]} = \mathbf{C}_{[s \times s]}\lambda\sigma.$$
 
@@
 \par To test our prediction that there would be little change between estimates of species with sufficient sizes, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data.
-In this case, we ran both models for 316 species of birds.
 Rufous Hummingbird (\textit{Selasphorus rufus}) was dropped from the distance modelling due to issues with divergent transitions \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
@@
 We also ran the single species removal and distance models for the same suite of birds.
-We then subtracted the estimate of the single-species cue rate or EDR from the estimate of the multi-species cue rate or EDR, and modelled the difference in estimated cue rates and EDRs $\Delta X$ as
-$$\Delta X \sim N(0,\sigma)$$
-$$\sigma \sim exp(1)$$
-
-If our prediction is true, we would expect to see a distribution of modelled differences centred about 0, indicating no difference in estimates of cue rate or EDR between single-species and multi-species models, for species for which we have sufficient data.
+We then modelled the multi species estimates of cue rate and EDR as functions of the single species estimates of cue rate and EDR, and obtained estimates of slope and y-intercept for this relationship.
+If our prediction is true, we would expect to see a slope of 1 and a y-intercept of 0, indicating a 1-to-1 relationship.
 
@@
 Because Stan allows draws from a posterior distribution even with no data, we drew values from the posterior distribution of all the held out species for each cross-validation fold, and used mean and standard deviation of those draws as the predicted cue rate or EDR.
-We then modelled the difference in those predicted cue rates or EDRs to the "true" cue rates or EDRs obtained from the multi-species models in Objective 1.
-If our prediction is true, then we would expect to have predicted cue rates and EDRs be similar to true cue rates and EDRs obtained from modelling with data.
+We then modelled the predictions of cue rate and EDR obtained with no data as a function of cue rate and EDR obtained with the full data, and obtained estimates of slope and y-intercept for this relationship.
+If our prediction is true, then we would expect to have a slope of 1 and a y-intercept of 0, indicating a 1-to-1 relationship.
 
@@
 Availability $p(t)$ is defined as the probability of a bird giving a cue during a survey run for $t$ minutes, and is calculated as
-$p(t) = 1 - e^{-t\phi}$
+
+\begin{equation}
+	p(t) = 1 - e^{-t\phi}
+\end{equation}
 where $\phi$ is the cue rate of the species.
 Perceptibility $q(r)$ is defined as the probability of an observer hearing or seeing a bird, given the bird gives a cue, with a survey radius of $r$ metres, and is calculated as
-$q(r) = \dfrac{\tau^2 \left\{1 - e^{-\dfrac{r^2}{\tau^2}}\right\}}{r^2}$
+\begin{equation}
+	q(r) = \dfrac{\tau^2 \left\{1 - e^{-\dfrac{r^2}{\tau^2}}\right\}}{r^2}
+\end{equation}
+
 where $\tau$ is the EDR of the species.
@@
 
-\par All code is available open-source at https://github.com/BrandonEdwards/multispecies-qpad-detectability and will be archived with a DOI upon acceptance of this paper.
+\par Code is available open-source at https://github.com/BrandonEdwards/multispecies-qpad-detectability and will be archived with a DOI upon acceptance of this paper.
 
@@
 
-\par We found slight evidence for an overall change in cue rate for a given species modelled by a single-species model vs a multi-species model.
-The mean modelled difference in cue rate was $-1.56 \times 10^{-3}$, indicating that estimates of cue rate from a multi-species model may be slightly less than estimates of cue rate from a single-species model.
+\par Estimates of cue rate obtained from a multi-species model closely followed estimates of cue rate obtained from a single-species model (y-intercept = $8.47 \times 10^{-4}$, slope = $0.99$; Figure \ref{fig:1vs1}A).
 Ten out of the 316 species considered for removal modelling had a percent change in cue rate of more than 10\%.
-These species were: Willow Ptarmigan (\textit{Lagopus lagopus}), Rusty Blackbird (\textit{Euphagus carolinus}), Red-bellied Sapsucker (\textit{Sphyrapicus ruber}), Prairie Falcon (\textit{Falco mexicanus}), Nuttall's Woodpecker (\textit{Picoides nuttallii}), Northern Goshawk (\textit{Accipiter atricapillus}), Ferruginous Hawk (\textit{Buteo regalis}), Common Redpoll (\textit{Acanthis flammea}), Calliope Hummingbird (\textit{Selasphorus calliope}), and Bohemian Waxwing (\textit{Bombycilla garrulus})(Figure \ref{fig:1vs1}A).
+These species were: Willow Ptarmigan (\textit{Lagopus lagopus}), Rusty Blackbird (\textit{Euphagus carolinus}), Red-bellied Sapsucker (\textit{Sphyrapicus ruber}), Prairie Falcon (\textit{Falco mexicanus}), Nuttall's Woodpecker (\textit{Picoides nuttallii}), Northern Goshawk (\textit{Accipiter atricapillus}), Ferruginous Hawk (\textit{Buteo regalis}), Common Redpoll (\textit{Acanthis flammea}), Calliope Hummingbird (\textit{Selasphorus calliope}), and Bohemian Waxwing (\textit{Bombycilla garrulus}).
 
-\par We found little evidence for a change in modelled EDR between single species and multi-species distance models. 
-The mean modelled difference in EDR was 0.97, but the 90\% credible interval overlapped zero.
-Two out of the 315 species had a percent change in EDR of more than 10\%.
-These species were Greater Prairie-chicken (\textit{Tympanuchus cupido}) and Chihuahuan Raven (\textit{Corvus cryptoleucus}) (Figure \ref{fig:1vs1}B).
+\par Estimates of EDR obtained from a multi-species model also closely followed estimates of EDR obtained from a single-species model. 
+With all species included, we obtained a y-intercept of $6.39$ and a slope of $0.91$.
+However, two out of the 315 species had a percent change in EDR of more than 10\%: Greater Prairie-chicken (\textit{Tympanuchus cupido}) and Chihuahuan Raven (\textit{Corvus cryptoleucus}) (Figure \ref{fig:1vs1}B).
+These species were particularly influential in this regression; after removing these two species, we obtained a y-intercept of $1.25$ and a slope of $0.98$, indicating a near 1-to-1 relationship for the remaining species.
 
@@
 
-\subsection{Predictions of Cue Rate and EDR for Undersampled Species}
+\par Out-of-sample predictions of cue rate tended to be underestimated compared to estimates obtained with full data (y-intercept = $0.17$, slope = $0.34$; Figure \ref{fig:species_cv}A).
+Interestingly, several of the species with the largest magnitude of difference in cue rate tended to be species associated with grasslands (e.g., Greater Prairie-chicken, predicted $-$ true $= -0.52$; Sprague's Pipit, predicted $-$ true $= -0.49$; Western Meadowlark, predicted $-$ true $= -0.36$, among several others).
 
-\par We used the multi-species removal model and distance model to generate predictions of cue rate and EDR, respectively, for seven species of undersampled birds from the original NA-POPS analysis (6 for distance modelling because Tricoloured Blackbird had sufficient distance sampling data).
+\par Out-of-sample predictions of EDR also tended to be underestimated compared to estimates obtained with full data (y-intercept = $4.46$, slope = $0.72$; Figure \ref{fig:species_cv}B).
+Of particular note was the fact that the model did not generate any out-of-sample predictions greater than 115 m, nor less than 69 m.
+In fact, the model was particularly conservative in generating estimates for each of the migrant x habitat groups, in that there was very little variation within each of the groups; any variation that did occur would have been due to differences in song pitch and body size (Figure \ref{fig:distance_cv_traits}).
 
-\par The multi-species removal model produced reasonable estimates of cue rates for all seven species considered here (Figure \ref{fig:predictions}A).
-Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush and Kirtland's Warbler) had reasonable precision associated with the estimates.
-One exception to this was LeConte's Thrasher, where the precision was rather low.
-The species for which we had no data (i.e., Bicknell's Thrush and Kirtland's Warbler) had estimates of cue rate similar to that of their closely-related species, demonstrating the shrinkage effect of the multi-species model.
+\subsection{Case Study: Predictions of Cue Rate and EDR for Undersampled Species}
 
-\par The multi-species distance model also produced reasonable estimates of EDR for all six species considered here (Figure \ref{fig:predictions}B).
-Species where the sample size requirements were relaxed (i.e., all except Bicknell's Thrush) had fairly precise estimates, except for LeConte's Thrasher and Lesser Prairie Chicken where the estimated EDR spanned greater than 50 m in either direction.
-The species for which we had no data (i.e., Bicknell's Thrush) had an estimate of EDR similar to what would be expected by a bird of its size, song pitch, and habitat association, demonstrating the shrinkage effect of the multi-species model.
-
-\section{Discussion}
-
-\par In this paper, we developed a multi-species model of detectability by expanding upon the QPAD methodology developed by \citet{solymos_calibrating_2013}.
-We applied the multi-species QPAD removal model and multi-species QPAD distance model to the large dataset of point counts collected by the NA-POPS project \citep{edwards_point_2023}, and compared estimates of cue rate and EDR from a multi-species context to estimates of cue rate and EDR from the previous single species context.
-Given the results of these comparisons, we predicted these components of detectability for seven species of North American landbird that are rare and undersampled.
-
-\par In both cases of QPAD removal modelling and QPAD distance modelling, we were able to generate estimates of detectability for species with no data, and precise estimates of detectability for data-sparse species.
-Overall, the shrinkage effect offered by the multi-species modelling approach allowed for more accurate estimates of cue rate and EDR to be derived in species with smaller sample sizes, without affected estimates in species with large sample sizes.
-The best model for predictive accuracy tended to depend on the species, but overall differences in the estimates were small enough that using a multi-species model should not produce extremely inaccurate estimates.
-
-\subsection{Predicting Detectability for Data-sparse Species}
-
-\par For five out of the seven species for which we generated new predictions of detectability in this study, we were interested in seeing how a multi-species QPAD removal model could estimate cue rate with few data, and how a multi-species QPAD distance model could estimate EDR with few data.
-
-\par Overall, the estimates of cue rate fell roughly where we would expect based on related species, for species for which we were simply relaxing the sample size requirements.
-For example, Harris's Sparrow (n = 65) had an estimated cue rate of 0.30 cues per minute, with a 95\% credible interval of [0.21, 0.42], compared to the overall mean cue rate of 0.37 cues per minute for all 32 other Passerellidae species in NA-POPS.
-Similarly, LeConte's Thrasher (n = 1) had an estimated cue rate of 0.32 cues per minute, with a 95\% credible interval of [0.18, 0.57].
-Though this was a much larger interval than Harris's Sparrow, it still falls within a reasonable distance from the overall mean cue rate of 0.36 cues per minute for all 6 other Mimidae species in NA-POPS.
-
-\par Estimates of EDR also fell roughly where we would expect based on species of similar traits, for the species for which we were relaxing samples sizes.
-For example, Harris's Sparrow (n = 13) had an estimated mean EDR of 65.6 m, with a 95\% credible interval of [53.1, 82.5].
-This range does fall below the mean EDR of 85.3 m for the other sparrows in NA-POPS, but Harris's Sparrow song pitch of approximately 3.8 Hz falls above the average song pitch of 3.0 Hz for the other sparrows in NA-POPS, which would contribute to the smaller EDR.
-On the other hand, LeConte's Thrasher (n = 1) had an estimated EDR of 134.2 m, with a 95\% credible interval of [81.8, 230.5], which falls roughly in line with the mean EDR of 113.2 m of the six other thrashers in NA-POPS.
-
-\par It is likely that the width of the credible intervals (i.e., the precision) of these relaxed sample-size species will depend on the number of data points available for that particular species, the number of "similar" species, and the number of data points available for those similar species.
-Additionally, it is important to note that we did not account for other sources of variation at the survey level, such as when the survey was conducted (i.e., time of day, time of year), or where the survey was conducted (e.g., in a forest, by a roadside).
-It is well known that the ability to detect a species can be very dependent on the time of day or year that an observer is surveying, as well as the environmental conditions of the survey \citep{edwards_point_2023}.
-Thus, some of the variation in these estimates of cue rate and EDR could be due to not accounting for survey-level variation that affects detectability, and so it is important that future iterations of this model incorporate both species-level effects on detectability \textit{and} survey-level effects on detectability.
-
-\subsection{Predicting Detectability for Species with No Data}
-
-\par We generated predictions of cue rate and EDR for both Bicknell’s Thrush and Kirtland’s Warbler.
-For both species, we had zero removal sampling data, and so cue rates were solely based on information sharing among phylogeny (and to a much lesser extent, migration status). 
-For Kirtland’s Warbler, we did have one data point for distance sampling, so predictions of EDR for Kirtland’s Warbler here are not entirely based on zero data points.
-Bicknell’s Thrush, however, had zero data points, and so EDRs for this species were based solely on information sharing among other species with similar traits.
+\par We used the multi-species removal model and distance model to generate predictions of cue rate and EDR, respectively, for seven species of undersampled birds from the original NA-POPS analysis (six for distance modelling because Tricolored Blackbird had sufficient distance sampling data; Figure \ref{fig:predictions}).
 
@@
 \par For distance modelling, the predicted EDR for Bicknell's Thrush (73.6 m) and the predicted EDR for Kirtland's Warbler (75.3 m) fell reasonably close to the overall mean EDR of 92.7 m for all other 11 Turdidae species and 61.2 m for all 44 other Parulidae species in NA-POPS, respectively.
-That is, in the absence of any data (or with very little data in the case of Kirtland’s Warbler), the estimated EDR of Bicknell’s Thrush and Kirtland’s Warbler was similar to those of other Thrushes and Warblers, respectively, but with departures from overall mean based on specific traits.
+That is, in the absence of any data (or with very little data in the case of Kirtland’s Warbler), the estimated EDR of Bicknell’s Thrush and Kirtland’s Warbler was similar to those of other Thrushes and Warblers, respectively, but with small departures from overall mean based on specific traits.
 
-\par In both the removal modelling and distance modelling, the magnitude of contribution of data from related species appeared to play a role in the width of the credible intervals. 
-For removal modelling, although both credible intervals were fairly wide, Bicknell’s Thrush had a 95\% credible interval of [0.16, 0.48], representing an approximate 53\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [0.18, 0.48], representing an approximate 49\% change on either end.
-For distance modelling, both credible intervals were also fairly wide, but Bicknell's Thrush had a 95\% credible interval of [40.5, 134.4], representing an approximate 58\% change on either end, whereas Kirtland's Warbler had a 95\% credible interval of [45.9, 125.6], representing an approximate 47\% change on either end.
-Although the difference in width of the intervals is small, this could be due to the greater number of other species contributing information to Kirtland's Warbler estimates than to Bicknell's Thrush.
-In the case of distance modelling, the one data point contributing to Kirtland's Warbler's EDR could have also played a part.
-With that in mind, it would likely be important to consider how many "related" species are available with information (and how many data points do each of those related species have) when deciding on making a prediction of cue rate.
-
-\subsection{Comparisons of Predicted Detectability to Other Studies}
-
-\par Using the predictions of cue rate and EDR for Bicknell’s Thrush and Kirtland’s Warbler, we can calculate availability and perceptibility, and hence detectability, for a given survey type, and compare to previous estimates of detectability obtained from other studies.
-
-\par For Bicknell's Thrush, using the equations above, the predicted cue rate of 0.28, predicted EDR of 73.3 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.34.
+\par For Bicknell's Thrush, using Equations 1 and 2, the predicted cue rate of 0.28, predicted EDR of 73.3 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.34.
 We can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respectively credible intervals) to get a range of detectabilities from 0.09 through 0.70.
@@
 
-\par For Kirtland's Warbler, using the equations above, the predicted cue rate of 0.29, predicted EDR of 75.3 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.36.
+\par For Kirtland's Warbler, Equations 1 and 2, the predicted cue rate of 0.29, predicted EDR of 75.3 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.36.
 Similar to our Bicknell's Thrush example, we can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respective credible intervals) to obtain a range of detectabilities from 0.12 through 0.67.
@@
 
+\section{Discussion}
+
+\par In this paper, we developed a multi-species model of detectability that accounts for phylogeny and traits, expanding upon the work from \citet{solymos_phylogeny_2018}.
+We applied the multi-species removal model and multi-species distance model to the large dataset of point counts collected by the NA-POPS project \citep{edwards_point_2023}, and compared estimates of cue rate and EDR from a multi-species context to estimates of cue rate and EDR from a single species context.
+We also assessed the multi-species models' ability to generate estimates of cue rate and EDR for species where we have few data points, and for species for which we have no data points.
+Finally, we examined a case study of seven undersampled species (in the NA-POPS database) to estimate cue rate and EDR using this multi-species model.
+
+\par Overall, the multi-species models generated very similar estimates of cue rate and EDR as single-species models, for species where we had sufficient data points.
+That is, we did not see too many significant changes in cue rate and EDR by modelling with a multi-species model versus modelling with a single-species model.
+This provided an appropriate baseline to continue our comparisons for data-sparse species.
+We saw a few exceptions in the distance modelling, in that estimates of EDR for Greater Prairie-chicken and Chihuahuan Raven in a multi-species context were much lower than that estimated from a single-species context.
+We note that these species are generally observed either in flight in highly specific regions (for Chihuahuan Raven) or in wide-open landscapes (for both). 
+Thus, it is possible that data for these species are biased high, such that a multi-species model brings the estimates down to a more realistic value.
+
+\par We did not see a large difference in precision of estimates between the single-species and multi-species model, except for a slight difference in cue rate estimates for sample sizes less than 1000.
+However, we note that our baseline single-species model was a Bayesian implementation of the QPAD models described in \citet{solymos_calibrating_2013}, and we used weakly informative $N(0,1)$ priors on the parameters for cue rate and EDR.
+In earlier versions of this study, we compared the precision of multi-species estimates of cue rate and EDR to the precision of the estimates obtained using maximum likelihood methods in \citet{edwards_point_2023}, and found a more pronounced increase in precision in the multi-species estimates.
+Therefore, while the multi-species implementation can somewhere improve precision of estimates for lower sample size, the use of even weakly-informed priors clearly goes a long way in improving precision.
+We therefore recommend future implementations of the NA-POPS detectability database to adopt a Bayesian modelling approach using at least weakly-informed priors.
+
+\subsection{Predicting Detectability for Data-sparse Species}
+
+\par In both cases of multi-species removal modelling and multi-species distance modelling, the k-fold cross validation showed that species with a smaller sample size tended to have better fitted estimates of cue rate and EDR from a multi-species model than from a single-species model.
+That is, for species with small sample size, a multi-species model that considers phylogeny (for cue rate) and traits (for EDR) should be able to provide reasonably accurate estimates of cue rate and EDR.
+
+\subsection{Predicting Detectability for Species with No Data}
+
+\par Interestingly, these multi-species models were not able to provide good estimates of cue rate or EDR for species where we had no data, as evidenced by our leave-species-out cross validation.
+For both multi-species removal and distance modelling, the predicted cue rates or EDRs for species that were left out of a given cross-validation fold were, on average, underestimated compared to predicted cue rates for EDRs for the same species with all its data.
+
+\par We noted that for removal modelling, the group of species that had severe underestimates in the predicted cue rate were almost all grassland birds.
+One potential mechanism behind this is how birds are recorded in removal sampling, in that birds are recorded in the first time bin that they are observed, and then "removed" from the available birds to sample from for the remainder of the survey.
+In grassland environments, where visual sightings are potentially more likely (and combined with less sound attenuation from vegetation), it is likely that the majority of these birds are being recorded in the first time bin, potentially inflating the cue rate when data is available.
+Then, when the species is modelled without the data, the cue rate is shrunk toward "similar" species that may occur in both open and closed environments, therefore shrinking the cue rate down to a lower value.
+Future studies could seek to disentangle the possible effects of habitat on cue rate; to go one step further, future studies should build upon the work of \citet{martin-schwarze_joint_2021} to assess how truly independent cue rate and EDR are.
+
+\par When considering EDR predictions, we noted that the multi-species models were not able to capture the variation of EDRs within the different trait groups.
+This was evidenced by the predicted EDRs only ever being between 69 - 115 m.
+When investigating further, Figure \ref{fig:distance_cv_traits} demonstrated a fairly clean deliniation of these different trait groupings, with little variation within each.
+With this in mind, while studies have shown that traits are useful for explaining variation in EDR, traits alone do not appear to be sufficient in predicting EDRs in similar birds, because traits cannot account for species-specific variation in EDR.
+Future studies should seek to continue understanding drivers and predictors of detection distance in birds, both related to and complementary to traits.
+
+\subsection{Opportunities to Expand Estimates of Detectability in North American Landbirds}
+
+\par Given the results of the k-fold cross-validation, we were comfortable with applying this model to species where we had only a few data points, such as Harris's Sparrow, Lesser Prairie-chicken, Spotted Owl, LeConte's Thrasher, and Tricolored Blackbird.
+We were able to show that if we had at least some data points for a given species, the fitted model with either phylogeny and migratory status for cue rate, or traits for EDR, can help provide additional information for these data sparse species.
+As such, we were able to generate estimates of cue rate and EDR for these data sparse species.
+The models presented here can be used to fill out the remaining missing species in the NA-POPS database, in order to estimate detectability for every species of North American landbird. 
+By making use of phylogeny and trait information, bird species with few data points can have estimates generated based on the few data points available, and information available from other species.
+
+\par We caution, however, the use of these models for species with no data points, given the result of the leave-species-out cross validation.
+For Bicknell's Thrush and Kirtland's Warbler, both estimates of detectability appeared to be underestimated compared to estimates found in the literature, which corroborates the consistent underestimation found from the cross-validation.
+Of course, in the absence of any detectability information at all, our models can at least provide a starting point for a researcher to consider, which can then be turned into ranges of potential detectability \citep{bennett_how_2024}.
+Regardless, we recommend continuing to make use of existing available data collected by previous studies on these data-sparse species prior to using these models to further bolster predictions.
+
+In both cases above, we also caution that detectability predictions for species with few data (e.g., < 1000 observations) should be identified as such, so that users know that the estimate depends on data from other similar species.
+
 \subsection{Model Performance}
 
-\par The multi-species QPAD Distance model had several issues related to initialization, sampling, and convergence throughout this study.
+\par The multi-species QPAD distance model had several issues related to initialization, sampling, and convergence throughout this study.
 For many iterations of the model development, generation of initial values using Stan proved difficult.
@@
 At the low end, an initial value of -2 would correspond to a detection distance of $e^{-2} \times 100 = 13.5$ m, which is low but somewhat plausible for a small-bodied, high-pitched species such as a hummingbird.
-At the high end, an initial value of 2 would correspond to a detection distance of $e^{2} \times 100 = 739$ m, which is high but still somewhat plausible for large-bodied, low-pitched species like a grouse or prairie-chickens.
-However, where Stan seemed to struggle is when these randomly assigned values for detection distances were effectively impossible for a large set of species.
-That is to say, if the values were initialized such that hummingbirds had a detection distance of 500 m or more, and grouse or prairie-chickens had a detection distance of 15 m or less, this would result in an impossible likelihood, forcing Stan to reject the initial values.
+At the high end, an initial value of 2 would correspond to a detection distance of $e^{2} \times 100 = 739$ m, which is high but still somewhat plausible for large-bodied, low-pitched species.
+However, the models would fail to start when these randomly assigned values for detection distances were effectively impossible for a large set of species.
+If the values were initialized such that hummingbirds had a detection distance of 500 m or more, and grouse or prairie-chickens had a detection distance of 15 m or less, this would result in an impossible likelihood, forcing Stan to reject the initial values.
 Our fix was to set initial values to be jittered mean detection distances originally estimated in \citet{edwards_point_2023}; for species without estimates, we used the mean EDR for all species.
-This approach seemed to work as the initial values were no longer rejected, but we caution that this could very well be a "bandaid solution" to an underlying issue with the model itself.
+This approach seemed to work as the initial values were no longer rejected, but we caution that this could very well be a "band-aid solution" to an underlying issue with the model itself.
 
-\par The distance models also seemed to struggle with sampling the posterior distribution, in that over a quarter of the iterations resulted in a divergent transition \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
+\par In early implementations of this study, the distance model struggled with sampling the posterior distribution, in that over a quarter of the iterations resulted in a divergent transition \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
 However, after several iterations of this model, we identified that the samplers were not appropriately sampling Rufous Hummingbird's EDR, in that there appeared to be a self-imposed barrier in being able to sample EDRs less than 25 m.
@@
 Thus, while we can generate reasonable estimates of the components of detectability (i.e., availability through cue rate and perceptibility through EDR) based on information borrowed from similar species (either through phylogeny or traits), this model as it stands may fail to glean more specific information that takes into account detectability peaks throughout the year, or detectability peaks given specific habitats.
-With this in mind, it may be beneficial to include temporal covariates at the survey level, similar to what was done in \citet{solymos_calibrating_2013, solymos_evaluating_2018, edwards_point_2023}, so that further information about cue rate may be shared among similar species giving cues at similar times of day or year.
-Similarly, it may be beneficial to include environmental covariates at the survey level, similar to what was done in \citet{solymos_calibrating_2013, edwards_point_2023}, so that further information about EDR may be shared among similar species being surveyed at similar habitats.
+With this in mind, it may be beneficial to include temporal covariates at the survey level, following \citet{solymos_calibrating_2013, solymos_evaluating_2018, edwards_point_2023}, so that further information about cue rate may be shared among similar species giving cues at similar times of day or year.
+It may also be beneficial to include environmental covariates at the survey level \citep{solymos_calibrating_2013, edwards_point_2023}, so that further information about EDR may be shared among similar species being surveyed at similar habitats.
 This could also provide an opportunity to disentangle species by survey interactions in detectability.
 
-\par The examples above additionally highlight an excellent opportunity to use a powerful feature of Bayesian models, which is the inclusion of expert knowledge as prior information. 
+\par The examples above additionally highlight an excellent opportunity to use a powerful capability of Bayesian models, which is the inclusion of expert knowledge as prior information. 
 For example, if a Bicknell’s Thrush expert knows that there is something particular about how often they sing or give a cue compared to other thrushes, or how disproportionately quiet or loud they are compared to species with similar traits, this can be coded directly into the prior distribution.
-Similarly, this may also provide a unique opportunity to use traditional ecological knowledge to further inform predictions of detectability.
+Similarly, this may also provide a unique opportunity to use traditional ecological knowledge to further inform predictions of detectability \citep{wardfear_sharper_2019}.
 For example, in the Anishinaabe language, bird taxonomy is based not only on physical features, but also behaviours such as flight, calls, morning rituals, and weather \citep{pitawanakwat_evening_2022}.
@@
 \subsection{Conservation Implications}
-\par Detectability frequently goes unaccounted for in conservation problems, often because the estimates of detection probability simply do not exist, or they are extremely uncertain \citep{bennett_how_2023}.
-This model provides an excellent opportunity for detection probabilities to be estimated with higher precision for several species of landbirds in North America, and also provides an opportunity to predict detection probabilities for species which have very little (or even no) data.
+\par Detectability frequently goes unaccounted for in conservation problems, often because the estimates of detection probability simply do not exist, or they are extremely uncertain \citep{bennett_how_2024}.
+This model provides an excellent opportunity for detection probabilities to be estimated with higher precision for several species of landbirds in North America, and also provides a method to predict detection probabilities for species which have very little (or even no) data.
 This means that conservation problems that rely on estimates of detectability such as Value of Information analyses \citep{canessa_when_2015, bennett_when_2018} or prioritizations \citep{hanson_prioritizr_2023} can make more accurate decisions.
-For organizations such as Partners in Flight, Boreal Avian Modelling project \citep{cumming_toward_2010}, and others that rely on detectability information for bird population size estimates, these improved estimates (and new estimates and predictions) provide a way forward for continuing to produce accurate estimates of population size that are based on data-driven detectability estimates.
-
-\par The models presented here can be used to fill out the remaining missing species in the NA-POPS database, in order to estimate detectability for every species of North American landbird.
-By making use of phylogeny and trait information, bird species with fewer than 75 data points can have estimates generated based on the few data points available, and information available from other species.
-As shown here, birds with no data points can have detectability predicted that falls closely in line with similar species.
-However, we caution that detectability predictions based on models using similar species should be noted as such, so that organizations know that the estimate is simply a "best guess" based on other similar species.
+For organizations such as Partners in Flight, Boreal Avian Modelling project \citep{cumming_toward_2010}, National Audubon Society, and others that explicitly rely on detectability information for bird population size estimates or abundance modelling using small point count datasets, these improved estimates (and new estimates and predictions) provide a way forward for continuing to produce accurate estimates of population size that are based on data-driven detectability estimates.
+For applications that tend to rely on known presences, such as threat assessments, our method opens new possibilities for incorporating detectability, and thus gaining a fuller understanding of the potential for undetected occurrences.
 
@@
 However, one benefit that we have shown here is that Bayesian models can still come up with reasonable predictions if there are informed enough priors on which to fall back.
-For taxa such as butterflies for which some species have reasonable data and most have very little data \citep{lewthwaite_geographical_2022}, a researcher may still be able to glean reasonable estimates of detectability with the little data available.
-We suggest that this type of model where information about detectability is shared among species be tested with taxon that generally have fewer data collected.
+For taxa such as butterflies for which some species have reasonable data and most have very little data \citep{lewthwaite_geographical_2022}, a researcher may still be able to glean reasonable estimates of detectability with the little data available, as long as detectability for some similar species has been examined.
+We suggest that this type of model where information about detectability is shared among species be tested with taxa that generally have fewer data collected.
 
@@
 \begin{figure}[h]
-	\includegraphics{../output/plots/cv_plot.png}
+	\includegraphics{../output/plots/kfold_cv_plot.png}
 	\caption{Mean difference in pointwise posterior predictive density for cue rate (A) and EDR (B) vs.
@@
 	\label{fig:cv}
+\end{figure}
+
+\begin{figure}[h]
+	\includegraphics{../output/plots/species_cv_plot.png}
+	\caption{Mean difference in pointwise posterior predictive density for cue rate (A) and EDR (B) vs.
+log species sample size.
+Points are mean difference by species.
+Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species model preference, whereas negative differences (coloured in blue for clarity) represent a single-species model preference.
+Grey points imply no differences.
+}
+	\label{fig:species_cv}
+\end{figure}
+
+\begin{figure}[h]
+	\includegraphics{../output/plots/species_cv_distance_traits.png}
+	\caption{Mean difference in pointwise posterior predictive density for cue rate (A) and EDR (B) vs.
+log species sample size.
+Points are mean difference by species.
+Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species model preference, whereas negative differences (coloured in blue for clarity) represent a single-species model preference.
+Grey points imply no differences.
+}
+	\label{fig:distance_cv_traits}
 \end{figure}

### 2024-07-31 article/article.tex
beginning stages of revisions

@@
 	\and
-	Gahbauer, Marcel\\
+	Gahbauer, Marcel A.\\
 	\and
@@
 	\and
-	Robinson, Barry\\
+	Robinson, Barry G.\\
 	\and
@@
 
-	Accurate assessments of population status and trends rely on accounting for probability of detection, making it an important metric in conservation, yet it is often unaccounted for or unavailable on a species level.
-	Although detection probability methods and estimates exist for many North American birds, probability estimates are most accurate for common, easy to detect species. 
-	There is therefore a need for methodology that provides detection estimates for species that are rare or undersampled, which generally do not have sufficient data for accurate estimates. 
+	Accurate assessments of population status and trends rely on accounting for probability of detection.
+	However, probability of detection is often unaccounted for or unavailable on a species level. 
+	Although methods to estimate detection probability have been developed and implemented for many North American birds, estimates are most accurate for common, easy to detect species. 
+	There is therefore a need for methodology that provides detection estimates for species that are rare or undersampled, which generally lack sufficient data for accurate estimates. 
 	Here, we take advantage of similarities in detection probabilities among phylogenetically related species and species with similar traits to estimate detection probabilities in seven species of rare North American birds, using a hierarchical Bayesian multi-species removal and distance models.
-	We first show that a multi-species model of detectability can improve precision of detectability estimates for species with low sample size, and even with species up to 1000 data points.
-	However, when there is no data available for a species, the multi-species models presented here tend to be overly conservative in predicting detectability.
-	Thus, we recommend using these models to 1) provide more precise estimates of detectability for species where there are sufficient data, 2) generate estimates of detectability for species for which we have sparse (but non-zero) data, and 3) augment previous estimates of detection probabilities in the NA-POPS database of detectability offsets for North American birds by combining this model with survey-level covariates.
+	We show that a multi-species model can improve precision of detectability estimates for species with low sample size, and even with species up to 1000 observations.
+	However, when there are no data available for a species, the multi-species models presented here tend to be overly conservative in predicting detectability. 
+	Thus, we recommend using these models to 1) provide more precise estimates of detectability for species where there are sufficient data, 2) generate estimates of detectability for species for which we have sparse (but non-zero) data, and 3) improve upon previous estimates of detection probabilities for North American birds by combining this model with survey-level covariates.
 
@@
 
-\par Accurate and precise estimates of detection probability (or detectability) are important both for estimating accurate population sizes, as well as for making effective conservation decisions.
+\par Accurate and precise estimates of detection probability (or detectability) are important for estimating accurate population sizes and making effective conservation decisions.
 In conservation management, prior knowledge of detectability can be used as a metric of sensitivity to evaluate when and how to monitor a system for a species, or when to act on current information \citep{canessa_when_2015, bennett_when_2018}.
-When estimating population size, detectability often comes into play as a way to convert observed counts into densities or estimates of true abundance \citep{solymos_calibrating_2013, johnson_defense_2008}.
-Despite its importance, detectability often goes unaccounted for in studies; and for many species, the ancillary data needed to estimate detectability does not exist \citep{bennett_how_2024}.
+When estimating population size, detectability is often used to convert observed counts (i.e., relative abundance) into densities or estimates of true abundance \citep{solymos_calibrating_2013, johnson_defense_2008}.
+Despite its importance, detectability is frequently unaccounted for in studies; and for many species, the ancillary data needed to estimate detectability do not exist \citep{bennett_how_2024}.
 
 \par The wealth of openly available bird observation data in North America means that estimates of detectability for birds are relatively easy to estimate, in comparison to most other taxonomic groups \citep{bennett_how_2024}.
-The NA-POPS project is an effort to estimate detection probabilities that account for variability in detection processes through environmental and temporal covariates for as many of North America’s landbird species as possible \citep{edwards_point_2023}.
+The NA-POPS project is an effort to estimate detection probabilities that accounts for variability in detection processes through the use of environmental and temporal covariates for as many of North America’s landbird species as possible \citep{edwards_point_2023}.
 To date, the project has estimated the two components of detectability--availability (the probability that a bird sings or gives a cue during the survey period) and perceptibility (the probability that an observer sees or hears a cueing bird)--for 319 species of birds, with an additional 19 species having estimates of either availability or perceptibility.
 This project expanded previous efforts by the Boreal Avian Modelling project \citep{cumming_toward_2010} to systematically estimate detectability in landbirds \citep{solymos_calibrating_2013, solymos_evaluating_2018}.
-Both the NA-POPS project and previous estimates by Boreal Avian Modelling project use QPAD \citep{solymos_calibrating_2013}, which is a flexible method to estimate availability and perceptibility given a harmonized dataset with the necessary survey-specific data (i.e., distance to bird from observer and time to first detection; \citep{barker_ecological_2015}).
+Both the NA-POPS project and previous estimates by Boreal Avian Modelling project use the QPAD approach \citep{solymos_calibrating_2013}, which is a flexible method to estimate availability and perceptibility given a harmonized dataset with the necessary survey-specific data (i.e., distance to bird from observer and time to first detection; \citep{barker_ecological_2015}).
 
 \par Despite the efforts of the NA-POPS project, approximately 25\% of the species in the Partners in Flight’s 2016 Landbird Conservation Plan \citep{rosenberg_partners_2016} have insufficient data to directly estimate detectability \citep{edwards_point_2023}.
-These species fall short of the threshold of 75 distinct observations recommended for removal and distance modelling \citep{buckland_introduction_2001, solymos_calibrating_2013}. 
-Many of the species that fell short of that threshold in the NA-POPS database are rare or undersampled species, that either have low population sizes, or occur in rare or inaccessible habitats.
+The data for these species fall short of the threshold of 75 distinct observations recommended for removal and distance modelling \citep{buckland_introduction_2001, solymos_calibrating_2013}. 
+Many of these are rare or undersampled species, that either have low population sizes, or occur in rare or inaccessible habitats.
 %Some examples of species not modelled include Kirtland’s Warbler (\textit{Setophaga kirtlandii}) and Bicknell’s Thrush (\textit{Catharus bicknelli}), both of which have low population sizes, and LeConte’s Thrasher (\textit{Toxostoma lecontei}) and Harris’s Sparrow (\textit{Zonotrichia querula}), both of which have rare habitat associations \citep{will_handbook_2020}.
@@
 \par Detectability in birds has been shown to be at least partially driven by variation in species traits.
-\cite{johnston_species_2014} showed that detection distance - and consequently perceptibility - varied with body mass and habitat association in a selection of European landbirds.
-In this case, larger body size tended to be associated with longer detection distances, which makes sense considering larger body sizes are associated with louder and lower-pitched songs in birds \citep{bowman_adaptive_1979, fletcher_acoustics_1999, ryan_role_1985}.
+\citet{johnston_species_2014} showed that detection distance - and consequently perceptibility - varied with body mass and habitat association in a selection of European landbirds.
+In this case, larger body size tended to be associated with longer detection distances, which makes sense considering that larger body sizes are typically associated with louder and lower-pitched songs in birds \citep{bowman_adaptive_1979, fletcher_acoustics_1999, ryan_role_1985}.
 Moreover, vegetation structure influences the attenuation of sound, whereby sound diminishes quicker in a closed-forested habitat than more open habitats, thus leading to variation in detectability among species associated with different habitats \citep{waide_tropical_1988, yip_sound_2017}.

### 2024-08-01 src/scripts/post-hoc/5-predictions-figure.R
reordering species to AOS order

@@ -3,19 +3,16 @@
 # Multi-species QPAD Detectability
 # posthoc/5-predictions-figure.R
 # Created December 2023
-# Last Updated April 2024
+# Last Updated August 2024
 
 ####### Import Libraries and External Files #######
 
 library(cmdstanr)
-library(bayesplot)
 library(plyr)
 library(ggplot2)
 library(ggpubr)
 library(ggrepel)
-library(gginnards)
 theme_set(theme_pubclean())
-bayesplot::color_scheme_set("red")
 
 ####### Read Data #################################
 
@@ -55,9 +52,10 @@ rem_summary$Scientific_BT <- gsub(x = rem_summary$Scientific_BT, pattern = " ", 
 rem_summary <- dplyr::left_join(x = rem_summary, y = traits[, c("Code", "Migrant")],
                                  by = "Code")
 
-sp <- c("LCTH", "LEPC", "HASP", "TRBL", "SPOW", "KIWA", "BITH")
+sp <- c("LEPC", "SPOW", "BITH", "LCTH", 'HASP', 'TRBL', 'KIWA')
 
 to_plot <- rem_summary[which(rem_summary$Code %in% sp), ]
+to_plot <- to_plot[match(sp, to_plot$Code), ]
 
 species_vars <- to_plot$variable
 
@@ -87,17 +85,17 @@ for (s in sp)
 
 other_sp_df <- other_sp_df[which(other_sp_df$Alpha > 0.001), ]
 
-rem_draws <- exp(rem_model$draws(species_vars))
-attributes(rem_draws)$dimnames$variable <- to_plot$Code
+to_plot$Code <- factor(to_plot$Code, levels = sp)
 
-(removal_plot <- bayesplot::mcmc_intervals(rem_draws) + 
-    xlab("Predicted Cue Rate") + 
-    ylab("Species") +
-    xlim(0,1)+
-    coord_flip())
-
-removal_plot$layers <- c(geom_point(data = other_sp_df, aes(x = mean, y = Species, alpha = Alpha), size = 0.5, position=position_jitter(height=0.2)),
-                         removal_plot$layers)
+removal_plot <- ggplot(data = to_plot, aes(x = Code, y = exp(mean))) +
+  geom_point(data = other_sp_df, aes(y = mean, x= Species, alpha = Alpha), size = 0.5, position=position_jitter(width = 0.2)) +
+  geom_point(size = 4, color = "darkred") +
[…]

### 2024-08-01 src/scripts/post-hoc/1-1vs1-figure.R
got rid of species labels and decreased size of figure

@@ -3,7 +3,7 @@
 # Multi-species QPAD Detectability
 # posthoc/1-1vs1-figure.R
 # Created December 2023
-# Last Updated June 2024
+# Last Updated August 2024
 
 ####### Import Libraries and External Files #######
 
@@ -97,7 +97,6 @@ rem_diff_model_draws <- diff_model_run$draws(format = "df")
     geom_abline(slope = 1, color = "red", linetype = 2) +
     xlim(0,1) +
     ylim(0,1) +
-    geom_text_repel(box.padding = 0.5, max.overlaps = Inf, aes(label = Label)) +
     xlab("Cue Rate (Single Species)") + 
     ylab("Cue Rate (Multi Species)") +
     NULL)
@@ -174,7 +173,6 @@ dis_diff_model_draws <- diff_model_run$draws(format = "df")
     geom_abline(slope = 1, color = "red", linetype = 2) +
     xlim(0,800) +
     ylim(0,800) +
-    geom_text_repel(box.padding = 0.5, max.overlaps = Inf, aes(label = Label)) +
     xlab("EDR (Single Species)") + 
     ylab("EDR (Multi Species)") +
     NULL)
@@ -197,11 +195,11 @@ write.table(edr_diff_model, file = "data/generated/tau_difference_model.csv", se
 
 
 tiff("output/plots/1vs1.tiff",
-     width = 6, height = 6, res = 600, units = "in")
-ggarrange(removal_plot, distance_plot, nrow = 2, labels = c("A", "B"))
+     width = 6, height = 3, res = 600, units = "in")
+ggarrange(removal_plot, distance_plot, nrow = 1, labels = c("A", "B"))
 dev.off()
 
 png("output/plots/1vs1.png",
-     width = 6, height = 6, res = 600, units = "in")
-ggarrange(removal_plot, distance_plot, nrow = 2, labels = c("A", "B"))
+     width = 6, height = 3, res = 600, units = "in")
+ggarrange(removal_plot, distance_plot, nrow = 1, labels = c("A", "B"))
 dev.off()

### 2024-08-01 src/scripts/post-hoc/4-species-cv-figure.R
Updated traits-based figure

@@ -167,7 +167,10 @@ distance_cv_traits_plot <- ggplot(data = dis_cv_df, aes(x = True_Exp, y = Predic
   geom_point(aes(color = Trait_Group)) +
   geom_abline(slope = 1, color = "red", linetype = 2) +
   xlim(60,150) + ylim(60,150) +
-  theme(legend.position = "right") +
+  theme(legend.position = "right",
+        legend.title = "Trait Group") +
+  xlab("EDR (Full Data)") +
+  ylab("EDR (No Data)") +
   NULL
 
 

### 2024-08-01 article/article.tex
close to final draft

@@
 	There is therefore a need for methodology that provides detection estimates for species that are rare or undersampled, which generally lack sufficient data for accurate estimates. 
-	Here, we take advantage of similarities in detection probabilities among phylogenetically related species and species with similar traits to estimate detection probabilities in seven species of rare North American birds, using a hierarchical Bayesian multi-species removal and distance models.
-	We show that a multi-species model can improve precision of detectability estimates for species with low sample size, and even with species up to 1000 observations.
-	However, when there are no data available for a species, the multi-species models presented here tend to be overly conservative in predicting detectability. 
+	Here, we take advantage of similarities in detection probabilities among phylogenetically-related species and species with similar traits to estimate detection probabilities in seven species of rare North American birds, using a hierarchical Bayesian multi-species removal and distance models.
+	We showed that a multi-species model improved precision of detectability estimates for species with low sample size, and even with species up to 1000 observations.
+	However, when there were no data available for a species, the multi-species models presented here tended to be overly conservative in predicting detectability. 
 	Thus, we recommend using these models to 1) provide more precise estimates of detectability for species where there are sufficient data, 2) generate estimates of detectability for species for which we have sparse (but non-zero) data, and 3) improve upon previous estimates of detection probabilities for North American birds by combining this model with survey-level covariates.
@@
 \par \citet{solymos_phylogeny_2018} recommended using these phylogenetic and trait relationships to predict components of detectability (i.e., availability through cue rate, and perceptibility through detection distance) more broadly.
-By leveraging the information provided by species with many data it may be possible to inform estimates for species with much fewer (including potentially zero) data points.
-This could be done in two ways: by making use of shrinkage factors and partial pooling, as done in Bayesian hierarchical models (Gelman and Vehtari 2021), and by accounting for phylogenetic relationships and the relative "relatedness" of different species, as done with Gaussian processes (Bernardo et al.
-1998; McElreath 2020).
+By leveraging the information provided by species with many data it may be possible to inform estimates for species with much fewer (including potentially zero) data points \citep{sollmann_hierarchical_2016, pacifici_guidelines_2014, zipkin_impacts_2009, white_conservation_2013}.
+This could be done in two ways: by making use of shrinkage factors and partial pooling, as done in Bayesian hierarchical models \citep{gelman_what_2021, gelman_bayesian_2006}, and by accounting for phylogenetic relationships and the relative "relatedness" of different species, as done with Gaussian processes \citep{bernardo_regression_1998, mcelreath_continous_2020}.
 
-\par The goal of this paper is to create a multi-species model of detectability that incorporates species traits and phylogeny, to both allow for the relaxation of the sample size requirements for estimating detectability, and to derive estimates of detectability for bird species without any data.
-To achieve this goal, we wanted to satisfy the following objectives and predictions:
+\par The goal of this paper is to present a multi-species model of detectability that incorporates species traits and phylogeny, to both allow for the relaxation of the sample size requirements for estimating detectability, and to derive estimates of detectability for bird species without any data.
+To achieve this goal, we wanted to satisfy the following objectives:
 \begin{enumerate}
-	\item Compare and contrast estimates of cue rate and EDR produced from a Bayesian hierarchical QPAD removal model and distance model, respectively, to estimates produced from single-species model which does not share information.
+	\item Compare and contrast estimates of cue rate and EDR produced from a Bayesian hierarchical QPAD removal model and distance model, respectively, to estimates produced from single-species model that does not share information.
 	We predicted that for both the Bayesian hierarchical QPAD removal model and Bayesian hierarchical QPAD distance model, estimates of cue rate and EDR, respectively, will differ very little from those estimated from a single-species model where no information is shared \cite{edwards_point_2023}, for species where we have sufficient sample size.
@@
 	We predicted that estimates of cue rate and EDR will be more precise when produced by a Bayesian hierarchical QPAD model than with a single-species QPAD model, but this increased precision will be most evident with smaller sample sizes.
-	\item Assess the capacity for a multi-species model to produces estimates of cue rate or EDR for species that are data sparse.
+	\item Assess the capacity for a multi-species model to produce estimates of cue rate or EDR for species that are data sparse.
 	We predicted that a multi-species model will be able to produce better fitting estimates of cue rate or EDR than a single species model, particularly for data sparse species.
@@
 
-Given these comparisons, we provide a case study of predicting detectability components for seven species for which we had little to no data: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird (\textit{Agelaius tricolor}), and Spotted Owl (\textit{Strix occidentalis}).
+Given these comparisons, we provide a case study of predicting detectability components for seven species for which we had little to no data: Lesser Prairie-chicken (\textit{Tympanuchus pallidicinctus}), Spotted Owl (\textit{Strix occidentalis}), Bicknell’s Thrush (\textit{Catharus bicknelli}), LeConte’s Thrasher (\textit{Toxostoma lecontei}), Harris’s Sparrow (\textit{Zonotrichia querula}), Tricolored Blackbird (\textit{Agelaius tricolor}), and Kirtland’s Warbler (\textit{Setophaga kirtlandii}).
 All seven of these bird species are listed on Partners in Flight’s watchlist, indicating their conservation importance \citep{will_handbook_2020}.
@@
 \par We accessed 2000 pseudo-posterior trees from \citet{jetz_global_2012} for all species with sufficient data for removal modelling ($>75$ observations as per \cite{edwards_point_2023,solymos_evaluating_2018,buckland_introduction_2001}).
-We also requested that the phylogenies include the following bird species with insufficient removal and/or distance data: LeConte's Thrasher, Bicknell's Thrush, Lesser Prairie-chicken, Harris's Sparrow, Kirtland's Warbler, Tricolored Blackbird, and Spotted Owl.
-All seven of these bird species are listed on Partner's in Flight's watchlist for various threats \citep{will_handbook_2020}. 
+We also included the following bird species with insufficient removal and/or distance data: Lesser Prairie-chicken, Spotted Owl, Bicknell's Thrush, LeConte's Thrasher, Harris's Sparrow, Tricolored Blackbird, and Kirtland's Warbler.
 This resulted in 2000 pseudo-posterior trees that contained 323 species.
 
-\par We generated a phylogenetic correlation matrix $\mathbf{C}$ using the same methods outlined in the Appendix of \citet{solymos_phylogeny_2018} that generate distances between species, and further processed using the R package "ape" \citet{paradis_ape_2019}.
+\par We generated a phylogenetic correlation matrix $\mathbf{C}$ using the same methods outlined in the Appendix of \citet{solymos_phylogeny_2018} that generate distances between species based on covariance, as processed using the R package "ape" \citep{paradis_ape_2019}.
 
@@
 Any bird species that has specific areas of their range that is only for breeding (i.e., not year-round throughout their range) was classified as "migratory".
+We note that for the purposes of this study, altitudinal and latitudinal migrants were classified as migratory, but irruptive species such as crossbills were not.
 Body mass for each bird species was obtained from the Elton traits dataset \citep{wilman_eltontraits_2014}. 
@@
 
-\par For removal modelling, we were only interested in the time bin in which the bird was recorded, and not the distance, and so we will sum counts from each sampling event $i$ over all distance bands.
-Thus, we will be considering counts $Y_{sij.} = \sum_{k=1}^{K}$. 
+\par For removal modelling, we are only interested in the time bin in which the bird was recorded, and not the distance, and so we sum counts from each sampling event $i$ over all distance bands. 
+Thus, we consider counts $Y_{sij.} = \sum_{k=1}^{K}$. 
 From \citet{solymos_calibrating_2013}, we model $Y_{sij.}$ as
+
 $$Y_{sij.} \sim multinomial\left(Y_{si..}, \mathbf{\Pi}_{si.}\right).$$
@@
 We set $\lambda = 0.76$, consistent with previous results \cite{solymos_phylogeny_2018}. 
+Of course, we could have chosen to draw estimates from a distribution of possible $\lambda$ as in \citet{solymos_phylogeny_2018}, but we chose not to for simplicity here. 
+We note that the suite of species modelled in \citet{solymos_evaluating_2018} was primarily Boreal birds, and so there could be slight differences in the phylogenetic strength of a larger suite of species such as what we model here.
 We also estimated the extra variance term and used an exponential prior $\sigma \sim exp(5)$.
@@
 
-\par For distance modelling, we are only interested in the distance to the cueing bird, and not the time bin that the bird was recorded in, and so we will sum counts from each sampling event $i$ over all time bands.
-Thus, we will be considering counts $Y_{si.k} = \sum_{j=1}^{J}$. 
+\par For distance modelling, we are only interested in the distance to the cueing bird, and not the time bin that the bird was recorded in, and so we sum counts from each sampling event $i$ over all time bands.
+Thus, we consider counts $Y_{si.k} = \sum_{j=1}^{J}$. 
 From \citet{solymos_calibrating_2013}, we model $Y_{si.k}$ as
@@
 
-\par To directly compare the multi-species models described above to its single-species equivalent (see below), we built two Bayesian single-species models in Stan.
+\par To directly compare the multi-species models described above to its single-species equivalent (see below), we built two Bayesian single-species models using the Stan Probabilistic Language \citep{stan_development_team_stan_2024}.
 The single species models used the same likelihood statements as above for the count data, except rather than cue rate and EDR being estimated with shared information, they are estimated independently from each other.
-That is, $\log \phi_s \sim N(0,1) \forall s$, and $\log \tau_s \sim N(0,1) \forall s$, thus making it equivalent to the Null model in \citet{edwards_point_2023}, just estimated in a Bayesian framework rather than maximum likelihood.
+That is, $\log \phi_s \sim N(0,1) \forall s$, and $\log \tau_s \sim N(0,1) \forall s$, thus making it equivalent to the Null model in \citet{edwards_point_2023}, but estimated in a Bayesian rather than maximum likelihood framework.
 
@@
 \par To test our prediction that there would be little change between estimates of species with sufficient sizes, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data.
+We did not include Woodhouse's Scrub-jay, Pacific Wren, or Sagebrush Sparrow in the analysis, because these species were not available in the phylogenies from \citet{jetz_global_2012}.
 Rufous Hummingbird (\textit{Selasphorus rufus}) was dropped from the distance modelling due to issues with divergent transitions \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
 Tricolored Blackbird was only included in the distance model because there was sufficient distance sampling data.
+This resulted in a total of 316 species for removal modelling, and 315 species for distance modelling.
 We also ran the single species removal and distance models for the same suite of birds.
@@
 \par To test our prediction about improved accuracy, we ran a k-fold cross-validation with 5 folds, stratified by species.
-That is, for each of the five cross-validation run, we held out 20\% of the data for each species for both the multi-species and single-species models. 
-We then derived multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR, for each cross-validation run, and used these parameter estimates to calculate the estimated log pointwise posterior density (LPPD) of each of the held out data points.
-We then calculated the difference in LPPD for removal and distance models, by subtracting the LPPD of the single-species model from the LPPD of the multi-species model (i.e., multi-species minus single-species).
+That is, for each of the five cross-validation runs, we held out 20\% of the data for each species for both the multi-species and single-species models. 
+We then derived multi-species estimates of cue rate and EDR and single-species estimates of cue rate and EDR, for each cross-validation run, and used these parameter estimates to calculate the estimated log pointwise posterior density (LPPD) of each of the held out data points, which tells us the likelihood of the data point under the estimated parameters.
+We then calculated the difference in LPPD for removal and distance models, by subtracting the pointwise LPPD of the single-species model from the LPPD of the multi-species model (i.e., multi-species minus single-species).
 Finally, we took the mean and standard error of the difference in LPPD for each species to create a species-specific mean and standard error of difference in LPPD.
-With this, positive intervals (i.e., where mean $\pm$ standard error is all positive) would indicate that the multi-species model was preferred, and negative intervals would indicate that the single-species model was preferred.
-If our prediction is true, then we would expect to see majority positive values of the species-specific mean LPPD.
+With this, positive intervals (i.e., where mean $\pm$ standard error is all positive) would indicate that the multi-species model was preferred, and negative intervals would indicate that the single-species model was preferred, because there would be an overall higher average likelihood of the data points given the parameters in the multi-species model than the single-species model with a positive interval.
+If our prediction is true, then we would expect to see majority positive values of the species-specific mean LPPD for species with lower sample sizes.
 
@@
 \par We ran the multi-species removal and distance models again, this time with the minimal data available for seven species of undersampled landbirds, and generated estimates of cue rate and EDR for these seven species.
-For tricolored blackbird, we had sufficient data points for distance modelling for Objectives 1 - 4, and so we only generated estimates of cue rate here.
+For Tricolored Blackbird, we had sufficient data points for distance modelling for Objectives 1 - 4, and so we only generated estimates of cue rate here.
 For two of the species, Kirtland's Warbler and Bicknell's Thrush, we used the predicted cue rate and EDR to calculate availability and perceptibility, and hence detectability, for a given survey type, and compared these metrics to previous estimates of detectability obtained from other studies.
@@
 Manipulation of phylogenetic trees, including finding a consensus tree for visualization and generating variance-covariance matrices based on the 2000 bootstrapped trees, was done using the \texttt{ape} R package \citep{paradis_ape_2019}.
-All Bayesian models were coded in the Stan Probabilistic Language \cite{stan_development_team_stan_2019}, and models were run using the \texttt{cmdstanr} R package \citep{gabry_cmdstanr_2022} and visualized using the \texttt{bayesplot} R package \citep{gabry_visualization_2019}.
+All Bayesian models were coded in Stan \cite{stan_development_team_stan_2024}, and models were run using the \texttt{cmdstanr} R package \citep{gabry_cmdstanr_2022} and visualized using the \texttt{bayesplot} R package \citep{gabry_visualization_2019}.
 Each of the single-species models and multi-species models, as well as the models created to assess performance, were run using 1000 warmup iterations and 2000 sampling iterations, each on 4 chains for a total of 8000 draws per model run.
@@
 \par Estimates of cue rate obtained from a multi-species model closely followed estimates of cue rate obtained from a single-species model (y-intercept = $8.47 \times 10^{-4}$, slope = $0.99$; Figure \ref{fig:1vs1}A).
-Ten out of the 316 species considered for removal modelling had a percent change in cue rate of more than 10\%.
-These species were: Willow Ptarmigan (\textit{Lagopus lagopus}), Rusty Blackbird (\textit{Euphagus carolinus}), Red-bellied Sapsucker (\textit{Sphyrapicus ruber}), Prairie Falcon (\textit{Falco mexicanus}), Nuttall's Woodpecker (\textit{Picoides nuttallii}), Northern Goshawk (\textit{Accipiter atricapillus}), Ferruginous Hawk (\textit{Buteo regalis}), Common Redpoll (\textit{Acanthis flammea}), Calliope Hummingbird (\textit{Selasphorus calliope}), and Bohemian Waxwing (\textit{Bombycilla garrulus}).
+Ten out of the 316 species considered for removal modelling had a percent difference in cue rate of more than 10\%.
+These species were: Willow Ptarmigan (\textit{Lagopus lagopus}), Calliope Hummingbird (\textit{Selasphorus calliope}), American Goshawk (\textit{Accipiter atricapillus}), Ferruginous Hawk (\textit{Buteo regalis}), Red-breasted Sapsucker (\textit{Sphyrapicus ruber}), Nuttall's Woodpecker (\textit{Picoides nuttallii}), Prairie Falcon (\textit{Falco mexicanus}), Bohemian Waxwing (\textit{Bombycilla garrulus}), Common Redpoll (\textit{Acanthis flammea}), Rusty Blackbird (\textit{Euphagus carolinus}).
 
@@
 With all species included, we obtained a y-intercept of $6.39$ and a slope of $0.91$.
-However, two out of the 315 species had a percent change in EDR of more than 10\%: Greater Prairie-chicken (\textit{Tympanuchus cupido}) and Chihuahuan Raven (\textit{Corvus cryptoleucus}) (Figure \ref{fig:1vs1}B).
+However, two out of the 315 species had a percent difference in EDR of more than 10\%: Greater Prairie-chicken (\textit{Tympanuchus cupido}) and Chihuahuan Raven (\textit{Corvus cryptoleucus}) (Figure \ref{fig:1vs1}B).
 These species were particularly influential in this regression; after removing these two species, we obtained a y-intercept of $1.25$ and a slope of $0.98$, indicating a near 1-to-1 relationship for the remaining species.
+
+\par Although it was not a focus of this study, Figure \ref{fig:params} shows the distributions of group-level parameters for both the multi-species removal and distance models. 
+Overall, we found little support for a migrant effect in the multi-species removal models (Figure \ref{fig:params}A), but some evidence for a migrant effect in the multi-species distance models (Figure \ref{fig:params}B).
+We found support for habitat, body mass, and song pitch effects in the multi-species distance models (Figure \ref{fig:params}B).
 
@@
 \par The k-fold cross validation revealed that the multi-species removal model had better predictive performance for more species (200 out of 316) than the single-species removal model (94 out of 316).
-Twenty-two out of 316 species showed no preference between either model.
+For 22 of 316 species, neither model was preferred.
 Mean overall difference in species-wise mean LPPD was $2.65 \times 10^{-4}$, indicating a small overall preference for the multi-species model.
 As expected, multi-species models tended to have better predictive capacity when sample size was low, with a negligible difference as sample size increased (Figure \ref{fig:cv}A).
-Among the ten species with greater than 10\% change in cue rate, three species (Bohemian Waxwing, Northern Goshawk, and Rusty Blackbird) had the multi-species model chosen as the better model, whereas the remaining seven species had the single-species model chosen as the better model.
+Among the ten species with greater than 10\% difference in cue rate, the multi-species model was better for three species (Bohemian Waxwing, American Goshawk, and Rusty Blackbird) whereas the single-species was better for the remaining seven species.
 
@@
 Similar to removal models, multi-species distance models tended to have better predictive capacity when sample size was low, with a negligible difference as sample size increased (Figure \ref{fig:cv}B).
-Among the two species with greater than 10\% change in EDR, Chihuahuan Raven had the single-species model chosen as the preferred model, and Greater Prairie-chicken had no preference.
+Of the two species with greater than 10\% difference in EDR, the single-species model was preferred for Chihuahuan Raven, while there was no preference for Greater Prairie-chicken.
 
@@
 
-\par As expected, the predicted cue rate for Bicknell’s Thrush (0.28 cues per minute) and the predicted cue rate for Kirtland’s Warbler (0.29 cues per minute) fell very close to the overall mean cue rate of 0.27 for all 11 other Turdidae species and 0.31 for all 44 other Parulidae species in NA-POPS, respectively.
-That is, in the absence of any data, the estimated cue rate of Bicknell’s Thrush and Kirtland’s Warbler was shrunk toward cue rates of most similar species.
+\par As expected, in the absence of any data, the estimated cue rate of Bicknell’s Thrush and Kirtland’s Warbler was shrunk toward cue rates of most similar species.
+For Bicknell's Thrush, the predicted cue rate was 0.28 cues per minute, which fell very close to the overall mean cue rate of 0.27 for the 11 other Turdidae species modelled here.
+Likewise, for Kirtland's Warbler, the predicted cue rate was 0.29 cues per minute, which fell very close to the overall mean cue rate of 0.31 for the 44 other Parulidae species modelled here.
 
-\par For distance modelling, the predicted EDR for Bicknell's Thrush (73.6 m) and the predicted EDR for Kirtland's Warbler (75.3 m) fell reasonably close to the overall mean EDR of 92.7 m for all other 11 Turdidae species and 61.2 m for all 44 other Parulidae species in NA-POPS, respectively.
-That is, in the absence of any data (or with very little data in the case of Kirtland’s Warbler), the estimated EDR of Bicknell’s Thrush and Kirtland’s Warbler was similar to those of other Thrushes and Warblers, respectively, but with small departures from overall mean based on specific traits.
+\par Similarly, in the absence of any data (or with very little data in the case of Kirtland’s Warbler), the estimated EDR of Bicknell’s Thrush and Kirtland’s Warbler was similar to those of other Thrushes and Warblers, respectively, but with small departures from overall mean based on specific traits.
+For Bicknell's Thrush, the predicted EDR of 73.6 m fell reasonably close to the overall mean EDR of 92.7 m for 11 other Turdidae species modelled here.
+For Kirtland's Warbler, the predicted EDR of 75.3 m fell reasonably close to the overall mean EDR of 61.2 m for the 44 other Parulidae species modelled here.
 
 \par For Bicknell's Thrush, using Equations 1 and 2, the predicted cue rate of 0.28, predicted EDR of 73.3 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.34.
-We can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respectively credible intervals) to get a range of detectabilities from 0.09 through 0.70.
+Using the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respectively credible intervals), the range of detectabilities was from 0.09 through 0.70.
 \citet{aubry_bicknells_2018}, however, estimate detection probabilities of $\geq$ 0.74 and $\leq$ 0.88, and so even our highest potential value of detectability falls short of their estimated range.
@@
 
-\par For Kirtland's Warbler, Equations 1 and 2, the predicted cue rate of 0.29, predicted EDR of 75.3 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.36.
-Similar to our Bicknell's Thrush example, we can also use the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respective credible intervals) to obtain a range of detectabilities from 0.12 through 0.67.
-A study by \citet{van_dyke_comparative_2022} found estimated detection probabilities for Kirtland’s Warbler that ranged from effectively 0, through to a detection probability of 0.73
-In that study, they compared estimates of detectability among different jack pine stands and red pine stands, where the detectability was higher in jack pine stands than the red pine stands.
+\par For Kirtland’s Warbler, using the predicted cue rate of 0.29, predicted EDR of 75.3 m, and an example survey design of 5 minutes recording all birds within 100 m, we estimate a detectability of 0.36.
+Similar to our Bicknell’s Thrush example, using the lowest predicted cue rate and EDR and the highest predicted cue rate and EDR (based on the respective credible intervals), the range of detectabilities was from 0.12 through 0.67.
+\citet{van_dyke_comparative_2022} found estimated detection probabilities for Kirtland’s Warbler that ranged from effectively 0, through to a detection probability of 0.73.
+In that study, they compared estimates of detectability among different jack pine (\textit{Pinus banksiana}) stands and red pine (\textit{Pinus resinosa}) stands, where the detectability was higher in jack pine stands than the red pine stands.
 
@@
 
-\par In this paper, we developed a multi-species model of detectability that accounts for phylogeny and traits, expanding upon the work from \citet{solymos_phylogeny_2018}.
+\par In this paper, we presented a multi-species model of detectability that accounts for phylogeny and traits, expanding upon the work from \citet{solymos_phylogeny_2018}.
 We applied the multi-species removal model and multi-species distance model to the large dataset of point counts collected by the NA-POPS project \citep{edwards_point_2023}, and compared estimates of cue rate and EDR from a multi-species context to estimates of cue rate and EDR from a single species context.
@@
 \par Overall, the multi-species models generated very similar estimates of cue rate and EDR as single-species models, for species where we had sufficient data points.
-That is, we did not see too many significant changes in cue rate and EDR by modelling with a multi-species model versus modelling with a single-species model.
+That is, we did not see many significant changes in cue rate and EDR by modelling with a multi-species model versus modelling with a single-species model.
 This provided an appropriate baseline to continue our comparisons for data-sparse species.
-We saw a few exceptions in the distance modelling, in that estimates of EDR for Greater Prairie-chicken and Chihuahuan Raven in a multi-species context were much lower than that estimated from a single-species context.
-We note that these species are generally observed either in flight in highly specific regions (for Chihuahuan Raven) or in wide-open landscapes (for both). 
-Thus, it is possible that data for these species are biased high, such that a multi-species model brings the estimates down to a more realistic value.
+This is perhaps unsurprising because the species with sufficient sample sizes effectively have more precise likelihoods, which would ``overwhelm" any hierarchical prior distribution assigned to the species, thus diminishing the species' pooling factor \citep{gelman_bayesian_2006}.
+We saw a few exceptions in the distance modelling, in that estimates of EDR for Greater Prairie-chicken and Chihuahuan Raven in a multi-species context were much lower than those estimated from a single-species context.
+We note that these species are generally observed either in flight in highly specific regions (for Chihuahuan Raven) or in open landscapes (for both).
+Thus, it is possible that data for these species are biased high, because an observer would be able to identify such a species even at extreme distances.
+In this case, a multi-species model brings the estimates down to a more realistic value for the species based on other similar species.
+Of course, this may neglect species-specific idiosyncrasies in detectability (but see subsection \ref{discussion-nodata}).
 
-\par We did not see a large difference in precision of estimates between the single-species and multi-species model, except for a slight difference in cue rate estimates for sample sizes less than 1000.
+\par We did not see a large difference in precision of estimates between the single-species and multi-species model, except for a slight difference in cue rate estimates for species with sample sizes less than 1000.
 However, we note that our baseline single-species model was a Bayesian implementation of the QPAD models described in \citet{solymos_calibrating_2013}, and we used weakly informative $N(0,1)$ priors on the parameters for cue rate and EDR.
 In earlier versions of this study, we compared the precision of multi-species estimates of cue rate and EDR to the precision of the estimates obtained using maximum likelihood methods in \citet{edwards_point_2023}, and found a more pronounced increase in precision in the multi-species estimates.
-Therefore, while the multi-species implementation can somewhere improve precision of estimates for lower sample size, the use of even weakly-informed priors clearly goes a long way in improving precision.
+Therefore, while the multi-species implementation can somewhat improve precision of estimates for lower sample size, the use of even weakly-informed priors clearly goes a long way in improving precision.
 We therefore recommend future implementations of the NA-POPS detectability database to adopt a Bayesian modelling approach using at least weakly-informed priors.
@@
 
-\subsection{Predicting Detectability for Species with No Data}
+\subsection{Predicting Detectability for Species with No Data}\label{discussion-nodata}
 
-\par Interestingly, these multi-species models were not able to provide good estimates of cue rate or EDR for species where we had no data, as evidenced by our leave-species-out cross validation.
+\par Interestingly, these multi-species models underestimated cue rate and EDR for species where we had no data, as evidenced by our leave-species-out cross validation.
 For both multi-species removal and distance modelling, the predicted cue rates or EDRs for species that were left out of a given cross-validation fold were, on average, underestimated compared to predicted cue rates for EDRs for the same species with all its data.
@@
 Then, when the species is modelled without the data, the cue rate is shrunk toward "similar" species that may occur in both open and closed environments, therefore shrinking the cue rate down to a lower value.
-Future studies could seek to disentangle the possible effects of habitat on cue rate; to go one step further, future studies should build upon the work of \citet{martin-schwarze_joint_2021} to assess how truly independent cue rate and EDR are.
+Future studies could seek to disentangle the possible effects of habitat on cue rate.
+In addition, future studies should build upon the work of \citet{martin-schwarze_joint_2021} to assess how truly independent cue rate and EDR are.
 
@@
 This was evidenced by the predicted EDRs only ever being between 69 - 115 m.
-When investigating further, Figure \ref{fig:distance_cv_traits} demonstrated a fairly clean deliniation of these different trait groupings, with little variation within each.
+When investigating further, Figure \ref{fig:distance_cv_traits} demonstrated a fairly clean delineation of these different trait groupings, with little variation within each.
 With this in mind, while studies have shown that traits are useful for explaining variation in EDR, traits alone do not appear to be sufficient in predicting EDRs in similar birds, because traits cannot account for species-specific variation in EDR.
-Future studies should seek to continue understanding drivers and predictors of detection distance in birds, both related to and complementary to traits.
+Future studies should seek to continue understanding drivers and predictors of detection distance in birds, both related to and complementary to traits, and think carefully about how to best group together species for purposes of multi-species modelling \citep{pacifici_guidelines_2014}.
 
@@
 
-\par Given the results of the k-fold cross-validation, we were comfortable with applying this model to species where we had only a few data points, such as Harris's Sparrow, Lesser Prairie-chicken, Spotted Owl, LeConte's Thrasher, and Tricolored Blackbird.
+\par Given the results of the k-fold cross-validation, we were comfortable with applying this model to species where we had only a few data points, such as Lesser Prairie-chicken, Spotted Owl, LeConte's Thrasher, Harris's Sparrow, and Tricolored Blackbird.
 We were able to show that if we had at least some data points for a given species, the fitted model with either phylogeny and migratory status for cue rate, or traits for EDR, can help provide additional information for these data sparse species.
@@
 
-\par We caution, however, the use of these models for species with no data points, given the result of the leave-species-out cross validation.
-For Bicknell's Thrush and Kirtland's Warbler, both estimates of detectability appeared to be underestimated compared to estimates found in the literature, which corroborates the consistent underestimation found from the cross-validation.
-Of course, in the absence of any detectability information at all, our models can at least provide a starting point for a researcher to consider, which can then be turned into ranges of potential detectability \citep{bennett_how_2024}.
+\par We caution, however, against the use of these models for species with no data points, given the result of the leave-species-out cross validation.
+Of course, in the absence of any detectability information at all, our models can at least provide a starting point for a researcher to consider, which can then be used to estimate reasonable ranges of potential detectability to help with management and survey planning \citep{bennett_how_2024}.
 Regardless, we recommend continuing to make use of existing available data collected by previous studies on these data-sparse species prior to using these models to further bolster predictions.
 
-In both cases above, we also caution that detectability predictions for species with few data (e.g., < 1000 observations) should be identified as such, so that users know that the estimate depends on data from other similar species.
+\par In the case of Bicknell's Thrush and Kirtland's Warbler, both estimates of detectability appeared to be underestimated compared to estimates found in the literature.
+This does corroborate the consistent underestimation found from the cross-validation.
+However, we note that species with low populations in specific habitats such as these species, highly-targeted surveys are often used such that reported detectability rates may be higher than for ``normal" (i.e., more broadly-sampled) species.
+With that in mind, it is important also to consider the applicability of this multi-species model to species where targeted surveys are more likely to occur.
+
+In all cases above, we also caution that detectability predictions for species with few data (e.g., $<$ 1000 observations) should be identified as such, so that users know that the estimate depends on data from other similar species.
 
@@
 \par In early implementations of this study, the distance model struggled with sampling the posterior distribution, in that over a quarter of the iterations resulted in a divergent transition \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
-However, after several iterations of this model, we identified that the samplers were not appropriately sampling Rufous Hummingbird's EDR, in that there appeared to be a self-imposed barrier in being able to sample EDRs less than 25 m.
+However, after several iterations of this model, we identified that the samplers were not appropriately sampling Rufous Hummingbird's EDR, in that there appeared to be a barrier in being able to sample EDRs less than 25 m.
 Once we removed Rufous Hummingbird from the analysis, all divergent transitions vanished.
@@
 This model provides an excellent opportunity for detection probabilities to be estimated with higher precision for several species of landbirds in North America, and also provides a method to predict detection probabilities for species which have very little (or even no) data.
-This means that conservation problems that rely on estimates of detectability such as Value of Information analyses \citep{canessa_when_2015, bennett_when_2018} or prioritizations \citep{hanson_prioritizr_2023} can make more accurate decisions.
+This means that conservation problems that rely on estimates of detectability such as Value of Information analyses \citep{canessa_when_2015, bennett_when_2018} or prioritizations \citep{hanson_prioritizr_2023} can make more effective decisions.
 For organizations such as Partners in Flight, Boreal Avian Modelling project \citep{cumming_toward_2010}, National Audubon Society, and others that explicitly rely on detectability information for bird population size estimates or abundance modelling using small point count datasets, these improved estimates (and new estimates and predictions) provide a way forward for continuing to produce accurate estimates of population size that are based on data-driven detectability estimates.
-For applications that tend to rely on known presences, such as threat assessments, our method opens new possibilities for incorporating detectability, and thus gaining a fuller understanding of the potential for undetected occurrences.
+For applications that tend to rely on known presences, such as threat assessments, our method opens new possibilities for incorporating detectability, and thus gaining a fuller understanding of  potential locations of undetected occurrences, and more effective allocation of survey effort.
 
@@
 Obviously, given the vast amount of bird data available, it was relatively easy to obtain estimates for enough data-rich species such that the few data-poor species could have enough information to borrow.
-However, one benefit that we have shown here is that Bayesian models can still come up with reasonable predictions if there are informed enough priors on which to fall back.
+However, one benefit that we have shown here is that Bayesian models can still derive reasonable predictions if there are informed enough priors.
 For taxa such as butterflies for which some species have reasonable data and most have very little data \citep{lewthwaite_geographical_2022}, a researcher may still be able to glean reasonable estimates of detectability with the little data available, as long as detectability for some similar species has been examined.
@@
 	\caption{1-to-1 plot comparing cue rates (A) and EDRs (B) estimated from the single-species model vs the multi-species model for each species.
-Labelled species are those which had a greater than 20\% difference in cue rate or EDR between the single-species and multi-species models.
-Insets are modelled differences with 95\% credible interval (shaded area) between cue rates (A) and EDRs (B) estimated from the multi-species model developed here, minus the single-species model in NA-POPS \citep{edwards_point_2023}.}
+Red dashed line denotes the 1-to-1 line.
+Black line is regression line with posterior draws in grey.
+Grey dots are individual species.}
 	\label{fig:1vs1}
+\end{figure}
+
+\begin{figure}[h]
+	\includegraphics{../output/plots/parameters_plot.png}
+	\caption{Group parameter posterior distributions for multi-species removal model (A) and multi-species distance model (B).}
+	\label{fig:params}
 \end{figure}
@@
 Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species model preference, whereas negative differences (coloured in blue for clarity) represent a single-species model preference.
-Grey points imply no differences.
+Grey points denote no differences.
 }
@@
 	\includegraphics{../output/plots/species_cv_plot.png}
-	\caption{Mean difference in pointwise posterior predictive density for cue rate (A) and EDR (B) vs.
-log species sample size.
-Points are mean difference by species.
-Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species model preference, whereas negative differences (coloured in blue for clarity) represent a single-species model preference.
-Grey points imply no differences.
-}
+	\caption{(A) Estimates of cue rate for a species (black dots) given full data versus predictions of cue rate for the same species given no data.
+(B) Estimates of effective detection radius (EDR) for a species (black dots) given full data versus predictions of EDR for the same species given no data.
+For both, red dashed line denotes the 1-to-1 line.
+Black line is regression line with posterior draws in grey.}
 	\label{fig:species_cv}
@@
 	\includegraphics{../output/plots/species_cv_distance_traits.png}
-	\caption{Mean difference in pointwise posterior predictive density for cue rate (A) and EDR (B) vs.
-log species sample size.
-Points are mean difference by species.
-Difference was calculated as multi-species minus single-species, and so positive differences (coloured in red for clarity) represent a multi-species model preference, whereas negative differences (coloured in blue for clarity) represent a single-species model preference.
-Grey points imply no differences.
-}
+	\caption{Estimates of effective detection radius (EDR) for a species (dots) given full data versus predictions of EDR for the same species given no data, focusing in on species with EDRs between 60 m and 150 m.
+Colours denote discrete trait group combinations of migratory strategy and habitat preference.
+Red dashed line denotes the 1-to-1 line.
+Black line is regression line with posterior draws in grey.}
 	\label{fig:distance_cv_traits}
@@
 	\includegraphics{../output/plots/predictions_figure.png}
-	\caption{(A) Cue rates (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for seven species of rare and undersampled North American landbirds.
-Species are 4 letter banding codes: SPOW = Spotted Owl (n = 7), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), TRBL = Tricoloured Blackbird (n = 74), HASP = Harris's Sparrow (n = 65), KIWA = Kirtland's Warbler (n = 0), and LEPC = Lesser Prairie-chicken (n = 45).
+	\caption{(A) Cue rates (red circles) and 90\% credible interval (red lines) for seven species of rare and undersampled North American landbirds.
+Species are 4 letter banding codes: LEPC = Lesser Prairie-chicken (n = 45), SPOW = Spotted Owl (n = 7), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), HASP = Harris's Sparrow (n = 65), TRBL = Tricoloured Blackbird (n = 74), and KIWA = Kirtland's Warbler (n = 0).
 Black points are cue rate estimates of other similar species with the same migration strategy, with the opacity of the point indicating the strength of the relationship.
-(B) Effective detection radius (points), 50\% credible interval (bolded line), and 95\% credible interval (thin line) for six species of rare and undersampled North American landbirds.
-Species are 4 letter banding codes: SPOW = Spotted Owl (n = 28), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), HASP = Harris's Sparrow (n = 13), KIWA = Kirtland's Warbler (n = 1), and LEPC = Lesser Prairie-chicken (n = 45).
+(B) Effective detection radius (red circles) and 90\% credible interval (red line) for six species of rare and undersampled North American landbirds.
+Species are 4 letter banding codes: LEPC = Lesser Prairie-chicken (n = 45), SPOW = Spotted Owl (n = 28), BITH = Bicknell's Thrush (n = 0), LCTH = LeConte's Thrasher (n = 1), HASP = Harris's Sparrow (n = 13), KIWA = Kirtland's Warbler (n = 1).
 Black points are EDR estimates of species with the same migration strategy and habitat preference, and with a mass and pitch within 40\% in either direction.}

### 2024-08-01 src/scripts/post-hoc/4-species-cv-figure.R
forgot to hit save

@@ -168,11 +168,10 @@ distance_cv_traits_plot <- ggplot(data = dis_cv_df, aes(x = True_Exp, y = Predic
   geom_abline(slope = 1, color = "red", linetype = 2) +
   xlim(60,150) + ylim(60,150) +
   theme(legend.position = "right",
-        legend.title = "Trait Group") +
+        legend.title = element_blank()) +
   xlab("EDR (Full Data)") +
   ylab("EDR (No Data)") +
   NULL
-
 
 ####### Output ####################################
 

### 2024-08-05 article/article.tex
final edits to draft 4

@@
 \title{Predicting detection probabilities for rare and undersampled North American landbirds}
-\author{
-	Edwards, Brandon P.M.\\
-	\and
-	Gahbauer, Marcel A.\\
-	\and
-	Grinde, Alexis\\
-	\and
-	Hope, David\\
-	\and
-	Knight, Elly C.\\
-	\and
-	Michel, Nicole L.\\
-	\and
-	Robinson, Barry G.\\
-	\and
-	Solymos, Peter\\
-	\and
-	Bennett, Joseph R.\\
-	\and
-	Smith, Adam C.\\
-}
-
+%\author{
+%	Edwards, Brandon P.M.\\
+%	\and
+%	Gahbauer, Marcel A.\\
+%	\and
+%	Grinde, Alexis\\
+%	\and
+%	Hope, David\\
+%	\and
+%	Knight, Elly C.\\
+%	\and
+%	Michel, Nicole L.\\
+%	\and
+%	Robinson, Barry G.\\
+%	\and
+%	Solymos, Peter\\
+%	\and
+%	Bennett, Joseph R.\\
+%	\and
+%	Smith, Adam C.\\
+%}
+\date{}
 \begin{document}
@@
 
-\par Accurate and precise estimates of detection probability (or detectability) are important for estimating accurate population sizes and making effective conservation decisions.
+\par Accurate and precise estimates of detection probability (or detectability) are important for estimating accurate population sizes and making effective conservation decisions \citep{bennett_how_2024}.
 In conservation management, prior knowledge of detectability can be used as a metric of sensitivity to evaluate when and how to monitor a system for a species, or when to act on current information \citep{canessa_when_2015, bennett_when_2018}.
@@
 
-\par The wealth of openly available bird observation data in North America means that estimates of detectability for birds are relatively easy to estimate, in comparison to most other taxonomic groups \citep{bennett_how_2024}.
+\par However, the wealth of openly-available observation data for birds in North America provides the opportunity to estimate detectability for a large number of species \citep{bennett_how_2024}. 
 The NA-POPS project is an effort to estimate detection probabilities that accounts for variability in detection processes through the use of environmental and temporal covariates for as many of North America’s landbird species as possible \citep{edwards_point_2023}.
 To date, the project has estimated the two components of detectability--availability (the probability that a bird sings or gives a cue during the survey period) and perceptibility (the probability that an observer sees or hears a cueing bird)--for 319 species of birds, with an additional 19 species having estimates of either availability or perceptibility.
-This project expanded previous efforts by the Boreal Avian Modelling project \citep{cumming_toward_2010} to systematically estimate detectability in landbirds \citep{solymos_calibrating_2013, solymos_evaluating_2018}.
-Both the NA-POPS project and previous estimates by Boreal Avian Modelling project use the QPAD approach \citep{solymos_calibrating_2013}, which is a flexible method to estimate availability and perceptibility given a harmonized dataset with the necessary survey-specific data (i.e., distance to bird from observer and time to first detection; \citep{barker_ecological_2015}).
+This project expanded previous efforts by the Boreal Avian Modelling project (\citet{cumming_toward_2010}, see also https://borealbirds.ca) to systematically estimate detectability for landbirds \citep{solymos_calibrating_2013, solymos_evaluating_2018}.
 
-\par Despite the efforts of the NA-POPS project, approximately 25\% of the species in the Partners in Flight’s 2016 Landbird Conservation Plan \citep{rosenberg_partners_2016} have insufficient data to directly estimate detectability \citep{edwards_point_2023}.
+\par Both the NA-POPS project and previous estimates by Boreal Avian Modelling project use the QPAD approach \citep{solymos_calibrating_2013}, which is a flexible method to estimate availability and perceptibility given a harmonized dataset with the necessary survey-specific data (i.e., distance to bird from observer and time to first detection; \citep{barker_ecological_2015}).
+In the QPAD approach, availability is calculated based on estimates of cue rate from removal models \citep{farnsworth_removal_2002, alldredge_time--detection_2007} and perceptibility is calculated based on estimates of effective detection radius (EDR) from distance sampling and the detection function \citep{buckland_introduction_2001}.
+
+\par Despite the efforts of the NA-POPS project, approximately 25\% of species in Partners in Flight’s 2016 Landbird Conservation Plan \citep{rosenberg_partners_2016} have insufficient data to directly estimate detectability \citep{edwards_point_2023}.
 The data for these species fall short of the threshold of 75 distinct observations recommended for removal and distance modelling \citep{buckland_introduction_2001, solymos_calibrating_2013}. 
@@
 %Some examples of species not modelled include Kirtland’s Warbler (\textit{Setophaga kirtlandii}) and Bicknell’s Thrush (\textit{Catharus bicknelli}), both of which have low population sizes, and LeConte’s Thrasher (\textit{Toxostoma lecontei}) and Harris’s Sparrow (\textit{Zonotrichia querula}), both of which have rare habitat associations \citep{will_handbook_2020}.
-Species with small population sizes and rare habitats are also often considered vulnerable, and thus detectability information for them would be particularly important to inform their conservation.
+Species with small population sizes and rare habitats are also often considered vulnerable \citep{davies_synergistic_2004, gray_effects_1989, segura_specialist_2007}, and thus detectability information would be particularly important to inform their conservation.
 
-\par Detectability in birds has been shown to be at least partially driven by variation in species traits.
+\par Evidence suggests detectability in birds is at least partially driven by variation in species traits.
 \citet{johnston_species_2014} showed that detection distance - and consequently perceptibility - varied with body mass and habitat association in a selection of European landbirds.
-In this case, larger body size tended to be associated with longer detection distances, which makes sense considering that larger body sizes are typically associated with louder and lower-pitched songs in birds \citep{bowman_adaptive_1979, fletcher_acoustics_1999, ryan_role_1985}.
+In this case, larger body size was associated with longer detection distances, likely because  larger body sizes are easier to detect visually and typically associated with louder and lower-pitched songs \citep{bowman_adaptive_1979, fletcher_acoustics_1999, ryan_role_1985}.
 Moreover, vegetation structure influences the attenuation of sound, whereby sound diminishes quicker in a closed-forested habitat than more open habitats, thus leading to variation in detectability among species associated with different habitats \citep{waide_tropical_1988, yip_sound_2017}.
@@
 \par \citet{solymos_phylogeny_2018} recommended using these phylogenetic and trait relationships to predict components of detectability (i.e., availability through cue rate, and perceptibility through detection distance) more broadly.
-By leveraging the information provided by species with many data it may be possible to inform estimates for species with much fewer (including potentially zero) data points \citep{sollmann_hierarchical_2016, pacifici_guidelines_2014, zipkin_impacts_2009, white_conservation_2013}.
+ By leveraging information provided by species with many detections in available datasets, it may be possible to inform detectability estimates for species with fewer (including potentially zero) detections \citep{sollmann_hierarchical_2016, pacifici_guidelines_2014, zipkin_impacts_2009, white_conservation_2013}.
 This could be done in two ways: by making use of shrinkage factors and partial pooling, as done in Bayesian hierarchical models \citep{gelman_what_2021, gelman_bayesian_2006}, and by accounting for phylogenetic relationships and the relative "relatedness" of different species, as done with Gaussian processes \citep{bernardo_regression_1998, mcelreath_continous_2020}.
 
-\par The goal of this paper is to present a multi-species model of detectability that incorporates species traits and phylogeny, to both allow for the relaxation of the sample size requirements for estimating detectability, and to derive estimates of detectability for bird species without any data.
-To achieve this goal, we wanted to satisfy the following objectives:
+\par The goal of this paper is to create a multi-species model of detectability that incorporates species traits and phylogeny, to both allow for the relaxation of sample size requirements for estimating detectability, and to derive estimates of detectability for bird species without any data.
+To achieve this goal, we aimed to satisfy the following objectives:
 \begin{enumerate}
-	\item Compare and contrast estimates of cue rate and EDR produced from a Bayesian hierarchical QPAD removal model and distance model, respectively, to estimates produced from single-species model that does not share information.
-	We predicted that for both the Bayesian hierarchical QPAD removal model and Bayesian hierarchical QPAD distance model, estimates of cue rate and EDR, respectively, will differ very little from those estimated from a single-species model where no information is shared \cite{edwards_point_2023}, for species where we have sufficient sample size.
-	\item Compare precision in estimates of cue rate and EDR produced from a Bayesian hierarchical QPAD removal model and distance model, respectively, to estimates produced from single-species model which does not share information.
-	We predicted that estimates of cue rate and EDR will be more precise when produced by a Bayesian hierarchical QPAD model than with a single-species QPAD model, but this increased precision will be most evident with smaller sample sizes.
+	\item Compare the mean and precision of estimates of cue rate and EDR produced from a multi-species, Bayesian hierarchical QPAD removal model and distance model, respectively, to estimates produced from single-species models that do not share information across species.
+	For species with a sufficient sample size, we predicted the mean and precision of cue rate and EDR estimates will be similar from multi- and single-species models \citep{edwards_point_2023}.
+	However, for species with relatively small sample sizes, we predict estimates for both cue rate and EDR will be more precise from the multi-species Bayesian model.
 	\item Assess the capacity for a multi-species model to produce estimates of cue rate or EDR for species that are data sparse.
@@
 \par The following subsections outline the modelling techniques used in this study as well as the data that were collected. 
-We start by describing the data that will be used in the models, and then introduce the Bayesian hierarchical QPAD removal and distance models.
+We start by describing data used in the models, and then introduce the Bayesian hierarchical QPAD removal and distance models.
 Finally, we describe the experiments conducted to assess model performance and model comparison, which then allow us to make predictions for the seven undersampled species of interest.
@@
 This consists of $712,138$ point counts conducted across $292$ projects in Canada and the United States, $422,514$ of which used removal sampling \citep{alldredge_time--detection_2007, farnsworth_removal_2002} and $522,820$ of which used distance sampling \citep{buckland_introduction_2001, buckland_distance_2015}.
-All of these point counts were recorded in the database in a standard way, using practices developed by the Boreal Avian Modelling project \citep{barker_ecological_2015}.
+ All point counts were recorded in the database using a standard method developed by the Boreal Avian Modelling project \citep{barker_ecological_2015}.
 
@@
 \subsubsection{Species Traits}
-\par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts of all the birds to model in this study \citep{billerman_birds_2022}. 
-Birds were classified as "closed habitat" if their primary habitat was forests, and "open habitat" if their primary habitat was grasslands or shrublands.
-Any bird species that has specific areas of their range that is only for breeding (i.e., not year-round throughout their range) was classified as "migratory".
+\par Species traits (song pitch, migratory vs resident, open vs. closed habitat) were obtained from Birds of the World accounts for all species considered in this study \citep{billerman_birds_2022}. 
+Species were classified as "closed habitat" if their primary habitat was forests, and "open habitat" if their primary habitat was grasslands or shrublands.
+Species with specific areas of their range that is only for breeding (i.e., not year-round throughout their range) was classified as "migratory".
 We note that for the purposes of this study, altitudinal and latitudinal migrants were classified as migratory, but irruptive species such as crossbills were not.
-Body mass for each bird species was obtained from the Elton traits dataset \citep{wilman_eltontraits_2014}. 
-Where body mass was not available through the Elton traits dataset, the Birds of the World account \citep{billerman_birds_2022} was used instead.
+Body mass for each bird species was obtained from the Elton traits dataset if available \citep{wilman_eltontraits_2014}, and through the Birds of the World account \citep{billerman_birds_2022} if not.
 
@@
 \par We are most interested in estimating $\phi_s$, the cue rate for species $s$. 
-To allow for the partial pooling of estimates for each species to be based on a phylogenetic relationship, we can use Gaussian processes, which effectively allow for "continuous" factors \citep{bernardo_regression_1998, mcelreath_continous_2020}.
+To allow for the partial pooling of estimates for each species to be based on a phylogenetic relationship, we can use Gaussian processes, which effectively allow for ``continuous" factors \citep{bernardo_regression_1998, mcelreath_continous_2020}.
 Let $\mathbf{\Phi}$ be the vector of cue rates (of length $s$). 
@@
 \par To test our prediction that there would be little change between estimates of species with sufficient sizes, we ran the Bayesian hierarchical QPAD Removal and Distance models on the set of species for which there was sufficient data.
-We did not include Woodhouse's Scrub-jay, Pacific Wren, or Sagebrush Sparrow in the analysis, because these species were not available in the phylogenies from \citet{jetz_global_2012}.
+We did not include Woodhouse's Scrub-jay (\textit{Aphelocoma woodhouseii}), Pacific Wren (\textit{Troglodytes pacificus}), or Sagebrush Sparrow (\textit{Artemisiospiza nevadensis})in the analysis, because these species were not available in the phylogenies from \citet{jetz_global_2012}.
 Rufous Hummingbird (\textit{Selasphorus rufus}) was dropped from the distance modelling due to issues with divergent transitions \citep{betancourt_diagnosing_2016, leimkuhler_simulating_2005}.
@@
 
-\subsubsection{Objective 2: Comparison of Precision}
 \par To test our prediction about improved precision, we plotted the species-specific estimates of standard deviations of cue rate and EDR, respectively, against the log sample size, for each of the single-species and multi-species model.
@@
 
-\subsubsection{Objective 3: Assessment of Estimates for Data-sparse Species}
+\subsubsection{Objective 2: Assessment of Estimates for Data-sparse Species}
 
@@
 
-\subsubsection{Objective 4: Assessment of Out-of-sample Predictions}
+\subsubsection{Objective 3: Assessment of Out-of-sample Predictions}
 \par To test our prediction that a multi-species model would be able to predict cue rates or EDRs for species with no data, we ran a 10-fold leave-species-out cross validation.
@@
 \par We ran the multi-species removal and distance models again, this time with the minimal data available for seven species of undersampled landbirds, and generated estimates of cue rate and EDR for these seven species.
-For Tricolored Blackbird, we had sufficient data points for distance modelling for Objectives 1 - 4, and so we only generated estimates of cue rate here.
+For Tricolored Blackbird, we had sufficient data points for distance modelling for Objectives 1 - 3, and so we only generated estimates of cue rate here.
 For two of the species, Kirtland's Warbler and Bicknell's Thrush, we used the predicted cue rate and EDR to calculate availability and perceptibility, and hence detectability, for a given survey type, and compared these metrics to previous estimates of detectability obtained from other studies.
@@
 Manipulation of phylogenetic trees, including finding a consensus tree for visualization and generating variance-covariance matrices based on the 2000 bootstrapped trees, was done using the \texttt{ape} R package \citep{paradis_ape_2019}.
-All Bayesian models were coded in Stan \cite{stan_development_team_stan_2024}, and models were run using the \texttt{cmdstanr} R package \citep{gabry_cmdstanr_2022} and visualized using the \texttt{bayesplot} R package \citep{gabry_visualization_2019}.
+All Bayesian models were coded in Stan \cite{stan_development_team_stan_2024}, and models were run using the \texttt{cmdstanr} R package \citep{gabry_cmdstanr_2023} and visualized using the \texttt{bayesplot} R package \citep{gabry_visualization_2019}.
 Each of the single-species models and multi-species models, as well as the models created to assess performance, were run using 1000 warmup iterations and 2000 sampling iterations, each on 4 chains for a total of 8000 draws per model run.
@@
 
-\par Code is available open-source at https://github.com/BrandonEdwards/multispecies-qpad-detectability and will be archived with a DOI upon acceptance of this paper.
+\par Code is available open-source at [REDACTED FOR PEER REVIEW] and will be archived with a DOI upon acceptance of this paper.
 
@@
 
+\par For both the removal and distance models, we found little evidence that a multi-species model improved the precision of estimates of cue rate or EDR, respectively.
+For removal modelling, the LOESS curves show some difference in sample sizes less than approximately $\exp\left\{7\right\} \approx 1097$, where the estimated LOESS curve for single-species standard deviation was slightly higher than the estimated LOESS curve for multi-species standard deviations (Figure \ref{fig:sd}A).
+For distance modelling, the curves were relatively equivalent throughout the range of sample sizes (Figure \ref{fig:sd}B).
+
 \par Although it was not a focus of this study, Figure \ref{fig:params} shows the distributions of group-level parameters for both the multi-species removal and distance models. 
@@
 We found support for habitat, body mass, and song pitch effects in the multi-species distance models (Figure \ref{fig:params}B).
-
-\subsection{Comparison of Precision between Single Species and Multi Species}
-
-\par For both the removal and distance models, we found little evidence that a multi-species model improved the precision of estimates of cue rate or EDR, respectively.
-For removal modelling, the LOESS curves show some difference in sample sizes less than approximately $\exp\left\{7\right\} \approx 1097$, where the estimated LOESS curve for single-species standard deviation was slightly higher than the estimated LOESS curve for multi-species standard deviations (Figure \ref{fig:sd}A).
-For distance modelling, the curves were relatively equivalent throughout the range of sample sizes (Figure \ref{fig:sd}B).
 
@@
 Then, when the species is modelled without the data, the cue rate is shrunk toward "similar" species that may occur in both open and closed environments, therefore shrinking the cue rate down to a lower value.
-Future studies could seek to disentangle the possible effects of habitat on cue rate.
-In addition, future studies should build upon the work of \citet{martin-schwarze_joint_2021} to assess how truly independent cue rate and EDR are.
+Future studies could seek to disentangle the possible effects of habitat on cue rate, and build upon the work of \citet{martin-schwarze_joint_2021} to assess how truly independent cue rate and EDR are.
+An additional mechanism could be the effect of wind on perceptibility.
+In grassland environments, wind direction and wind speed tends to alter the detection radius of birds, in that the wind will ``carry" the song toward (or away from) the observer \citep{rigby_factors_2019}.
+Future iterations of this model could consider including weather-related variables, particularly for species whose perceptibility are most affected by factors such as wind.
 
@@
 This model provides an excellent opportunity for detection probabilities to be estimated with higher precision for several species of landbirds in North America, and also provides a method to predict detection probabilities for species which have very little (or even no) data.
-This means that conservation problems that rely on estimates of detectability such as Value of Information analyses \citep{canessa_when_2015, bennett_when_2018} or prioritizations \citep{hanson_prioritizr_2023} can make more effective decisions.
+This means that conservation problems that rely on estimates of detectability such as Value of Information analyses \citep{canessa_when_2015, bennett_when_2018} or prioritizations \citep{hanson_prioritizr_2022} can make more effective decisions.
 For organizations such as Partners in Flight, Boreal Avian Modelling project \citep{cumming_toward_2010}, National Audubon Society, and others that explicitly rely on detectability information for bird population size estimates or abundance modelling using small point count datasets, these improved estimates (and new estimates and predictions) provide a way forward for continuing to produce accurate estimates of population size that are based on data-driven detectability estimates.
@@
 For taxa such as butterflies for which some species have reasonable data and most have very little data \citep{lewthwaite_geographical_2022}, a researcher may still be able to glean reasonable estimates of detectability with the little data available, as long as detectability for some similar species has been examined.
-We suggest that this type of model where information about detectability is shared among species be tested with taxa that generally have fewer data collected.
+We suggest that this type of model---where information about detectability is shared among species---be tested with taxa that generally have fewer data collected, and we encourage researchers to continue to make use of existing sources of data \citep{binley_minimizing_2023} to inform these models.
 
@@
 \begin{figure}[h]
+	\includegraphics{../output/plots/sd_comp.png}
+	\caption{LOESS curves of standard deviations of cue rate (A) and EDR (B) for species modelled by a single-species model (blue lines) and multi-species model (red lines).}
+	\label{fig:sd}
+\end{figure}
+
+\begin{figure}[h]
 	\includegraphics{../output/plots/parameters_plot.png}
@@
 	\label{fig:params}
-\end{figure}
-
-\begin{figure}[h]
-	\includegraphics{../output/plots/sd_comp.png}
-	\caption{LOESS curves of standard deviations of cue rate (A) and EDR (B) for species modelled by a single-species model (blue lines) and multi-species model (red lines).}
-	\label{fig:sd}
 \end{figure}

### 2024-08-06 article/article.tex
update article latex code for thesis template

@@
-\documentclass[12pt]{article}
+\chapter{Predicting detection probabilities for rare and undersampled North American landbirds}
 
-\usepackage{setspace}
-\doublespacing
-
-\usepackage{natbib}
-\bibliographystyle{apalike}
-
-\usepackage[margin=1.0in]{geometry}
-
-\usepackage{amsmath}
-
-\usepackage{lineno}
-
-\usepackage{graphicx}
-\linenumbers
-
-%opening
-\title{Predicting detection probabilities for rare and undersampled North American landbirds}
-%\author{
-%	Edwards, Brandon P.M.\\
-%	\and
-%	Gahbauer, Marcel A.\\
-%	\and
-%	Grinde, Alexis\\
-%	\and
-%	Hope, David\\
-%	\and
-%	Knight, Elly C.\\
-%	\and
-%	Michel, Nicole L.\\
-%	\and
-%	Robinson, Barry G.\\
-%	\and
-%	Solymos, Peter\\
-%	\and
-%	Bennett, Joseph R.\\
-%	\and
-%	Smith, Adam C.\\
-%}
-\date{}
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
+\section{abstract}
 
@@
 	Thus, we recommend using these models to 1) provide more precise estimates of detectability for species where there are sufficient data, 2) generate estimates of detectability for species for which we have sparse (but non-zero) data, and 3) improve upon previous estimates of detection probabilities for North American birds by combining this model with survey-level covariates.
-
-\end{abstract}
 
@@
 
-\bibliography{refs}
-
 \section{Figures}
@@
 \begin{figure}[h]
-	\includegraphics{../output/plots/1vs1.png}
+	\includegraphics{../ch4-multispecies-detectability/output/plots/1vs1.png}
 	\caption{1-to-1 plot comparing cue rates (A) and EDRs (B) estimated from the single-species model vs the multi-species model for each species.
@@
 \begin{figure}[h]
-	\includegraphics{../output/plots/sd_comp.png}
+	\includegraphics{../ch4-multispecies-detectability/output/plots/sd_comp.png}
 	\caption{LOESS curves of standard deviations of cue rate (A) and EDR (B) for species modelled by a single-species model (blue lines) and multi-species model (red lines).}
@@
 \begin{figure}[h]
-	\includegraphics{../output/plots/parameters_plot.png}
+	\includegraphics{../ch4-multispecies-detectability/output/plots/parameters_plot.png}
 	\caption{Group parameter posterior distributions for multi-species removal model (A) and multi-species distance model (B).}
@@
 \begin{figure}[h]
-	\includegraphics{../output/plots/kfold_cv_plot.png}
+	\includegraphics{../ch4-multispecies-detectability/output/plots/kfold_cv_plot.png}
 	\caption{Mean difference in pointwise posterior predictive density for cue rate (A) and EDR (B) vs.
@@
 \begin{figure}[h]
-	\includegraphics{../output/plots/species_cv_plot.png}
+	\includegraphics{../ch4-multispecies-detectability/output/plots/species_cv_plot.png}
 	\caption{(A) Estimates of cue rate for a species (black dots) given full data versus predictions of cue rate for the same species given no data.
@@
 \begin{figure}[h]
-	\includegraphics{../output/plots/species_cv_distance_traits.png}
+	\includegraphics{../ch4-multispecies-detectability/output/plots/species_cv_distance_traits.png}
 	\caption{Estimates of effective detection radius (EDR) for a species (dots) given full data versus predictions of EDR for the same species given no data, focusing in on species with EDRs between 60 m and 150 m.
@@
 \begin{figure}[h]
-	\includegraphics{../output/plots/predictions_figure.png}
+	\includegraphics{../ch4-multispecies-detectability/output/plots/predictions_figure.png}
 	\caption{(A) Cue rates (red circles) and 90\% credible interval (red lines) for seven species of rare and undersampled North American landbirds.
